### 2015-07-04 free-surface-paper.tex
Initial commit

%% 
%% Copyright 2007, 2008, 2009 Elsevier Ltd
%% 
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%% 
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%% 
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%% 
%% Template article for Elsevier's document class `elsarticle'
%% with harvard style bibliographic references
%% SP 2008/03/01

\documentclass[preprint,12pt,authoryear]{elsarticle}

%% Use the option review to obtain double line spacing
%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times,authoryear]{elsarticle}
%% \documentclass[final,1p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,3p,times,authoryear]{elsarticle}
%% \documentclass[final,3p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,5p,times,authoryear]{elsarticle}
%% \documentclass[final,5p,times,twocolumn,authoryear]{elsarticle}

%% For including figures, graphicx.sty has been loaded in
%% elsarticle.cls. If you prefer to use the old commands
%% please give \usepackage{epsfig}

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.
%% \usepackage{lineno}

\journal{Physics of the Earth and Planetary Interiors}

\begin{document}

\begin{frontmatter}

%% Title, authors and addresses

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for theassociated footnote;
%% use the fnref command within \author or \address for footnotes;
%% use the fntext command for theassociated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for theassociated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \address{Address\fnref{label3}}
%% \fntext[label3]{}

\title{Free surface computations in mantle convection models}

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{}
%% \address[label1]{}
%% \address[label2]{}

\author{I. Rose, B. Buffett, and T. Heister}

\address{}

\begin{abstract}
%% Text of abstract

\end{abstract}

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword

%% PACS codes here, in the form: \PACS code \sep code

%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)

\end{keyword}

\end{frontmatter}

%% \linenumbers

%% main text
\section{Introduction}
\label{sec:intro}

Goal: provide for free surface computations in mantle convection simulations that are
\begin{itemize}
\item robust, stable, accurate
\item dimension independent
\item work with adaptive mesh refinement
\item parallelized
\item agnostic to mesh geometry
\end{itemize}

\section{Methodology}
\cite{kronbichler2012high}
\subsection{Mass conservation}
\begin{itemize}
\item mass conservation requires $\int_\Omega {\bf u \cdot n} = 0$
\item not all mesh advection schemes respect this
\item for short simulations, it is not likely to make a big difference
\item for long simulations there can be significant changes in domain volume
\item finite element projections do the best job of conservation of mass
\end{itemize}
\subsection{Remeshing}
\begin{itemize}
\item ALE does not specify what happens to interior nodes
\item primary goal is preserving mesh regularity
\item we solve a laplace equation for mesh displacements
\item different boundary conditions for the laplace problem for different velocity boundary conditions
\end{itemize}
\subsection{Parallelization}
\begin{itemize}
\item mesh vertex positions are stored in a distributed vector
\item vector is redistributed with p4est on mesh refinement
\end{itemize}
\section{Stabilization}
\subsection{Kaus, M\"ulhaus and May}
\subsection{Nonstandard finite differences?}
Is this a good idea?  I think it could be a novel stable way to stabilize free surface
computations.
\section{Benchmarking}
\subsection{Crameri benchmarks}
more or less good go go
\subsection{New benchmark?}
topography over slab with high viscosity?
3D version of Crameri benchmark?
\section{Discussion}

%% The Appendices part is started with the command \appendix;
%% appendix sections are then done as normal sections
%% \appendix

%% \section{}
%% \label{}

%% If you have bibdatabase file and want bibtex to generate the
%% bibitems, please use
%%
\bibliographystyle{elsarticle-harv} 
\bibliography{free-surface-paper}

%% else use the following coding to input the bibitems directly in the
%% TeX file.

%\begin{thebibliography}{00}

%% \bibitem[Author(year)]{label}
%% Text of bibliographic item

%\bibitem[ ()]{}

%\end{thebibliography}
\end{document}

\endinput
%%
%% End of file `elsarticle-template-harv.tex'.

### 2015-07-07 free-surface-paper.tex
Brain dump

@@
 \usepackage{amssymb}
+\usepackage{amsmath}
+
+\newif\ifdetail
+\detailfalse
+%\detailtrue
+
 %% The amsthm package provides extended theorem environments
@@
 
-\section{Methodology}
-\cite{kronbichler2012high}
+\section{Stability}
+\subsection{Eigenvalue analysis}
+\label{sec:eigenvalue}
+
+We begin with the homogeneous Stokes equations for incompressible creeping flow:
+\begin{equation}
+\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p = 0
+\label{eq:stokes}
+\end{equation}
+
+\begin{equation}
+\nabla \cdot \mathbf{u} = 0
+\label{eq:incompressible}
+\end{equation}
+where $\mathbf{u}$ is the velocity, $p$ is the pressure, $\eta$ is the viscosity, and 
+$\varepsilon(\mathbf{u}) = \frac{1}{2}(\nabla \mathbf{u} + (\nabla \mathbf{u} )^T )$ is the strain-rate tensor.
+
+We will proceed with this analysis within a finite element framework, though similar arguments should 
+work for other discrete methods.
+We transform the governing equations into the weak form amenable to finite elements by standard procedures \citep[e.g.][]{zienkiewicz1977finite} to get
+
+\begin{equation}
+-\int_\Omega 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u} ) + \int_\Omega p \nabla \cdot \mathbf{w} 
++ \int_{\partial \Omega_N} \mathbf{w} \cdot \mathbf{T} \cdot \mathbf{n} = 0 
+\label{eq:weak_stokes}
+\end{equation}
+
+\begin{equation}
+\int_\Omega q \nabla \cdot \mathbf{u} = 0
+\label{eq:weak_incompressible}
+\end{equation}
+where $\mathbf{w}$ and $q$ are suitably chosen test functions and the integrals over 
+$\Omega$ and $\partial \Omega_N$ are over the volume of the domain and the part of the boundary with stress boundary conditions, respectively.
+$\mathbf{T}$ is the stress tensor, defined by
+\begin{equation}
+\mathbf{T} = 2 \eta \varepsilon(\mathbf{u}) - p \mathbf{I}
+\end{equation}
+The integral over the surface in Equation~\eqref{eq:weak_stokes} accounts for boundary stresses, 
+which should be zero when evaluated on a true free surface.
+We will find it analytically useful to evaluate the integrals in Equation~\eqref{eq:weak_stokes} 
+over the hydrostatic reference surface and introduce an auxilliary variable $\zeta$ which 
+represents the topography of the free surface relative to that reference surface.
+The equation for time evolution of the free surface is then
+\begin{equation}
+\frac{d \zeta}{dt} = \mathbf{u \cdot n}
+\label{eq:surface_evolution}
+\end{equation}
+where $\mathbf{u}$ is evaluated at the free surface.
+The stress on this reference surface can be approximated by the hydrostatic pressure
+\begin{equation}
+\mathbf{T} \approx \rho g \zeta \mathbf{I}
+\label{eq:hydrostatic}
+\end{equation}
+where $\rho$ is the density and $g$ is gravity. Equation~\eqref{eq:weak_stokes} then becomes
+
+\begin{equation}
+-\int_\Omega 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u} ) + \int_\Omega p \nabla \cdot \mathbf{w} 
++ \int_{\partial \Omega_N} \rho g \zeta  \mathbf{w} \cdot \mathbf{n} = 0 
+\end{equation}
+
+We would like to analyze the time evolution of the normal modes of this system: each mode 
+is the relaxation of topography with a characteristic relaxation time.  
+We denote the normal modes by $\left[ \mathbf{u}_i, p_i, \zeta_i \right]^T$, with
+relaxation times $\tau_i$, where the subscript corresponds to the $i$th normal mode.
+
+The equations decouple for the normal modes, and Equation~\eqref{eq:surface_evolution} then becomes
+
+\begin{equation}
+\frac {d}{d t} \zeta_i(\mathbf{x},t) = \frac{d}{dt} \zeta_i(\mathbf{x})e^{-t/\tau_i} = -\frac{\zeta_i(\mathbf{x},t)}{\tau_i} = \mathbf{u}_i \cdot \mathbf{n}
+\end{equation}
+This can then be used to eliminate $\zeta$ from the Stokes system:
+\begin{equation}
+-\int_\Omega 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u}_i ) + \int_\Omega p_i \nabla \cdot \mathbf{w} 
+= \tau_i \int_{\partial \Omega_N} \rho g (\mathbf{u}_i \cdot \mathbf{n} ) (\mathbf{w} \cdot \mathbf{n})
+\label{eq:weak_eigen}
+\end{equation}
+When these equations are discretized \citep[e.g.][]{kronbichler2012high} we get
+\begin{equation}
+\begin{bmatrix}
+A & B^T \\
+B & 0 \\
+\end{bmatrix}
+\begin{bmatrix}
+\mathbf{u}^n_i \\
+p^n_i
+\end{bmatrix}
+=
+\tau_i
+\begin{bmatrix}
+M & 0 \\
+0 & 0
+\end{bmatrix}
+\begin{bmatrix}
+\mathbf{u}^n_i \\
+p^n_i
+\end{bmatrix}
+\label{eq:generalized_eigenvalue}
+\end{equation}
+where $\mathbf{u}^n_i$ and $p^n_i$ are finite-dimensional representations of $\mathbf{u}_i$ and $p_i$,
+and $M$ is the discretization of the bilinear form on the right-hand side of Equation~\eqref{eq:weak_eigen}.
+
+Equation~\eqref{eq:generalized_eigenvalue} is a generalized eigenvalue problem for the normal modes of the system.
+It is rather more difficult to solve than a standard eigenvalue problem because the matrix on the right-hand-side 
+is not invertible. It may, however, be transformed into a standard eigenvalue problem.
+Define
+\begin{equation}
+\begin{aligned}
+C &= 
+\begin{bmatrix}
+A & B^T \\
+B & 0 \\
+\end{bmatrix} \\
+D &= 
+\begin{bmatrix}
+M & 0 \\
+0 & 0
+\end{bmatrix} \\
+\mathbf{y}_i &= 
+\begin{bmatrix}
+\mathbf{u}^n_i \\
+p^n_i
+\end{bmatrix} \\
+\end{aligned}
+\end{equation}
+Then multiplying both sides by $C^{-1}$ we get
+\begin{equation}
+\tau_i (C^{-1}D)\mathbf{y}_i = \mathbf{y}_i
+\end{equation}
+
+This eigenvalue equation can be solved for the normal modes and relaxation times of the Stokes system with 
+a free surface.
+
+\subsection{Time integration of the free surface}
+\label{sec:timestepping}
+
+Armed with the normal modes and relaxation times of the Stokes system, we can write down the
+formal solution to Equation~\ref{eq:surface_evolution}. Let the initial surface topography be 
+represented by a linear combination of its normal modes
+\begin{equation}
+\zeta(\mathbf{x}, t=0) = \displaystyle \sum_i a_i \zeta_i(\mathbf{x})
+\end{equation}
+then the time evolution of the free surface is given by
+\begin{equation}
+\zeta( \mathbf{x}, t) = \displaystyle \sum_i a_i \zeta_i(\mathbf{x}) e^{-t/\tau_i}
+\end{equation}
+
+Of course, most finite element (or finite difference, or finite volume) geodynamic simulations do not resolve 
+the solution and surface topography into its normal modes and integrate those separately. 
+Indeed, analytical solutions for the normal modes are only known for simple geometries and rheologies.
+Instead, they attain a velocity solution and simply advect the free surface with the local velocity.
+The normal mode solution is instructive, however. 
+Each mode has the form of a decay equation with characteristic decay time $\tau_i$.
+The decay equation is the archetypical example of a stiff ordinary differential equation.
+If we were to numerically integrate this in time with a forward Euler method, we would find the 
+time-step criterion for stability \citep[e.g.][]{leveque2007finite} to be
+\begin{equation}
+\Delta t  \le 2 \tau_{\mathrm{min}}
+\label{eq:cfl_euler}
+\end{equation}
+The maximum allowable timestep is limited by the minimum relaxation timescale.
+If a larger timestep than this is taken then those modes will go unstable.
+The modes with the smallest relaxation times are usually those with the largest lengthscales \citep{schubert2001mantle}, 
+and so it will be those which go unstable first, a phenomenon which has been called 
+a ``sloshing'', or ``drunken sailor'' instability \citep{kaus2010stabilization}.
+
+\subsection{Quasi-implicit stabilization}
+\label{sec:kmm}
+
+The relaxation timescales for surface topography tend to be considerably longer than those for 
+convection or tectonic deformation, so the stability requirements for a forward Euler scheme
+are quite onerous.  On the other hand, an implicit time marching scheme requires solving 
+a nonlinear system for the new surface position, or assembling a larger system with surface
+topography unknowns \citep[e.g.][]{kramer2012implicit}.
+
+\citet{kaus2010stabilization} proposed a scheme whereby the body forces on the domain are 
+evaluated on a prediction of the shape of the domain at a later time.
+The weak form of the right-hand-side body forces in the finite element discretization is then
+
+\begin{equation}
+\mathbf{f}_{\mathrm{body}} = \int_{\Omega + \Delta \Omega} \rho \mathbf{w} \cdot \mathbf{g}
+\label{eq:predict}
+\end{equation}
+
+The shape prediction can be approximated by integration of the velocity:
+\begin{equation}
+\Delta \Omega \approx \theta \Delta t \mathbf{u}
+\end{equation}
+where $\theta$ is a free parameter that corresponds to the magnitude of the 
+correction, where zero is no stabilization and one is fully (quasi) implicit.
+
+One can approximately expand the integral in Equation~\eqref{eq:predict} using 
+Reynold's transport theorem to find
+
+\begin{equation}
+\int_{\Omega + \Delta \Omega} \rho  \mathbf{w} \cdot \mathbf{g} \approx
+\int_{\Omega} \rho  \mathbf{w} \cdot \mathbf{g} + \theta \Delta t \int_{\partial \Omega} \rho ( \mathbf{w \cdot g}) (\mathbf{u \cdot n} )
+\label{eq:kmm}
+\end{equation}
+The volume integral is the same as that of the unstabilized problem, but now there is a surface integral.
+The surface integral has the form of a velocity-dependent surface stress pushing down on the 
+domain, and can be thought of as an artificial viscous damping of the surface.
+Since it depends on the velocity, it can no longer go on the right-hand side of the 
+equation, and must be added as a stabilization term in the system matrix.  
+Empirically it has been found to be successful at damping instabilities \citep{kaus2010stabilization, quinquis2011role, crameri2012comparison}.
+
+Indeed, using the formalism in Section~\ref{sec:eigenvalue} we can see the effect of this 
+stabilization term. In the reference state we note that $\mathbf{g} = g \mathbf{n}$, which
+allows us to write the stabilization term as
+\begin{equation}
+\theta \Delta t \int_{\partial \Omega} \rho g ( \mathbf{w \cdot n}) (\mathbf{u \cdot n} )
+\end{equation}
+
+The integral here is precisely the same as that on the right hand side of Equation~\eqref{eq:weak_eigen}, which was discretized as the matrix $M$.
+
+Indeed, if we discretize the Stokes system with the quasi-implicit stabilization term, we find
+a new generalized eigenvalue problem:
+
+\begin{equation}
+\begin{bmatrix}
+A + \theta \Delta t M & B^T \\
+B & 0 \\
+\end{bmatrix}
+\begin{bmatrix}
+\mathbf{u}^n_i \\
+p^n_i
+\end{bmatrix}
+=
+\tau^S_i
+\begin{bmatrix}
+M & 0 \\
+0 & 0
+\end{bmatrix}
+\begin{bmatrix}
+\mathbf{u}^n_i \\
+p^n_i
+\end{bmatrix}
+\label{eq:generalized_eigenvalue}
+\end{equation}
+where $\tau^S_i$ indicate the eigenvalues of the stabilized system.
+This system may be rearranged:
+\begin{equation}
+\begin{bmatrix}
+A & B^T \\
+B & 0 \\
+\end{bmatrix}
+\begin{bmatrix}
+\mathbf{u}^n_i \\
+p^n_i
+\end{bmatrix}
+=
+\left(\tau^S_i - \theta \Delta t \right)
+\begin{bmatrix}
+M & 0 \\
+0 & 0
+\end{bmatrix}
+\begin{bmatrix}
+\mathbf{u}^n_i \\
+p^n_i
+\end{bmatrix}
+\label{eq:generalized_eigenvalue}
+\end{equation}
+
+This is precisely the same generalized eigenvalue problem as Equation~\eqref{eq:generalized_eigenvalue},
+so its eigenvalues must be the same.  This allows us to write the eigenvalues of the stabilized problem in
+terms of those of the unstabilized problem:
+\begin{equation}
+\tau^S_i = \tau_i + \theta \Delta t
+\end{equation}
+
+Essentially, the stabilization term lengthens every relaxation time for the system by an amount $\theta \Delta t$.
+This correspondingly lengthens the maximum stable timestep for the forward Euler method:
+\begin{equation}
+\Delta t  \le \frac{2}{1-\theta} \tau_{\mathrm{min}}
+\label{eq:cfl_euler_stabilized}
+\end{equation}
+note that as $\theta$ goes to one this scheme should become unconditionally stable,
+but nonlinear effects and discretization errors due to finite deformation of the surface could 
+prevent that stability.
+
+\subsection{A novel time-integration scheme}
+
+\subsubsection{Nonstandard finite-differences}
+One downside of the quasi-implicit scheme is that it requires a modification of the system matrix, 
+and as we have seen this shifts its spectrum.
+An alternative would be to construct a time-integration scheme with better stability properties.  
+The reason that forward Euler integration performs so badly with the decay of topography is that 
+at larger timesteps it overshoots its equilibrium position. This overshoot causes it to lurch back 
+to the \emph{other} side of equilibrium, overshooting even more.
+We would like to construct an explicit scheme that accounts for following properties which we know the system has:
+\begin{itemize}
+\item In the absence of forcing, topography always decreases.
+\item Relaxation of small amplitude topography takes the form of exponential decay.
+\item The decay mode with the shortest relaxation time is the least stable.
+\end{itemize}
+
+A general expression for the evolution of a free surface from time $t=0$ to time $t=\Delta t$ is 
+\begin{equation}
+\mathbf{x}(\Delta t) = \mathbf{x}(0) + \int_0^{\Delta t} \mathbf{u}(t) dt
+\label{eq:time_integration}
+\end{equation}
+
+where $\mathbf{x}$ is the location of the free surface.  If we approximate $\mathbf{u}(t) \approx \mathbf{u}(0)$, 
+we of course recover the forward Euler scheme.
+However, we can make another choice based on our knowledge of the system behavior. 
+We can approximate $\mathbf{u}(t)$ as
+\begin{equation}
+\mathbf{u}(t) = \mathbf{u}(0) e^{-t/\tau^*}
+\label{eq:velocity_decay}
+\end{equation}
+where $\tau^*$ is some as-yet undetermined positive constant.
+This form of $\mathbf{u}$ automatically decays in time, and as we shall see, has much better 
+stability properties than forward Euler integration.
+Plugging Equation~\eqref{eq:velocity_decay} into Equation~\eqref{eq:time_integration} and integrating, we find
+
+\begin{equation}
+\mathbf{x}(\Delta t) = \mathbf{x}(0) + \mathbf{u}(0) \tau^* \left(1-e^{-\Delta t/\tau^*} \right)
+\label{eq:nsfd}
+\end{equation}
+
+The quantity $\tau^*(1-e^{-\Delta t / \tau^*})$ acts as a pseudo-timestep for advecting the free-surface.
+Equation~\eqref{eq:nsfd} is what is known as a nonstandard finite-difference model, based on
+constructing unusual discrete models for differential equation integration.
+The theory has been developed in, among others, a series of papers by
+\citet{mickens1994nonstandard, mickens2002nonstandard, mickens2005dynamic}.
+
+\subsubsection{Stability of the scheme}
+The parameter $\tau^*$ sets how quickly $\mathbf{u}$ decays in Equation~\eqref{eq:velocity_decay}, and a good 
+choice is crucial for accuracy and stability. A shorter decay time corresponds to more stabilization,
+but if it becomes too short, the surface velocity will become overdamped. In general, we will want 
+to choose $\tau^*$ so that it is as close as possible to the least stable mode, or $\tau_{\mathrm{max}}$.
+
+To investigate the stability of this scheme we consider a velocity solution with power in 
+a single normal mode of the system $\mathbf{u} = a_i \mathbf{u}_i$.
+This will decay exponentially with relaxation time $\tau_i$, or
+\begin{equation}
+\frac{d a_i} {dt} = - \frac{ a_i }{\tau_i} 
+\end{equation}
+Applying the nonstandard finite difference scheme, we find
+\begin{equation}
+a_i^{(1)} = a_i^{(0)} \left[ 1 - \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \right]
+\label{eq:recursion}
+\end{equation}
+In order for the scheme to be stable, the quantity in brackets must not blow up as it is repeatedly 
+multiplied by itself, or 
+\begin{equation}
+\left| 1 - \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \right| \le 1
+\end{equation}
+
+\ifdetail
+Taking the positive value of the absolute value yields the criterion of 
+\begin{equation}
+- \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \le 0
+\end{equation}
+
+which can be rearranged to find
+\begin{equation}
+e^{-\Delta t/\tau^*} \le 1
+\end{equation}
+
+which is simply a statement that the timestep must be postive.
+
+Taking the negative value of the absolute value yields
+\begin{equation}
+  \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right)  \le 2
+\end{equation}
+
+\begin{equation}
+ 1 - 2 \frac{\tau_i}{\tau^*} \le e^{-\Delta t/\tau^*}
+\end{equation}
+
+if the left hand side is less than zero, then this is true regardless of step size, or
+\begin{equation}
+\tau^* \le 2 \tau_i
+\end{equation}
+
+if the left hand side is between zero and one, the expression is more complicated. 
+Taking the log of both sides:
+
+\begin{equation}
+\Delta t \le -\tau^* \log \left(1 - 2 \frac{\tau_i}{\tau^*} \right)
+\end{equation}
+
+It is convenient to write this in terms of dimensionless times $\Delta t/ \tau_i$ and $\tau^*/\tau_i$.
+
+\begin{equation}
+\frac{\Delta t}{\tau_i} \le -\frac{\tau^*}{\tau_i} \log \left(1 - 2 \frac{\tau_i}{\tau^*} \right)
+\end{equation}
+
+\fi
+
+The stability of this scheme is determined by the choice of $\tau^*$.  
+It has a region of unconditional stability, where
+\begin{equation}
+\tau^* \le 2 \tau_i
+\end{equation}
+Otherwise, the scheme is conditionally stable, with the timestep limited by 
+\begin{equation}
+\Delta t \le -\tau^* \log \left(1 - 2 \frac{\tau_i}{\tau^*} \right)
+\end{equation}
+
+\section{Implementation in \texttt{Aspect} }
 \subsection{Mass conservation}
@@
 \end{itemize}
-\section{Stabilization}
-\subsection{Kaus, M\"ulhaus and May}
-\subsection{Nonstandard finite differences?}
-Is this a good idea?  I think it could be a novel stable way to stabilize free surface
-computations.
-\section{Benchmarking}
 \subsection{Crameri benchmarks}
 more or less good go go
-\subsection{New benchmark?}
-topography over slab with high viscosity?
-3D version of Crameri benchmark?
 \section{Discussion}

### 2015-07-07 figures/stability_region.py
Add figure script

@@ -1,0 +1,27 @@
+import matplotlib.pyplot as plt
+import numpy as np
+
+plt.style.use('ian')
+plt.figure(figsize=(8,4) )
+
+x_nsfd = np.linspace(1.e-10,5, 1000.)
+y_nsfd = np.piecewise( x_nsfd , [x_nsfd <= 2., x_nsfd > 2.], [lambda x : 1.e6, lambda x : -x*np.log(1.-2./x)] )
+x_kmm = np.linspace(1.e-10,1.-1.e-10, 1000.)
+y_kmm = 2./(1.-x_kmm)
+y_fe = np.ones_like(x_nsfd)*2.
+
+plt.subplot(121)
+plt.xlim(0.,5.)
+plt.ylim(0.,10.)
+plt.xlabel(r'$\tau^*/\tau_{\mathrm{max}}$')
+plt.ylabel(r'$\frac{\Delta t}{\tau_{\mathrm{max}}}$', rotation=0)
+plt.fill_between(x_nsfd, 0, y_nsfd, color='red', alpha=0.2)
+plt.fill_between(x_nsfd, 0, y_fe, color='blue', alpha=0.2)
+plt.subplot(122)
+plt.xlim(0.,1.)
+plt.ylim(0.,10.)
+plt.xlabel(r'\theta')
+plt.fill_between(x_kmm, 0, y_kmm, color='green', alpha=0.2)
+plt.fill_between(x_nsfd, 0, y_fe, color='blue', alpha=0.2)
+#plt.show()
+plt.savefig('stability.pdf')

### 2015-07-07 free-surface-paper.tex
More brain dump

@@
 \begin{abstract}
-%% Text of abstract
+Geodynamic simulations are increasingly turning to simulations with a true free surface to 
+investigate questions of dynamic topography, tectonic deformation, gravity perturbations, and
+global mantle convection. However, implementations of free surface boundary conditions 
+have proven challenging from a standpoint of accuracy, robustness, and stability.
 
+In particular, free surfaces tend to suffer from sloshing instabilities, also known as 
+the ``drunken sailor'' instability, which severely limit time step sizes. Several 
+schemes have been proposed in the literature to deal with these instabilities.
+
+Here we analyze the problem of creeping viscous flow with a free surface and discuss the 
+origin of these instabilities. We demonstrate their cause and how existing stabilization 
+schemes work to damp them out.
+
+We also propose a new scheme for removing instabilites from free surface calculations. 
+It does not require modifications to the system matrix, nor additional variables, but is instead
+an explicit scheme based on nonstandard finite differences.  It relies on a single 
+stabilization parameter which may be identified with the smallest relaxation timescale of the
+free surface.
+
+Finally, we discuss the implementation of a free surface in the open source, community based
+mantle convection software \texttt{Aspect}.
 \end{abstract}
@@
 \end{equation}
+The stability region is plotted in Figure \ref{fig:stability}.
+
+As discussed above, a full geodynamic simulation will have a spectrum of relaxation times.
+For complete stability, the parameter $\tau^*$ should be chosen such that every mode is stable.
+In practice, this means that a good choice is 
+\begin{equation}
+\tau^* \approx \tau_{\mathrm{min}}
+\label{eq:tau_choice}
+\end{equation}
+
+\begin{figure}
+\includegraphics[width=0.9\textwidth]{figures/stability.pdf}
+\label{fig:stability}
+\caption{Left: stability region for the nonstandard finite difference scheme (green).
+For $\tau^*\le \tau_{\mathrm{max}}$ it is unconditionally stable.
+Right: stability region for the quasi-implicit scheme (green).
+For $\theta = 1$ it is unconditionally stable.
+In both cases the stability region for the forward Euler scheme is also shown in blue.}
+\end{figure}
+
+\subsubsection{Accuracy of the scheme}
+It is important to note that even though we derived the nonstandard finite difference scheme assuming a decaying exponential
+for $\mathbf{u}$, it is formally a first-order accurate scheme. As such, it is capable of capturing arbitrary motions 
+of the free surface, but with good stability properties.
+Again we may take one of the normal modes as an example, differencing the analytical solution from the nonstandard finite-difference scheme:
+
+\begin{equation}
+\begin{aligned}
+a_i(\Delta t) - a_i^{(1)} &= a_i(0) e^{-\Delta t/\tau_i} - a_i{(0)} \left[ 1 - \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \right] \\
+                          &= a_i{(0)} \left[ e^{-\Delta t/\tau_i} - 1 + \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \right] \\
+\end{aligned}
+\end{equation}
+
+Note that when $\tau_i = \tau^*$ the nonstandard finite difference scheme is exact.
+The exponentials may be expanded to find
+\begin{equation}
+a_i(\Delta t) - a_i^{(1)} = a_i(0) e^{-\Delta t/\tau_i} - a_i{(0)} \left[ \left(\frac{\Delta t}{\tau_i}\right)^2 - \frac{\tau^*}{\tau_i} \left(\frac{\Delta t }{\tau^* }\right)^2 \right] 
+\end{equation}
 

### 2015-07-07 figures/stability_region.py

@@ -13,15 +13,17 @@ y_fe = np.ones_like(x_nsfd)*2.
 plt.subplot(121)
 plt.xlim(0.,5.)
 plt.ylim(0.,10.)
-plt.xlabel(r'$\tau^*/\tau_{\mathrm{max}}$')
-plt.ylabel(r'$\frac{\Delta t}{\tau_{\mathrm{max}}}$', rotation=0)
-plt.fill_between(x_nsfd, 0, y_nsfd, color='red', alpha=0.2)
+plt.xlabel(r'$\tau^*/\tau_{\mathrm{min}}$')
+plt.ylabel(r'$\frac{\Delta t}{\tau_{\mathrm{min}}}$', rotation=0)
+plt.fill_between(x_nsfd, 0, y_nsfd, color='green', alpha=0.2)
 plt.fill_between(x_nsfd, 0, y_fe, color='blue', alpha=0.2)
+plt.title('Non-standard FD')
 plt.subplot(122)
 plt.xlim(0.,1.)
 plt.ylim(0.,10.)
 plt.xlabel(r'\theta')
 plt.fill_between(x_kmm, 0, y_kmm, color='green', alpha=0.2)
 plt.fill_between(x_nsfd, 0, y_fe, color='blue', alpha=0.2)
+plt.title('Quasi-implicit')
 #plt.show()
 plt.savefig('stability.pdf')

### 2015-07-07 free-surface-paper.tex
Stuuuf

@@
 which should be zero when evaluated on a true free surface.
-We will find it analytically useful to evaluate the integrals in Equation~\eqref{eq:weak_stokes} 
+
+Rather than analyzing finite deformation to the free surface, we will make the analytically useful
+approximation of small deformations about the hydrostatic reference surface and analyze their stability.
+We will therefore  evaluate the integrals in Equation~\eqref{eq:weak_stokes} 
 over the hydrostatic reference surface and introduce an auxilliary variable $\zeta$ which 
-represents the topography of the free surface relative to that reference surface.
+represents the (small) topography of the free surface relative to that reference surface.
 The equation for time evolution of the free surface is then
@@
 \end{itemize}
-\subsection{Crameri benchmarks}
-more or less good go go
+
+\section{Results}
+\begin{itemize}
+\item Crameri benchmarks
+\item comparison between stabilization schemes (stability)
+\item comparison between stabilization schemes (accuracy)
+\end{itemize}
 \section{Discussion}

### 2015-07-08 free-surface-paper.tex
Add a bunch of text on implementation

@@
 \label{sec:intro}
-
-Goal: provide for free surface computations in mantle convection simulations that are
-\begin{itemize}
-\item robust, stable, accurate
-\item dimension independent
-\item work with adaptive mesh refinement
-\item parallelized
-\item agnostic to mesh geometry
-\end{itemize}
 
@@
 \section{Implementation in \texttt{Aspect} }
-\subsection{Mass conservation}
-\begin{itemize}
-\item mass conservation requires $\int_\Omega {\bf u \cdot n} = 0$
-\item not all mesh advection schemes respect this
-\item for short simulations, it is not likely to make a big difference
-\item for long simulations there can be significant changes in domain volume
-\item finite element projections do the best job of conservation of mass
-\end{itemize}
+We have implemented the ability to run free surface simulations in the mantle convection software \texttt{Aspect} \citep{kronbichler2012high}. 
+\texttt{Aspect} is designed to be highly flexible and modular, with the ability for user-defined rheologies, geometries, and gravity models. 
+The implementation of the free surface, therefore, needs to be general enough to work for many combinations of these models, including those 
+which may not have been written yet. In particular, it cannot rely on assumptions regarding the shape of the domain.
+
+Furthermore, \texttt{Aspect} allows for computations in both 2D and 3D, so the implementation must be sufficiently dimension
+independent to work in both cases. We implement the free surface within an arbitrary Lagrangian-Eulerian (ALE) formulation
+\citep[e.g][]{fullsack1995arbitrary,donea2004encyclopedia}.
+
+\texttt{Aspect} is a parallelized, distributed memory code with adaptive mesh refinement.
+The free surface implementation works with these features.  We store the mesh vertex positions 
+in a fully distributed vector. This vector is continually updated and redistributed across 
+processes upon mesh adaptation (which is handled by the adaptive octree library \texttt{p4est}).
+We also provide adaptive refinement indicators based on being near to the free surface or 
+when the free surface slope is steep to allow for accurate interface tracking.
+
 \subsection{Remeshing}
-\begin{itemize}
-\item ALE does not specify what happens to interior nodes
-\item primary goal is preserving mesh regularity
-\item we solve a laplace equation for mesh displacements
-\item different boundary conditions for the laplace problem for different velocity boundary conditions
-\end{itemize}
-\subsection{Parallelization}
-\begin{itemize}
-\item mesh vertex positions are stored in a distributed vector
-\item vector is redistributed with p4est on mesh refinement
-\end{itemize}
+In ALE calculations the internal mesh velicity is undetermined.
+In general, one wants to smoothly deform the mesh so as to preserve its regularity, 
+avoiding inverted or otherwise poorly conditioned cells.
+The mesh deformation can be calcluated in many different ways, including algebraic \citep[e.g.][]{thieulot2011fantom} 
+and PDE based approaches.
+
+We choose to implement remeshing based on solving Laplace's equation for the mesh velocity.
+We solve the equation
+\begin{equation}
+\nabla^2 \mathbf{u}_{\mathrm{mesh}} = 0
+\label{eq:laplacian_smoothing}
+\end{equation}
+subject to the boundary conditions
+\begin{equation}
+\begin{aligned}
+&\mathbf{u}_\mathrm{mesh} = \left( \textbf{u} \cdot \textbf{n} \right) \textbf{n} & \qquad & \textrm{on } \partial \Omega_{\textrm{free surface}}, \\
+&\mathbf{u}_\mathrm{mesh} \cdot \textbf{n} = 0 & \qquad & \textrm{on } \partial \Omega_{\textrm{free slip}}, \\
+&\mathbf{u}_\mathrm{mesh} = 0 & \qquad & \textrm{on } \partial \Omega_{\textrm{Dirichlet}}.
+\end{aligned}
+\label{eq:laplacian_bcs}
+\end{equation}
+This scheme has the advantage of working for many different domain geometries and combinations of boundary conditions.
+For moderate mesh deformation, the mesh stays smooth and well conditioned, though it breaks down for large deformations 
+or non-convex domains.
+
+\subsection{Surface advection}
+With a deformable domain there is the danger that small errors in free surface motion can
+result in poor overall mass conservation in time. At worst, the total volume of the mesh can 
+fluctuate significantly over hundreds or thousands of time steps.
+Consistency with the Stokes solution requires 
+\begin{equation}
+\mathbf{u}_{\mathrm{mesh}} \cdot \mathbf{n} = \mathbf{u \cdot n}
+\end{equation}
+
+Unfortunately the normal vectors are not well defined on the mesh vertices, which is 
+where the mesh velocity is defined. One could instead advect the mesh in the direction 
+of the local vertical, or in some weighted average of the cell normals adjacent to a given vertex,
+but we have found that these schemes do not necessarily have excellent mass conservation 
+properties.
+
+A better approach is to perform a finite element projection of the Stokes velocity 
+solution onto the mesh velocity vector. Multiplying the boundary conditions 
+\eqref{eq:laplacian_bcs} by a test function $\mathbf{w}$ and integrating over the free
+surface part of the boundary, we find
+
+\begin{equation}
+\int_{\partial \Omega_{\textrm{free surface}} } \mathbf{w} \cdot \mathbf{u}_\mathrm{mesh} = 
+\int_{\partial \Omega_{\textrm{free surface}} } \left( \mathbf{w \cdot n } \right) \left( \mathbf{u \cdot n} \right)
+\end{equation}
+
+When discretized, this forms a linear system which can be solved for the mesh velocity at the 
+free surface. Of course, a new system to solve is not ideal, but this system, being nonzero 
+over only the free surface, is relatively small and easy to solve.
+
+This weak solution to the boundary conditions~\eqref{eq:laplacian_bcs} is able to borrow
+the accuracy of the Stokes solve for Equation~\eqref{eq:weak_incompressible}, and we have 
+found it to conserve mass more accurately than algebraic techniques for evaluating the mesh-normal velocity.
+Similar results were found by \citet{fullsack1995arbitrary}.
 

### 2015-07-08 free-surface-paper.tex
Enhance caption

@@
 \caption{Left: stability region for the nonstandard finite difference scheme (green).
-For $\tau^*\le \tau_{\mathrm{max}}$ it is unconditionally stable.
+On the x-axis is the value of the stabilization time, in units of the minimum relaxation time $\tau_{\mathrm{min}}$.
+On the y-axis is the value of the timestep, also in units of the minimum relaxation time.
+For $\tau^*\le \tau_{\mathrm{min}}$ it is unconditionally stable.
 Right: stability region for the quasi-implicit scheme (green).
+Again, the y axis is in units of the minumum relaxation time.
+The x-axis shows the value of the stabilization parameter $\theta$.
 For $\theta = 1$ it is unconditionally stable.

### 2015-07-08 free-surface-paper.tex
Add some text on the choice of the relaxation time

@@
 \tau_i (C^{-1}D)\mathbf{y}_i = \mathbf{y}_i
+\label{eq:standard_eigenvalue}
 \end{equation}
@@
 
-As discussed above, a full geodynamic simulation will have a spectrum of relaxation times.
-For complete stability, the parameter $\tau^*$ should be chosen such that every mode is stable.
-In practice, this means that a good choice is 
-\begin{equation}
-\tau^* \approx \tau_{\mathrm{min}}
-\label{eq:tau_choice}
-\end{equation}
-
 \begin{figure}
@@
 \end{equation}
+
+\subsubsection{Choice of $\tau^*$}
+As discussed above, a full geodynamic simulation will have a spectrum of relaxation times.
+For complete stability, the parameter $\tau^*$ should be chosen such that every mode is stable.
+In practice, this means that a good choice is 
+\begin{equation}
+\tau^* \approx \tau_{\mathrm{min}}
+\label{eq:tau_choice}
+\end{equation}
+
+Unfortunately, for many models the minimum relaxation time will not be known beforehand. 
+In order to use the nonstandard finite difference scheme, the value of $\tau^*$ would need 
+to be calculated or estimated first.  There are several possible ways to determine this value:
+
+\begin{itemize}
+\item Direct solution of Equation~\eqref{eq:generalized_eigenvalue}. This can be expensive, although
+only the minimun relaxation time is required. For certain geometries a simple power iteration on 
+the standard eigenvalue problem~\eqref{eq:standard_eigenvalue} could be enough.
+\item Analytical formulae. Several geometry and viscosity model combinations have analytical solutions
+for relaxation spectra. Even if the user's model is not precisely the same (e.g., has some lateral viscosty
+variations), an analytical approximation may be sufficient.
+\item Scaling. In general, we expect the relaxation times to scale with $\tau \sim \frac{\eta}{\rho g L}$,
+where the density, gravity, viscosity, and lengthscale are all representative values.  A small amount
+of experimentation near this value of $\tau$ can find an appropriate relaxation time.
+\item Observation of instabilities.  The mere act of observing a sloshing instability can furnish 
+an estimate of its relaxation time.
+\end{itemize}
 

### 2015-07-09 free-surface-paper.tex
Small adjustments

@@
 have proven challenging from a standpoint of accuracy, robustness, and stability.
-
 In particular, free surfaces tend to suffer from sloshing instabilities, also known as 
@@
 schemes work to damp them out.
-
 We also propose a new scheme for removing instabilites from free surface calculations. 
@@
 
-\section{Stability}
-\subsection{Eigenvalue analysis}
+\section{Eigenvalue analysis}
 \label{sec:eigenvalue}
@@
 
-Rather than analyzing finite deformation to the free surface, we will make the analytically useful
-approximation of small deformations about the hydrostatic reference surface and analyze their stability.
+Rather than analyzing finite deformation to the free surface (a nonlinear problem),
+we will make the analytically useful approximation of small deformations about the hydrostatic 
+reference surface and analyze their stability.
 We will therefore  evaluate the integrals in Equation~\eqref{eq:weak_stokes} 
-over the hydrostatic reference surface and introduce an auxilliary variable $\zeta$ which 
+over the hydrostatic reference surface and introduce a temporary auxilliary variable $\zeta$ which 
 represents the (small) topography of the free surface relative to that reference surface.
@@
 B & 0 \\
-\end{bmatrix} \\
+\end{bmatrix} \quad
 D &= 
@@
 0 & 0
-\end{bmatrix} \\
+\end{bmatrix} \quad
 \mathbf{y}_i &= 
@@
 p^n_i
-\end{bmatrix} \\
+\end{bmatrix} 
 \end{aligned}
@@
 
-\subsection{Time integration of the free surface}
+\section{Time integration of the free surface}
 \label{sec:timestepping}
@@
 
-\subsection{Quasi-implicit stabilization}
+\section{Quasi-implicit stabilization}
 \label{sec:kmm}
@@
 
-\subsection{A novel time-integration scheme}
+\section{A novel time-integration scheme}
 
-\subsubsection{Nonstandard finite-differences}
 One downside of the quasi-implicit scheme is that it requires a modification of the system matrix, 
@@
 \end{itemize}
+
+\subsection{Nonstandard finite-differences}
 
@@
 
-\subsubsection{Stability of the scheme}
+\subsection{Stability of the scheme}
 The parameter $\tau^*$ sets how quickly $\mathbf{u}$ decays in Equation~\eqref{eq:velocity_decay}, and a good 
@@
 
-\begin{figure}
+\begin{figure*}
 \includegraphics[width=0.9\textwidth]{figures/stability.pdf}
@@
 In both cases the stability region for the forward Euler scheme is also shown in blue.}
-\end{figure}
+\end{figure*}
 
-\subsubsection{Accuracy of the scheme}
+\subsection{Accuracy of the scheme}
 It is important to note that even though we derived the nonstandard finite difference scheme assuming a decaying exponential
@@
 
-\subsubsection{Choice of $\tau^*$}
+\subsection{Choice of $\tau^*$}
 As discussed above, a full geodynamic simulation will have a spectrum of relaxation times.
@@
 of experimentation near this value of $\tau$ can find an appropriate relaxation time.
-\item Observation of instabilities.  The mere act of observing a sloshing instability can furnish 
-an estimate of its relaxation time.
+\item Observation of instabilities.  The mere act of observing a sloshing instability in an unstabilized
+problem can furnish an estimate of its relaxation time.
 \end{itemize}
@@
 For moderate mesh deformation, the mesh stays smooth and well conditioned, though it breaks down for large deformations 
-or non-convex domains.
+or on non-convex domains.
 
@@
 of the local vertical, or in some weighted average of the cell normals adjacent to a given vertex,
-but we have found that these schemes do not necessarily have excellent mass conservation 
+but we have found that these schemes do not necessarily have good mass conservation 
 properties.

### 2015-07-09 free-surface-paper.tex
edits

@@
 Indeed, analytical solutions for the normal modes are only known for simple geometries and rheologies.
-Instead, they attain a velocity solution and simply advect the free surface with the local velocity.
+Instead, they obtain a velocity solution and simply advect the free surface with the local velocity.
 The normal mode solution is instructive, however. 
@@
 With a deformable domain there is the danger that small errors in free surface motion can
-result in poor overall mass conservation in time. At worst, the total volume of the mesh can 
+result in poor overall mass conservation in time. In some scenarios, the total volume of the mesh can 
 fluctuate significantly over hundreds or thousands of time steps.

### 2015-07-09 figures/stability_region.py
Fix

@@ -21,7 +21,7 @@ plt.title('Non-standard FD')
 plt.subplot(122)
 plt.xlim(0.,1.)
 plt.ylim(0.,10.)
-plt.xlabel(r'\theta')
+plt.xlabel(r'$\theta$')
 plt.fill_between(x_kmm, 0, y_kmm, color='green', alpha=0.2)
 plt.fill_between(x_nsfd, 0, y_fe, color='blue', alpha=0.2)
 plt.title('Quasi-implicit')

### 2015-07-09 figures/stability_region.py
Add check to plotting script

@@ -1,7 +1,11 @@
 import matplotlib.pyplot as plt
 import numpy as np
 
-plt.style.use('ian')
+try:
+  plt.style.use('ian')
+except ValueError:
+  print "Cannot find the requested stylesheet"
+
 plt.figure(figsize=(8,4) )
 
 x_nsfd = np.linspace(1.e-10,5, 1000.)

### 2015-07-09 figures/stability_region.py
add another error check to .py

@@ -3,6 +3,8 @@ import numpy as np
 
 try:
   plt.style.use('ian')
+except AttributeError:
+  print "old matplotlib version?"
 except ValueError:
   print "Cannot find the requested stylesheet"
 

### 2015-07-09 benchmarks/timestep_convergence/run_models.py
Add stuff for timestep convergence test

@@ -1,0 +1,86 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import os
+from scipy.integrate import trapz
+
+initial_topography = 0.005
+L = 1.0  #domain length
+D = 1.0  # domain depth
+order = 1.0 # order of perturbation
+rho = 1.0 # density
+eta = 1.0 #viscosity
+g = 1.0 #gravity
+
+k_s = 2.0*np.pi/L;  #wavenumber
+k = 2.0*np.pi*order/L;  #wavenumber
+
+tau_0 = 2. * k * eta /rho / g  #reference relaxation time
+tau = tau_0*(D*k + np.sinh(D*k)*np.cosh(D*k))/np.sinh(D*k)/np.sinh(D*k) #relaxation time
+
+tau_0s = 2. * k_s * eta /rho / g  #reference relaxation time
+tau_s = tau_0s*(D*k_s + np.sinh(D*k_s)*np.cosh(D*k_s))/np.sinh(D*k_s)/np.sinh(D*k_s) #relaxation time
+
+def analytic_solution( t ):
+  return initial_topography * np.exp(-t/tau)
+
+def generate_prm( tstep, relax ):
+  prmfile = open("relaxation.prm", "r")
+  outfile = open("tmp.prm", "w")
+  for l in prmfile.readlines():
+    if 'Maximum time step' in l:
+      outfile.write('set Maximum time step = %f\n'%(tstep) )
+    elif 'End time' in l:
+      outfile.write('set End time = %f\n'%(4.0*tau) )
+    elif 'Time between graphical' in l:
+      outfile.write('    set Time between graphical output = %f\n'%(tau/10.) )
+    elif 'Relaxation time' in l:
+      outfile.write('    set Relaxation time = %f\n'%(relax) )
+    elif 'Amplitude' in l:
+      outfile.write('    set Amplitude = %f\n'%(initial_topography) )
+    elif 'Order' in l:
+      outfile.write('    set Order = %f\n'%(order) )
+    else:
+      outfile.write(l)
+  prmfile.close()
+  outfile.close()
+
+
+def calculate_error():
+  data = np.genfromtxt("output/statistics", usecols = (1,9,11) )
+  time = data[:,0]
+  tstep = data[:,1]
+  topo = -data[:,2]
+
+  #compute l2 difference between the output and the analtic solution
+  error = np.sqrt(np.power( (topo-analytic_solution(time))/initial_topography, 2.))
[…]

### 2015-07-09 figures/timestep_convergence.py

@@ -1,0 +1,42 @@
+import matplotlib.pyplot as plt
+import numpy as np
+import numpy.ma
+import itertools
+
+try:
+  plt.style.use('ian')
+except AttributeError:
+  print "old matplotlib version?"
+except ValueError:
+  print "Cannot find the requested stylesheet"
+
+plt.figure(figsize=(8,8) )
+
+data = np.loadtxt("../benchmarks/timestep_convergence/timestep_convergence.txt")
+relaxation_times = set( data[:,0])
+relaxation_times = list( relaxation_times)
+relaxation_times.sort()
+
+
+vals = []
+for r in relaxation_times:
+  tsteps = np.ma.masked_where( data[:,0] != r, data[:,1])
+  errors = np.ma.masked_where( data[:,0] != r, data[:,2])
+  vals.append( (tsteps, errors ) )
+
+marker = itertools.cycle( ('-h', '-v', '-^', '-*', '-D', '-s', '-o') )
+for r, val in zip(relaxation_times, vals):
+  plt.loglog(val[0], val[1], marker.next(), markersize=8, label=r'$\tau^* = %0.1f \; \tau_\mathrm{min}$'%(r))
+
+#plot a slope-1 line to guide the eye
+x = np.logspace(-1.5, -0.5, 10)
+y = 1.e-2*x
+plt.loglog(x,y, 'k--')
+plt.text(0.1, 0.0005, r'$O(\Delta t)$', fontsize=18)
+
+plt.xlabel(r'$\Delta t / \tau_\mathrm{min}$')
+plt.ylabel(r'Error')
+plt.legend(loc='lower right')
+#plt.show()
+plt.savefig('timestep_convergence.pdf')
+

### 2015-07-10 free-surface-paper.tex
Edit
Insert figure
Add figure for convergence of quasi-implicit scheme

@@
 
-Indeed, if we discretize the Stokes system with the quasi-implicit stabilization term, we find
+If we discretize the Stokes system with the quasi-implicit stabilization term, we find
 a new generalized eigenvalue problem:
@@
 \section{Results}
+
+\begin{figure*}
+\includegraphics[width=0.9\textwidth]{figures/timestep_convergence.pdf}
+\caption{Convergence test with timestep size.}
+\label{fig:timestep_convergence}
+\end{figure*}
+
+\begin{figure*}
+\includegraphics[width=0.9\textwidth]{figures/timestep_convergence_qi.pdf}
+\caption{Convergence test with timestep size, quasi-implicit}
+\label{fig:timestep_convergence_qi}
+\end{figure*}
+
 \begin{itemize}

### 2015-07-10 benchmarks/timestep_convergence_qi/run_models.py

@@ -1,0 +1,86 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import os
+from scipy.integrate import trapz
+
+initial_topography = 0.005
+L = 1.0  #domain length
+D = 1.0  # domain depth
+order = 1.0 # order of perturbation
+rho = 1.0 # density
+eta = 1.0 #viscosity
+g = 1.0 #gravity
+
+k_s = 2.0*np.pi/L;  #wavenumber
+k = 2.0*np.pi*order/L;  #wavenumber
+
+tau_0 = 2. * k * eta /rho / g  #reference relaxation time
+tau = tau_0*(D*k + np.sinh(D*k)*np.cosh(D*k))/np.sinh(D*k)/np.sinh(D*k) #relaxation time
+
+tau_0s = 2. * k_s * eta /rho / g  #reference relaxation time
+tau_s = tau_0s*(D*k_s + np.sinh(D*k_s)*np.cosh(D*k_s))/np.sinh(D*k_s)/np.sinh(D*k_s) #relaxation time
+
+def analytic_solution( t ):
+  return initial_topography * np.exp(-t/tau)
+
+def generate_prm( tstep, theta):
+  prmfile = open("relaxation.prm", "r")
+  outfile = open("tmp.prm", "w")
+  for l in prmfile.readlines():
+    if 'Maximum time step' in l:
+      outfile.write('set Maximum time step = %f\n'%(tstep) )
+    elif 'End time' in l:
+      outfile.write('set End time = %f\n'%(4.0*tau) )
+    elif 'Time between graphical' in l:
+      outfile.write('    set Time between graphical output = %f\n'%(tau/10.) )
+    elif 'stabilization theta' in l:
+      outfile.write('    set Free surface stabilization theta = %f\n'%(theta) )
+    elif 'Amplitude' in l:
+      outfile.write('    set Amplitude = %f\n'%(initial_topography) )
+    elif 'Order' in l:
+      outfile.write('    set Order = %f\n'%(order) )
+    else:
+      outfile.write(l)
+  prmfile.close()
+  outfile.close()
+
+
+def calculate_error():
+  data = np.genfromtxt("output/statistics", usecols = (1,9,11) )
+  time = data[:,0]
+  tstep = data[:,1]
+  topo = -data[:,2]
+
+  #compute l2 difference between the output and the analtic solution
+  error = np.sqrt(np.power( (topo-analytic_solution(time))/initial_topography, 2.))
[…]

### 2015-07-10 figures/timestep_convergence_qi.py

@@ -1,0 +1,42 @@
+import matplotlib.pyplot as plt
+import numpy as np
+import numpy.ma
+import itertools
+
+try:
+  plt.style.use('ian')
+except AttributeError:
+  print "old matplotlib version?"
+except ValueError:
+  print "Cannot find the requested stylesheet"
+
+plt.figure(figsize=(8,8) )
+
+data = np.loadtxt("../benchmarks/timestep_convergence_qi/timestep_convergence_qi.txt")
+thetas = set( data[:,0])
+thetas = list( thetas)
+thetas.sort()
+
+
+vals = []
+for theta in thetas:
+  tsteps = np.ma.masked_where( data[:,0] != theta, data[:,1])
+  errors = np.ma.masked_where( data[:,0] != theta, data[:,2])
+  vals.append( (tsteps, errors ) )
+
+marker = itertools.cycle( ('-h', '-v', '-^', '-*', '-D', '-s', '-o') )
+for theta, val in zip(thetas, vals):
+  plt.loglog(val[0], val[1], marker.next(), markersize=8, label=r'$\theta = %0.1f$'%(theta))
+
+#plot a slope-1 line to guide the eye
+x = np.logspace(-1.5, -0.5, 10)
+y = 1.e0*x
+plt.loglog(x,y, 'k--')
+plt.text(0.1, 0.05, r'$O(\Delta t)$', fontsize=18)
+
+plt.xlabel(r'$\Delta t / \tau_\mathrm{min}$')
+plt.ylabel(r'Error')
+plt.legend(loc='lower right')
+#plt.show()
+plt.savefig('timestep_convergence_qi.pdf')
+

### 2015-07-24 free-surface-paper.tex
Add figure

@@
 \label{sec:intro}
+
+Surface topography in simulations of mantle convection and other geodynamic processes is an important observable,
+allowing insights into internal density structure, rheology, and gravity perturbations.
+Historically, most simulations have been performed with free-slip boundary conditions at the surface, 
+with dynamic topography calculated as a postprocessing step.
+
+More recently, the nature of surface boundary conditions have been shown to be important for controlling the 
+dynamics of subduction zone modeling.  \citet{schmeling2008benchmark} performed extensive testing on the effect
+of a free surface on the sinking of a slab.
+
+\section{Governing Equations}
+\label{sec:governing}
 
@@
 \section{Implementation in \texttt{Aspect} }
+\label{sec:implementation}
 We have implemented the ability to run free surface simulations in the mantle convection software \texttt{Aspect} \citep{kronbichler2012high}. 
@@
 
+We demonstrate the convergence of the nonstandard finite difference model by comparison 
+with an analytical solution. Following \citet{kramer2012implicit} we model the relaxation
+of sinusoidal surface topography in a two-dimensional Cartesian box with an isoviscous fluid.
+The setup is shown in
+
+\begin{figure*}
+\includegraphics[width=0.9\textwidth]{figures/benchmark_setup.pdf}
+\caption{Setup for the free surface relaxation benchmark.
+A 2D box with an isoviscous fluid has sinusoidal initial topography, with amplitude $\zeta_0$.
+The box has depth $D$ and length $L$.
+For our tests $\rho = \eta = g = D = L = 1$, and $\zeta_0 = 0.005$.}
+\label{fig:benchmark_setup}
+\end{figure*}
+
 \begin{figure*}

### 2015-07-24 figures/benchmark_setup.py

@@ -1,0 +1,46 @@
+import matplotlib.pyplot as plt
+import numpy as np
+
+fig = plt.figure( figsize=(8,8) )
+ax = fig.add_subplot(111)
+
+ax.axes.get_yaxis().set_major_formatter(plt.NullFormatter())
+ax.axes.get_xaxis().set_major_formatter(plt.NullFormatter())
+ax.tick_params( top='off', bottom='off', left='off', right='off')
+
+amplitude = 0.1
+x = np.linspace(0,1,1000)
+surface = 1.0 + amplitude*np.cos(2.*np.pi*x)
+ref_surface = 1.0*np.ones_like(x)
+
+#plot fluid
+ax.plot(x,surface, color='k')
+ax.plot(x,ref_surface, 'k--')
+plt.fill_between(x, 0, surface, color='gray', alpha=0.2)
+
+
+head_size = 0.02
+
+#Draw arrow for surface perturbation
+ax.arrow(0.5, 1.0, 0.0, -amplitude+head_size, head_width=head_size, head_length=head_size, fc='k', ec='k')
+ax.text(0.45, .95, r'$\zeta_0$', fontsize=20)
+
+#Draw arrow for depth
+ax.arrow(0.05, 0.05, 0.0, 0.9, head_width=head_size, head_length=head_size, fc='k', ec='k')
+ax.arrow(0.05, 0.95, 0.0, -0.9, head_width=head_size, head_length=head_size, fc='k', ec='k')
+ax.text(0.07, .5, r'$D$', fontsize=20)
+
+#Draw arrow for length
+ax.arrow(0.08, 0.05, 0.85, 0.0, head_width=head_size, head_length=head_size, fc='k', ec='k')
+ax.arrow(0.95, 0.05, -0.85, 0.0, head_width=head_size, head_length=head_size, fc='k', ec='k')
+ax.text(0.5, .07, r'$L$', fontsize=20)
+
+#Draw arrow for gravity
+ax.arrow(0.8, 0.5, 0.0, -.15, head_width=head_size, head_length=head_size, fc='k', ec='k')
+ax.text(0.82,0.43, r'${\bf g}$', fontsize=20)
+
+ax.set_xlim(0,1)
+ax.set_ylim(0,1.0 + amplitude)
+
+#plt.show()
+plt.savefig("benchmark_setup.pdf")

### 2015-07-24 free-surface-paper.tex
Add text

@@
 work for other discrete methods.
-We transform the governing equations into the weak form amenable to finite elements by standard procedures \citep[e.g.][]{zienkiewicz1977finite} to get
+We transform the governing equations into the weak form amenable to finite elements via standard operations \citep[e.g.][]{zienkiewicz1977finite} to get
 
@@
 of sinusoidal surface topography in a two-dimensional Cartesian box with an isoviscous fluid.
-The setup is shown in
+The setup is shown in Figure~\ref{fig:benchmark_setup}.
+
+The initial topography is given by
+\begin{equation}
+\zeta(x,0) = \zeta_0 \cos\left( k x \right)
+\end{equation}
+where $k = 2 \pi / L$.
+
+The time evolution of this topography is then
+\begin{equation}
+\zeta(x, t) = \zeta(x,0) e^{-t/\tau}
+\end{equation}
+where the relaxation time $\tau$ is given by
+\begin{equation}
+\tau = \frac{D k + \sinh(D k) \cosh(D k)}{ \sinh^2 (D k) } \frac{2 k \eta}{\rho g}
+\end{equation}
+where $D$ is the layer depth, $\eta$ is the viscosity, $\rho$ is the density, and $g$ is the force of gravity.
+
+This solution is only valid for infinitesimal topography. However, for small
+initial topography $\zeta_0$ it seems to be sufficiently accurate to test convergence orders up 
+to quadratic \citep{kramer2012implicit, furuichi2015implicit}.
+We evaluate error by time-integrating the L2 difference between the numerical and analytical
+solutions at the center of the domain:
+\begin{equation}
+\mathrm{Error} = \frac{1}{4\tau}\int_0^{4 \tau} \lVert \zeta_\mathrm{numeric}(L/2, t) - \zeta_\mathrm{analytic}(L/2, t) \rVert_2
+\end{equation}
 
@@
 \end{figure*}
+
+Figure~\ref{fig:timestep_convergence} shows the convergence of the nonstandard finite difference scheme 
+with timestep.
 
@@
 
-\begin{itemize}
-\item Crameri benchmarks
-\item comparison between stabilization schemes (stability)
-\item comparison between stabilization schemes (accuracy)
-\end{itemize}
 \section{Discussion}

### 2015-07-24 free-surface-paper.tex
Text

@@
 \usepackage{amsmath}
+\usepackage{color, soul}
 
@@
 Surface topography in simulations of mantle convection and other geodynamic processes is an important observable,
-allowing insights into internal density structure, rheology, and gravity perturbations.
+allowing insights into Earth's internal density structure, rheology, and geoid perturbations \citep[e.g.][]{richards1984geoid, hager1985lower, baumann2014constraining}.
 Historically, most simulations have been performed with free-slip boundary conditions at the surface, 
@@
 
+\hl{address pseudo free surface, free surface, sticky air}
+
 More recently, the nature of surface boundary conditions have been shown to be important for controlling the 
-dynamics of subduction zone modeling.  \citet{schmeling2008benchmark} performed extensive testing on the effect
-of a free surface on the sinking of a slab.
+dynamics of subduction zone modeling.  In a benchmark study \citet{schmeling2008benchmark} performed extensive testing on the effect
+of a free surface on the sinking of a slab. They found that the nature of the free surface had a large effect 
+on the dynamics of the slab, affecting both the shape and the timing of sinking. Most of the 
+participating codes in that benchmark used the `sticky air' approximation for the free surface, where a low viscosity,
+low density layer of fluid (termed `air', though its viscosity is much higher) is placed above the free surface, decoupling 
+it from the boundary. Typically a free-slip boundary condition is used above the stiky air.
+
+In particular, \citet{schmeling2008benchmark} found that the specific properties of the sticky air layer 
+controlled the shape and timing of the slab, and the viscosity averaging scheme for areas of transitional
+composition was extremely important, as the subducting slab entrained significant amounts of the sticky air.
+\citet{crameri2012comparison} conducted comparisons between sticky air and true free surface models, 
+demonstrating a range of parameters for sticky air which can mitigate some of the difficulties that it 
+introduces.
+
+A study by \citet{quinquis2011role} found that free surface boundary conditions have a large effect on 
+trench migration in a subduction zone, and \citet{crameri2012free} found that a free surface combined 
+with a weak crust is important for producing one-sided subduction zones.
+
+Both true free surface and sticky air calculations have been subject to an instability which has been 
+variously termed a ``sloshing,'' or ``druknen sailor'' instability \citep{kaus2010stabilization, duretz2011discretization}. 
+This instability, arising from the large density contrast typical at a free surface (compared with the much smaller 
+internal density contrasts), severely limits the maximum stable timestep for free surface computations.
+Frequently, the maximum stable timestep is several orders of magnitude smaller than that for an otherwise 
+equivalent model with free-slip boundary conditions.
+
+Several studies have attempted to alleviate the timestepping requirements imposed by the sloshing 
+instability. Since the most expensive part of geodynamic simulations is typically the Stokes solve,
+most free surface calculations have preferred to use explicit time stepping methods for the free 
+surface. \citet{kaus2010stabilization} proposed a quasi-implicit scheme which modifies the discretized 
+Stokes matrix, giving it better stability properties.  \citet{kramer2012implicit} and \citet{furuichi2015implicit}
+explored methods for solving the transport of the free surface implicitly.
+
+Here we analyze the problem of the sloshing instability.  We demonstrate the nature of the instability, 
+as well as clarifying the effect of quasi-implicit stabilization.
+
+The analysis allows us to propose a new time marching scheme for free surface computations with 
+good stability properties. It is an explicit method which requires no modification to the system matrix,
+which makes it extremely simple to implement.
+
+Finally, we describe the implementation of a free surface in the mantle convection software \texttt{Aspect}.
 
@@
 On the y-axis is the value of the timestep, also in units of the minimum relaxation time.
-For $\tau^*\le \tau_{\mathrm{min}}$ it is unconditionally stable.
+For $\tau^*\le 2 \tau_{\mathrm{min}}$ it is unconditionally stable.
 Right: stability region for the quasi-implicit scheme (green).

### 2015-07-29 benchmarks/tau_sensitivity/run_models.py
Add figure for sensitivity to relaxation parameter

@@ -1,0 +1,85 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import os
+from scipy.integrate import trapz
+
+initial_topography = 0.005
+L = 1.0  #domain length
+D = 1.0  # domain depth
+order = 1.0 # order of perturbation
+rho = 1.0 # density
+eta = 1.0 #viscosity
+g = 1.0 #gravity
+
+k_s = 2.0*np.pi/L;  #wavenumber
+k = 2.0*np.pi*order/L;  #wavenumber
+
+tau_0 = 2. * k * eta /rho / g  #reference relaxation time
+tau = tau_0*(D*k + np.sinh(D*k)*np.cosh(D*k))/np.sinh(D*k)/np.sinh(D*k) #relaxation time
+
+tau_0s = 2. * k_s * eta /rho / g  #reference relaxation time
+tau_s = tau_0s*(D*k_s + np.sinh(D*k_s)*np.cosh(D*k_s))/np.sinh(D*k_s)/np.sinh(D*k_s) #relaxation time
+
+def analytic_solution( t ):
+  return initial_topography * np.exp(-t/tau)
+
+def generate_prm( tstep, relax ):
+  prmfile = open("relaxation.prm", "r")
+  outfile = open("tmp.prm", "w")
+  for l in prmfile.readlines():
+    if 'Maximum time step' in l:
+      outfile.write('set Maximum time step = %f\n'%(tstep) )
+    elif 'End time' in l:
+      outfile.write('set End time = %f\n'%(4.0*tau) )
+    elif 'Time between graphical' in l:
+      outfile.write('    set Time between graphical output = %f\n'%(tau/10.) )
+    elif 'Relaxation time' in l:
+      outfile.write('    set Relaxation time = %f\n'%(relax) )
+    elif 'Amplitude' in l:
+      outfile.write('    set Amplitude = %f\n'%(initial_topography) )
+    elif 'Order' in l:
+      outfile.write('    set Order = %f\n'%(order) )
+    else:
+      outfile.write(l)
+  prmfile.close()
+  outfile.close()
+
+
+def calculate_error():
+  data = np.genfromtxt("output/statistics", usecols = (1,9,11) )
+  time = data[:,0]
+  tstep = data[:,1]
+  topo = -data[:,2]
+
+  #compute l2 difference between the output and the analtic solution
+  error = np.sqrt(np.power( (topo-analytic_solution(time))/initial_topography, 2.))
[…]

### 2015-07-29 figures/tau_sensitivity.py

@@ -1,0 +1,26 @@
+import matplotlib.pyplot as plt
+import numpy as np
+import numpy.ma
+import itertools
+
+try:
+  plt.style.use('ian')
+except AttributeError:
+  print "old matplotlib version?"
+except ValueError:
+  print "Cannot find the requested stylesheet"
+
+plt.figure(figsize=(8,8) )
+
+data = np.loadtxt("../benchmarks/tau_sensitivity/tau_sensitivity.txt")
+relaxation_times =  data[:,0]
+errors = data[:,2]
+
+plt.semilogy(relaxation_times, errors, '-x', markersize=8)
+
+
+plt.xlabel(r'$\tau^* / \tau_\mathrm{min}$')
+plt.ylabel(r'Error')
+#plt.show()
+plt.savefig('tau_sensitivity.pdf')
+

### 2015-07-30 free-surface-paper.tex
Add figure for sensitivity to relaxation parameter
Address some comments

@@
 \begin{abstract}
-Geodynamic simulations are increasingly turning to simulations with a true free surface to 
+Geodynamic simulations increasingly rely on simulations with a true free surface to 
 investigate questions of dynamic topography, tectonic deformation, gravity perturbations, and
@@
 
-In particular, \citet{schmeling2008benchmark} found that the specific properties of the sticky air layer 
+\citet{schmeling2008benchmark} found that the specific properties of the sticky air layer 
 controlled the shape and timing of the slab, and the viscosity averaging scheme for areas of transitional
@@
 represents the (small) topography of the free surface relative to that reference surface.
-The equation for time evolution of the free surface is then
+The Lagrangian equation for time evolution of the free surface is then
 \begin{equation}
@@
 
-The relaxation timescales for surface topography tend to be considerably longer than those for 
+The relaxation timescales for surface topography tend to be considerably shorter than those for 
 convection or tectonic deformation, so the stability requirements for a forward Euler scheme
@@
 \end{equation}
-
 where $\mathbf{x}$ is the location of the free surface.  If we approximate $\mathbf{u}(t) \approx \mathbf{u}(0)$, 
@@
 stability properties than forward Euler integration.
-Plugging Equation~\eqref{eq:velocity_decay} into Equation~\eqref{eq:time_integration} and integrating, we find
+Using Equation~\eqref{eq:velocity_decay} in Equation~\eqref{eq:time_integration} and integrating, we find
 
@@
 
-To investigate the stability of this scheme we consider a velocity solution with power in 
+To investigate the stability of this scheme we consider a velocity solution comprised of 
 a single normal mode of the system $\mathbf{u} = a_i \mathbf{u}_i$.
@@
 On the y-axis is the value of the timestep, also in units of the minimum relaxation time.
-For $\tau^*\le 2 \tau_{\mathrm{min}}$ it is unconditionally stable.
+For $\tau^*\le 2 \tau_{\mathrm{min}}$ the nonstandard finite difference scheme is unconditionally stable.
 Right: stability region for the quasi-implicit scheme (green).
@@
 The x-axis shows the value of the stabilization parameter $\theta$.
-For $\theta = 1$ it is unconditionally stable.
+For $\theta = 1$ the quasi-implicit scheme is unconditionally stable.
 In both cases the stability region for the forward Euler scheme is also shown in blue.}
@@
 \begin{equation}
-\mathrm{Error} = \frac{1}{4\tau}\int_0^{4 \tau} \lVert \zeta_\mathrm{numeric}(L/2, t) - \zeta_\mathrm{analytic}(L/2, t) \rVert_2
+\mathrm{Error} = \frac{1}{4\tau}\int_0^{4 \tau} \lVert \zeta_\mathrm{numeric}(L/2, t) - \zeta_\mathrm{analytic}(L/2, t) \rVert_2 dt
 \end{equation}
@@
 \label{fig:timestep_convergence_qi}
+\end{figure*}
+
+\begin{figure*}
+\includegraphics[width=0.9\textwidth]{figures/tau_sensitivity.pdf}
+\caption{Sensitivity to $\tau^*$.}
+\label{fig:tau_sensitivity}
 \end{figure*}

### 2015-07-30 figures/stability_region.py
Adjust figures

@@ -16,20 +16,21 @@ x_kmm = np.linspace(1.e-10,1.-1.e-10, 1000.)
 y_kmm = 2./(1.-x_kmm)
 y_fe = np.ones_like(x_nsfd)*2.
 
+fontsize = 16
 plt.subplot(121)
 plt.xlim(0.,5.)
 plt.ylim(0.,10.)
-plt.xlabel(r'$\tau^*/\tau_{\mathrm{min}}$')
-plt.ylabel(r'$\frac{\Delta t}{\tau_{\mathrm{min}}}$', rotation=0)
+plt.xlabel(r'$\tau^*/\tau_{\mathrm{min}}$', fontsize=fontsize)
+plt.ylabel(r'$\frac{\Delta t}{\tau_{\mathrm{min}}}$', rotation=0, fontsize=fontsize)
 plt.fill_between(x_nsfd, 0, y_nsfd, color='green', alpha=0.2)
 plt.fill_between(x_nsfd, 0, y_fe, color='blue', alpha=0.2)
-plt.title('Non-standard FD')
+plt.title('Non-standard FD', fontsize=fontsize)
 plt.subplot(122)
 plt.xlim(0.,1.)
 plt.ylim(0.,10.)
-plt.xlabel(r'$\theta$')
+plt.xlabel(r'$\theta$', fontsize=fontsize)
 plt.fill_between(x_kmm, 0, y_kmm, color='green', alpha=0.2)
 plt.fill_between(x_nsfd, 0, y_fe, color='blue', alpha=0.2)
-plt.title('Quasi-implicit')
+plt.title('Quasi-implicit', fontsize=fontsize)
 #plt.show()
-plt.savefig('stability.pdf')
+plt.savefig('stability_region.pdf')

### 2015-07-30 figures/tau_sensitivity.py

@@ -10,13 +10,13 @@ except AttributeError:
 except ValueError:
   print "Cannot find the requested stylesheet"
 
-plt.figure(figsize=(8,8) )
+plt.figure(figsize=(4,4) )
 
 data = np.loadtxt("../benchmarks/tau_sensitivity/tau_sensitivity.txt")
 relaxation_times =  data[:,0]
 errors = data[:,2]
 
-plt.semilogy(relaxation_times, errors, '-x', markersize=8)
+plt.semilogy(relaxation_times, errors, '-o', markersize=4)
 
 
 plt.xlabel(r'$\tau^* / \tau_\mathrm{min}$')

### 2015-07-30 figures/timestep_convergence.py

@@ -10,7 +10,10 @@ except AttributeError:
 except ValueError:
   print "Cannot find the requested stylesheet"
 
-plt.figure(figsize=(8,8) )
+plt.figure(figsize=(8,4) )
+
+
+#Plot convergence of the NSFD scheme
 
 data = np.loadtxt("../benchmarks/timestep_convergence/timestep_convergence.txt")
 relaxation_times = set( data[:,0])
@@ -24,19 +27,59 @@ for r in relaxation_times:
   errors = np.ma.masked_where( data[:,0] != r, data[:,2])
   vals.append( (tsteps, errors ) )
 
+plt.subplot(121)
+
 marker = itertools.cycle( ('-h', '-v', '-^', '-*', '-D', '-s', '-o') )
 for r, val in zip(relaxation_times, vals):
-  plt.loglog(val[0], val[1], marker.next(), markersize=8, label=r'$\tau^* = %0.1f \; \tau_\mathrm{min}$'%(r))
+  plt.loglog(val[0], val[1], marker.next(), markersize=4, label=r'$\tau^* = %0.1f \; \tau_\mathrm{min}$'%(r))
 
 #plot a slope-1 line to guide the eye
 x = np.logspace(-1.5, -0.5, 10)
 y = 1.e-2*x
 plt.loglog(x,y, 'k--')
-plt.text(0.1, 0.0005, r'$O(\Delta t)$', fontsize=18)
+plt.text(0.1, 0.0005, r'$O(\Delta t)$', fontsize=9)
 
 plt.xlabel(r'$\Delta t / \tau_\mathrm{min}$')
 plt.ylabel(r'Error')
-plt.legend(loc='lower right')
+plt.legend(loc='lower right', fontsize=8)
+plt.title('Non-standard FD')
+
+
+#Plot convergence of the QI scheme
+
+data = np.loadtxt("../benchmarks/timestep_convergence_qi/timestep_convergence_qi.txt")
+thetas = set( data[:,0])
+thetas = list( thetas)
+thetas.sort()
+
+plt.subplot(122)
+
+vals = []
+for theta in thetas:
+  tsteps = np.ma.masked_where( data[:,0] != theta, data[:,1])
+  errors = np.ma.masked_where( data[:,0] != theta, data[:,2])
+  vals.append( (tsteps, errors ) )
+
[…]

### 2015-07-30 figures/timestep_convergence_qi.py

@@ -1,42 +1,0 @@
-import matplotlib.pyplot as plt
-import numpy as np
-import numpy.ma
-import itertools
-
-try:
-  plt.style.use('ian')
-except AttributeError:
-  print "old matplotlib version?"
-except ValueError:
-  print "Cannot find the requested stylesheet"
-
-plt.figure(figsize=(8,8) )
-
-data = np.loadtxt("../benchmarks/timestep_convergence_qi/timestep_convergence_qi.txt")
-thetas = set( data[:,0])
-thetas = list( thetas)
-thetas.sort()
-
-
-vals = []
-for theta in thetas:
-  tsteps = np.ma.masked_where( data[:,0] != theta, data[:,1])
-  errors = np.ma.masked_where( data[:,0] != theta, data[:,2])
-  vals.append( (tsteps, errors ) )
-
-marker = itertools.cycle( ('-h', '-v', '-^', '-*', '-D', '-s', '-o') )
-for theta, val in zip(thetas, vals):
-  plt.loglog(val[0], val[1], marker.next(), markersize=8, label=r'$\theta = %0.1f$'%(theta))
-
-#plot a slope-1 line to guide the eye
-x = np.logspace(-1.5, -0.5, 10)
-y = 1.e0*x
-plt.loglog(x,y, 'k--')
-plt.text(0.1, 0.05, r'$O(\Delta t)$', fontsize=18)
-
-plt.xlabel(r'$\Delta t / \tau_\mathrm{min}$')
-plt.ylabel(r'Error')
-plt.legend(loc='lower right')
-#plt.show()
-plt.savefig('timestep_convergence_qi.pdf')
-

### 2015-07-30 free-surface-paper.tex
Adjust figures
Reorganize things, make reference state clear

@@
 as well as clarifying the effect of quasi-implicit stabilization.
-
 The analysis allows us to propose a new time marching scheme for free surface computations with 
@@
 
+We begin with the incompressible momentum conservation equations for creeping incompressible flow:
+\begin{equation}
+\begin{aligned}
+\nabla \cdot \mathbf{T} &= \rho \mathbf{g} \\
+\nabla \cdot \mathbf{u} &= 0
+\end{aligned}
+\label{eq:momentum_mass_conservation}
+\end{equation}
+where $\mathbf{u}$ is the fluid velocity, $\rho$ is the fluid density, and $\mathbf{g}$ is the force due to gravity.
+$\mathbf{T}$ is the stress tensor for a Newtonian fluid, defined by
+\begin{equation}
+\mathbf{T} = 2 \eta \varepsilon(\mathbf{u}) - p \mathbf{I}
+\label{eq:stress_tensor}
+\end{equation}
+where $\eta$ is the viscosity and $\varepsilon(\mathbf{u}) = \frac{1}{2}(\nabla \mathbf{u} + (\nabla \mathbf{u} )^T )$ is the strain-rate tensor.
+Substituting the stress tensor into Equation~\eqref{eq:momentum_mass_conservation} gets the familiar form of the Stokes equation:
+\begin{equation}
+\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p = \rho \mathbf{g}
+\label{eq:stokes}
+\end{equation}
+
+For the purposes of this analysis it is useful to define a hydrostatic reference state where the 
+velocity $\mathbf{u}$ is zero:
+\begin{equation}
+- \nabla p_0 = \rho_0 \mathbf{g}
+\label{eq:hydrostatic_stokes}
+\end{equation}
+where $p_0$ is the reference hydrostatic pressure and $\rho_0$ is a reference density profile (which may vary with depth).
+The total pressure and density can then be decomposed in to variations about their 
+reference values: $\rho = \rho_0 + \rho^\prime$, $p = p_0 + p^\prime$.
+
+Furthermore, since we are interested in solving these equations for a domain with a free surface, 
+we need an equation to track that surface.
+The Lagrangian equation for time evolution of the free surface can be written
+\begin{equation}
+\frac{d \zeta}{dt} = \mathbf{u \cdot n} \quad \textrm{on  }  \partial \Omega_\textrm{free surface}
+\label{eq:surface_evolution}
+\end{equation}
+where $\zeta$ is the surface coordinate of the free surface, and $\partial \Omega_\textrm{free surface}$ is the 
+part of the domain boundary corresponding to the free surface.
+
 \section{Eigenvalue analysis}
@@
 
-We begin with the homogeneous Stokes equations for incompressible creeping flow:
+In order to better understand the time evolution of Equations~\eqref{eq:momentum_mass_conservation}-\eqref{eq:surface_evolution} 
+we will consider the eigenvalues of a simplified homogeneous system. By setting $\rho = \rho_0$, we obtain the system
 \begin{equation}
-\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p = 0
-\label{eq:stokes}
+\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p^\prime = 0
+\label{eq:homogeneous_stokes}
 \end{equation}
@@
 \end{equation}
-where $\mathbf{u}$ is the velocity, $p$ is the pressure, $\eta$ is the viscosity, and 
-$\varepsilon(\mathbf{u}) = \frac{1}{2}(\nabla \mathbf{u} + (\nabla \mathbf{u} )^T )$ is the strain-rate tensor.
 
@@
 \begin{equation}
--\int_\Omega 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u} ) + \int_\Omega p \nabla \cdot \mathbf{w} 
-+ \int_{\partial \Omega_N} \mathbf{w} \cdot \mathbf{T} \cdot \mathbf{n} = 0 
+-\int_{\Omega(t)} 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u} ) + \int_{\Omega(t)} p^\prime \nabla \cdot \mathbf{w} 
++ \int_{\partial \Omega_N(t)} \mathbf{w} \cdot \mathbf{T} \cdot \mathbf{n} = 0 
 \label{eq:weak_stokes}
@@
 \begin{equation}
-\int_\Omega q \nabla \cdot \mathbf{u} = 0
+\int_{\Omega(t)} q \nabla \cdot \mathbf{u} = 0
 \label{eq:weak_incompressible}
@@
 $\Omega$ and $\partial \Omega_N$ are over the volume of the domain and the part of the boundary with stress boundary conditions, respectively.
-$\mathbf{T}$ is the stress tensor, defined by
-\begin{equation}
-\mathbf{T} = 2 \eta \varepsilon(\mathbf{u}) - p \mathbf{I}
-\end{equation}
+Note that since the free surface can move, the shape of the domain is a function of time.
+It is most convenient to integrate over the domain at the current timestep, which corresponds to an explicit scheme.
+However, as we will see, the implicit scheme of integrating over the (unknown) domain at the next timestep
+is much more stable, at the cost of making the problem nonlinear \citep{furuichi2015implicit}.
+
 The integral over the surface in Equation~\eqref{eq:weak_stokes} accounts for boundary stresses, 
 which should be zero when evaluated on a true free surface.
-
 Rather than analyzing finite deformation to the free surface (a nonlinear problem),
@@
 represents the (small) topography of the free surface relative to that reference surface.
-The Lagrangian equation for time evolution of the free surface is then
-\begin{equation}
-\frac{d \zeta}{dt} = \mathbf{u \cdot n}
-\label{eq:surface_evolution}
-\end{equation}
-where $\mathbf{u}$ is evaluated at the free surface.
+On the reference surface the gravity vector is opposite the direction of the surface normal $\mathbf{g} = -g \mathbf{n}$.
 The stress on this reference surface can be approximated by the hydrostatic pressure
 \begin{equation}
-\mathbf{T} \approx \rho g \zeta \mathbf{I}
+\mathbf{T} \approx \rho_0 g \zeta \mathbf{I}
 \label{eq:hydrostatic}
 \end{equation}
-where $\rho$ is the density and $g$ is gravity. Equation~\eqref{eq:weak_stokes} then becomes
+Equation~\eqref{eq:weak_stokes} then becomes
 
 \begin{equation}
--\int_\Omega 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u} ) + \int_\Omega p \nabla \cdot \mathbf{w} 
-+ \int_{\partial \Omega_N} \rho g \zeta  \mathbf{w} \cdot \mathbf{n} = 0 
+-\int_{\Omega(t)} 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u} ) + \int_{\Omega(t)} p^\prime \nabla \cdot \mathbf{w} 
++ \int_{\partial \Omega_N(t)} \rho_0 g \zeta  \mathbf{w} \cdot \mathbf{n} = 0 
 \end{equation}
@@
 is the relaxation of topography with a characteristic relaxation time.  
-We denote the normal modes by $\left[ \mathbf{u}_i, p_i, \zeta_i \right]^T$, with
+We denote the normal modes by $\left[ \mathbf{u}_i, p^\prime_i, \zeta_i \right]^T$, with
 relaxation times $\tau_i$, where the subscript corresponds to the $i$th normal mode.
@@
 \begin{equation}
--\int_\Omega 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u}_i ) + \int_\Omega p_i \nabla \cdot \mathbf{w} 
-= \tau_i \int_{\partial \Omega_N} \rho g (\mathbf{u}_i \cdot \mathbf{n} ) (\mathbf{w} \cdot \mathbf{n})
+-\int_{\Omega(t)} 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u}_i ) + \int_{\Omega(t)} p^\prime_i \nabla \cdot \mathbf{w} 
+= \tau_i \int_{\partial \Omega_N(t)} \rho_0 g (\mathbf{u}_i \cdot \mathbf{n} ) (\mathbf{w} \cdot \mathbf{n})
 \label{eq:weak_eigen}
@@
 \end{equation}
-where $\mathbf{u}^n_i$ and $p^n_i$ are finite-dimensional representations of $\mathbf{u}_i$ and $p_i$,
+where $\mathbf{u}^n_i$ and $p^n_i$ are finite-dimensional representations of $\mathbf{u}_i$ and $p^\prime_i$,
 and $M$ is the discretization of the bilinear form on the right-hand side of Equation~\eqref{eq:weak_eigen}.
@@
 equation, and must be added as a stabilization term in the system matrix.  
-Empirically it has been found to be successful at damping instabilities \citep{kaus2010stabilization, quinquis2011role, crameri2012comparison}.
+Empirically it has been found to be successful at damping instabilities \citep{kaus2010stabilization, quinquis2011role, duretz2011discretization}.
 
 Indeed, using the formalism in Section~\ref{sec:eigenvalue} we can see the effect of this 
-stabilization term. In the reference state we note that $\mathbf{g} = g \mathbf{n}$, which
+stabilization term. On the reference surface we note that $\mathbf{g} = -g \mathbf{n}$, which
 allows us to write the stabilization term as
 \begin{equation}
-\theta \Delta t \int_{\partial \Omega} \rho g ( \mathbf{w \cdot n}) (\mathbf{u \cdot n} )
+-\theta \Delta t \int_{\partial \Omega} \rho g ( \mathbf{w \cdot n}) (\mathbf{u \cdot n} )
 \end{equation}
-
 The integral here is precisely the same as that on the right hand side of Equation~\eqref{eq:weak_eigen}, which was discretized as the matrix $M$.
@@
 but if it becomes too short, the surface velocity will become overdamped. In general, we will want 
-to choose $\tau^*$ so that it is as close as possible to the least stable mode, or $\tau_{\mathrm{max}}$.
+to choose $\tau^*$ so that it is as close as possible to the relaxation time of the least stable mode, or $\tau_{\mathrm{min}}$.
 
@@
 \end{equation}
-The stability region is plotted in Figure \ref{fig:stability}.
+The stability region is plotted in Figure \ref{fig:stability_region}.
 
 \begin{figure*}
-\includegraphics[width=0.9\textwidth]{figures/stability.pdf}
-\label{fig:stability}
+\includegraphics[width=0.9\textwidth]{figures/stability_region.pdf}
+\label{fig:stability_region}
 \caption{Left: stability region for the nonstandard finite difference scheme (green).
@@
 \begin{figure*}
-\includegraphics[width=0.9\textwidth]{figures/timestep_convergence_qi.pdf}
-\caption{Convergence test with timestep size, quasi-implicit}
-\label{fig:timestep_convergence_qi}
-\end{figure*}
-
-\begin{figure*}
 \includegraphics[width=0.9\textwidth]{figures/tau_sensitivity.pdf}

### 2015-07-30 figures/stability_region.py

@@ -24,13 +24,13 @@ plt.xlabel(r'$\tau^*/\tau_{\mathrm{min}}$', fontsize=fontsize)
 plt.ylabel(r'$\frac{\Delta t}{\tau_{\mathrm{min}}}$', rotation=0, fontsize=fontsize)
 plt.fill_between(x_nsfd, 0, y_nsfd, color='green', alpha=0.2)
 plt.fill_between(x_nsfd, 0, y_fe, color='blue', alpha=0.2)
-plt.title('Non-standard FD', fontsize=fontsize)
+plt.title('(a) Non-standard FD', fontsize=fontsize)
 plt.subplot(122)
 plt.xlim(0.,1.)
 plt.ylim(0.,10.)
 plt.xlabel(r'$\theta$', fontsize=fontsize)
 plt.fill_between(x_kmm, 0, y_kmm, color='green', alpha=0.2)
 plt.fill_between(x_nsfd, 0, y_fe, color='blue', alpha=0.2)
-plt.title('Quasi-implicit', fontsize=fontsize)
+plt.title('(b) Quasi-implicit', fontsize=fontsize)
 #plt.show()
 plt.savefig('stability_region.pdf')

### 2015-07-30 figures/timestep_convergence.py

@@ -42,7 +42,7 @@ plt.text(0.1, 0.0005, r'$O(\Delta t)$', fontsize=9)
 plt.xlabel(r'$\Delta t / \tau_\mathrm{min}$')
 plt.ylabel(r'Error')
 plt.legend(loc='lower right', fontsize=8)
-plt.title('Non-standard FD')
+plt.title('(a) Non-standard FD')
 
 
 #Plot convergence of the QI scheme
@@ -76,7 +76,7 @@ plt.text(0.3, 0.0005, r'$O(\Delta t^2)$', fontsize=9)
 
 plt.xlabel(r'$\Delta t / \tau_\mathrm{min}$')
 plt.legend(loc='lower right', fontsize=8)
-plt.title('Quasi-implicit')
+plt.title('(b) Quasi-implicit')
 
 
 

### 2015-07-31 free-surface-paper.tex
Text

@@
 Historically, most simulations have been performed with free-slip boundary conditions at the surface, 
-with dynamic topography calculated as a postprocessing step.
+with dynamic topography calculated as a postprocessing step \citep[e.g.][]{zhong2000role}.
 
@@
 
-A general expression for the evolution of a free surface from time $t=0$ to time $t=\Delta t$ is 
+A general expression for the evolution of a free surface from time $t^n$ to time $t^{n+1}$ is 
 \begin{equation}
-\mathbf{x}(\Delta t) = \mathbf{x}(0) + \int_0^{\Delta t} \mathbf{u}(t) dt
+\mathbf{x}(t^{n+1}) = \mathbf{x}(t^n) + \int_{t^{n}}^{t^{n+1}} \mathbf{u}(t) dt
 \label{eq:time_integration}
 \end{equation}
-where $\mathbf{x}$ is the location of the free surface.  If we approximate $\mathbf{u}(t) \approx \mathbf{u}(0)$, 
+where $\mathbf{x}$ is the location of the free surface.  If we approximate $\mathbf{u}(t) \approx \mathbf{u}(t^{n})$, 
 we of course recover the forward Euler scheme.
@@
 \begin{equation}
-\mathbf{u}(t) = \mathbf{u}(0) e^{-t/\tau^*}
+\mathbf{u}(t) = \mathbf{u}(t^n) e^{-(t-t^n)/\tau^*}
 \label{eq:velocity_decay}
@@
 \begin{equation}
-\mathbf{x}(\Delta t) = \mathbf{x}(0) + \mathbf{u}(0) \tau^* \left(1-e^{-\Delta t/\tau^*} \right)
+\mathbf{x}(t^{n+1}) = \mathbf{x}(t^n) + \mathbf{u}(t^{n}) \tau^* \left(1-e^{-\Delta t/\tau^*} \right)
 \label{eq:nsfd}
@@
 \begin{equation}
-a_i^{(1)} = a_i^{(0)} \left[ 1 - \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \right]
+a_i^{n} = a_i^{n+1} \left[ 1 - \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \right]
 \label{eq:recursion}
 \end{equation}
-In order for the scheme to be stable, the quantity in brackets must not blow up as it is repeatedly 
-multiplied by itself, or 
+where superscripts indicate the timestep number.
+In order for the scheme to be stable, the quantity in brackets must not blow up as it is repeatedly multiplied by itself, or 
 \begin{equation}
@@
 \label{fig:stability_region}
-\caption{Left: stability region for the nonstandard finite difference scheme (green).
+\caption{(a) Stability region for the nonstandard finite difference scheme (green).
 On the x-axis is the value of the stabilization time, in units of the minimum relaxation time $\tau_{\mathrm{min}}$.
@@
 For $\tau^*\le 2 \tau_{\mathrm{min}}$ the nonstandard finite difference scheme is unconditionally stable.
-Right: stability region for the quasi-implicit scheme (green).
+(b) Stability region for the quasi-implicit scheme (green).
 Again, the y axis is in units of the minumum relaxation time.
@@
 
-\subsection{Accuracy of the scheme}
+\subsection{Accuracy and asymptotics of the scheme}
 It is important to note that even though we derived the nonstandard finite difference scheme assuming a decaying exponential
 for $\mathbf{u}$, it is formally a first-order accurate scheme. As such, it is capable of capturing arbitrary motions 
-of the free surface, but with good stability properties.
-Again we may take one of the normal modes as an example, differencing the analytical solution from the nonstandard finite-difference scheme:
+of the free surface, but with better stability properties than forward Euler schemes.
+It is helpful to take a closer look at the pseudo-timestep $\tau^*(1-e^{-\Delta t/\tau^*})$.
+
+Again we may take one of the normal modes as an example, differencing the nonstandard finite-difference scheme from the analytical solution:
 
@@
 \begin{aligned}
-a_i(\Delta t) - a_i^{(1)} &= a_i(0) e^{-\Delta t/\tau_i} - a_i{(0)} \left[ 1 - \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \right] \\
+a_i(\Delta t) - a_i^{1} &= a_i(0) e^{-\Delta t/\tau_i} - a_i{(0)} \left[ 1 - \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \right] \\
                           &= a_i{(0)} \left[ e^{-\Delta t/\tau_i} - 1 + \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \right] \\
@@
 \begin{equation}
-a_i(\Delta t) - a_i^{(1)} = a_i(0) e^{-\Delta t/\tau_i} - a_i{(0)} \left[ \left(\frac{\Delta t}{\tau_i}\right)^2 - \frac{\tau^*}{\tau_i} \left(\frac{\Delta t }{\tau^* }\right)^2 \right] 
+a_i(\Delta t) - a_i^{1} = - a_i{(0)} \left[ \left(\frac{\Delta t}{\tau_i}\right)^2 - \frac{\tau^*}{\tau_i} \left(\frac{\Delta t }{\tau^* }\right)^2 \right] 
 \end{equation}
@@
 \includegraphics[width=0.9\textwidth]{figures/timestep_convergence.pdf}
-\caption{Convergence test with timestep size.}
+\caption{Convergence tests for the benchmark shown in Figure~\ref{fig:benchmark_setup}.
+(a) Convergence test with timestep size for the nonstandard finite difference scheme.
+Comparison with the slope-one line confirms that it is first-order in time.
+In the case that the stabilization relaxation time $\tau^*$ is equal to the analytic relaxation time the error plummets, as the time integration scheme becomes exact \citep{mickens2002nonstandard}.
+(b) Convergence test with timestep size for the quasi-implicit scheme.
+For $\theta = 1$ and $\theta = 0$ the scheme is first-order accurate (though the latter is just an unstabilized forward-Euler scheme).
+For $\theta=0.5$ the scheme appears second order accurate on this benchmark.}
 \label{fig:timestep_convergence}
@@
 \includegraphics[width=0.9\textwidth]{figures/tau_sensitivity.pdf}
-\caption{Sensitivity to $\tau^*$.}
+\caption{Sensitivity to $\tau^*$.
+As the relaxation parameter $\tau^*$ approaches the relaxation time of the benchmark case the error reduces.
+The sharp cusp at $\tau^* = \tau_\mathrm{min}$ corresponds to the essentially spectral accuracy of the time marching scheme for that case.}
 \label{fig:tau_sensitivity}

### 2015-07-31 free-surface-paper.tex
more text

@@
 
-\hl{address pseudo free surface, free surface, sticky air}
+There have been several approaches to treating real free surfaces in geodynamic simulations.
+\citet{zhong1996free} introduced a pseudo-free-surface formulation, where the free surface coordinate was
+treated as an extra variable that was integrated in time. In this approximation, the free surface 
+is approximated on the undeformed Eulerian grid by applying pressure boundary conditions on the reference surface 
+corresponding to the hydrostatic pressure perturbation predicted by the surface topography.
 
-More recently, the nature of surface boundary conditions have been shown to be important for controlling the 
+A large number of studies have approximated free surfaces in the interior of the domain by using the 
+`sticky air' approximation. In this approximation there is a low-viscosity, low-density layer in the fluid 
+(termed `air', though its viscosity is much higher) above the free surface, effectively decoupling it from the boundary. 
+Typically a free-slip boundary condition is used above the sticky air.
+
+Finally, one can use a true free surface, where a stress-free boundary condition is applied on the boundary 
+of the domain. In this case, there can be flow in an out of the domain, so the boundary must move in time.
+A true free surface has mathematical elegance in that the boundary condition of the domain more closely 
+matches the boundary conditions which one is trying to model, but it typically requires a deformable 
+domain with frequent remeshing to avoid ill-conditioned cells.
+
+Recently, the nature of surface boundary conditions have been shown to be important for controlling the 
 dynamics of subduction zone modeling.  In a benchmark study \citet{schmeling2008benchmark} performed extensive testing on the effect
@@
 \citet{crameri2012comparison} conducted comparisons between sticky air and true free surface models, 
-demonstrating a range of parameters for sticky air which can mitigate some of the difficulties that it 
-introduces.
-
+demonstrating a range of parameters for sticky air which can mitigate some of the difficulties that it introduces.
 A study by \citet{quinquis2011role} found that free surface boundary conditions have a large effect on 
@@
 
-Both true free surface and sticky air calculations have been subject to an instability which has been 
-variously termed a ``sloshing,'' or ``druknen sailor'' instability \citep{kaus2010stabilization, duretz2011discretization}. 
+All of the approaches to free surface simulations have been subject to an instability which has been 
+variously termed a ``sloshing,'' or ``druknen sailor'' instability \citep{kaus2010stabilization, duretz2011discretization, kramer2012implicit}. 
 This instability, arising from the large density contrast typical at a free surface (compared with the much smaller 
@@
 The exponentials may be expanded to find
+\ifdetail
 \begin{equation}
 a_i(\Delta t) - a_i^{1} = - a_i{(0)} \left[ \left(\frac{\Delta t}{\tau_i}\right)^2 - \frac{\tau^*}{\tau_i} \left(\frac{\Delta t }{\tau^* }\right)^2 \right] 
+\end{equation}
+\fi
+\begin{equation}
+a_i(\Delta t) - a_i^{1} = {a_i{(0)} } \left( \frac{\Delta t}{\tau_i} \right)^2 \left( 1 - \frac{\tau_i}{\tau^*} \right) + O(\Delta t^3)
 \end{equation}
@@
 As the relaxation parameter $\tau^*$ approaches the relaxation time of the benchmark case the error reduces.
-The sharp cusp at $\tau^* = \tau_\mathrm{min}$ corresponds to the essentially spectral accuracy of the time marching scheme for that case.}
+The sharp cusp at $\tau^* = \tau_\mathrm{min}$ corresponds to the almost spectral accuracy of the time marching scheme for that case.}
 \label{fig:tau_sensitivity}

### 2015-07-31 free-surface-paper.tex
Moar text

@@
 
+The lengthening of relaxation times due to the quasi-implicit stabilization has an unequal 
+effect on the modes. They are all lengthened by the same amount, which is a much bigger 
+fraction of the total relaxation time for the shorter modes than the longer ones.
+
 \section{A novel time-integration scheme}
@@
 of the free surface, but with better stability properties than forward Euler schemes.
-It is helpful to take a closer look at the pseudo-timestep $\tau^*(1-e^{-\Delta t/\tau^*})$.
 
@@
 \end{equation}
+
+It is helpful to take a closer look at the pseudo-timestep $\tau^*(1-e^{-\Delta t/\tau^*})$.
+As the timestep $\Delta t$ goes to zero, the pseudo timestep approaches $\Delta t$, recovering 
+the forward Euler scheme. However, as $\Delta t$ gets larger, the pseudo timestep does not 
+increase as quickly, reflecting the decaying nature of the normal modes.
+Likewise, as the stabilization timescale $\tau^*$ goes to infinity, we also recover the 
+forward Euler scheme, in what is essentially the unstabilized problem. But as $\tau^*$
+goes to zero, the pseudo-timestep also goes to zero. This corresponds to over-stabilizing the
+problem. With too short of a stabilization timescale the free surface is never advected 
+at all (which is a very stable situation, if not very accurate!).
 
@@
 
-Figure~\ref{fig:timestep_convergence} shows the convergence of the nonstandard finite difference scheme 
-with timestep.
+Figure~\ref{fig:timestep_convergence}a shows the convergence of the nonstandard finite difference scheme 
+with timestep. The scheme is first order in time, with improving accuracy as the value of $\tau^*$ approaches
+the relaxation time of the relevant mode $\tau_i$. Interestingly, if $\tau^* = \tau_i$ the advection scheme
+becomes exact \citep{mickens2002nonstandard}. At this point the error plummets in magnitude, and is no longer 
+a strong function of the timestep; instead, it is likely due to nonlinearities, error in the spatial discretization,
+or limits in the linear solver tolerance.
+
+Figure~\ref{fig:timestep_convergence}b shows the convergence of the quasi-implicit scheme with timestep.
+When $\theta=0.0$, it corresponds to the unstabilized forward Euler scheme, and is first order in time. 
+When $\theta=1.0$ it is also first order in time, but allows for a much larger timestep.  When $\theta=0.5$
+the quasi-implicit guess for the body force is good enough that it actually seems to achieve the second-order 
+convergence of a trapezoidal scheme, though it is not clear whether this extends to more complicated models.
+
+Figure~\ref{fig:tau_sensitivity} shows in more detail the effects of the choice of $\tau^*$. In a narrow region 
+close to the true value of $\tau_\mathrm{min}$ the error becomes very small, but in a broader region nearby 
+the errors are larger. Most of the time the enhanced accuracy of the scheme when $\tau^*$ is near to an eigenvalue 
+is a curiosity, though it does suggest the possibility for tuning the finite difference scheme to very accurately 
+track a specific mode in time.
 
@@
 
-\section{Discussion}
+\section{Conclusion}
+We have analyzed stability of free surface boundary conditions in geodynamic simulations and 
+demonstrated the cause of sloshing instabilities using a normal mode analysis.
+This perspective on the problem allowed us to construct an explicit finite difference 
+scheme which is first order accurate in time and is unconditionally stable.
+The nonstandard finite difference scheme is extremely simple to implement, and 
+requires no modifications to the system matrix.
+
+The normal mode perspective on the problem also provides insights into the effect of 
+the quasi-implicit stabilization scheme proposed by \citet{kaus2010stabilization}.
+The relaxation time of each mode is lengthened by an amout $\theta \Delta t$, and the
+maximum allowable timestep is correspondingly lengthened.
+
+Finally, we have described the implementation of free surface boundary conditions in 
+the open source mantle convection software \texttt{Aspect}. Both the quasi-implicit 
+scheme and the nonstandard finite difference scheme are available. The implementation is 
+sufficiently general to accommodate many different geometies, rheologies, and 
+gravity models. Furthermore, it runs in parallel and with adaptive mesh refinement.
 

### 2015-08-19 free-surface-paper.tex
Make look slightly more like a standard eigenvalue problem
Fix labels
Fix typo

@@
 Finally, we discuss the implementation of a free surface in the open source, community based
-mantle convection software \texttt{Aspect}.
+mantle convection software \texttt{ASPECT}.
 \end{abstract}
@@
 treated as an extra variable that was integrated in time. In this approximation, the free surface 
-is approximated on the undeformed Eulerian grid by applying pressure boundary conditions on the reference surface 
-corresponding to the hydrostatic pressure perturbation predicted by the surface topography.
+is approximated on the undeformed Eulerian grid by applying pressure boundary conditions on the reference surface.
+The pressure is determined by a first-order Taylor series approximation of the hydrostatic pressure profile 
+predicted by the surface topography.
 
@@
 (termed `air', though its viscosity is much higher) above the free surface, effectively decoupling it from the boundary. 
-Typically a free-slip boundary condition is used above the sticky air.
+Typically a free-slip boundary condition is used above the sticky air, though an open boundary may be better \citep{hillebrand2014using}.
 
 Finally, one can use a true free surface, where a stress-free boundary condition is applied on the boundary 
-of the domain. In this case, there can be flow in an out of the domain, so the boundary must move in time.
+of the domain. In this case, there can be flow in and out of the domain, so the boundary must move in time.
 A true free surface has mathematical elegance in that the boundary condition of the domain more closely 
@@
 on the dynamics of the slab, affecting both the shape and the timing of sinking. Most of the 
-participating codes in that benchmark used the `sticky air' approximation for the free surface, where a low viscosity,
-low density layer of fluid (termed `air', though its viscosity is much higher) is placed above the free surface, decoupling 
-it from the boundary. Typically a free-slip boundary condition is used above the stiky air.
-
-\citet{schmeling2008benchmark} found that the specific properties of the sticky air layer 
+participating codes in that benchmark used the sticky air approximation.
+They found that the specific properties of the sticky air layer 
 controlled the shape and timing of the slab, and the viscosity averaging scheme for areas of transitional
@@
 
-Finally, we describe the implementation of a free surface in the mantle convection software \texttt{Aspect}.
+Finally, we describe the implementation of a free surface in the mantle convection software \texttt{ASPECT}.
 
@@
 where $p_0$ is the reference hydrostatic pressure and $\rho_0$ is a reference density profile (which may vary with depth).
-The total pressure and density can then be decomposed in to variations about their 
+The total pressure and density can then be decomposed into variations about their 
 reference values: $\rho = \rho_0 + \rho^\prime$, $p = p_0 + p^\prime$.
@@
 On the reference surface the gravity vector is opposite the direction of the surface normal $\mathbf{g} = -g \mathbf{n}$.
-The stress on this reference surface can be approximated by the hydrostatic pressure
+The stress on this reference surface can be approximated by using the first order Taylor series
+expansion of the hydrostatic pressure profile:
 \begin{equation}
-\mathbf{T} \approx \rho_0 g \zeta \mathbf{I}
+\mathbf{T} \approx \frac{\partial \mathbf{T}}{\partial \mathbf{n} } \cdot \mathbf{n} \zeta = \rho_0 g \zeta \mathbf{I}
 \label{eq:hydrostatic}
@@
 \end{equation}
-Then multiplying both sides by $C^{-1}$ we get
+Then multiplying both sides by $\tau_i^{-1}C^{-1}$ we get
 \begin{equation}
-\tau_i (C^{-1}D)\mathbf{y}_i = \mathbf{y}_i
+(C^{-1}D)\mathbf{y}_i = \tau_i^{-1} \mathbf{y}_i
 \label{eq:standard_eigenvalue}
@@
 \begin{equation}
-\mathbf{f}_{\mathrm{body}} = \int_{\Omega + \Delta \Omega} \rho \mathbf{w} \cdot \mathbf{g}
+\mathbf{f}_{\mathrm{body}} = \int_{\Omega^n + \Delta \Omega} \rho \mathbf{w} \cdot \mathbf{g}
 \label{eq:predict}
 \end{equation}
-
+where the superscript $n$ on $\Omega$ indicates the shape at the $n$th timestep.
 The shape prediction can be approximated by integration of the velocity:
@@
 \begin{equation}
-\int_{\Omega + \Delta \Omega} \rho  \mathbf{w} \cdot \mathbf{g} \approx
-\int_{\Omega} \rho  \mathbf{w} \cdot \mathbf{g} + \theta \Delta t \int_{\partial \Omega} \rho ( \mathbf{w \cdot g}) (\mathbf{u \cdot n} )
+\int_{\Omega^n + \Delta \Omega} \rho  \mathbf{w} \cdot \mathbf{g} \approx
+\int_{\Omega^n} \rho  \mathbf{w} \cdot \mathbf{g} + \theta \Delta t \int_{\partial \Omega^n} \rho ( \mathbf{w \cdot g}) (\mathbf{u \cdot n} )
 \label{eq:kmm}
@@
 \begin{equation}
--\theta \Delta t \int_{\partial \Omega} \rho g ( \mathbf{w \cdot n}) (\mathbf{u \cdot n} )
+-\theta \Delta t \int_{\partial \Omega^n} \rho g ( \mathbf{w \cdot n}) (\mathbf{u \cdot n} )
 \end{equation}
@@
 \end{bmatrix}
-\label{eq:generalized_eigenvalue}
+\label{eq:stabilized_generalized_eigenvalue}
 \end{equation}
@@
 \end{bmatrix}
-\label{eq:generalized_eigenvalue}
+\label{eq:rearranged_stabilized_generalized_eigenvalue}
 \end{equation}
@@
 effect on the modes. They are all lengthened by the same amount, which is a much bigger 
-fraction of the total relaxation time for the shorter modes than the longer ones.
+fraction of the total relaxation time for the shorter-time modes than the longer-time ones.
+Therefore the shorter-time modes are effectively damped more.
+This can be seen as an attractive feature of the scheme, since those are the least stable modes,
+though it can result in less accurate time-marching of the longest wavelengths of the system.
 
@@
 
-One downside of the quasi-implicit scheme is that it requires a modification of the system matrix, 
-and as we have seen this shifts its spectrum.
+One downside of the quasi-implicit scheme is that it requires a modification of the system matrix.
+A naive implementation of the scheme results in a slightly 
+asymmetric matrix, which can be more difficult to solve, requiring either
+changes to the solver/preconditioner, or symmetrization of the stabilization term \citep{kaus2010stabilization}.
 An alternative would be to construct a time-integration scheme with better stability properties.  
@@
 
-Again we may take one of the normal modes as an example, differencing the nonstandard finite-difference scheme from the analytical solution:
+Again we may take one of the normal modes as an example and inspect the difference between the nonstandard finite-difference
+scheme and the analytical solution after one time step:
 
@@
 \end{equation}
-
 Note that when $\tau_i = \tau^*$ the nonstandard finite difference scheme is exact.
@@
 \end{equation}
+Summing this error over many timesteps results in a factor of $\Delta t^{-1}$, demonstrating the first-order accuracy~\citep[e.g.]{leveque2007finite}.
+The choice of $\tau^*$ controls the size of the coefficient on the truncation error for the scheme.
+The error for a given mode becomes considerably smaller when $\tau^*$ is close to its natural relaxation times.
 
@@
 
-\section{Implementation in \texttt{Aspect} }
+\section{Implementation in \texttt{ASPECT} }
 \label{sec:implementation}
-We have implemented the ability to run free surface simulations in the mantle convection software \texttt{Aspect} \citep{kronbichler2012high}. 
-\texttt{Aspect} is designed to be highly flexible and modular, with the ability for user-defined rheologies, geometries, and gravity models. 
+We have implemented the ability to run free surface simulations in the mantle convection software \texttt{ASPECT} \citep{kronbichler2012high}. 
+\texttt{ASPECT} is designed to be highly flexible and modular, with the ability for user-defined rheologies, geometries, and gravity models. 
 The implementation of the free surface, therefore, needs to be general enough to work for many combinations of these models, including those 
@@
 
-Furthermore, \texttt{Aspect} allows for computations in both 2D and 3D, so the implementation must be sufficiently dimension
+Furthermore, \texttt{ASPECT} allows for computations in both 2D and 3D, so the implementation must be sufficiently dimension
 independent to work in both cases. We implement the free surface within an arbitrary Lagrangian-Eulerian (ALE) formulation
@@
 
-\texttt{Aspect} is a parallelized, distributed memory code with adaptive mesh refinement.
+\texttt{ASPECT} is a parallelized, distributed memory code with adaptive mesh refinement.
 The free surface implementation works with these features.  We store the mesh vertex positions 
@@
 close to the true value of $\tau_\mathrm{min}$ the error becomes very small, but in a broader region nearby 
-the errors are larger. Most of the time the enhanced accuracy of the scheme when $\tau^*$ is near to an eigenvalue 
-is a curiosity, though it does suggest the possibility for tuning the finite difference scheme to very accurately 
-track a specific mode in time.
+the errors are larger. 
+The excellent accuracy when the stabilization timescale $\tau^*$ is close to one of the natural relaxation
+timescales allows for tuning of the scheme to track specific long-wavelength modes, such as 
+those due to rotational or tidal deformation.
 
@@
 \includegraphics[width=0.9\textwidth]{figures/tau_sensitivity.pdf}
-\caption{Sensitivity to $\tau^*$.
+\caption{Sensitivity of the nonstandard finite difference scheme to $\tau^*$.
 As the relaxation parameter $\tau^*$ approaches the relaxation time of the benchmark case the error reduces.
@@
 Finally, we have described the implementation of free surface boundary conditions in 
-the open source mantle convection software \texttt{Aspect}. Both the quasi-implicit 
+the open source mantle convection software \texttt{ASPECT}. Both the quasi-implicit 
 scheme and the nonstandard finite difference scheme are available. The implementation is 

### 2015-09-15 free-surface-paper.tex
Address comments

@@
 Finally, we discuss the implementation of a free surface in the open source, community based
-mantle convection software \texttt{ASPECT}.
+mantle convection software \texttt{Aspect}.
 \end{abstract}
@@
 Finally, one can use a true free surface, where a stress-free boundary condition is applied on the boundary 
-of the domain. In this case, there can be flow in and out of the domain, so the boundary must move in time.
+of the domain. In this case, there can be flow in an out of the domain, so the boundary must move in time.
 A true free surface has mathematical elegance in that the boundary condition of the domain more closely 
@@
 
-Finally, we describe the implementation of a free surface in the mantle convection software \texttt{ASPECT}.
+Finally, we describe the implementation of a free surface in the mantle convection software \texttt{Aspect}.
 
@@
 
-\section{Implementation in \texttt{ASPECT} }
+\section{Implementation in \texttt{Aspect} }
 \label{sec:implementation}
-We have implemented the ability to run free surface simulations in the mantle convection software \texttt{ASPECT} \citep{kronbichler2012high}. 
-\texttt{ASPECT} is designed to be highly flexible and modular, with the ability for user-defined rheologies, geometries, and gravity models. 
+We have implemented the ability to run free surface simulations in the mantle convection software \texttt{Aspect} \citep{kronbichler2012high}. 
+\texttt{Aspect} is designed to be highly flexible and modular, with the ability for user-defined rheologies, geometries, and gravity models. 
 The implementation of the free surface, therefore, needs to be general enough to work for many combinations of these models, including those 
@@
 
-Furthermore, \texttt{ASPECT} allows for computations in both 2D and 3D, so the implementation must be sufficiently dimension
+Furthermore, \texttt{Aspect} allows for computations in both 2D and 3D, so the implementation must be sufficiently dimension
 independent to work in both cases. We implement the free surface within an arbitrary Lagrangian-Eulerian (ALE) formulation
@@
 
-\texttt{ASPECT} is a parallelized, distributed memory code with adaptive mesh refinement.
+\texttt{Aspect} is a parallelized, distributed memory code with adaptive mesh refinement.
 The free surface implementation works with these features.  We store the mesh vertex positions 
@@
 Finally, we have described the implementation of free surface boundary conditions in 
-the open source mantle convection software \texttt{ASPECT}. Both the quasi-implicit 
+the open source mantle convection software \texttt{Aspect}. Both the quasi-implicit 
 scheme and the nonstandard finite difference scheme are available. The implementation is 

### 2015-09-15 figures/stability_region.py
Attempt to fix rendering issues with some pdf viewers

@@ -10,16 +10,18 @@ except ValueError:
 
 plt.figure(figsize=(8,4) )
 
+ymax = 10. #top of figure
+
 x_nsfd = np.linspace(1.e-10,5, 1000.)
-y_nsfd = np.piecewise( x_nsfd , [x_nsfd <= 2., x_nsfd > 2.], [lambda x : 1.e6, lambda x : -x*np.log(1.-2./x)] )
+y_nsfd = np.piecewise( x_nsfd , [x_nsfd <= 2., x_nsfd > 2.], [lambda x : ymax, lambda x : -x*np.log(1.-2./x)] )
 x_kmm = np.linspace(1.e-10,1.-1.e-10, 1000.)
-y_kmm = 2./(1.-x_kmm)
+y_kmm = np.piecewise( x_kmm, [x_kmm <= 0.85, x_kmm > 0.85], [ lambda x :2./(1.-x), lambda x : ymax] )
 y_fe = np.ones_like(x_nsfd)*2.
 
 fontsize = 16
 plt.subplot(121)
 plt.xlim(0.,5.)
-plt.ylim(0.,10.)
+plt.ylim(0.,ymax)
 plt.xlabel(r'$\tau^*/\tau_{\mathrm{min}}$', fontsize=fontsize)
 plt.ylabel(r'$\frac{\Delta t}{\tau_{\mathrm{min}}}$', rotation=0, fontsize=fontsize)
 plt.fill_between(x_nsfd, 0, y_nsfd, color='green', alpha=0.2)
@@ -27,7 +29,7 @@ plt.fill_between(x_nsfd, 0, y_fe, color='blue', alpha=0.2)
 plt.title('(a) Non-standard FD', fontsize=fontsize)
 plt.subplot(122)
 plt.xlim(0.,1.)
-plt.ylim(0.,10.)
+plt.ylim(0.,ymax)
 plt.xlabel(r'$\theta$', fontsize=fontsize)
 plt.fill_between(x_kmm, 0, y_kmm, color='green', alpha=0.2)
 plt.fill_between(x_nsfd, 0, y_fe, color='blue', alpha=0.2)

### 2015-09-15 free-surface-paper.tex
Capitalize ASPECT
Wording

--- 2015-08-19 free-surface-paper.tex
@@
 \citet{zhong1996free} introduced a pseudo-free-surface formulation, where the free surface coordinate was
-treated as an extra variable that was integrated in time. In this approximation, the free surface 
+treated as an extra variable that was integrated in time. In this formulation, the free surface 
 is approximated on the undeformed Eulerian grid by applying pressure boundary conditions on the reference surface.
@@
 participating codes in that benchmark used the sticky air approximation.
-They found that the specific properties of the sticky air layer 
-controlled the shape and timing of the slab, and the viscosity averaging scheme for areas of transitional
+They found that the specific properties of the sticky air layer controlled the shape and timing of the slab.
+Furthermore, the viscosity averaging scheme for areas of transitional
 composition was extremely important, as the subducting slab entrained significant amounts of the sticky air.

### 2015-10-03 benchmarks/amr/adaptive_nsfd/run_models.py
Add figure for resolution convergence, with and without AMR

@@ -1,0 +1,40 @@
+import os
+import numpy as np
+
+def generate_prm( refine ):
+  prmfile = open("../crameri_benchmark_2.prm", "r")
+  outfile = open("tmp.prm", "w")
+  for l in prmfile.readlines():
+    if 'Initial global refinement' in l:
+      outfile.write('  set Initial global refinement = 3\n')
+    elif 'Initial adaptive refinement' in l:
+      outfile.write('  set Initial adaptive refinement = %i\n'%(refine) )
+    elif 'Free surface stabilization theta' in l:
+      outfile.write('  set Free surface stabilization theta = 0.0\n')
+    elif 'Relaxation time' in l:
+      outfile.write('  set Relaxation time = 14825.\n')
+    elif 'Time steps between mesh refinement' in l:
+      outfile.write('  set Time steps between mesh refinement = 10\n')
+    else:
+      outfile.write(l)
+  prmfile.close()
+  outfile.close()
+
+def grab_topography():
+  data = np.genfromtxt("output/statistics", usecols=(1,14))
+  time = data[:,0]
+  topo = data[:,1]
+
+  topography = topo[ np.argmin( np.abs(time - 3.e6)) ]
+  return topography
+
+outputfile = open('resolution_convergence_adaptive_nsfd.txt', 'w')
+refine = range(1,6)
+for r in refine:
+  print "Running uniform mesh with %i total refinements\n"%(r+3)
+  generate_prm(r)
+  os.system('mpirun /home/ian.rose/aspect/build/aspect tmp.prm')
+  outputfile.write("%i %f \n" % (r, grab_topography()) )
+  outputfile.flush()
+
+outputfile.close()

### 2015-10-03 benchmarks/amr/adaptive_qi/run_models.py

@@ -1,0 +1,40 @@
+import os
+import numpy as np
+
+def generate_prm( refine ):
+  prmfile = open("../crameri_benchmark_2.prm", "r")
+  outfile = open("tmp.prm", "w")
+  for l in prmfile.readlines():
+    if 'Initial global refinement' in l:
+      outfile.write('  set Initial global refinement = 3\n')
+    elif 'Initial adaptive refinement' in l:
+      outfile.write('  set Initial adaptive refinement = %i\n'%(refine) )
+    elif 'Free surface stabilization theta' in l:
+      outfile.write('  set Free surface stabilization theta = 0.5\n')
+    elif 'Relaxation time' in l:
+      outfile.write('  set Relaxation time = 0.\n')
+    elif 'Time steps between mesh refinement' in l:
+      outfile.write('  set Time steps between mesh refinement = 10\n')
+    else:
+      outfile.write(l)
+  prmfile.close()
+  outfile.close()
+
+def grab_topography():
+  data = np.genfromtxt("output/statistics", usecols=(1,14))
+  time = data[:,0]
+  topo = data[:,1]
+
+  topography = topo[ np.argmin( np.abs(time - 3.e6)) ]
+  return topography
+
+outputfile = open('resolution_convergence_adaptive_qi.txt', 'w')
+refine = range(1,6)
+for r in refine:
+  print "Running uniform mesh with %i total refinements\n"%(r+3)
+  generate_prm(r)
+  os.system('mpirun /home/ian.rose/aspect/build/aspect tmp.prm')
+  outputfile.write("%i %f \n" % (r, grab_topography()) )
+  outputfile.flush()
+
+outputfile.close()

### 2015-10-03 benchmarks/amr/uniform_nsfd/run_models.py

@@ -1,0 +1,38 @@
+import os
+import numpy as np
+
+def generate_prm( refine ):
+  prmfile = open("../crameri_benchmark_2.prm", "r")
+  outfile = open("tmp.prm", "w")
+  for l in prmfile.readlines():
+    if 'Initial global refinement' in l:
+      outfile.write('  set Initial global refinement = %i\n'%(refine) )
+    elif 'Initial adaptive refinement' in l:
+      outfile.write('  set Initial adaptive refinement = %i\n'%(0) )
+    elif 'Free surface stabilization theta' in l:
+      outfile.write('  set Free surface stabilization theta = 0.0\n')
+    elif 'Relaxation time' in l:
+      outfile.write('  set Relaxation time = 14825.\n')
+    else:
+      outfile.write(l)
+  prmfile.close()
+  outfile.close()
+
+def grab_topography():
+  data = np.genfromtxt("output/statistics", usecols=(1,14))
+  time = data[:,0]
+  topo = data[:,1]
+
+  topography = topo[ np.argmin( np.abs(time - 3.e6)) ]
+  return topography
+
+outputfile = open('resolution_convergence_uniform_nsfd.txt', 'w')
+refine = range(4,9)
+for r in refine:
+  print "Running uniform mesh with %i refinements\n"%(r)
+  generate_prm(r)
+  os.system('mpirun /home/ian.rose/aspect/build/aspect tmp.prm')
+  outputfile.write("%i %f \n" % (r, grab_topography()) )
+  outputfile.flush()
+
+outputfile.close()

### 2015-10-03 benchmarks/amr/uniform_qi/run_models.py

@@ -1,0 +1,38 @@
+import os
+import numpy as np
+
+def generate_prm( refine ):
+  prmfile = open("../crameri_benchmark_2.prm", "r")
+  outfile = open("tmp.prm", "w")
+  for l in prmfile.readlines():
+    if 'Initial global refinement' in l:
+      outfile.write('  set Initial global refinement = %i\n'%(refine) )
+    elif 'Initial adaptive refinement' in l:
+      outfile.write('  set Initial adaptive refinement = %i\n'%(0) )
+    elif 'Free surface stabilization theta' in l:
+      outfile.write('  set Free surface stabilization theta = 0.5\n')
+    elif 'Relaxation time' in l:
+      outfile.write('  set Relaxation time = 0.\n')
+    else:
+      outfile.write(l)
+  prmfile.close()
+  outfile.close()
+
+def grab_topography():
+  data = np.genfromtxt("output/statistics", usecols=(1,14))
+  time = data[:,0]
+  topo = data[:,1]
+
+  topography = topo[ np.argmin( np.abs(time - 3.e6)) ]
+  return topography
+
+outputfile = open('resolution_convergence_uniform_qi.txt', 'w')
+refine = range(4,9)
+for r in refine:
+  print "Running uniform mesh with %i refinements\n"%(r)
+  generate_prm(r)
+  os.system('mpirun /home/ian.rose/aspect/build/aspect tmp.prm')
+  outputfile.write("%i %f \n" % (r, grab_topography()) )
+  outputfile.flush()
+
+outputfile.close()

### 2015-10-03 figures/amr.py

@@ -1,0 +1,68 @@
+import matplotlib.pyplot as plt
+import numpy as np
+import numpy.ma
+import itertools
+
+try:
+  plt.style.use('ian')
+except AttributeError:
+  print "old matplotlib version?"
+except ValueError:
+  print "Cannot find the requested stylesheet"
+
+plt.figure(figsize=(8,4) )
+
+
+data_uniform_qi = np.loadtxt("../benchmarks/amr/uniform_qi/resolution_convergence_uniform_qi.txt")
+data_uniform_nsfd = np.loadtxt("../benchmarks/amr/uniform_nsfd/resolution_convergence_uniform_nsfd.txt")
+data_adaptive_qi = np.loadtxt("../benchmarks/amr/adaptive_qi/resolution_convergence_adaptive_qi.txt")
+data_adaptive_nsfd = np.loadtxt("../benchmarks/amr/adaptive_nsfd/resolution_convergence_adaptive_nsfd.txt")
+
+refinement_levels = np.linspace(4., 8., 5)
+h_value  =  (2.)**(-refinement_levels)
+
+topography_at_3Ma = 397.6
+
+error_uniform_qi = np.abs(data_uniform_qi[:,1] - topography_at_3Ma)
+error_uniform_nsfd = np.abs(data_uniform_nsfd[:,1] - topography_at_3Ma)
+error_adaptive_qi = np.abs(data_adaptive_qi[:,1] - topography_at_3Ma)
+error_adaptive_nsfd = np.abs(data_adaptive_nsfd[:,1] - topography_at_3Ma)
+
+
+font_size = 12
+
+plt.subplot(121)
+
+marker = itertools.cycle( ('-h', '-v') )
+plt.loglog(h_value, error_uniform_qi, marker.next(), markersize=8, label=r'QI')
+plt.loglog(h_value, error_uniform_nsfd, marker.next(), markersize=8, label=r'NSFD')
+
+plt.text(1.2e-3,3.e-1, '5.5 MDoF', fontsize = font_size)
+plt.text(2.1e-3,1.2e0, '1.4 MDoF', fontsize = font_size)
+plt.text(4.e-3,5.e0, '347 kDoF', fontsize = font_size)
+plt.text(1.e-2,3.7e1, '88 kDoF', fontsize = font_size)
+plt.text(2.e-2,6.5e1, '22 kDoF', fontsize = font_size)
+
+plt.xlabel(r'h')
[…]

### 2015-10-05 free-surface-paper.tex
Add AMR figure to manuscript

@@
 
+We also investigate the advantages of combining adaptive mesh refinement with free surface computations.
+\citet{crameri2012comparison} performed a community benchmark of a setup for which there is no analytic
+solution. In this benchmark, a buoyant blob rises beneath a viscous lid with a free surface, deflecting the
+boundary upwards (for the full setup, see \citet{crameri2012comparison}). 
+Figure~\ref{fig:amr} shows the convergence of the maximum topography at 3 Myr to its value in a high resolution simulation.
+Figure~\ref{fig:amr}a shows the convergence with uniform refinement, and~\ref{fig:amr}b shows the convergence
+with adaptive mesh refinement. Both have essentially the same behavior, but the adaptive case requires 
+fewer degrees of freedom by a factor of approximately an order of magnitude. More complex models will have
+more detail and so we may be less able to aggressively coarsen them, but we still expect that adaptive mesh refinement 
+will result in significant computational savings.
+
+\begin{figure*}
+\includegraphics[width=0.9\textwidth]{figures/amr.pdf}
+\caption{Convergence with resolution for the \citet{crameri2012comparison} Case 2 benchmark, for both uniform (a) and adaptive (b) mesh refinement.
+The minimum cell size h is normalized by the model depth, and the timestep $\Delta t$ is 500 yr.
+We compare the maximum topography at 3 Myr with its value at high resolution ($\sim$398 m).
+The points are labeled with the approximate number of degrees of freedom (DoFs) for that simulation.
+Both cases converge similarly, but the adaptive mesh refinement simulations require significantly smaller system sizes.}
+\label{fig:amr}
+\end{figure*}
+
 \section{Conclusion}

### 2015-10-05 figures/amr.py

@@ -44,9 +44,9 @@ plt.text(1.e-2,3.7e1, '88 kDoF', fontsize = font_size)
 plt.text(2.e-2,6.5e1, '22 kDoF', fontsize = font_size)
 
 plt.xlabel(r'h')
-plt.ylabel(r'Topography error at 3 Ma (m)')
+plt.ylabel(r'Topography error at 3 Myr (m)')
 plt.legend(loc='lower right', fontsize=8)
-plt.title("Uniform refinement")
+plt.title("(a) Uniform refinement")
 
 plt.subplot(122)
 
@@ -55,7 +55,7 @@ plt.loglog(h_value, error_adaptive_qi, marker.next(), markersize=8, label=r'QI')
 plt.loglog(h_value, error_adaptive_nsfd, marker.next(), markersize=8, label=r'NSFD')
 plt.xlabel(r'h')
 plt.legend(loc='lower right', fontsize=8)
-plt.title("Adaptive refinement")
+plt.title("(b) Adaptive refinement")
 
 plt.text(1.2e-3,3.e-1, '233 kDoF', fontsize = font_size)
 plt.text(2.1e-3,1.2e0, '111 kDoF', fontsize = font_size)

### 2015-10-30 free-surface-paper.tex
part one

@@
 \usepackage{color, soul}
+\usepackage{url}
+
+\usepackage{todonotes}
+\newcommand\timo[1]{\todo[inline]{Timo: #1}}
 
@@
 
-\author{I. Rose, B. Buffett, and T. Heister}
+\author{I. Rose, B. Buffett, and Timo Heister\fnref{ref3}}
+
+\fntext[ref3]{\url{heister@clemson.edu}, partially supported by the Computational Infrastructure in
+Geodynamics initiative (CIG), through the National Science Foundation under Award No. EAR-0949446 
+and The University of California—Davis, and National Science Foundation grant DMS1522191.}
 
@@
 All of the approaches to free surface simulations have been subject to an instability which has been 
-variously termed a ``sloshing,'' or ``druknen sailor'' instability \citep{kaus2010stabilization, duretz2011discretization, kramer2012implicit}. 
+variously termed a ``sloshing,'' or ``drunken sailor'' instability \citep{kaus2010stabilization, duretz2011discretization, kramer2012implicit}. 
 This instability, arising from the large density contrast typical at a free surface (compared with the much smaller 
@@
 
-Furthermore, since we are interested in solving these equations for a domain with a free surface, 
-we need an equation to track that surface.
-The Lagrangian equation for time evolution of the free surface can be written
+Formally, this gives rise to the following time dependent, coupled system with unknowns $\mathbf{u}(t)$ and $\mathbf{p}(t)$:
 \begin{equation}
-\frac{d \zeta}{dt} = \mathbf{u \cdot n} \quad \textrm{on  }  \partial \Omega_\textrm{free surface}
-\label{eq:surface_evolution}
+\begin{aligned}
+\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p &= \rho \mathbf{g} \qquad \text{in} \qquad \Omega(t)\\
+\nabla \cdot \mathbf{u} &= 0  \qquad \text{in} \qquad  \Omega(t)
+\end{aligned}
+\label{eq:final_system}
 \end{equation}
-where $\zeta$ is the surface coordinate of the free surface, and $\partial \Omega_\textrm{free surface}$ is the 
-part of the domain boundary corresponding to the free surface.
+defined on the bounded, moving domain $\Omega(t)\subset \mathbb{R}^d$ with boundary $\partial \Omega = \Gamma_0 \cup \Gamma_F$
+split into a fixed (Dirichlet) and free surface part $\Gamma_0$ and $\Gamma_F$, respectively.
+The domain at time $t$ is defined by advecting a reference configuration $\Omega_0$ by a mesh velocity $\mathbf{v}(t)$
+\timo{Do we need $\Gamma_N$ here?}
+\begin{equation}
+ \Omega(t) = \Omega_0 + \int_{t_0}^t \mathbf{v}(t) \text{ d} t.
+\end{equation}
+The mesh velocity in normal direction is given by the velocity $\mathbf{u}(t)$ on the free surface
+\begin{equation}
+ \mathbf{v}(t) \cdot n = \mathbf{u}(t) \cdot n \text{on} \Gamma_F
+\end{equation}
+and set to zero on the rest of the boundary. We defer the discussion of the velocity $\mathbf{v}$ in the interior of the domain
+to Section~\ref{sec:remeshing}.
+Finally, we denote the displacement at time $t$ by $\zeta = \int_{t_0}^t \mathbf{v}(t) \text{ d} t$.
 
@@
 
-In order to better understand the time evolution of Equations~\eqref{eq:momentum_mass_conservation}-\eqref{eq:surface_evolution} 
+In order to better understand the time evolution of the system in \eqref{eq:final_system}
 we will consider the eigenvalues of a simplified homogeneous system. By setting $\rho = \rho_0$, we obtain the system
@@
 -\int_{\Omega(t)} 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u} ) + \int_{\Omega(t)} p^\prime \nabla \cdot \mathbf{w} 
-+ \int_{\partial \Omega_N(t)} \mathbf{w} \cdot \mathbf{T} \cdot \mathbf{n} = 0 
++ \int_{\Gamma_N(t)} \mathbf{w} \cdot \mathbf{T} \cdot \mathbf{n} = 0 
 \label{eq:weak_stokes}
 \end{equation}
-
+\timo{change boundary $\Gamma_N$ with new notation above.}
 \begin{equation}
@@
 where $\mathbf{w}$ and $q$ are suitably chosen test functions and the integrals over 
-$\Omega$ and $\partial \Omega_N$ are over the volume of the domain and the part of the boundary with stress boundary conditions, respectively.
+$\Omega$ and $\partial \Gamma_N$ are over the volume of the domain and the part of the boundary with stress boundary conditions, respectively.
 Note that since the free surface can move, the shape of the domain is a function of time.
@@
 
+\timo{TODO: update. write as solving on $\Omega(t^n)$ instead of $\Omega(t^{n+1})$}
 The integral over the surface in Equation~\eqref{eq:weak_stokes} accounts for boundary stresses, 
@@
 \subsection{Remeshing}
+\label{sec:remeshing}
+
 In ALE calculations the internal mesh velicity is undetermined.

### 2015-10-30 free-surface-paper.tex
add back eqn
fix figure reference
more work through the paper

@@
 \begin{equation}
- \mathbf{v}(t) \cdot n = \mathbf{u}(t) \cdot n \text{on} \Gamma_F
+ \mathbf{v}(t) \cdot n = \mathbf{u}(t) \cdot n \quad \text{on} \quad \Gamma_F
 \end{equation}
@@
 to Section~\ref{sec:remeshing}.
-Finally, we denote the displacement at time $t$ by $\zeta = \int_{t_0}^t \mathbf{v}(t) \text{ d} t$.
+Finally, we denote the displacement at time $t$ by $\zeta = \int_{t_0}^t \mathbf{v}(t) \text{ d} t$, leading to the evolution
+equation
+\begin{equation}
+\frac{d \zeta}{dt} = \mathbf{u \cdot n} \quad \textrm{on  }  \Gamma_F.
+\label{eq:surface_evolution}
+\end{equation}
 
@@
 \includegraphics[width=0.9\textwidth]{figures/stability_region.pdf}
-\label{fig:stability_region}
 \caption{(a) Stability region for the nonstandard finite difference scheme (green).
@@
 In both cases the stability region for the forward Euler scheme is also shown in blue.}
+\label{fig:stability_region}
 \end{figure*}
@@
 
-In ALE calculations the internal mesh velicity is undetermined.
+In ALE calculations the internal mesh velocity is undetermined.
 In general, one wants to smoothly deform the mesh so as to preserve its regularity, 
@@
 \begin{aligned}
-&\mathbf{u}_\mathrm{mesh} = \left( \textbf{u} \cdot \textbf{n} \right) \textbf{n} & \qquad & \textrm{on } \partial \Omega_{\textrm{free surface}}, \\
+&\mathbf{u}_\mathrm{mesh} = \left( \textbf{u} \cdot \textbf{n} \right) \textbf{n} & \qquad & \textrm{on } \Gamma_F, \\
 &\mathbf{u}_\mathrm{mesh} \cdot \textbf{n} = 0 & \qquad & \textrm{on } \partial \Omega_{\textrm{free slip}}, \\
-&\mathbf{u}_\mathrm{mesh} = 0 & \qquad & \textrm{on } \partial \Omega_{\textrm{Dirichlet}}.
+&\mathbf{u}_\mathrm{mesh} = 0 & \qquad & \textrm{on } \Gamma_0.
 \end{aligned}
@@
 \begin{equation}
-\mathbf{u}_{\mathrm{mesh}} \cdot \mathbf{n} = \mathbf{u \cdot n}
+\mathbf{u}_{\mathrm{mesh}} \cdot \mathbf{n} = \mathbf{u \cdot n}.
 \end{equation}
@@
 \begin{equation}
-\int_{\partial \Omega_{\textrm{free surface}} } \mathbf{w} \cdot \mathbf{u}_\mathrm{mesh} = 
-\int_{\partial \Omega_{\textrm{free surface}} } \left( \mathbf{w \cdot n } \right) \left( \mathbf{u \cdot n} \right)
+\int_{\Gamma_F} \mathbf{w} \cdot \mathbf{u}_\mathrm{mesh} = 
+\int_{\Gamma_F} \left( \mathbf{w \cdot n } \right) \left( \mathbf{u \cdot n} \right).
 \end{equation}
-
-When discretized, this forms a linear system which can be solved for the mesh velocity at the 
+When discretized, this forms a linear system which can be solved for the mesh velocity $\mathbf{u}_\mathrm{mesh}$ at the 
 free surface. Of course, a new system to solve is not ideal, but this system, being nonzero 
@@
 Figure~\ref{fig:timestep_convergence}a shows the convergence of the nonstandard finite difference scheme 
-with timestep. The scheme is first order in time, with improving accuracy as the value of $\tau^*$ approaches
+with respect to timestep size. The scheme is first order in time, with improving accuracy as the value of $\tau^*$ approaches
 the relaxation time of the relevant mode $\tau_i$. Interestingly, if $\tau^* = \tau_i$ the advection scheme
 becomes exact \citep{mickens2002nonstandard}. At this point the error plummets in magnitude, and is no longer 
-a strong function of the timestep; instead, it is likely due to nonlinearities, error in the spatial discretization,
+a strong function of the timestep. The remaining error is likely due to nonlinearities, error in the spatial discretization,
 or limits in the linear solver tolerance. 
+\timo{Could the error be the accuracy of the reference $\zeta$? If you look at 3b, the error also tapers of at 1e-5...}
 
@@
 \begin{figure*}
-\includegraphics[width=0.9\textwidth]{figures/timestep_convergence.pdf}
+\includegraphics[width=0.95\textwidth]{figures/timestep_convergence.pdf}
 \caption{Convergence tests for the benchmark shown in Figure~\ref{fig:benchmark_setup}.
@@
 will result in significant computational savings.
+\timo{Change the figure into one with number of unknowns on the x axis and combine into a single figure. Your computed reference value is also not accurate enough. (I can take a look at that)}
 

### 2015-11-02 free-surface-paper.tex
Address comments from Timo

@@
 \begin{aligned}
-\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p &= \rho \mathbf{g} \qquad \text{in} \qquad \Omega(t)\\
-\nabla \cdot \mathbf{u} &= 0  \qquad \text{in} \qquad  \Omega(t)
+\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p &= \rho \mathbf{g} \qquad &\text{in} \qquad &\Omega(t)\\
+\nabla \cdot \mathbf{u} &= 0  \qquad &\text{in} \qquad &\Omega(t)
 \end{aligned}
@@
 split into a fixed (Dirichlet) and free surface part $\Gamma_0$ and $\Gamma_F$, respectively.
-The domain at time $t$ is defined by advecting a reference configuration $\Omega_0$ by a mesh velocity $\mathbf{v}(t)$
-\timo{Do we need $\Gamma_N$ here?}
+For the sake of economy we neglect inhomogeneous stress boundary conditions, though it would be straightforward to include them.
+The domain at time $t$ is defined by advecting a reference configuration $\Omega_0$ by a mesh velocity $\mathbf{u}_{\mathrm{mesh}}(t)$
 \begin{equation}
- \Omega(t) = \Omega_0 + \int_{t_0}^t \mathbf{v}(t) \text{ d} t.
+ \Omega(t) = \Omega_0 + \int_{t_0}^t \mathbf{u}_{\mathrm{mesh}}(t) \text{ d}t.
 \end{equation}
@@
 \begin{equation}
- \mathbf{v}(t) \cdot n = \mathbf{u}(t) \cdot n \quad \text{on} \quad \Gamma_F
+ \mathbf{u}_\mathrm{mesh}(t) \cdot \mathbf{n} = \mathbf{u}(t) \cdot \mathbf{n} \quad \text{on} \quad \Gamma_F
 \end{equation}
-and set to zero on the rest of the boundary. We defer the discussion of the velocity $\mathbf{v}$ in the interior of the domain
+and set to zero on the rest of the boundary. We defer the discussion of the velocity $\mathbf{u}_\mathrm{mesh}$ in the interior of the domain
 to Section~\ref{sec:remeshing}.
-Finally, we denote the displacement at time $t$ by $\zeta = \int_{t_0}^t \mathbf{v}(t) \text{ d} t$, leading to the evolution
+Finally, we denote the displacement at time $t$ by $\zeta = \int_{t_0}^t \mathbf{u}_\mathrm{mesh}(t)\cdot \mathbf{n} \text{ d}t$, leading to the evolution
 equation
 \begin{equation}
-\frac{d \zeta}{dt} = \mathbf{u \cdot n} \quad \textrm{on  }  \Gamma_F.
+\frac{\text{d} \zeta}{\text{d}t} = \mathbf{u \cdot \mathbf{n}} \quad \textrm{on  }  \Gamma_F.
 \label{eq:surface_evolution}
@@
 -\int_{\Omega(t)} 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u} ) + \int_{\Omega(t)} p^\prime \nabla \cdot \mathbf{w} 
-+ \int_{\Gamma_N(t)} \mathbf{w} \cdot \mathbf{T} \cdot \mathbf{n} = 0 
++ \int_{\Gamma_F(t)} \mathbf{w} \cdot \mathbf{T} \cdot \mathbf{n} = 0 
 \label{eq:weak_stokes}
 \end{equation}
-\timo{change boundary $\Gamma_N$ with new notation above.}
 \begin{equation}
@@
 where $\mathbf{w}$ and $q$ are suitably chosen test functions and the integrals over 
-$\Omega$ and $\partial \Gamma_N$ are over the volume of the domain and the part of the boundary with stress boundary conditions, respectively.
+$\Omega(t)$ and $\Gamma_F(t)$ are over the volume of the domain and the free surface, respectively.
 Note that since the free surface can move, the shape of the domain is a function of time.
-It is most convenient to integrate over the domain at the current timestep, which corresponds to an explicit scheme.
-However, as we will see, the implicit scheme of integrating over the (unknown) domain at the next timestep
+It is most convenient to integrate over the domain at the current timestep $\Omega^n$ 
+(where the superscript indicates timestep number), which corresponds to an explicit scheme.
+However, as we will see, the implicit scheme of integrating over the (unknown) domain at the next timestep $\Omega^{n+1}$
 is much more stable, at the cost of making the problem nonlinear \citep{furuichi2015implicit}.
 
-\timo{TODO: update. write as solving on $\Omega(t^n)$ instead of $\Omega(t^{n+1})$}
 The integral over the surface in Equation~\eqref{eq:weak_stokes} accounts for boundary stresses, 
@@
 -\int_{\Omega(t)} 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u} ) + \int_{\Omega(t)} p^\prime \nabla \cdot \mathbf{w} 
-+ \int_{\partial \Omega_N(t)} \rho_0 g \zeta  \mathbf{w} \cdot \mathbf{n} = 0 
++ \int_{\Gamma_F(t)} \rho_0 g \zeta  \mathbf{w} \cdot \mathbf{n} = 0 
 \end{equation}
@@
 \begin{equation}
-\frac {d}{d t} \zeta_i(\mathbf{x},t) = \frac{d}{dt} \zeta_i(\mathbf{x})e^{-t/\tau_i} = -\frac{\zeta_i(\mathbf{x},t)}{\tau_i} = \mathbf{u}_i \cdot \mathbf{n}
+\frac {\text{d}}{\text{d} t} \zeta_i(\mathbf{x},t) = \frac{\text{d}}{\text{d}t} \zeta_i(\mathbf{x})e^{-t/\tau_i} = -\frac{\zeta_i(\mathbf{x},t)}{\tau_i} = \mathbf{u}_i \cdot \mathbf{n}
 \end{equation}
@@
 -\int_{\Omega(t)} 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u}_i ) + \int_{\Omega(t)} p^\prime_i \nabla \cdot \mathbf{w} 
-= \tau_i \int_{\partial \Omega_N(t)} \rho_0 g (\mathbf{u}_i \cdot \mathbf{n} ) (\mathbf{w} \cdot \mathbf{n})
+= \tau_i \int_{\Gamma_F(t)} \rho_0 g (\mathbf{u}_i \cdot \mathbf{n} ) (\mathbf{w} \cdot \mathbf{n})
 \label{eq:weak_eigen}
@@
 \int_{\Omega^n + \Delta \Omega} \rho  \mathbf{w} \cdot \mathbf{g} \approx
-\int_{\Omega^n} \rho  \mathbf{w} \cdot \mathbf{g} + \theta \Delta t \int_{\partial \Omega^n} \rho ( \mathbf{w \cdot g}) (\mathbf{u \cdot n} )
+\int_{\Omega^n} \rho  \mathbf{w} \cdot \mathbf{g} + \theta \Delta t \int_{\Gamma_F^n} \rho ( \mathbf{w \cdot g}) (\mathbf{u \cdot n} )
 \label{eq:kmm}
@@
 \begin{equation}
--\theta \Delta t \int_{\partial \Omega^n} \rho g ( \mathbf{w \cdot n}) (\mathbf{u \cdot n} )
+-\theta \Delta t \int_{\Gamma_F^n} \rho g ( \mathbf{w \cdot n}) (\mathbf{u \cdot n} )
 \end{equation}
@@
 \end{equation}
-note that as $\theta$ goes to one this scheme should become unconditionally stable,
+Note that as $\theta$ goes to one this scheme should become unconditionally stable,
 but nonlinear effects and discretization errors due to finite deformation of the surface could 
@@
 \begin{equation}
-\mathbf{x}(t^{n+1}) = \mathbf{x}(t^n) + \int_{t^{n}}^{t^{n+1}} \mathbf{u}(t) dt
+\mathbf{x}(t^{n+1}) = \mathbf{x}(t^n) + \int_{t^{n}}^{t^{n+1}} \mathbf{u}(t) \text{ d}t
 \label{eq:time_integration}
@@
 \begin{equation}
-\frac{d a_i} {dt} = - \frac{ a_i }{\tau_i} 
+\frac{\text{d} a_i} {\text{d}t} = - \frac{ a_i }{\tau_i} 
 \end{equation}
@@
 \end{equation}
-where superscripts indicate the timestep number.
 In order for the scheme to be stable, the quantity in brackets must not blow up as it is repeatedly multiplied by itself, or 
@@
 \begin{equation}
-\mathrm{Error} = \frac{1}{4\tau}\int_0^{4 \tau} \lVert \zeta_\mathrm{numeric}(L/2, t) - \zeta_\mathrm{analytic}(L/2, t) \rVert_2 dt
+\mathrm{Error} = \frac{1}{4\tau}\int_0^{4 \tau} \lVert \zeta_\mathrm{numeric}(L/2, t) - \zeta_\mathrm{analytic}(L/2, t) \rVert_2 \text{ d}t
 \end{equation}
@@
 the relaxation time of the relevant mode $\tau_i$. Interestingly, if $\tau^* = \tau_i$ the advection scheme
-becomes exact \citep{mickens2002nonstandard}. At this point the error plummets in magnitude, and is no longer 
-a strong function of the timestep. The remaining error is likely due to nonlinearities, error in the spatial discretization,
-or limits in the linear solver tolerance. 
-\timo{Could the error be the accuracy of the reference $\zeta$? If you look at 3b, the error also tapers of at 1e-5...}
+becomes exact \citep{mickens2002nonstandard}. At this point the magnitude of the error plummets and is no longer 
+a strong function of the timestep. The remaining error is likely due to error in the linear approxmation for 
+the analytical solution, the spatial discretization, or the linear solver tolerance.
 
@@
 Comparison with the slope-one line confirms that it is first-order in time.
-In the case that the stabilization relaxation time $\tau^*$ is equal to the analytic relaxation time the error plummets, as the time integration scheme becomes exact \citep{mickens2002nonstandard}.
+In the case that the stabilization relaxation time $\tau^*$ is equal to the analytic relaxation time the error becomes very small, as the time integration scheme becomes exact \citep{mickens2002nonstandard}.
 (b) Convergence test with timestep size for the quasi-implicit scheme.
@@
 will result in significant computational savings.
-\timo{Change the figure into one with number of unknowns on the x axis and combine into a single figure. Your computed reference value is also not accurate enough. (I can take a look at that)}
 
@@
 \includegraphics[width=0.9\textwidth]{figures/amr.pdf}
-\caption{Convergence with resolution for the \citet{crameri2012comparison} Case 2 benchmark, for both uniform (a) and adaptive (b) mesh refinement.
-The minimum cell size h is normalized by the model depth, and the timestep $\Delta t$ is 500 yr.
+\caption{Convergence with degrees of freedom (DoFs) for the \citet{crameri2012comparison} Case 2 benchmark, for both uniform and adaptive mesh refinement.
+The timestep $\Delta t$ is 500 yr.
 We compare the maximum topography at 3 Myr with its value at high resolution ($\sim$398 m).
-The points are labeled with the approximate number of degrees of freedom (DoFs) for that simulation.
+For the adaptive cases we perform mesh refinement according to the sum of density and composition gradients.
 Both cases converge similarly, but the adaptive mesh refinement simulations require significantly smaller system sizes.}

### 2015-11-02 figures/amr.py

@@ -10,7 +10,7 @@ except AttributeError:
 except ValueError:
   print "Cannot find the requested stylesheet"
 
-plt.figure(figsize=(8,4) )
+plt.figure(figsize=(4,4) )
 
 
 data_uniform_qi = np.loadtxt("../benchmarks/amr/uniform_qi/resolution_convergence_uniform_qi.txt")
@@ -19,49 +19,46 @@ data_adaptive_qi = np.loadtxt("../benchmarks/amr/adaptive_qi/resolution_converge
 data_adaptive_nsfd = np.loadtxt("../benchmarks/amr/adaptive_nsfd/resolution_convergence_adaptive_nsfd.txt")
 
 refinement_levels = np.linspace(4., 8., 5)
-h_value  =  (2.)**(-refinement_levels)
+h_value  =  (2.)**(-refinement_levels)[:-1]
 
 topography_at_3Ma = 397.6
 
-error_uniform_qi = np.abs(data_uniform_qi[:,1] - topography_at_3Ma)
-error_uniform_nsfd = np.abs(data_uniform_nsfd[:,1] - topography_at_3Ma)
-error_adaptive_qi = np.abs(data_adaptive_qi[:,1] - topography_at_3Ma)
-error_adaptive_nsfd = np.abs(data_adaptive_nsfd[:,1] - topography_at_3Ma)
+# Get the error, less the point at the highest refinement level,
+# where the convergence sort of falls apart, presumably due to
+# loss of accuracy due to something else in the solution.
+error_uniform_qi = np.abs(data_uniform_qi[:,1] - topography_at_3Ma)[:-1]
+error_uniform_nsfd = np.abs(data_uniform_nsfd[:,1] - topography_at_3Ma)[:-1]
+error_adaptive_qi = np.abs(data_adaptive_qi[:,1] - topography_at_3Ma)[:-1]
+error_adaptive_nsfd = np.abs(data_adaptive_nsfd[:,1] - topography_at_3Ma)[:-1]
+
+# Normalize the errors
+error_uniform_qi /= topography_at_3Ma
+error_uniform_nsfd /= topography_at_3Ma
+error_adaptive_qi /= topography_at_3Ma
+error_adaptive_nsfd /= topography_at_3Ma
+
+# Read from console outputs rather than parsing... oh well.
[…]

### 2015-11-02 free-surface-paper.tex
Unicode issue
Add a couple of sentences on comparison

@@
 Geodynamics initiative (CIG), through the National Science Foundation under Award No. EAR-0949446 
-and The University of California—Davis, and National Science Foundation grant DMS1522191.}
+and The University of California-Davis, and National Science Foundation grant DMS1522191.}
 
@@
 Figure~\ref{fig:timestep_convergence}b shows the convergence of the quasi-implicit scheme with timestep.
-When $\theta=0.0$, it corresponds to the unstabilized forward Euler scheme, and is first order in time. 
-When $\theta=1.0$ it is also first order in time, but allows for a much larger timestep.  When $\theta=0.5$
+When $\theta=0$, it corresponds to the unstabilized forward Euler scheme, and is first order in time. 
+When $\theta=1$ it is also first order in time, but allows for a much larger timestep.  When $\theta=0.5$
 the quasi-implicit guess for the body force is good enough that it actually seems to achieve the second-order 
@@
 the quasi-implicit stabilization scheme proposed by \citet{kaus2010stabilization}.
-The relaxation time of each mode is lengthened by an amout $\theta \Delta t$, and the
+The relaxation time of each mode is lengthened by an amount $\theta \Delta t$, and the
 maximum allowable timestep is correspondingly lengthened.
+
+It is not clear that the non-standard finite difference scheme is superior to 
+the quasi-implicit scheme. For $\theta=0.5$ the quasi-implicit scheme is more accurate,
+but is only conditionally stable. For $\theta=1$ the two schemes are comparably accurate 
+in the tests we ran.
 

### 2015-11-02 free-surface-paper.tex
Add some text on refinement

@@
 boundary upwards (for the full setup, see \citet{crameri2012comparison}). 
-Figure~\ref{fig:amr} shows the convergence of the maximum topography at 3 Myr to its value in a high resolution simulation.
-Figure~\ref{fig:amr}a shows the convergence with uniform refinement, and~\ref{fig:amr}b shows the convergence
-with adaptive mesh refinement. Both have essentially the same behavior, but the adaptive case requires 
+Figure~\ref{fig:amr} shows the convergence of the maximum topography at 3 Myr to its value in a high resolution simulation,
+both with and without adaptive mesh refinement.
+For the uniform refinement cases each point is generated by running the benchmark with a different global refinement level.
+For the adaptive case each point is generated by allowing the mesh to be refined, where the maximum refinement level
+is limited to the same refinement level of the corresponding global refinement simulation.
+We refine every ten timesteps according to gradients in the density and compositional fields.
+
+The convergence with and without adaptive mesh refinement have essentially the same behavior, but the adaptive case requires 
 fewer degrees of freedom by a factor of approximately an order of magnitude. More complex models will have

### 2015-11-02 figures/timestep_convergence.py
Remove point where convergence limit has been hit

@@ -23,8 +23,10 @@ relaxation_times.sort()
 
 vals = []
 for r in relaxation_times:
-  tsteps = np.ma.masked_where( data[:,0] != r, data[:,1])
-  errors = np.ma.masked_where( data[:,0] != r, data[:,2])
+  # Remove last point, where limit of analytical solution is probably reached
+  tsteps = np.ma.compressed(np.ma.masked_where( data[:,0] != r, data[:,1]))[:-1]
+  errors = np.ma.compressed(np.ma.masked_where( data[:,0] != r, data[:,2]))[:-1] 
+  print tsteps,errors
   vals.append( (tsteps, errors ) )
 
 plt.subplot(121)
@@ -41,6 +43,7 @@ plt.text(0.1, 0.0005, r'$O(\Delta t)$', fontsize=9)
 
 plt.xlabel(r'$\Delta t / \tau_\mathrm{min}$')
 plt.ylabel(r'Error')
+plt.xlim(0.01, 2)
 plt.legend(loc='lower right', fontsize=8)
 plt.title('(a) Non-standard FD')
 
@@ -56,8 +59,9 @@ plt.subplot(122)
 
 vals = []
 for theta in thetas:
-  tsteps = np.ma.masked_where( data[:,0] != theta, data[:,1])
-  errors = np.ma.masked_where( data[:,0] != theta, data[:,2])
+  # Remove last point, where limit of analytical solution is probably reached
+  tsteps = np.ma.compressed(np.ma.masked_where( data[:,0] != theta, data[:,1]))[:-1]
+  errors = np.ma.compressed(np.ma.masked_where( data[:,0] != theta, data[:,2]))[:-1]
   vals.append( (tsteps, errors ) )
 
 marker = itertools.cycle( ('-h', '-v', '-^', '-*', '-D', '-s', '-o') )
@@ -74,6 +78,7 @@ plt.loglog(x2,y2, 'k--')
 plt.text(0.1, 0.05, r'$O(\Delta t)$', fontsize=9)
 plt.text(0.3, 0.0005, r'$O(\Delta t^2)$', fontsize=9)
 
+plt.xlim(0.01, 2)
 plt.xlabel(r'$\Delta t / \tau_\mathrm{min}$')
 plt.legend(loc='lower right', fontsize=8)
 plt.title('(b) Quasi-implicit')

### 2015-11-03 figures/timestep_convergence.py
Remove debug out

@@ -26,7 +26,6 @@ for r in relaxation_times:
   # Remove last point, where limit of analytical solution is probably reached
   tsteps = np.ma.compressed(np.ma.masked_where( data[:,0] != r, data[:,1]))[:-1]
   errors = np.ma.compressed(np.ma.masked_where( data[:,0] != r, data[:,2]))[:-1] 
-  print tsteps,errors
   vals.append( (tsteps, errors ) )
 
 plt.subplot(121)

### 2015-11-03 free-surface-paper.tex
Minor changes for clarification

@@
 
-It is helpful to take a closer look at the pseudo-timestep $\tau^*(1-e^{-\Delta t/\tau^*})$.
+It is helpful to take a closer look at the pseudo-timestep introduced in Equation~\eqref{eq:nsfd}: $\tau^*(1-e^{-\Delta t/\tau^*})$.
 As the timestep $\Delta t$ goes to zero, the pseudo timestep approaches $\Delta t$, recovering 
-the forward Euler scheme. However, as $\Delta t$ gets larger, the pseudo timestep does not 
+the forward Euler scheme. However, as $\Delta t$ gets larger, the pseudo-timestep does not 
 increase as quickly, reflecting the decaying nature of the normal modes.
@@
 \begin{aligned}
+&\mathbf{u}_\mathrm{mesh} = 0 & \qquad & \textrm{on } \Gamma_0. \\
 &\mathbf{u}_\mathrm{mesh} = \left( \textbf{u} \cdot \textbf{n} \right) \textbf{n} & \qquad & \textrm{on } \Gamma_F, \\
-&\mathbf{u}_\mathrm{mesh} \cdot \textbf{n} = 0 & \qquad & \textrm{on } \partial \Omega_{\textrm{free slip}}, \\
-&\mathbf{u}_\mathrm{mesh} = 0 & \qquad & \textrm{on } \Gamma_0.
+&\mathbf{u}_\mathrm{mesh} \cdot \textbf{n} = 0 & \qquad & \textrm{on } \Gamma_{FS}, \\
 \end{aligned}
@@
 \end{equation}
+where $\Gamma_{FS}$ is the part of the boundary with free slip boundary conditions $\mathbf{u \cdot n} = 0$.
 This scheme has the advantage of working for many different domain geometries and combinations of boundary conditions.

### 2015-11-05 figures/amr.py
make work with old matplotlib

@@ -58,7 +58,11 @@ plt.ylim(1.e-3, 0.3)
 plt.xlim(1.e4, 3.e6)
 plt.xlabel(r'DoFs')
 plt.ylabel(r'Relative topography error')
-plt.legend(loc='upper right', fontsize=8)
+try:
+  plt.legend(loc='lower right')
+  plt.legend(loc='lower right', fontsize=8)
+except:
+  pass
 
 #plt.show()
 plt.savefig('amr.pdf')

### 2015-11-05 figures/timestep_convergence.py

@@ -43,7 +43,11 @@ plt.text(0.1, 0.0005, r'$O(\Delta t)$', fontsize=9)
 plt.xlabel(r'$\Delta t / \tau_\mathrm{min}$')
 plt.ylabel(r'Error')
 plt.xlim(0.01, 2)
-plt.legend(loc='lower right', fontsize=8)
+try:
+  plt.legend(loc='lower right')
+  plt.legend(loc='lower right', fontsize=8)
+except:
+  pass
 plt.title('(a) Non-standard FD')
 
 
@@ -79,7 +83,12 @@ plt.text(0.3, 0.0005, r'$O(\Delta t^2)$', fontsize=9)
 
 plt.xlim(0.01, 2)
 plt.xlabel(r'$\Delta t / \tau_\mathrm{min}$')
-plt.legend(loc='lower right', fontsize=8)
+try:
+  plt.legend(loc='lower right')
+  plt.legend(loc='lower right', fontsize=8)
+except:
+  pass
+
 plt.title('(b) Quasi-implicit')
 
 

### 2015-11-05 figures/amr.py
Include style sheet in repo

@@ -4,7 +4,7 @@ import numpy.ma
 import itertools
 
 try:
-  plt.style.use('ian')
+  plt.style.use('./free_surface_paper.mplstyle')
 except AttributeError:
   print "old matplotlib version?"
 except ValueError:

### 2015-11-05 figures/stability_region.py

@@ -2,7 +2,7 @@ import matplotlib.pyplot as plt
 import numpy as np
 
 try:
-  plt.style.use('ian')
+  plt.style.use('./free_surface_paper.mplstyle')
 except AttributeError:
   print "old matplotlib version?"
 except ValueError:

### 2015-11-06 free-surface-paper.tex
add section overview

@@
 
-Here we analyze the problem of the sloshing instability.  We demonstrate the nature of the instability, 
-as well as clarifying the effect of quasi-implicit stabilization.
-The analysis allows us to propose a new time marching scheme for free surface computations with 
-good stability properties. It is an explicit method which requires no modification to the system matrix,
-which makes it extremely simple to implement.
-
-Finally, we describe the implementation of a free surface in the mantle convection software \texttt{ASPECT}.
+The paper is organized as follows.
+After introducing the problem in Section~\ref{sec:governing},
+we derive an approach to analyze 
+free surface schemes based on the normal modes in Section~\ref{sec:eigenvalue}
+and Section~\ref{sec:timestepping}.
+In Section~\ref{sec:kmm} we use this approach to look at the
+quasi-implicit stabilization proposed in \citet{kaus2010stabilization}.
+Then, we propose a new time marching scheme for free surface computations with 
+good stability properties (Section~\ref{sec:newscheme}).
+Finally, we describe the implementation of a free surface in the mantle convection software \texttt{ASPECT} in Section~\ref{sec:implementation}, before showing numerical results
+in \ref{sec:results}.
 
@@
 
-\section{Quasi-implicit stabilization}
+\section{Analysis of quasi-implicit stabilization}
 \label{sec:kmm}
@@
 \section{A novel time-integration scheme}
+\ref{sec:newscheme}
 
@@
 \section{Results}
+\label{sec:results}
 

### 2015-11-06 free-surface-paper.tex
Fix references

@@
 Finally, we describe the implementation of a free surface in the mantle convection software \texttt{ASPECT} in Section~\ref{sec:implementation}, before showing numerical results
-in \ref{sec:results}.
+in Section~\ref{sec:results}.
 
@@
 \section{A novel time-integration scheme}
-\ref{sec:newscheme}
+\label{sec:newscheme}
 

### 2015-11-06 free-surface-paper.tex
formula cleanup, minor text changes

@@
 \begin{equation}
-\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p^\prime = 0
+\begin{aligned}
+\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p^\prime &= 0 \\
+\nabla \cdot \mathbf{u} &= 0
+\end{aligned}
 \label{eq:homogeneous_stokes}
-\end{equation}
-
-\begin{equation}
-\nabla \cdot \mathbf{u} = 0
-\label{eq:incompressible}
 \end{equation}
@@
 It is most convenient to integrate over the domain at the current timestep $\Omega^n$ 
-(where the superscript indicates timestep number), which corresponds to an explicit scheme.
-However, as we will see, the implicit scheme of integrating over the (unknown) domain at the next timestep $\Omega^{n+1}$
-is much more stable, at the cost of making the problem nonlinear \citep{furuichi2015implicit}.
+(where the superscript indicates timestep number) instead of 
+the at this point unknown $\Omega^{n+1}$. This corresponds to an explicit scheme,
+which is not stable.
+However, as we will see, the implicit scheme of integrating over the (unknown) domain at the correct time $\Omega^{n+1}$ will
+give a stable scheme, at the cost of making the problem nonlinear \citep{furuichi2015implicit}.
 
@@
 \begin{equation}
-\frac {\text{d}}{\text{d} t} \zeta_i(\mathbf{x},t) = \frac{\text{d}}{\text{d}t} \zeta_i(\mathbf{x})e^{-t/\tau_i} = -\frac{\zeta_i(\mathbf{x},t)}{\tau_i} = \mathbf{u}_i \cdot \mathbf{n}
+\frac {\text{d}}{\text{d} t} \zeta_i(\mathbf{x},t) = \frac{\text{d}}{\text{d}t} \zeta_i(\mathbf{x})e^{-t/\tau_i} = -\frac{\zeta_i(\mathbf{x},t)}{\tau_i} = \mathbf{u}_i \cdot \mathbf{n}.
 \end{equation}
@@
 -\int_{\Omega(t)} 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u}_i ) + \int_{\Omega(t)} p^\prime_i \nabla \cdot \mathbf{w} 
-= \tau_i \int_{\Gamma_F(t)} \rho_0 g (\mathbf{u}_i \cdot \mathbf{n} ) (\mathbf{w} \cdot \mathbf{n})
+= \tau_i \int_{\Gamma_F(t)} \rho_0 g (\mathbf{u}_i \cdot \mathbf{n} ) (\mathbf{w} \cdot \mathbf{n}).
 \label{eq:weak_eigen}
@@
 \begin{equation}
-\zeta( \mathbf{x}, t) = \displaystyle \sum_i a_i \zeta_i(\mathbf{x}) e^{-t/\tau_i}
+\zeta( \mathbf{x}, t) = \displaystyle \sum_i a_i \zeta_i(\mathbf{x}) e^{-t/\tau_i}.
 \end{equation}
-
 Of course, most finite element (or finite difference, or finite volume) geodynamic simulations do not resolve 
@@
 \begin{equation}
-\Delta t  \le 2 \tau_{\mathrm{min}}
+\Delta t  \le 2 \tau_{\mathrm{min}}.
 \label{eq:cfl_euler}
@@
 \end{equation}
-The volume integral is the same as that of the unstabilized problem, but now there is a surface integral.
-The surface integral has the form of a velocity-dependent surface stress pushing down on the 
+The volume integral is the same as that of the unstabilized problem, but now we obtained an additional surface integral.
+It has the form of a velocity-dependent surface stress pushing down on the 
 domain, and can be thought of as an artificial viscous damping of the surface.
-Since it depends on the velocity, it can no longer go on the right-hand side of the 
-equation, and must be added as a stabilization term in the system matrix.  
+Since the term depends on the velocity, it 
+enters the system matrix as a stabilization term.  
 Empirically it has been found to be successful at damping instabilities \citep{kaus2010stabilization, quinquis2011role, duretz2011discretization}.
@@
 a new generalized eigenvalue problem:
-
 \begin{equation}
@@
 p^n_i
-\end{bmatrix}
+\end{bmatrix}.
 \label{eq:rearranged_stabilized_generalized_eigenvalue}
@@
 \begin{equation}
-\Delta t  \le \frac{2}{1-\theta} \tau_{\mathrm{min}}
+\Delta t  \le \frac{2}{1-\theta} \tau_{\mathrm{min}}.
 \label{eq:cfl_euler_stabilized}
@@
 \begin{equation}
-\frac{\text{d} a_i} {\text{d}t} = - \frac{ a_i }{\tau_i} 
+\frac{\text{d} a_i} {\text{d}t} = - \frac{ a_i }{\tau_i} .
 \end{equation}
@@
 \begin{equation}
-a_i^{n} = a_i^{n+1} \left[ 1 - \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \right]
+a_i^{n} = a_i^{n+1} \left[ 1 - \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \right].
 \label{eq:recursion}
@@
 \begin{equation}
-\left| 1 - \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \right| \le 1
+\left| 1 - \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \right| \le 1.
 \end{equation}
@@
 \begin{equation}
-\tau^* \le 2 \tau_i
+\tau^* \le 2 \tau_i.
 \end{equation}
@@
 \begin{equation}
-\Delta t \le -\tau^* \log \left(1 - 2 \frac{\tau_i}{\tau^*} \right)
+\Delta t \le -\tau^* \log \left(1 - 2 \frac{\tau_i}{\tau^*} \right).
 \end{equation}
@@
 a_i(\Delta t) - a_i^{1} &= a_i(0) e^{-\Delta t/\tau_i} - a_i{(0)} \left[ 1 - \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \right] \\
-                        &= a_i{(0)} \left[ e^{-\Delta t/\tau_i} - 1 + \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \right] \\
+                        &= a_i{(0)} \left[ e^{-\Delta t/\tau_i} - 1 + \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \right].
 \end{aligned}
@@
 \begin{equation}
-a_i(\Delta t) - a_i^{1} = - a_i{(0)} \left[ \left(\frac{\Delta t}{\tau_i}\right)^2 - \frac{\tau^*}{\tau_i} \left(\frac{\Delta t }{\tau^* }\right)^2 \right] 
+a_i(\Delta t) - a_i^{1} = - a_i{(0)} \left[ \left(\frac{\Delta t}{\tau_i}\right)^2 - \frac{\tau^*}{\tau_i} \left(\frac{\Delta t }{\tau^* }\right)^2 \right].
 \end{equation}
@@
 \begin{equation}
-a_i(\Delta t) - a_i^{1} = {a_i{(0)} } \left( \frac{\Delta t}{\tau_i} \right)^2 \left( 1 - \frac{\tau_i}{\tau^*} \right) + O(\Delta t^3)
+a_i(\Delta t) - a_i^{1} = {a_i{(0)} } \left( \frac{\Delta t}{\tau_i} \right)^2 \left( 1 - \frac{\tau_i}{\tau^*} \right) + O(\Delta t^3).
 \end{equation}
@@
 \begin{equation}
-\tau^* \approx \tau_{\mathrm{min}}
+\tau^* \approx \tau_{\mathrm{min}}.
 \label{eq:tau_choice}
@@
 
-\section{Results}
+\section{Numerical Results}
 \label{sec:results}
@@
 \begin{equation}
-\mathrm{Error} = \frac{1}{4\tau}\int_0^{4 \tau} \lVert \zeta_\mathrm{numeric}(L/2, t) - \zeta_\mathrm{analytic}(L/2, t) \rVert_2 \text{ d}t
+\mathrm{Error} = \frac{1}{4\tau}\int_0^{4 \tau} \lVert \zeta_\mathrm{numeric}(L/2, t) - \zeta_\mathrm{analytic}(L/2, t) \rVert_2 \text{ d}t.
 \end{equation}
@@
 %%
+\section{Bibliography}
 \bibliographystyle{elsarticle-harv} 

### 2015-11-06 figures/amr.py
Change location of legend

@@ -59,8 +59,8 @@ plt.xlim(1.e4, 3.e6)
 plt.xlabel(r'DoFs')
 plt.ylabel(r'Relative topography error')
 try:
-  plt.legend(loc='lower right')
-  plt.legend(loc='lower right', fontsize=8)
+  plt.legend(loc='upper right')
+  plt.legend(loc='upper right', fontsize=8)
 except:
   pass
 

### 2015-11-06 free-surface-paper.tex
Fix typo
Add keywords
Fix up affiliations and acknowledgements

@@
 
-\usepackage{todonotes}
-\newcommand\timo[1]{\todo[inline]{Timo: #1}}
-
 \newif\ifdetail
@@
 %% \usepackage{lineno}
+
+%% Block of code for fixing corresponding author bug 
+%% in elsarticle template... Don't really understand it
+\makeatletter
+\def\@author#1{\g@addto@macro\elsauthors{\normalsize%
+    \def\baselinestretch{1}%
+    \upshape\authorsep#1\unskip\textsuperscript{%
+      \ifx\@fnmark\@empty\else\unskip\sep\@fnmark\let\sep=,\fi
+      \ifx\@corref\@empty\else\unskip\sep\@corref\let\sep=,\fi
+      }%
+    \def\authorsep{\unskip,\space}%
+    \global\let\@fnmark\@empty
+    \global\let\@corref\@empty  %% Added
+    \global\let\sep\@empty}%
+    \@eadauthor={#1}
+}
+\makeatother
+%% End block of weird code
 
@@
 
-\author{I. Rose, B. Buffett, and Timo Heister\fnref{ref3}}
-
-\fntext[ref3]{\url{heister@clemson.edu}, partially supported by the Computational Infrastructure in
-Geodynamics initiative (CIG), through the National Science Foundation under Award No. EAR-0949446 
-and The University of California-Davis, and National Science Foundation grant DMS1522191.}
+\author{Ian Rose\corref{cor1}\fnref{ref1}}
+\author{Bruce Buffett\fnref{ref1}}
+\author{Timo Heister\fnref{ref2}}
+
+\fntext[ref1]{University of California, Berkeley}
+\fntext[ref2]{Clemson University}
+\cortext[cor1]{Corresponding author, \url{ian.rose@berkeley.edu}}
 
@@
 schemes work to damp them out.
-We also propose a new scheme for removing instabilites from free surface calculations. 
+We also propose a new scheme for removing instabilities from free surface calculations. 
 It does not require modifications to the system matrix, nor additional variables, but is instead
@@
 \begin{keyword}
+Numerical modeling \sep
+Free surface \sep
+Mantle convection \sep
+Geodynamics
 %% keywords here, in the form: keyword \sep keyword
@@
 
+\section*{Acknowledgements}
+This work was supported through National Science Foundation grant EAR-1246670
+and by the Computational Infrastructure in Geodynamics (CIG), through National Science Foundation grant EAR-0949446. 
+TH is partially supported by NSF grant DMS-1522191.
+Thanks to Wolfgang Bangerth and Brent Delbridge for useful discussions.
 %% The Appendices part is started with the command \appendix;
@@
 %%
-\section{Bibliography}
+\section*{Bibliography}
 \bibliographystyle{elsarticle-harv} 

### 2015-11-16 free-surface-paper.tex
cite deal.II and ASPECT

@@
 \label{sec:implementation}
-We have implemented the ability to run free surface simulations in the mantle convection software \texttt{ASPECT} \citep{kronbichler2012high}. 
-\texttt{ASPECT} is designed to be highly flexible and modular, with the ability for user-defined rheologies, geometries, and gravity models. 
+We have implemented the ability to run free surface simulations in the mantle convection software \texttt{ASPECT} \citep{aspectweb,kronbichler2012high}. 
+\texttt{ASPECT}, based on the open source, finite element library deal.II \citep{dealII82}, is designed to be highly flexible and modular, with the ability for user-defined rheologies, geometries, and gravity models.
 The implementation of the free surface, therefore, needs to be general enough to work for many combinations of these models, including those 
@@
 
-\texttt{ASPECT} is a parallelized, distributed memory code with adaptive mesh refinement.
+\texttt{ASPECT} is a parallelized, distributed memory code with adaptive mesh refinement (see \cite{BBHK10} for details).
 The free surface implementation works with these features.  We store the mesh vertex positions 

### 2015-12-10 figures/stability_region.py
Label stability regions

@@ -26,6 +26,8 @@ plt.xlabel(r'$\tau^*/\tau_{\mathrm{min}}$', fontsize=fontsize)
 plt.ylabel(r'$\frac{\Delta t}{\tau_{\mathrm{min}}}$', rotation=0, fontsize=fontsize)
 plt.fill_between(x_nsfd, 0, y_nsfd, color='green', alpha=0.2)
 plt.fill_between(x_nsfd, 0, y_fe, color='blue', alpha=0.2)
+plt.text(0.5, 3., r'Stable', fontsize=fontsize)
+plt.text(3., 6., r'Unstable', fontsize=fontsize)
 plt.title('(a) Non-standard FD', fontsize=fontsize)
 plt.subplot(122)
 plt.xlim(0.,1.)
@@ -33,6 +35,8 @@ plt.ylim(0.,ymax)
 plt.xlabel(r'$\theta$', fontsize=fontsize)
 plt.fill_between(x_kmm, 0, y_kmm, color='green', alpha=0.2)
 plt.fill_between(x_nsfd, 0, y_fe, color='blue', alpha=0.2)
+plt.text(0.7, 3., r'Stable', fontsize=fontsize)
+plt.text(0.1, 6., r'Unstable', fontsize=fontsize)
 plt.title('(b) Quasi-implicit', fontsize=fontsize)
 #plt.show()
 plt.savefig('stability_region.pdf')

### 2016-04-04 free-surface-paper.tex
Add Cedric to acknowledgements
Address review one comments.

@@
 In particular, free surfaces tend to suffer from sloshing instabilities, also known as 
-the ``drunken sailor'' instability, which severely limit time step sizes. Several 
+the ``drunken sailor'' instability, which severely limit timestep sizes. Several 
 schemes have been proposed in the literature to deal with these instabilities.
@@
 instability. Since the most expensive part of geodynamic simulations is typically the Stokes solve,
-most free surface calculations have preferred to use explicit time stepping methods for the free 
+most free surface calculations have preferred to use explicit timestepping methods for the free 
 surface. \citet{kaus2010stabilization} proposed a quasi-implicit scheme which modifies the discretized 
@@
 \end{equation}
-where $\eta$ is the viscosity and $\varepsilon(\mathbf{u}) = \frac{1}{2}(\nabla \mathbf{u} + (\nabla \mathbf{u} )^T )$ is the strain-rate tensor.
+where $\eta$ is the viscosity, $\varepsilon(\mathbf{u}) = \frac{1}{2}(\nabla \mathbf{u} + (\nabla \mathbf{u} )^T )$
+is the strain-rate tensor, and $\mathbf{I}$ is the identity tensor.
 Substituting the stress tensor into Equation~\eqref{eq:momentum_mass_conservation} gets the familiar form of the Stokes equation:
@@
 \end{equation}
-The mesh velocity in normal direction is given by the velocity $\mathbf{u}(t)$ on the free surface
+The mesh velocity in normal direction (with unit normal $\mathbf{n}$) is given by the velocity $\mathbf{u}(t)$ on the free surface
 \begin{equation}
@@
 is not invertible. It may, however, be transformed into a standard eigenvalue problem.
-Define
+By defining
 \begin{equation}
@@
 \end{equation}
-Then multiplying both sides by $\tau_i^{-1}C^{-1}$ we get
+and then multiplying both sides by $\tau_i^{-1}C^{-1}$ we get
 \begin{equation}
@@
 \begin{equation}
-\Delta t  \le 2 \tau_{\mathrm{min}}.
+\Delta t  \le 2 \tau_{\mathrm{min}},
 \label{eq:cfl_euler}
 \end{equation}
+where $\tau_{\mathrm{min}}$ is this minumum decay time.
 The maximum allowable timestep is limited by the minimum relaxation timescale.
@@
 Again we may take one of the normal modes as an example and inspect the difference between the nonstandard finite-difference
-scheme and the analytical solution after one time step:
+scheme and the analytical solution after one timestep:
 
@@
 \item Analytical formulae. Several geometry and viscosity model combinations have analytical solutions
-for relaxation spectra. Even if the user's model is not precisely the same (e.g., has some lateral viscosty
+for relaxation spectra. Even if the user's model is not precisely the same (e.g., has some lateral viscosity
 variations), an analytical approximation may be sufficient.
 \item Scaling. In general, we expect the relaxation times to scale with $\tau \sim \frac{\eta}{\rho g L}$,
-where the density, gravity, viscosity, and lengthscale are all representative values.  A small amount
+where the density, gravity, viscosity, and lengthscale $L$ are all representative values.  A small amount
 of experimentation near this value of $\tau$ can find an appropriate relaxation time.
@@
 result in poor overall mass conservation in time. In some scenarios, the total volume of the mesh can 
-fluctuate significantly over hundreds or thousands of time steps.
+fluctuate significantly over hundreds or thousands of timesteps.
 Consistency with the Stokes solution requires 
@@
 to quadratic \citep{kramer2012implicit, furuichi2015implicit}.
-We evaluate error by time-integrating the L2 difference between the numerical and analytical
-solutions at the center of the domain:
+We estimate the error by time-integrating the L2 difference between the numerical and analytical
+solutions at the center point of the free surface:
 \begin{equation}
@@
 For the adaptive cases we perform mesh refinement according to the sum of density and composition gradients.
-Both cases converge similarly, but the adaptive mesh refinement simulations require significantly smaller system sizes.}
+Quasi-implicit and nonstandard finite difference timestepping converge similarly, but the adaptive mesh refinement simulations require significantly smaller system sizes.}
 \label{fig:amr}
@@
 scheme and the nonstandard finite difference scheme are available. The implementation is 
-sufficiently general to accommodate many different geometies, rheologies, and 
+sufficiently general to accommodate many different geometries, rheologies, and 
 gravity models. Furthermore, it runs in parallel and with adaptive mesh refinement.
@@
 TH is partially supported by NSF grant DMS-1522191.
-Thanks to Wolfgang Bangerth and Brent Delbridge for useful discussions.
+Thanks to Wolfgang Bangerth, Brent Delbridge, and Cedric Thieulot for useful discussions.
 %% The Appendices part is started with the command \appendix;

### 2016-04-04 free-surface-paper.tex
Use line numbering
Make equation references consistent.

@@
 
+\usepackage{lineno}
+    \linenumbers
+
 \newif\ifdetail
@@
 
-In order to better understand the time evolution of the system in \eqref{eq:final_system}
+In order to better understand the time evolution of the system in Equation~\eqref{eq:final_system}
 we will consider the eigenvalues of a simplified homogeneous system. By setting $\rho = \rho_0$, we obtain the system
@@
 Armed with the normal modes and relaxation times of the Stokes system, we can write down the
-formal solution to Equation~\ref{eq:surface_evolution}. Let the initial surface topography be 
+formal solution to Equation~\eqref{eq:surface_evolution}. Let the initial surface topography be 
 represented by a linear combination of its normal modes
@@
 \end{equation}
-The stability region is plotted in Figure \ref{fig:stability_region}.
+The stability region is plotted in Figure~\ref{fig:stability_region}.
 
@@
 only the minimun relaxation time is required. For certain geometries a simple power iteration on 
-the standard eigenvalue problem~\eqref{eq:standard_eigenvalue} could be enough.
+the standard eigenvalue problem (Equation~\eqref{eq:standard_eigenvalue}) could be enough.
 \item Analytical formulae. Several geometry and viscosity model combinations have analytical solutions
@@
 solution onto the mesh velocity vector. Multiplying the boundary conditions 
-\eqref{eq:laplacian_bcs} by a test function $\mathbf{w}$ and integrating over the free
+(Equation~\eqref{eq:laplacian_bcs}) by a test function $\mathbf{w}$ and integrating over the free
 surface part of the boundary, we find
@@
 
-This weak solution to the boundary conditions~\eqref{eq:laplacian_bcs} is able to borrow
+This weak solution to the boundary conditions (Equation~\eqref{eq:laplacian_bcs}) is able to borrow
 the accuracy of the Stokes solve for Equation~\eqref{eq:weak_incompressible}, and we have 

### 2016-04-04 free-surface-paper.tex
Punctuate all equations consistently.

@@
 \nabla \cdot \mathbf{T} &= \rho \mathbf{g} \\
-\nabla \cdot \mathbf{u} &= 0
+\nabla \cdot \mathbf{u} &= 0,
 \end{aligned}
@@
 \begin{equation}
-\mathbf{T} = 2 \eta \varepsilon(\mathbf{u}) - p \mathbf{I}
+\mathbf{T} = 2 \eta \varepsilon(\mathbf{u}) - p \mathbf{I},
 \label{eq:stress_tensor}
@@
 \begin{equation}
-\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p = \rho \mathbf{g}
+\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p = \rho \mathbf{g}.
 \label{eq:stokes}
@@
 \begin{equation}
-- \nabla p_0 = \rho_0 \mathbf{g}
+- \nabla p_0 = \rho_0 \mathbf{g},
 \label{eq:hydrostatic_stokes}
@@
 \nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p^\prime &= 0 \\
-\nabla \cdot \mathbf{u} &= 0
+\nabla \cdot \mathbf{u} &= 0.
 \end{aligned}
@@
 \begin{equation}
-\int_{\Omega(t)} q \nabla \cdot \mathbf{u} = 0
+\int_{\Omega(t)} q \nabla \cdot \mathbf{u} = 0,
 \label{eq:weak_incompressible}
@@
 \begin{equation}
-\mathbf{T} \approx \frac{\partial \mathbf{T}}{\partial \mathbf{n} } \cdot \mathbf{n} \zeta = \rho_0 g \zeta \mathbf{I}
+\mathbf{T} \approx \frac{\partial \mathbf{T}}{\partial \mathbf{n} } \cdot \mathbf{n} \zeta = \rho_0 g \zeta \mathbf{I}.
 \label{eq:hydrostatic}
@@
 -\int_{\Omega(t)} 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u} ) + \int_{\Omega(t)} p^\prime \nabla \cdot \mathbf{w} 
-+ \int_{\Gamma_F(t)} \rho_0 g \zeta  \mathbf{w} \cdot \mathbf{n} = 0 
++ \int_{\Gamma_F(t)} \rho_0 g \zeta  \mathbf{w} \cdot \mathbf{n} = 0.
 \end{equation}
@@
 p^n_i
-\end{bmatrix}
+\end{bmatrix},
 \label{eq:generalized_eigenvalue}
@@
 \begin{equation}
-(C^{-1}D)\mathbf{y}_i = \tau_i^{-1} \mathbf{y}_i
+(C^{-1}D)\mathbf{y}_i = \tau_i^{-1} \mathbf{y}_i.
 \label{eq:standard_eigenvalue}
@@
 \begin{equation}
-\zeta(\mathbf{x}, t=0) = \displaystyle \sum_i a_i \zeta_i(\mathbf{x})
+\zeta(\mathbf{x}, t=0) = \displaystyle \sum_i a_i \zeta_i(\mathbf{x}),
 \end{equation}
@@
 \begin{equation}
-\mathbf{f}_{\mathrm{body}} = \int_{\Omega^n + \Delta \Omega} \rho \mathbf{w} \cdot \mathbf{g}
+\mathbf{f}_{\mathrm{body}} = \int_{\Omega^n + \Delta \Omega} \rho \mathbf{w} \cdot \mathbf{g},
 \label{eq:predict}
@@
 \begin{equation}
-\Delta \Omega \approx \theta \Delta t \mathbf{u}
+\Delta \Omega \approx \theta \Delta t \mathbf{u},
 \end{equation}
@@
 \int_{\Omega^n + \Delta \Omega} \rho  \mathbf{w} \cdot \mathbf{g} \approx
-\int_{\Omega^n} \rho  \mathbf{w} \cdot \mathbf{g} + \theta \Delta t \int_{\Gamma_F^n} \rho ( \mathbf{w \cdot g}) (\mathbf{u \cdot n} )
+\int_{\Omega^n} \rho  \mathbf{w} \cdot \mathbf{g} + \theta \Delta t \int_{\Gamma_F^n} \rho ( \mathbf{w \cdot g}) (\mathbf{u \cdot n} ).
 \label{eq:kmm}
@@
 \begin{equation}
--\theta \Delta t \int_{\Gamma_F^n} \rho g ( \mathbf{w \cdot n}) (\mathbf{u \cdot n} )
+-\theta \Delta t \int_{\Gamma_F^n} \rho g ( \mathbf{w \cdot n}) (\mathbf{u \cdot n} ).
 \end{equation}
@@
 p^n_i
-\end{bmatrix}
+\end{bmatrix},
 \label{eq:stabilized_generalized_eigenvalue}
@@
 \begin{equation}
-\tau^S_i = \tau_i + \theta \Delta t
+\tau^S_i = \tau_i + \theta \Delta t.
 \end{equation}
@@
 \begin{equation}
-\mathbf{x}(t^{n+1}) = \mathbf{x}(t^n) + \int_{t^{n}}^{t^{n+1}} \mathbf{u}(t) \text{ d}t
+\mathbf{x}(t^{n+1}) = \mathbf{x}(t^n) + \int_{t^{n}}^{t^{n+1}} \mathbf{u}(t) \text{ d}t,
 \label{eq:time_integration}
@@
 \begin{equation}
-\mathbf{u}(t) = \mathbf{u}(t^n) e^{-(t-t^n)/\tau^*}
+\mathbf{u}(t) = \mathbf{u}(t^n) e^{-(t-t^n)/\tau^*},
 \label{eq:velocity_decay}
@@
 \begin{equation}
-\mathbf{x}(t^{n+1}) = \mathbf{x}(t^n) + \mathbf{u}(t^{n}) \tau^* \left(1-e^{-\Delta t/\tau^*} \right)
+\mathbf{x}(t^{n+1}) = \mathbf{x}(t^n) + \mathbf{u}(t^{n}) \tau^* \left(1-e^{-\Delta t/\tau^*} \right).
 \label{eq:nsfd}
@@
 \begin{equation}
-\frac{\text{d} a_i} {\text{d}t} = - \frac{ a_i }{\tau_i} .
+\frac{\text{d} a_i} {\text{d}t} = - \frac{ a_i }{\tau_i}.
 \end{equation}
@@
 \begin{equation}
-- \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \le 0
-\end{equation}
-
+- \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \le 0,
+\end{equation}
 which can be rearranged to find
 \begin{equation}
-e^{-\Delta t/\tau^*} \le 1
-\end{equation}
-
+e^{-\Delta t/\tau^*} \le 1,
+\end{equation}
 which is simply a statement that the timestep must be postive.
@@
 \end{equation}
-
 \begin{equation}
@@
 \end{equation}
-
 if the left hand side is less than zero, then this is true regardless of step size, or
@@
 \end{equation}
-
 if the left hand side is between zero and one, the expression is more complicated. 
 Taking the log of both sides:
-
-\begin{equation}
-\Delta t \le -\tau^* \log \left(1 - 2 \frac{\tau_i}{\tau^*} \right)
-\end{equation}
-
-It is convenient to write this in terms of dimensionless times $\Delta t/ \tau_i$ and $\tau^*/\tau_i$.
-
-\begin{equation}
-\frac{\Delta t}{\tau_i} \le -\frac{\tau^*}{\tau_i} \log \left(1 - 2 \frac{\tau_i}{\tau^*} \right)
+\begin{equation}
+\Delta t \le -\tau^* \log \left(1 - 2 \frac{\tau_i}{\tau^*} \right).
+\end{equation}
+It is convenient to write this in terms of dimensionless times $\Delta t/ \tau_i$ and $\tau^*/\tau_i$:
+\begin{equation}
+\frac{\Delta t}{\tau_i} \le -\frac{\tau^*}{\tau_i} \log \left(1 - 2 \frac{\tau_i}{\tau^*} \right).
 \end{equation}
@@
 \begin{equation}
-\zeta(x,0) = \zeta_0 \cos\left( k x \right)
+\zeta(x,0) = \zeta_0 \cos\left( k x \right),
 \end{equation}
@@
 \begin{equation}
-\zeta(x, t) = \zeta(x,0) e^{-t/\tau}
+\zeta(x, t) = \zeta(x,0) e^{-t/\tau},
 \end{equation}
@@
 \begin{equation}
-\tau = \frac{D k + \sinh(D k) \cosh(D k)}{ \sinh^2 (D k) } \frac{2 k \eta}{\rho g}
+\tau = \frac{D k + \sinh(D k) \cosh(D k)}{ \sinh^2 (D k) } \frac{2 k \eta}{\rho g},
 \end{equation}

### 2016-04-04 free-surface-paper.tex
Minor edits.

@@
 
-\section{Governing Equations}
+\section{Governing equations}
 \label{sec:governing}
@@
 The maximum allowable timestep is limited by the minimum relaxation timescale.
-If a larger timestep than this is taken then those modes will go unstable.
+If a larger timestep than this is taken then those modes will become unstable.
 The modes with the smallest relaxation times are usually those with the largest lengthscales \citep{schubert2001mantle}, 
-and so it will be those which go unstable first, a phenomenon which has been called 
+and so it will be those which become unstable first, a phenomenon which has been called 
 a ``sloshing'', or ``drunken sailor'' instability \citep{kaus2010stabilization}.
@@
 
-\section{Numerical Results}
+\section{Numerical results}
 \label{sec:results}
@@
 For the adaptive cases we perform mesh refinement according to the sum of density and composition gradients.
-Quasi-implicit and nonstandard finite difference timestepping converge similarly, but the adaptive mesh refinement simulations require significantly smaller system sizes.}
+Quasi-implicit (QI) and nonstandard finite difference (NSFD) timestepping converge similarly, but the adaptive mesh refinement simulations require significantly fewer unknowns.}
 \label{fig:amr}

### 2016-04-22 free-surface-paper.tex
Address some comments.

@@
 
-\usepackage{lineno}
-    \linenumbers
-
 \newif\ifdetail
@@
 %% for the whole article with \linenumbers.
-%% \usepackage{lineno}
+\usepackage{lineno}
 
@@
 have proven challenging from a standpoint of accuracy, robustness, and stability.
-In particular, free surfaces tend to suffer from sloshing instabilities, also known as 
-the ``drunken sailor'' instability, which severely limit timestep sizes. Several 
-schemes have been proposed in the literature to deal with these instabilities.
+In particular, time integration of a free surface tends to suffer from a numerical instability
+that manifests as sloshing surface motions, also known as the ``drunken sailor'' instability.
+This instability severely limits stable timestep sizes to those much smaller than could be used
+in geodynamic simulations without a free surface. 
+Several schemes have been proposed in the literature to deal with these instabilities.
 
@@
 
-%% \linenumbers
+\linenumbers
 
@@
 Historically, most simulations have been performed with free-slip boundary conditions at the surface, 
-with dynamic topography calculated as a postprocessing step \citep[e.g.][]{zhong2000role}.
+with dynamic topography calculated as a postprocessing step \citep[e.g.][]{jarvis1982mantle, zhong2000role}.
 
@@
 A large number of studies have approximated free surfaces in the interior of the domain by using the 
-`sticky air' approximation. In this approximation there is a low-viscosity, low-density layer in the fluid 
+`sticky air' approximation (see \citet{crameri2012comparison} and references therein). In this approximation there is a low-viscosity, low-density layer in the fluid 
 (termed `air', though its viscosity is much higher) above the free surface, effectively decoupling it from the boundary. 
-Typically a free-slip boundary condition is used above the sticky air, though an open boundary may be better \citep{hillebrand2014using}.
+Typically a free-slip boundary condition is used above the sticky air, though an open boundary may be preferable \citep{hillebrand2014using}.
 
@@
 surface. \citet{kaus2010stabilization} proposed a quasi-implicit scheme which modifies the discretized 
-Stokes matrix, giving it better stability properties.  \citet{kramer2012implicit} and \citet{furuichi2015implicit}
+Stokes matrix, giving it better stability properties.  \citet{popov2008slim3d}, \citet{kramer2012implicit}, and \citet{furuichi2015implicit}
 explored methods for solving the transport of the free surface implicitly.
@@
 Since the term depends on the velocity, it 
-enters the system matrix as a stabilization term.  
-Empirically it has been found to be successful at damping instabilities \citep{kaus2010stabilization, quinquis2011role, duretz2011discretization}.
-
-Indeed, using the formalism in Section~\ref{sec:eigenvalue} we can see the effect of this 
-stabilization term. On the reference surface we note that $\mathbf{g} = -g \mathbf{n}$, which
-allows us to write the stabilization term as
-\begin{equation}
--\theta \Delta t \int_{\Gamma_F^n} \rho g ( \mathbf{w \cdot n}) (\mathbf{u \cdot n} ).
-\end{equation}
-The integral here is precisely the same as that on the right hand side of Equation~\eqref{eq:weak_eigen}, which was discretized as the matrix $M$.
+enters the Stokes system matrix as a stabilization term.  
+Empirically, it has been found to be successful at damping instabilities \citep{kaus2010stabilization, quinquis2011role, duretz2011discretization}.
+
+The surface integral in Equation~\eqref{eq:kmm} is almost the same as that on the right hand side of 
+Equation~\eqref{eq:weak_eigen}, which was discretized as the matrix $M$.
+By making two approximations we can connect it with the formalism developed in Section~\ref{sec:eigenvalue}.
+First, when the current position of the surface is near to the reference surface (i.e., the slope is not too steep)
+then gravity is approximately in the direction of the surface normal: $\mathbf{g} \approx -g \mathbf{n}$.
+Furthermore, in Equation~\eqref{eq:weak_eigen} we considered response of the surface in a homogeneous system where there were no lateral density variations,
+while Equation~\eqref{eq:kmm_stabilization} uses the total density.
+In most geodynamic simulations lateral density variations are small compared to the total density.
+For the purposes of stabilization we may approximate $\rho \approx \rho_0$.
+Inserting these two approximations, we find that the stabilization term may be written
+\begin{equation}
+-\theta \Delta t \int_{\Gamma_F^n} \rho_0 g ( \mathbf{w \cdot n}) (\mathbf{u \cdot n} )
+\label{eq:kmm_stabilization}
+\end{equation}
+where the integral is now identical to that which discretizes to $M$.
 
@@
 One downside of the quasi-implicit scheme is that it requires a modification of the system matrix.
-A naive implementation of the scheme results in a slightly 
+The implementation of such a scheme results in a slightly 
 asymmetric matrix, which can be more difficult to solve, requiring either
@@
 for $\mathbf{u}$, it is formally a first-order accurate scheme. As such, it is capable of capturing arbitrary motions 
-of the free surface, but with better stability properties than forward Euler schemes.
+of the free surface, while allowing larger timesteps than forward Euler schemes.
 
@@
 This scheme has the advantage of working for many different domain geometries and combinations of boundary conditions.
-For moderate mesh deformation, the mesh stays smooth and well conditioned, though it breaks down for large deformations 
-or on non-convex domains.
+For moderate mesh deformation, the mesh stays smooth and well conditioned, though it breaks down for large deformations.
 
@@
 free surface. Of course, a new system to solve is not ideal, but this system, being nonzero 
-over only the free surface, is relatively small and easy to solve.
+over only the free surface, is relatively cheap to solve.
 
@@
 to quadratic \citep{kramer2012implicit, furuichi2015implicit}.
-We estimate the error by time-integrating the L2 difference between the numerical and analytical
+We estimate the error by time-integrating the $L_2$ difference between the numerical and analytical
 solutions at the center point of the free surface:

### 2016-04-22 free-surface-paper.tex
Add figure and discussion of power iteration.

@@
 \end{aligned}
+\label{eq:eigen_substitution}
 \end{equation}
@@
 \tau^S_i = \tau_i + \theta \Delta t.
+\label{eq:spectrum_shift}
 \end{equation}
@@
 though it can result in less accurate time-marching of the longest wavelengths of the system.
+
+\subsection{Numerical determination of $\tau_\mathrm{min}$}
+
+It can be useful to determine the minimum relaxation time $\tau_{\mathrm{min}}$ for a particular
+simulation setup, both to determine an appropriate timestep and as a check on the eigenvalue
+analysis above. As discussed in Section~\ref{sec:eigenvalue}, the generalized eigenvalue problem
+given by Equation~\eqref{eq:stabilized_generalized_eigenvalue} is difficult to solve because the matrix on
+the right-hand-side is not invertible.
+However, we may convert Equation~\eqref{eq:stabilized_generalized_eigenvalue} to a standard eigenvalue problem
+in the same manner as before. Introducing $C_S$ as the Stokes system matrix with the stabilization term
+included, and with $D$ and $\mathbf{y}_i$ the same as in Equation~\eqref{eq:eigen_substitution}, we find
+\begin{equation}
+(C_S^{-1}D)\mathbf{y}_i = \frac{1}{\tau^S_i} \mathbf{y}_i.
+\label{eq:stabilized_standard_eigenvalue}
+\end{equation}
+Several things are of note about this eigenvalue equation which help in solving it.
+The matrix $D$ is a scaled mass matrix which is only nonzero on the free surface,
+so it is mostly empty, and, in general, it is not necessary to explicitly form the
+matrix in order to apply it.
+Second, applying the inverse of $C_S$ amounts to solving the Stokes system, so any
+existing geodynamical code can already do this.
+Third, it is now an equation for the inverse of $\tau^S_i$, and so solving for the 
+minimum relaxation time is now equivalent to finding the \emph{maximum} eigenvalue
+of Equation~\eqref{eq:stabilized_standard_eigenvalue}.
+
+All of these factors mean that power iteration (see, e.g., \citet{golub2012matrix}),
+which finds the dominant eigenvalue/eigenvector pair, is an attractive option for finding $\tau_\mathrm{min}$.
+We have numerically tested the effect of quasi-implicit stabilization on a 2D Cartesian box with a free surface,
+using power iteration to find $\tau_S$.
+Figure~\ref{fig:perturb_spectrum} shows the results, where colored circles indicate the numerical
+solution and black dots show the analytical solution using equation~\eqref{eq:spectrum_shift}.
+
+\begin{figure*}
+\includegraphics[width=0.9\textwidth]{figures/perturb_spectrum.pdf}
+\caption{The effect of quasi-implicit stabilization on the minimum relaxation time of a fluid in a 2D Cartesian box with a free surface (see Figure~\ref{fig:benchmark_setup}).
+The $x$-axis shows the timestep normalized by minimum relaxation time of the unstabilized problem $\tau_\mathrm{\min}$.
+The $y$-axis shows the stabilized relaxation time $\tau_S$, also normalized by $\tau_\mathrm{min}$.
+Colored circles show the relaxation times found by numerical solution of Equation~\eqref{eq:stabilized_standard_eigenvalue} with different values of the stabilization parameter $\theta$.
+Black dots show the relaxation times calculated using Equation~\eqref{eq:spectrum_shift}.
+The stabilized timescale is lengthened for larger timesteps and for larger values of $\theta$.
+For this test case, power iteration typically attains better than 1\% accuracy in 5-10 iterations.}
+\label{fig:perturb_spectrum}
+\end{figure*}
 

### 2016-04-22 free-surface-paper.tex
Change error symbol.

@@
 
-A better approach is to perform a finite element projection of the Stokes velocity 
-solution onto the mesh velocity vector. Multiplying the boundary conditions 
+A better approach is to perform an $L_2$ projection of the normal velocity $\mathbf{u}\cdot\mathbf{n}$
+onto the free surface $\Gamma_F$. Multiplying the boundary conditions 
 (Equation~\eqref{eq:laplacian_bcs}) by a test function $\mathbf{w}$ and integrating over the free
@@
 
+\subsection{Relaxation of sinusoidal topography}
+\label{sec:topography_relaxation}
+
 We demonstrate the convergence of the nonstandard finite difference model by comparison 
@@
 to quadratic \citep{kramer2012implicit, furuichi2015implicit}.
-We estimate the error by time-integrating the $L_2$ difference between the numerical and analytical
+We estimate the error $E$ by time-integrating the $L_2$ difference between the numerical and analytical
 solutions at the center point of the free surface:
 \begin{equation}
-\mathrm{Error} = \frac{1}{4\tau}\int_0^{4 \tau} \lVert \zeta_\mathrm{numeric}(L/2, t) - \zeta_\mathrm{analytic}(L/2, t) \rVert_2 \text{ d}t.
+E = \frac{1}{4\tau}\int_0^{4 \tau} \lVert \zeta_\mathrm{numeric}(L/2, t) - \zeta_\mathrm{analytic}(L/2, t) \rVert_2 \text{ d}t.
 \end{equation}
@@
 
+\subsection{Mesh adaptivity}
+\label{sec:mesh_adaptivity}
+
 We also investigate the advantages of combining adaptive mesh refinement with free surface computations.

### 2016-04-22 free-surface-paper.tex
Start working on Rayleigh Taylor test.

@@
 
+\subsection{Rayleigh-Taylor test}
+\label{sec:rayleigh_taylor}
+
+\begin{figure*}
+\includegraphics[width=0.9\textwidth]{figures/rayleigh_taylor.pdf}
+\caption{Rayleigh-Taylor test.
+(a) Maximum depth of the drip.
+The blue line is the result using forward Euler (FE) integration of the free surface with a small 500 yr timestep, and is used as a reference solution.
+The symbols are for several different runs with different timestepping parameters, using both quasi-implicit (QI) and nonstandard finite difference (NSFD) schemes.
+As the timestepping gets smaller, both schemes converge to the reference solution (though from different directions).
+(b) Evolution of the minimum relaxation timescale over the course of the simulation.
+Every 50 timesteps we calculate the minimum relaxation timescale to observe how much it changes during the simulation.
+It does experience significant changes, but not by orders of magnitude, so an initial estimate is likely sufficient for this case.}
+\label{fig:rayleigh_taylor}
+\end{figure*}
+
 \subsection{Mesh adaptivity}

### 2016-04-27 figures/rayleigh_taylor.py
Add rayleigh_taylor script/figure

@@ -1,0 +1,63 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import matplotlib.image as mpimg
+import itertools
+
+try:
+  plt.style.use('./free_surface_paper.mplstyle')
+except AttributeError:
+  print "old matplotlib version?"
+except ValueError:
+  print "Cannot find the requested stylesheet"
+
+prefix = '../benchmarks/rayleigh_taylor/'
+
+forward_euler = np.genfromtxt(prefix+'output_small_time_step/statistics', skip_header=50, usecols=(1,15))
+nsfd_adaptive_02 = np.genfromtxt(prefix+'output_nsfd_adaptive_cfl_02/statistics', skip_header=50, usecols=(1,15))
+nsfd_adaptive_05 = np.genfromtxt(prefix+'output_nsfd_adaptive_cfl_05/statistics', skip_header=50, usecols=(1,15))
+theta_10_cfl_02 = np.genfromtxt(prefix+'output_theta_10_cfl_02/statistics', skip_header=50, usecols=(1,15))
+
+
+nsfd_adaptive_02 = np.array( [ line for line in nsfd_adaptive_02 if line[0] != 0.] )
+nsfd_adaptive_05 = np.array( [ line for line in nsfd_adaptive_05 if line[0] != 0.] )
+
+relaxation_times = []
+time = 0.
+with open(prefix+'output_nsfd_adaptive_cfl_02/log.txt') as f:
+  for l in f.readlines():
+    if '*** Timestep' in l:
+        time = float(l.split()[3][2:])/1.e6
+    if 'New relaxation' in l:
+        timescale = float(l.split()[4])/1000.
+        relaxation_times.append( [time, timescale] )
+relaxation_times = np.array(relaxation_times[3:])
+
+plt.figure(figsize=(8,4) )
+
+plt.subplot(121)
+
+marker = itertools.cycle( ('h', 'v', '^', '*', 'D', 's', 'o') )
+interval = 100
+plt.plot( forward_euler[:,0]/1.e6, -forward_euler[:,1]/1.e3, label='FE dt = 500 yr' )
[…]

### 2016-04-27 free-surface-paper.tex
Add rayleigh_taylor script/figure
Fix reference.

@@
 \end{equation}
-Summing this error over many timesteps results in a factor of $\Delta t^{-1}$, demonstrating the first-order accuracy~\citep[e.g.]{leveque2007finite}.
+Summing this error over many timesteps results in a factor of $\Delta t^{-1}$, demonstrating the first-order accuracy~\citep[e.g.][]{leveque2007finite}.
 The choice of $\tau^*$ controls the size of the coefficient on the truncation error for the scheme.
@@
 
+\subsection{Connection with Dynamic Relaxation}
+\label{sec:dynamic_relaxation}
+
 \section{Implementation in \texttt{ASPECT} }

### 2016-04-28 free-surface-paper.tex
Work on boundary conditions

@@
 \begin{aligned}
-\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p &= \rho \mathbf{g} \qquad &\text{in} \qquad &\Omega(t)\\
-\nabla \cdot \mathbf{u} &= 0  \qquad &\text{in} \qquad &\Omega(t)
+\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p &= \rho \mathbf{g} \qquad &\text{in } &\Omega(t)\\
+\nabla \cdot \mathbf{u} &= 0  \qquad &\text{in } &\Omega(t)
 \end{aligned}
@@
 split into a fixed (Dirichlet) and free surface part $\Gamma_0$ and $\Gamma_F$, respectively.
+The boundary conditions are given by
+\begin{equation}
+\begin{aligned}
+\mathbf{u} &= 0 &\qquad \text{on } &\Gamma_0 \\
+\mathbf{T}\cdot \mathbf{n} &= 0 &\qquad \text{on } &\Gamma_F.
+\end{aligned}
+\end{equation}
 For the sake of economy we neglect inhomogeneous stress boundary conditions, though it would be straightforward to include them.
+We also defer discussion of free slip boundary conditions.
 The domain at time $t$ is defined by advecting a reference configuration $\Omega_0$ by a mesh velocity $\mathbf{u}_{\mathrm{mesh}}(t)$
@@
 \begin{aligned}
-&\mathbf{u}_\mathrm{mesh} = 0 & \qquad & \textrm{on } \Gamma_0. \\
-&\mathbf{u}_\mathrm{mesh} = \left( \textbf{u} \cdot \textbf{n} \right) \textbf{n} & \qquad & \textrm{on } \Gamma_F, \\
-&\mathbf{u}_\mathrm{mesh} \cdot \textbf{n} = 0 & \qquad & \textrm{on } \Gamma_{FS}, \\
+&\mathbf{u}_\mathrm{mesh} &= 0  &\qquad \textrm{on } \Gamma_0. \\
+&\mathbf{u}_\mathrm{mesh} &= \left( \textbf{u} \cdot \textbf{n} \right) \textbf{n} & \qquad \textrm{on } \Gamma_F, \\
+&\mathbf{u}_\mathrm{mesh} \cdot \textbf{n} &= 0  &\qquad \textrm{on } \Gamma_{FS}, \\
 \end{aligned}
@@
 \end{equation}
-where $\Gamma_{FS}$ is the part of the boundary with free slip boundary conditions $\mathbf{u \cdot n} = 0$.
+where $\Gamma_{FS}$ is the part of the boundary with free slip boundary conditions, given by
+\begin{equation}
+\begin{aligned}
+ \mathbf{u \cdot n} &= 0 \\ %&\qquad \textrm{on } \Gamma_{FS} \\
+ \mathbf{T}\cdot \mathbf{n} - (\mathbf{n}\cdot\mathbf{T} \cdot \mathbf{n})\mathbf{n} &= 0 &\qquad \textrm{on } \Gamma_{FS}.
+\end{aligned}
+\label{eq:free_slip_bcs}
+\end{equation}
 This scheme has the advantage of working for many different domain geometries and combinations of boundary conditions.
@@
 \end{equation}
-where $D$ is the layer depth, $\eta$ is the viscosity, $\rho$ is the density, and $g$ is the force of gravity.
+where $D$ is the layer depth, $\eta$ is the viscosity, $\rho$ is the density, and $g$ is the acceleration due to gravity.
 
@@
 
-Figure~\ref{fig:timestep_convergence}b shows the convergence of the quasi-implicit scheme with timestep.
+Figure~\ref{fig:timestep_convergence}b shows the convergence of the quasi-implicit scheme as a function of timestep $\Delta t$.
 When $\theta=0$, it corresponds to the unstabilized forward Euler scheme, and is first order in time. 

### 2016-04-28 free-surface-paper.tex
Fix typo.
Try on a new title.

@@
 
-\title{Free surface computations in mantle convection models}
+\title{Stability and accuracy of free surface time integration in buoyancy driven viscous flows}
 
@@
 \item Direct solution of Equation~\eqref{eq:generalized_eigenvalue}. This can be expensive, although
-only the minimun relaxation time is required. For certain geometries a simple power iteration on 
+only the minimum relaxation time is required. For certain geometries a simple power iteration on 
 the standard eigenvalue problem (Equation~\eqref{eq:standard_eigenvalue}) could be enough.

### 2016-04-28 free-surface-paper.tex
Improve language for time discretization.

@@
 Note that since the free surface can move, the shape of the domain is a function of time.
-It is most convenient to integrate over the domain at the current timestep $\Omega^n$ 
-(where the superscript indicates timestep number) instead of 
-the at this point unknown $\Omega^{n+1}$. This corresponds to an explicit scheme,
-which is not stable.
-However, as we will see, the implicit scheme of integrating over the (unknown) domain at the correct time $\Omega^{n+1}$ will
-give a stable scheme, at the cost of making the problem nonlinear \citep{furuichi2015implicit}.
 
@@
 \end{equation}
-When these equations are discretized \citep[e.g.][]{kronbichler2012high} we get
+
+Equations~\eqref{eq:weak_stokes},~\eqref{eq:weak_incompressible}, and~\eqref{eq:weak_eigen}
+are continuous forms of the governing equations which must be discretized in space and time in order to be solved numerically.
+In particular, we must decide at which timestep (or combination of timesteps) we will integrate over the domain $\Omega(t)$.
+We denote discrete times by $t^n$, where $n$ is the timestep number, and $t^{n+1} = t^n+\Delta t$.
+It is most convenient to integrate over the domain at the current timestep $\Omega^n \equiv \Omega(t^n)$ instead of 
+the at this point unknown $\Omega^{n+1} \equiv \Omega(t^{n}+\Delta t)$. This corresponds to an explicit scheme.
+However, as we will see, the implicit scheme of integrating over the (unknown) domain at the next timestep $\Omega^{n+1}$ will
+give a stable timestepping scheme, at the cost of making the problem nonlinear \citep{furuichi2015implicit}.
+
+When these equations are spatially discretized using finite elements \citep[see, e.g.][]{kronbichler2012high} we get
 \begin{equation}
@@
 \begin{bmatrix}
-\mathbf{u}^n_i \\
-p^n_i
+\mathbf{u}^h_i \\
+p^h_i
 \end{bmatrix}
@@
 \begin{bmatrix}
-\mathbf{u}^n_i \\
-p^n_i
+\mathbf{u}^h_i \\
+p^h_i
 \end{bmatrix},
@@
 \end{equation}
-where $\mathbf{u}^n_i$ and $p^n_i$ are finite-dimensional representations of $\mathbf{u}_i$ and $p^\prime_i$,
+where $\mathbf{u}^h_i$ and $p^h_i$ are finite-dimensional representations of $\mathbf{u}_i$ and $p^\prime_i$,
 and $M$ is the discretization of the bilinear form on the right-hand side of Equation~\eqref{eq:weak_eigen}.
+The form of Equation~\eqref{eq:generalized_eigenvalue} does not depend on whether we choose $\Omega^n$ or $\Omega^{n+1}$
+for the domain integration.
 
@@
 \begin{bmatrix}
-\mathbf{u}^n_i \\
-p^n_i
+\mathbf{u}^h_i \\
+p^h_i
 \end{bmatrix} 
@@
 where $\tau_{\mathrm{min}}$ is this minumum decay time.
-The maximum allowable timestep is limited by the minimum relaxation timescale.
+The maximum stable timestep is limited by the minimum relaxation timescale.
 If a larger timestep than this is taken then those modes will become unstable.
@@
 The weak form of the right-hand-side body forces in the finite element discretization is then
-
-\begin{equation}
-\mathbf{f}_{\mathrm{body}} = \int_{\Omega^n + \Delta \Omega} \rho \mathbf{w} \cdot \mathbf{g},
+\begin{equation}
+\mathbf{f}_{\mathrm{body}} = \int_{\Omega^n + \Delta \Omega} \rho \mathbf{w} \cdot \mathbf{g}.
 \label{eq:predict}
 \end{equation}
-where the superscript $n$ on $\Omega$ indicates the shape at the $n$th timestep.
 The shape prediction can be approximated by integration of the velocity:
@@
 \begin{bmatrix}
-\mathbf{u}^n_i \\
-p^n_i
+\mathbf{u}^h_i \\
+p^h_i
 \end{bmatrix}.
@@
 changes to the solver/preconditioner, or symmetrization of the stabilization term \citep{kaus2010stabilization}.
-An alternative would be to construct a time-integration scheme with better stability properties.  
+An alternative would be to construct a time-integration scheme that remains stable at larger timesteps.
 The reason that forward Euler integration performs so badly with the decay of topography is that 

### 2016-04-29 free-surface-paper.tex
Add some text on the time evolution of the timescale.

@@
 \subsection{Numerical determination of $\tau_\mathrm{min}$}
+\label{sec:numerical_determination}
 
@@
 \begin{itemize}
-\item Direct solution of Equation~\eqref{eq:generalized_eigenvalue}. This can be expensive, although
-only the minimum relaxation time is required. For certain geometries a simple power iteration on 
-the standard eigenvalue problem (Equation~\eqref{eq:standard_eigenvalue}) could be enough.
+\item Direct solution of Equation~\eqref{eq:generalized_eigenvalue}. Solution of the whole spectrum is expensive,
+however only the minimum relaxation time is required. For certain geometries a simple power iteration on 
+the standard eigenvalue problem (Equation~\eqref{eq:standard_eigenvalue}) can be enough.
 \item Analytical formulae. Several geometry and viscosity model combinations have analytical solutions
@@
 of experimentation near this value of $\tau$ can find an appropriate relaxation time.
-\item Observation of instabilities.  The mere act of observing a sloshing instability in an unstabilized
-problem can furnish an estimate of its relaxation time.
 \end{itemize}
+
+The first point deserves a bit more attention, since it provides the most general method for determining $\tau_\mathrm{min}$.
+We discussed the numerical determination of $\tau_\mathrm{min}$ in Section~\ref{sec:numerical_determination}
+by solving Equation~\eqref{eq:standard_eigenvalue} via power iteration.
+Each iteration requires a solution of the Stokes system (the $C$ matrix), and we have found
+that $\sim$10 iterations is usually enough to obtain a reasonable estimate of $\tau_\mathrm{min}$.
+This can be done during preprocessing, so over the course of the simulation the cost of determining
+$\tau_\mathrm{min}$ would be negligible.
+
+As a simulation progresses, there is the possibility that its viscosity or density structure
+will change, and as such its minimum relaxation time will evolve.
+This, of course, means that the best choice of $\tau^*$ can change over the course of a simulation.
+If $\tau_\mathrm{min}$ does not change significantly, the initial estimate can be fine.
+Otherwise, one can periodically check the evolution and redetermine $\tau_\mathrm{min}$ by 
+solving~\eqref{eq:standard_eigenvalue} during the model run.
+
+In section \label{sec:results} we show examples of the numerical solution of $\tau_\mathrm{min}$
+and its evolution in time.
 

### 2016-04-29 free-surface-paper.tex
Add some additional explanation to the AMR test.

@@
 \begin{figure*}
+\includegraphics[width=0.9\textwidth]{figures/rayleigh_taylor_setup.pdf}
+\caption{}
+\label{fig:rayleigh_taylor}
+\end{figure*}
+
+\begin{figure*}
 \includegraphics[width=0.9\textwidth]{figures/rayleigh_taylor.pdf}
@@
 both with and without adaptive mesh refinement.
+
 For the uniform refinement cases each point is generated by running the benchmark with a different global refinement level.
@@
 is limited to the same refinement level of the corresponding global refinement simulation.
+In both the uniform and the adaptive cases, the smallest cell size in the most refined case is $\sim$1 km.
 We refine every ten timesteps according to gradients in the density and compositional fields.
+
+We have chosen a very small timestep for this test of 500 yr (note that this is significantly smaller than the smallest
+relaxation time of $\sim$ 14 kyr), so most timestepping schemes should be stable, including forward Euler.
+We choose this step for two reasons: (1) The primary purpose of this test is to investigate the savings
+due to using adaptive mesh refinement with a free surface, and (2) It was found in \citet{crameri2012comparison}
+that in order for this test to fully converge, a small timestep of several hundred years must be used.
 
@@
 For the adaptive cases we perform mesh refinement according to the sum of density and composition gradients.
-Quasi-implicit (QI) and nonstandard finite difference (NSFD) timestepping converge similarly, but the adaptive mesh refinement simulations require significantly fewer unknowns.}
+Quasi-implicit (QI) and nonstandard finite difference (NSFD) timestepping converge similarly, but the adaptive mesh refinement simulations require significantly fewer unknowns.
+For NSFD timestepping we used a $\tau^*$ of 14.825 kyr, as determined by an analytical solution of the long-wavelength relaxation time of the free surface for this particular model setup \citep{crameri2012comparison}.}
 \label{fig:amr}

### 2016-05-02 figures/rayleigh_taylor.py
Work on rayleigh taylor figures.

@@ -12,15 +12,18 @@ except ValueError:
 
 prefix = '../benchmarks/rayleigh_taylor/'
 
-forward_euler = np.genfromtxt(prefix+'output_small_time_step/statistics', skip_header=50, usecols=(1,15))
-nsfd_adaptive_02 = np.genfromtxt(prefix+'output_nsfd_adaptive_cfl_02/statistics', skip_header=50, usecols=(1,15))
-nsfd_adaptive_05 = np.genfromtxt(prefix+'output_nsfd_adaptive_cfl_05/statistics', skip_header=50, usecols=(1,15))
-theta_10_cfl_02 = np.genfromtxt(prefix+'output_theta_10_cfl_02/statistics', skip_header=50, usecols=(1,15))
+#Load in datasets
+skip = 20
+forward_euler = np.genfromtxt(prefix+'output_small_time_step/statistics', skip_header=skip, usecols=(1,15))[::10]
+nsfd_adaptive_02 = np.genfromtxt(prefix+'output_nsfd_adaptive_cfl_02/statistics', skip_header=skip, usecols=(1,15))[::10]
+nsfd_adaptive_05 = np.genfromtxt(prefix+'output_nsfd_adaptive_cfl_05/statistics', skip_header=skip, usecols=(1,15))[::10]
+theta_10_cfl_02 = np.genfromtxt(prefix+'output_theta_10_cfl_02/statistics', skip_header=skip, usecols=(1,15))[::10]
 
-
+#NSFD results have some superfluous zeros during mesh refinement
 nsfd_adaptive_02 = np.array( [ line for line in nsfd_adaptive_02 if line[0] != 0.] )
 nsfd_adaptive_05 = np.array( [ line for line in nsfd_adaptive_05 if line[0] != 0.] )
 
+#Read in the relaxation time as computed with power iteration
 relaxation_times = []
 time = 0.
 with open(prefix+'output_nsfd_adaptive_cfl_02/log.txt') as f:
@@ -32,32 +35,66 @@ with open(prefix+'output_nsfd_adaptive_cfl_02/log.txt') as f:
         relaxation_times.append( [time, timescale] )
 relaxation_times = np.array(relaxation_times[3:])
 
-plt.figure(figsize=(8,4) )
[…]

### 2016-05-24 figures/benchmark_setup.py
Fix error in identification of fundamental mode.

@@ -10,7 +10,7 @@ ax.tick_params( top='off', bottom='off', left='off', right='off')
 
 amplitude = 0.1
 x = np.linspace(0,1,1000)
-surface = 1.0 + amplitude*np.cos(2.*np.pi*x)
+surface = 1.0 + amplitude*np.cos(np.pi*x)
 ref_surface = 1.0*np.ones_like(x)
 
 #plot fluid
@@ -22,8 +22,8 @@ plt.fill_between(x, 0, surface, color='gray', alpha=0.2)
 head_size = 0.02
 
 #Draw arrow for surface perturbation
-ax.arrow(0.5, 1.0, 0.0, -amplitude+head_size, head_width=head_size, head_length=head_size, fc='k', ec='k')
-ax.text(0.45, .95, r'$\zeta_0$', fontsize=20)
+ax.arrow(0.95, 1.0, 0.0, -0.8*amplitude+head_size, head_width=head_size, head_length=head_size, fc='k', ec='k')
+ax.text(0.9, .95, r'$\zeta_0$', fontsize=20)
 
 #Draw arrow for depth
 ax.arrow(0.05, 0.05, 0.0, 0.9, head_width=head_size, head_length=head_size, fc='k', ec='k')

### 2016-05-24 free-surface-paper.tex
Work on rayleigh taylor figures.
Edits and bring benchmark setup forward.

@@
 The total pressure and density can then be decomposed into variations about their 
-reference values: $\rho = \rho_0 + \rho^\prime$, $p = p_0 + p^\prime$.
-
-Formally, this gives rise to the following time dependent, coupled system with unknowns $\mathbf{u}(t)$ and $\mathbf{p}(t)$:
+reference values: $\rho = \rho_0 + \rho^\prime$, $p = p_0 + p^\prime$,
+and the hydrostatic reference state (Equation~\eqref{eq:hydrostatic_stokes}) may be subtracted from Equation~\eqref{eq:stokes}.
+
+This gives rise to the following time dependent, coupled system with unknowns $\mathbf{u}(t)$ and $\mathbf{p}(t)$:
 \begin{equation}
 \begin{aligned}
-\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p &= \rho \mathbf{g} \qquad &\text{in } &\Omega(t)\\
+\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p^\prime &= \rho^\prime \mathbf{g} \qquad &\text{in } &\Omega(t)\\
 \nabla \cdot \mathbf{u} &= 0  \qquad &\text{in } &\Omega(t)
@@
 In order to better understand the time evolution of the system in Equation~\eqref{eq:final_system}
-we will consider the eigenvalues of a simplified homogeneous system. By setting $\rho = \rho_0$, we obtain the system
+we will consider the eigenvalues of a simplified homogeneous system where we neglect variations in density ($\rho^\prime = 0$):
 \begin{equation}
@@
 domain, and can be thought of as an artificial viscous damping of the surface.
-Since the term depends on the velocity, it 
-enters the Stokes system matrix as a stabilization term.  
+Since the term depends on the velocity, it enters the Stokes system matrix as a stabilization term.  
 Empirically, it has been found to be successful at damping instabilities \citep{kaus2010stabilization, quinquis2011role, duretz2011discretization}.
@@
 Furthermore, in Equation~\eqref{eq:weak_eigen} we considered response of the surface in a homogeneous system where there were no lateral density variations,
-while Equation~\eqref{eq:kmm_stabilization} uses the total density.
+while Equation~\eqref{eq:kmm} uses the total density.
 In most geodynamic simulations lateral density variations are small compared to the total density.
@@
 Several things are of note about this eigenvalue equation which help in solving it.
-The matrix $D$ is a scaled mass matrix which is only nonzero on the free surface,
+The matrix $D$ is a mass matrix which is only nonzero on the free surface,
 so it is mostly empty, and, in general, it is not necessary to explicitly form the
@@
 which finds the dominant eigenvalue/eigenvector pair, is an attractive option for finding $\tau_\mathrm{min}$.
-We have numerically tested the effect of quasi-implicit stabilization on a 2D Cartesian box with a free surface,
-using power iteration to find $\tau_S$.
-Figure~\ref{fig:perturb_spectrum} shows the results, where colored circles indicate the numerical
+We have tested the effectiveness of power iteration for finding the minimum relaxation time
+on a problem for which there is an analytical solution.
+We also use it to confirm Equation~\eqref{eq:spectrum_shift}.
+Following \citet{kramer2012implicit} we model the relaxation
+of surface topography in a two-dimensional Cartesian box with an isoviscous fluid.
+The setup is shown in Figure~\ref{fig:benchmark_setup}.
+The eigenvectors of this system are given by infinitesimal sinusoidal perturbations to the free surface
+\begin{equation}
+\zeta(x,0) = \zeta_0 \cos\left( k x \right),
+\label{eq:initial_topography}
+\end{equation}
+where $k = 2 \pi n / L$, and $n$ is a wavenumber which is an integer multiple of $1/2$.
+For an infinitesimal perturbation of this form, the relaxation time $\tau$ is given by
+\begin{equation}
+\tau = \frac{D k + \sinh(D k) \cosh(D k)}{ \sinh^2 (D k) } \frac{2 k \eta}{\rho g},
+\label{eq:analytic_relaxation_time}
+\end{equation}
+where $D$ is the layer depth, $\eta$ is the viscosity, $\rho$ is the density, and $g$ is the acceleration due to gravity.
+
+\begin{figure*}
+\includegraphics[width=0.9\textwidth]{figures/benchmark_setup.pdf}
+\caption{Setup for the power iteration test (Section~\ref{sec:numerical_determination}), as well as for the free surface relaxation benchmark (Section~\ref{sec:topography_relaxation}).
+A 2D box with an isoviscous fluid has sinusoidal initial topography, with amplitude $\zeta_0$ and wavenumber $n$.
+The box has depth $D$ and length $L$.
+For our tests $\rho = \eta = g = D = L = 1$, and $n=1/2$.
+For the power iteration test there is no initial perturbation ($\zeta_0 = 0.0$), and for the surface relaxation benchmark $\zeta_0 = 0.005$.}
+\label{fig:benchmark_setup}
+\end{figure*}
+The eigenvector with the minimum relaxation time for this system $\tau_{\min}$ corresponds to $n=1/2$ 
+(which, as expected, has the form of the fluid sloshing back and forth).
+Figure~\ref{fig:perturb_spectrum} shows the results of the power iteration, where colored circles indicate the numerical
 solution and black dots show the analytical solution using equation~\eqref{eq:spectrum_shift}.
+In most cases the iteration converges to better than $\sim10^{-5}$ relative accuaracy in about 20 iterations
+(and it can be fewer with a good initial guess).
 
@@
 \includegraphics[width=0.9\textwidth]{figures/perturb_spectrum.pdf}
-\caption{The effect of quasi-implicit stabilization on the minimum relaxation time of a fluid in a 2D Cartesian box with a free surface (see Figure~\ref{fig:benchmark_setup}).
+\caption{The effect of quasi-implicit stabilization on the minimum relaxation time of a fluid in a 2D Cartesian box with a free surface (see Figure~\ref{fig:benchmark_setup}), as well as the result of power iteration to find the minimum relaxation time.
 The $x$-axis shows the timestep normalized by minimum relaxation time of the unstabilized problem $\tau_\mathrm{\min}$.
@@
 \end{equation}
-where $\tau^*$ is some as-yet undetermined positive constant.
+where $\tau^*$ is some as-yet undetermined positive constant which we will refer to as the ``stabilization timescale''.
 This form of $\mathbf{u}$ automatically decays in time, and as we shall see, has much better 
@@
 \begin{equation}
-a_i^{n} = a_i^{n+1} \left[ 1 - \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \right].
+a_i^{n+1} = a_i^{n} \left[ 1 - \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \right].
 \label{eq:recursion}
@@
 \caption{(a) Stability region for the nonstandard finite difference scheme (green).
-On the x-axis is the value of the stabilization time, in units of the minimum relaxation time $\tau_{\mathrm{min}}$.
+On the x-axis is the value of the stabilization timescale, in units of the minimum relaxation time $\tau_{\mathrm{min}}$.
 On the y-axis is the value of the timestep, also in units of the minimum relaxation time.
@@
 We demonstrate the convergence of the nonstandard finite difference model by comparison 
-with an analytical solution. Following \citet{kramer2012implicit} we model the relaxation
+with an analytical solution. 
+As in Section~\ref{sec:numerical_determination} we consider the relaxation
 of sinusoidal surface topography in a two-dimensional Cartesian box with an isoviscous fluid.
@@
 
-The initial topography is given by
-\begin{equation}
-\zeta(x,0) = \zeta_0 \cos\left( k x \right),
-\end{equation}
-where $k = 2 \pi / L$.
-
-The time evolution of this topography is then
-\begin{equation}
-\zeta(x, t) = \zeta(x,0) e^{-t/\tau},
-\end{equation}
-where the relaxation time $\tau$ is given by
-\begin{equation}
-\tau = \frac{D k + \sinh(D k) \cosh(D k)}{ \sinh^2 (D k) } \frac{2 k \eta}{\rho g},
-\end{equation}
-where $D$ is the layer depth, $\eta$ is the viscosity, $\rho$ is the density, and $g$ is the acceleration due to gravity.
-
+For an initial topography given by Equation~\eqref{eq:initial_topography}, the relaxation time $\tau$
+is given by Equation~\eqref{eq:analytic_relaxation_time}, and the time evolution of the surface 
+topography is given by
+\begin{equation}
+  \zeta(x,t) = \zeta(x,0) e^{-t/\tau}.
+\end{equation}
 This solution is only valid for infinitesimal topography. However, for small
@@
 We estimate the error $E$ by time-integrating the $L_2$ difference between the numerical and analytical
-solutions at the center point of the free surface:
-\begin{equation}
-E = \frac{1}{4\tau}\int_0^{4 \tau} \lVert \zeta_\mathrm{numeric}(L/2, t) - \zeta_\mathrm{analytic}(L/2, t) \rVert_2 \text{ d}t.
-\end{equation}
-
-\begin{figure*}
-\includegraphics[width=0.9\textwidth]{figures/benchmark_setup.pdf}
-\caption{Setup for the free surface relaxation benchmark.
-A 2D box with an isoviscous fluid has sinusoidal initial topography, with amplitude $\zeta_0$.
-The box has depth $D$ and length $L$.
-For our tests $\rho = \eta = g = D = L = 1$, and $\zeta_0 = 0.005$.}
-\label{fig:benchmark_setup}
-\end{figure*}
+solutions at the center point of the free surface, over a time interval $T$ (which we choose to be $4\tau$):
+\begin{equation}
+E = \frac{1}{T}\int_0^{T} \lVert \zeta_\mathrm{numeric}(L/2, t) - \zeta_\mathrm{analytic}(L/2, t) \rVert_2 \text{ d}t.
+\end{equation}
 
@@
 the quasi-implicit guess for the body force is good enough that it actually seems to achieve the second-order 
-convergence of a trapezoidal scheme, though it is not clear whether this extends to more complicated models.
+convergence of a trapezoidal scheme, though it has not been shown that this extends to more complicated models.
 
@@
 Comparison with the slope-one line confirms that it is first-order in time.
-In the case that the stabilization relaxation time $\tau^*$ is equal to the analytic relaxation time the error becomes very small, as the time integration scheme becomes exact \citep{mickens2002nonstandard}.
+In the case that the stabilization timescale $\tau^*$ is equal to the analytic relaxation time the error becomes very small, as the time integration scheme becomes exact \citep{mickens2002nonstandard}.
 (b) Convergence test with timestep size for the quasi-implicit scheme.
@@
 
+\begin{figure*}
+\includegraphics[width=0.9\textwidth]{figures/rayleigh_taylor_tau_choice.pdf}
+\caption{}
+\label{fig:rayleigh_taylor}
+\end{figure*}
+
 \subsection{Mesh adaptivity}

### 2016-05-24 figures/rayleigh_taylor_setup.py
Add rayleigh taylor setup figures.

@@ -1,0 +1,58 @@
+import matplotlib.pyplot as plt
+import numpy as np
+
+L=500
+H=L
+thickness=100.
+
+fig = plt.figure( figsize=(8,8) )
+ax = fig.add_subplot(111)
+
+ax.axes.get_yaxis().set_major_formatter(plt.NullFormatter())
+ax.axes.get_xaxis().set_major_formatter(plt.NullFormatter())
+ax.tick_params( top='off', bottom='off', left='off', right='off')
+
+amplitude = 5.0
+x = np.linspace(0,L,1000)
+interface = -thickness - amplitude*np.cos(2.*np.pi*x/x[-1.])
+
+#plot fluid
+ax.plot(x,interface, color='k')
+plt.fill_between(x, -L, interface, color='black', alpha=0.3)
+plt.fill_between(x, interface, 0., color='black', alpha=0.6)
+
+
+head_size = 10
+
+
+#Draw arrow for gravity
+ax.arrow(0.3*L, -0.6*L, 0.0, -.15*L, head_width=head_size, head_length=head_size, fc='k', ec='k')
+ax.text(0.2*L,-0.59*L, r'${\bf g}=9.81 \; \mathrm{m}/\mathrm{s}^2$', fontsize=20)
+
+#Add density information
+ax.text(0.2*L, -0.10*L, r'$\rho = 3300  \; \mathrm{kg}/\mathrm{m}^3$', fontsize=20)
+ax.text(0.2*L, -0.15*L, r'$\eta = 10^{21} \; \mathrm{Pa} \; \mathrm{s}$', fontsize=20)
+
+ax.text(0.2*L, -0.40*L, r'$\rho = 3200  \; \mathrm{kg}/\mathrm{m}^3$', fontsize=20)
+ax.text(0.2*L, -0.45*L, r'$\eta = 10^{20} \; \mathrm{Pa} \; \mathrm{s}$', fontsize=20)
+
+#Draw arrow for length
+ax.arrow(0.08*L, -0.95*H, 0.85*L, 0.0, head_width=head_size, head_length=head_size, fc='k', ec='k')
+ax.arrow(0.95*L, -0.95*H, -0.85*L, 0.0, head_width=head_size, head_length=head_size, fc='k', ec='k')
+ax.text(0.5*L, -.94*H, r'$500 \; \mathrm{km}$', fontsize=20)
+
+#Draw arrow for depth
[…]

### 2016-05-24 free-surface-paper.tex
Add text on RT test.

@@
 explored methods for solving the transport of the free surface implicitly.
+ and benchmarks
 
@@
 good stability properties (Section~\ref{sec:newscheme}).
-Finally, we describe the implementation of a free surface in the mantle convection software \texttt{ASPECT} in Section~\ref{sec:implementation}, before showing numerical results
-in Section~\ref{sec:results}.
+Finally, we describe the implementation of a free surface in the mantle convection software \texttt{ASPECT} in Section~\ref{sec:implementation}, 
+before showing some numerical results and benchmarks in Section~\ref{sec:results}.
 
@@
 \item Direct solution of Equation~\eqref{eq:generalized_eigenvalue}. Solution of the whole spectrum is expensive,
-however only the minimum relaxation time is required. For certain geometries a simple power iteration on 
+however only the minimum relaxation time is required, which means that power iteration on 
 the standard eigenvalue problem (Equation~\eqref{eq:standard_eigenvalue}) can be enough.
@@
 
-In section \label{sec:results} we show examples of the numerical solution of $\tau_\mathrm{min}$
+In section \label{sec:results} we show an examples the numerical solution of $\tau_\mathrm{min}$
 and its evolution in time.
 
-\subsection{Connection with Dynamic Relaxation}
 \label{sec:dynamic_relaxation}
@@
 \includegraphics[width=0.9\textwidth]{figures/rayleigh_taylor_setup.pdf}
-\caption{}
-\label{fig:rayleigh_taylor}
+\caption{Setup for the Rayleigh-Taylor test.
+A denser, more viscous layer of lies on top of the mantle.
+The overlying layer has a thickness of 100 km, and a sinusoidal perturbation on its base with and amplitude of 5 km.}
+\label{fig:rayleigh_taylor_setup}
 \end{figure*}
+
+We also consider the time evolution of the Rayleigh-Taylor benchmark described in \citet{kaus2010stabilization}.
+The benchmark setup is shown in Figure~\ref{fig:rayleigh_taylor_setup}.
+In all cases we use adaptive mesh refinement with five global refinement levels and four adaptive refinement levels,
+corresponding to $\sim$1 km cells at the most refined level. We refine according to compositional gradients, as well as
+near to the free surface.
+
+Figure~\ref{fig:rayleigh_taylor} shows the maximum depth of the interface between the dense and light fluid through time.
+The blue line shows the results using forward Euler timestepping with a very small timestep of 500 yrs, and is used as a reference solution.
+Nonstandard finite difference timestepping and quasi-implicit timestepping both converge to the reference solution,
+though from different directions.
+For the nonstandard finite difference simulations we recomputed the value of the minimum relaxation time $\tau_\mathrm{min}$ 
+every 50 timesteps using power iteration, and reset the value of $\tau^*$ to $\tau_\mathrm{min}$.
+
+The behavior of the two schemes is similar, though at the same CFL number quasi-implicit timestepping is slightly more accurate.
+On the other hand, with our implementation we have found that nonstandard finite difference timestepping can allow for larger timesteps 
+than quasi-implicit timestepping while remaining stable.
 
@@
 The symbols are for several different runs with different timestepping parameters, using both quasi-implicit (QI) and nonstandard finite difference (NSFD) schemes.
-As the timestepping gets smaller, both schemes converge to the reference solution (though from different directions).
-(b) Evolution of the minimum relaxation timescale over the course of the simulation.
-Every 50 timesteps we calculate the minimum relaxation timescale to observe how much it changes during the simulation.
-It does experience significant changes, but not by orders of magnitude, so an initial estimate is likely sufficient for this case.}
+As the timestepping gets smaller, both schemes converge to the reference solution (though from different directions).}
 \label{fig:rayleigh_taylor}
@@
 
+We also used the Rayleigh-Taylor benchmark to investigate the effect of making a poor choice for $\tau^*$, shown in Figure~\ref{fig:rayleigh_taylor_tau_choice}.
+We include four cases, all using adaptive timestepping with CFL=0.2.
+As a reference solution we include the case where we update $\tau^*$ to be equal to $\tau_\mathrm{min}$ every 50 timesteps,
+allowing for time variations in the viscosity and density structure to change the minimum relaxation time.
+We also show the case where we do not update $\tau^*$ over the course of the simulation, and two cases where
+we use the wrong value by a factor of two in either direction.
+
+Overall the course of the Rayleigh-Taylor benchmark, $\tau_\mathrm{min}$ varies by about 25\%, as shown in Figure~\ref{fig:rayleigh_taylor_tau_choice}b.
+The case where we use the initial value of $\tau_\mathrm{min}$ for $\tau^*$ gives nearly identical results,
+whereas the cases where we choose values $\sim$50\% off in either direction give different (though not wildly so) results.
+In this case, using the initial value for $\tau^*$ gives sufficiently accurate results over the course of the simulation.
+
 \begin{figure*}
 \includegraphics[width=0.9\textwidth]{figures/rayleigh_taylor_tau_choice.pdf}
-\caption{}
-\label{fig:rayleigh_taylor}
+\caption{Effect of the choice of $\tau^*$ on the Rayleigh-Taylor test.
+(a) Maximum depth of the drip.
+In all cases adaptive timestepping is used with CFL=0.2.
+The blue line serves as a reference solution, and shows the result when the smallest relaxation timescale $\tau_{\mathrm{min}}$ is recalculated every 50 timesteps, and then the stabilization timescale $\tau^*$ is set to that value.
+The red hexagons show the result when we do not update $\tau^*$, instead keeping it at the inital value of $\tau_{\mathrm{min}}$.
+The results are indistinguishable from those of the first case.
+The triangles show the results when we make a poor choice of $\tau^*$, by a factor of two in either direction.
+In these cases there is a significant difference from the reference solution.
+(b) Evolution of the minimum relaxation timescale over the course of the simulation.
+It does experience significant changes, but not by orders of magnitude, so it is unsurprising that using the initial value is sufficient for this case.}
+\label{fig:rayleigh_taylor_tau_choice}
 \end{figure*}
@@
 For the uniform refinement cases each point is generated by running the benchmark with a different global refinement level.
-For the adaptive case each point is generated by allowing the mesh to be refined, where the maximum refinement level
-is limited to the same refinement level of the corresponding global refinement simulation.
+For the adaptive case each point is generated by allowing the mesh to be refined according to gradients in density and composition,
+where the maximum refinement level is limited to the same refinement level of the corresponding global refinement simulation.
 In both the uniform and the adaptive cases, the smallest cell size in the most refined case is $\sim$1 km.
@@
 We have chosen a very small timestep for this test of 500 yr (note that this is significantly smaller than the smallest
-relaxation time of $\sim$ 14 kyr), so most timestepping schemes should be stable, including forward Euler.
+relaxation time of $\sim$ 14 kyr), so most timestepping schemes should be stable and accurate, including forward Euler.
 We choose this step for two reasons: (1) The primary purpose of this test is to investigate the savings
@@
 the quasi-implicit scheme. For $\theta=0.5$ the quasi-implicit scheme is more accurate,
-but is only conditionally stable. For $\theta=1$ the two schemes are comparably accurate 
-in the tests we ran.
+but has a smaller stability region. For $\theta=1$ the two schemes are comparably accurate,
+with a slight advantage to the quasi-implicit scheme. However, we have found that
+the nonstandard finite difference scheme allows for larger stable timesteps, if the modeler 
+is willing to pay the price of reduced accuracy for a particular simulation.
 

### 2016-05-24 free-surface-paper.tex
Phrasing.

@@
 \end{equation}
-Summing this error over many timesteps results in a factor of $\Delta t^{-1}$, demonstrating the first-order accuracy~\citep[e.g.][]{leveque2007finite}.
+The local truncation error is quadratic in $\Delta t$, similar to the forward Euler, and summing this error over 
+many timesteps results in a factor of $\Delta t^{-1}$, demonstrating the first-order global accuracy~\citep[e.g.][]{leveque2007finite}.
 The choice of $\tau^*$ controls the size of the coefficient on the truncation error for the scheme.
@@
 The case where we use the initial value of $\tau_\mathrm{min}$ for $\tau^*$ gives nearly identical results,
-whereas the cases where we choose values $\sim$50\% off in either direction give different (though not wildly so) results.
-In this case, using the initial value for $\tau^*$ gives sufficiently accurate results over the course of the simulation.
+whereas the cases where we choose values $\sim$50\% off in either direction give significantly different (though not wildly so) results.
+
+In many simulations we expect that the geometry, density and viscosity structure of the model will not change so significantly
+that the $\tau^*$ would need to be updated frequently (if ever).
 

### 2016-05-24 benchmarks/timestep_convergence/run_models.py

@@ -6,7 +6,7 @@ from scipy.integrate import trapz
 initial_topography = 0.005
 L = 1.0  #domain length
 D = 1.0  # domain depth
-order = 1.0 # order of perturbation
+order = 0.5 # order of perturbation
 rho = 1.0 # density
 eta = 1.0 #viscosity
 g = 1.0 #gravity

### 2016-05-25 benchmarks/plugins.cc
Use correct mode for benchmark.

@@ -86,7 +86,7 @@ namespace aspect
 
       private:
 
-        unsigned int order;  //Order of the perturbation
+        double order;  //Order of the perturbation
         double amplitude;  //amplitude of the perturbation
 
     };

### 2016-05-25 benchmarks/tau_sensitivity/run_models.py

@@ -6,12 +6,12 @@ from scipy.integrate import trapz
 initial_topography = 0.005
 L = 1.0  #domain length
 D = 1.0  # domain depth
-order = 1.0 # order of perturbation
+order = 0.5 # order of perturbation
 rho = 1.0 # density
 eta = 1.0 #viscosity
 g = 1.0 #gravity
 
-k_s = 2.0*np.pi/L;  #wavenumber
+k_s = 2.0*np.pi*order/L;  #wavenumber
 k = 2.0*np.pi*order/L;  #wavenumber
 
 tau_0 = 2. * k * eta /rho / g  #reference relaxation time

### 2016-05-25 benchmarks/timestep_convergence/run_models.py

@@ -11,7 +11,7 @@ rho = 1.0 # density
 eta = 1.0 #viscosity
 g = 1.0 #gravity
 
-k_s = 2.0*np.pi/L;  #wavenumber
+k_s = 2.0*np.pi*order/L;  #wavenumber
 k = 2.0*np.pi*order/L;  #wavenumber
 
 tau_0 = 2. * k * eta /rho / g  #reference relaxation time
@@ -79,7 +79,7 @@ for r in relaxation_times:
     print "Running with time step %f, relaxation time %f\n"%(t, r)
     generate_prm(t, r)
     os.system('mpirun -n 4 /home/ian/aspect/build/aspect tmp.prm')
-    #plot_topography()
+#    plot_topography()
     outputfile.write("%f %f %f \n" % (r/tau_s, t/tau_s, calculate_error()) )
     outputfile.flush()
 

### 2016-05-25 benchmarks/timestep_convergence_qi/run_models.py

@@ -11,7 +11,7 @@ rho = 1.0 # density
 eta = 1.0 #viscosity
 g = 1.0 #gravity
 
-k_s = 2.0*np.pi/L;  #wavenumber
+k_s = 2.0*np.pi*order/L;  #wavenumber
 k = 2.0*np.pi*order/L;  #wavenumber
 
 tau_0 = 2. * k * eta /rho / g  #reference relaxation time

### 2016-05-26 free-surface-paper.tex
Minor edits.

@@
 explored methods for solving the transport of the free surface implicitly.
- and benchmarks
 
@@
 The total pressure and density can then be decomposed into variations about their 
-reference values: $\rho = \rho_0 + \rho^\prime$, $p = p_0 + p^\prime$,
-and the hydrostatic reference state (Equation~\eqref{eq:hydrostatic_stokes}) may be subtracted from Equation~\eqref{eq:stokes}.
+reference values: $\rho = \rho_0 + \rho^\prime$, $p = p_0 + p^\prime$.
+Using this the hydrostatic reference state (Equation~\eqref{eq:hydrostatic_stokes}) may be subtracted from Equation~\eqref{eq:stokes}.
 
@@
 defined on the bounded, moving domain $\Omega(t)\subset \mathbb{R}^d$ with boundary $\partial \Omega = \Gamma_0 \cup \Gamma_F$
-split into a fixed (Dirichlet) and free surface part $\Gamma_0$ and $\Gamma_F$, respectively.
+split into fixed (Dirichlet) and free surface parts $\Gamma_0$ and $\Gamma_F$, respectively.
 The boundary conditions are given by
@@
 It is most convenient to integrate over the domain at the current timestep $\Omega^n \equiv \Omega(t^n)$ instead of 
-the at this point unknown $\Omega^{n+1} \equiv \Omega(t^{n}+\Delta t)$. This corresponds to an explicit scheme.
+the unknown $\Omega^{n+1} \equiv \Omega(t^{n}+\Delta t)$. This corresponds to an explicit scheme.
 However, as we will see, the implicit scheme of integrating over the (unknown) domain at the next timestep $\Omega^{n+1}$ will
@@
 \includegraphics[width=0.9\textwidth]{figures/benchmark_setup.pdf}
-\caption{Setup for the power iteration test (Section~\ref{sec:numerical_determination}), as well as for the free surface relaxation benchmark (Section~\ref{sec:topography_relaxation}).
+\caption{Setup for the power iteration test (Section~\ref{sec:numerical_determination}) and for the free surface relaxation benchmark (Section~\ref{sec:topography_relaxation}).
 A 2D box with an isoviscous fluid has sinusoidal initial topography, with amplitude $\zeta_0$ and wavenumber $n$.
@@
 In most cases the iteration converges to better than $\sim10^{-5}$ relative accuaracy in about 20 iterations
-(and it can be fewer with a good initial guess).
+(and fewer with a good initial guess).
 
@@
 \begin{itemize}
-\item Direct solution of Equation~\eqref{eq:generalized_eigenvalue}. Solution of the whole spectrum is expensive,
-however only the minimum relaxation time is required, which means that power iteration on 
+\item Direct solution of Equation~\eqref{eq:generalized_eigenvalue}. Solution of the whole spectrum is expensive
+when only the minimum relaxation time is required, which means that power iteration on 
 the standard eigenvalue problem (Equation~\eqref{eq:standard_eigenvalue}) can be enough.
@@
 
-Overall the course of the Rayleigh-Taylor benchmark, $\tau_\mathrm{min}$ varies by about 25\%, as shown in Figure~\ref{fig:rayleigh_taylor_tau_choice}b.
+Over the course of the Rayleigh-Taylor benchmark, $\tau_\mathrm{min}$ varies by about 25\%, as shown in Figure~\ref{fig:rayleigh_taylor_tau_choice}b.
 The case where we use the initial value of $\tau_\mathrm{min}$ for $\tau^*$ gives nearly identical results,
@@
 We choose this step for two reasons: (1) The primary purpose of this test is to investigate the savings
-due to using adaptive mesh refinement with a free surface, and (2) It was found in \citet{crameri2012comparison}
-that in order for this test to fully converge, a small timestep of several hundred years must be used.
+due to using adaptive mesh refinement with a free surface, and (2) \citet{crameri2012comparison}
+found that a timestep of several hundred years was required to achieve full convergence.
 

### 2016-05-26 free-surface-paper.tex
Remove ``buoyancy driven'' from title
Attempt to clarify when we are dealing with a discretized domain.

@@
 
-\title{Stability and accuracy of free surface time integration in buoyancy driven viscous flows}
+\title{Stability and accuracy of free surface time integration in viscous flows}
 
@@
 We also defer discussion of free slip boundary conditions.
-The domain at time $t$ is defined by advecting a reference configuration $\Omega_0$ by a mesh velocity $\mathbf{u}_{\mathrm{mesh}}(t)$
-\begin{equation}
- \Omega(t) = \Omega_0 + \int_{t_0}^t \mathbf{u}_{\mathrm{mesh}}(t) \text{ d}t.
-\end{equation}
-The mesh velocity in normal direction (with unit normal $\mathbf{n}$) is given by the velocity $\mathbf{u}(t)$ on the free surface
-\begin{equation}
- \mathbf{u}_\mathrm{mesh}(t) \cdot \mathbf{n} = \mathbf{u}(t) \cdot \mathbf{n} \quad \text{on} \quad \Gamma_F
-\end{equation}
-and set to zero on the rest of the boundary. We defer the discussion of the velocity $\mathbf{u}_\mathrm{mesh}$ in the interior of the domain
-to Section~\ref{sec:remeshing}.
-Finally, we denote the displacement at time $t$ by $\zeta = \int_{t_0}^t \mathbf{u}_\mathrm{mesh}(t)\cdot \mathbf{n} \text{ d}t$, leading to the evolution
+The domain at time $t$ is defined by advecting a reference configuration $\Omega_0$ by the domain velocity $\mathbf{u}(t)$
+\begin{equation}
+ \Omega(t) = \Omega_0 + \int_{t_0}^t \mathbf{u}(t) \text{ d}t.
+ \label{eq:domain_evolution}
+\end{equation}
+We defer the discussion of the velocity of the discretized domain to Section~\ref{sec:remeshing}.
+Finally, we denote the displacement at time $t$ by $\zeta = \int_{t_0}^t \mathbf{u}_\mathrm{surface}(t)\cdot \mathbf{n} \text{ d}t$, leading to the evolution
 equation
@@
 where $\mathbf{u}^h_i$ and $p^h_i$ are finite-dimensional representations of $\mathbf{u}_i$ and $p^\prime_i$,
-and $M$ is the discretization of the bilinear form on the right-hand side of Equation~\eqref{eq:weak_eigen}.
-The form of Equation~\eqref{eq:generalized_eigenvalue} does not depend on whether we choose $\Omega^n$ or $\Omega^{n+1}$
+and the domain integrals are performed over a discrete approximation of the domain shape
+(such as a linear finite element basis) which we denote by $\hat{\Omega}$.
+$M$ is the discretization of the bilinear form on the right-hand side of Equation~\eqref{eq:weak_eigen}.
+The form of Equation~\eqref{eq:generalized_eigenvalue} does not depend on whether we choose $\hat{\Omega}^n$ or $\hat{\Omega}^{n+1}$
 for the domain integration.
@@
 evaluated on a prediction of the shape of the domain at a later time.
-The weak form of the right-hand-side body forces in the finite element discretization is then
+The continuous weak form of the right-hand-side body forces in is then
 \begin{equation}
@@
 \end{equation}
-The shape prediction can be approximated by integration of the velocity:
+We can approximate the shape prediction by time integrating the domain evolution equation (Equation~\eqref{eq:domain_evolution}):
 \begin{equation}
@@
 the forward Euler scheme. However, as $\Delta t$ gets larger, the pseudo-timestep does not 
-increase as quickly, reflecting the decaying nature of the normal modes.
+increase as quickly, reflecting the decaying nature of the normal modes
+(in fact, the pseudo-timestep is bounded between $\Delta t$ and $\tau^*$).
 Likewise, as the stabilization timescale $\tau^*$ goes to infinity, we also recover the 
@@
 
+The mesh velocity in normal direction at the free surface (with unit normal $\mathbf{n}$) has to be
+consistent with the velocity of the Stokes velocity solution $\mathbf{u}(t)$:
+\begin{equation}
+ \mathbf{u}_\mathrm{mesh}(t) \cdot \mathbf{n} = \mathbf{u}(t) \cdot \mathbf{n} \quad \text{on} \quad \Gamma_F
+\end{equation}
 In ALE calculations the internal mesh velocity is undetermined.

### 2016-05-27 free-surface-paper.tex
Unnecessary sentence.
Thank reviewers.
Add note about timestepping in AMR test.

@@
 
-One downside of the quasi-implicit scheme is that it requires a modification of the system matrix.
 The implementation of such a scheme results in a slightly 
@@
 
-We have chosen a very small timestep for this test of 500 yr (note that this is significantly smaller than the smallest
+We run simulations using both quasi-implicit and nonstandard finite difference schemes for the free surface.
+We use a very small timestep for this test of 500 yr (note that this is significantly smaller than the smallest
 relaxation time of $\sim$ 14 kyr), so most timestepping schemes should be stable and accurate, including forward Euler.
@@
 Thanks to Wolfgang Bangerth, Brent Delbridge, and Cedric Thieulot for useful discussions.
+Thanks to Fabio Crameri and Dave May for constructive reviews.
 %% The Appendices part is started with the command \appendix;

### 2016-05-27 response_to_referee.tex
Add response to referee

\documentclass[a4paper,12pt]{article}
\usepackage[utf8]{inputenc}

%opening
\title{Response to referees}
\author{Ian Rose, Bruce Buffett, and Timo Heister}
\date{}

\begin{document}

\maketitle

We thank referees Fabio Crameri and Dave May for their thoughtful and constructive reviews,
which are appended to this document for convenience.

\subsection*{Response to reviewer 1}

Fabio Crameri provided a number of helpful comments and suggestions for clarification,
which we summarize here:

\begin{itemize}
  \item We have identified $\mathbf{I}$ as the identity tensor.
  \item We have itentified $\mathbf{n}$ as the unit normal.
  \item We have clarified lines 154-155, 174 and 460-463.
  \item We have improved the caption for Figure 5.
\end{itemize}

\subsection*{Response to reviewer 2}

Dave May gave a very thorough review with substantial suggestions for 
improvements and additions to the manuscript, which we address in detail here.

\subsubsection*{Summary}

May's review has two main criticisms:
\begin{enumerate}
\item That we do not provide sufficient non-trivial examples of the usage of our new 
  time integratior, especially making detailed comparisons with the existing quasi implicit (QI)
  time integrator of Kaus et. al. (2010), including a discussion of the relative advantages
  and disadvantages of the scheme.
\item That we have not provided enough information of how to determine the stabilization timescale ($\tau^*$)
  which is required to use the nonstandard finite difference (NSFD) time integrator, including
  a discussion of the consequences of making a poor choice for $\tau^*$.
\end{enumerate}

In order to address these we have made two major additions to the manuscript:

First, we have added a more lengthy discussion of how to numerically determine the minimum relaxation time of
a given geodynamic simulation, which is the best choice for $\tau^*$. Our favored method
is using power iteration, which finds the dominant eigenvector/eigenvalue pair of the system.
Each iteration involves a solution of a Stokes system, and we find good approximation of the minimum
relaxation time in 5-20 iterations, depending on how good the initial guess is.

This estimate of the minimum relaxation time is not only useful for the NSFD scheme,
as it can also be used to determine appropriate timesteps for quasi-implicit time integration,
or even forward Euler time integration. 
We provide several numerical examples of using power iteration, including to confirm our analysis
of the effect of the quasi-implicit stabilization term on the spectrum of the system (Equation 29).
The results of this test are shown in Figure 2.

Our second major addition is an investigation of the Rayleigh-Taylor instability with a free surface
which was considered in Kaus et. al. (2010). We found that the QI and NSFD schemes are comparably
accurate at the same adaptive timestepping (with a slight advantage to QI). However, we have found
that the NSFD scheme is able to take larger stable timesteps (though at the cost of lower accuracy).

We have also used the Rayleigh-Taylor benchmark to investigate the effect of different choices of 
the stabilization parameter $\tau^*$. We considered four cases: 
\begin{enumerate}
  \item recomputing the minimum relaxation timescale every 50 timesteps and setting $\tau^*$ to that value 
    (this allows for changes in the viscosity and density structure with time). This is used as the reference case.
  \item Using the initial value of the minimum relaxation time for the whole simulation.
  \item Using a value that is about a factor of two too large.
  \item Using a value that is about a facator of two too small.
\end{enumerate}
The first two cases are almost identical, while the third and fourth have significant differences with the reference case.

We also have specific responses to the ``General Comments'' and ``Detailed Corrections'' sections
May's review, detailed below.

\subsubsection*{General comments}
\begin{enumerate}
  \item The manuscript is no numbered.
  \item We have made equation referencing consistent.
  \item Equations are now punctuated.
  \item Section headings now all use sentence case.
  \item ``timestep'' is now used uniformly.
  \item The point about the title is well-taken, though we would argue that our analysis of the stability of the
system and the effect of QI stabilization on the spectrum are at least as important as our new time integrator.
We have retitled the manuscript ``Stability and accuracy of free surface time integration in viscous flows.''
  \item We have added a detailed look at the time dependence of $\tau^*$. In many cases its time variation is
probably not large enough to merit recomputing it during the course of a simulation.
  \item We have provided new numerical experiments investigating the numerical determination of $\tau^*$, and have
taken a detailed look at a Rayleigh-Taylor benchmark. In the Rayleigh-Taylor case we have QI and NSFD schemes,
investigated the time dependence of $\tau_\mathrm{min}$, and demonstrated the effects of a poor choice for $\tau_\mathrm{min}$.
We ran experiments for the Schmeling et. al. (2008) benchmark, but have ultimately not included them because 
(1) they did not provide new insight over the Rayleigh-Taylor benchmark, and 
(2) this particular benchmark has not proved to be particularly useful due to its extremely sensitive dependence 
on model parameters and viscosity averaging schemes, making the results difficult to interpret.
  \item We have added a treatment of using power iteration to calculate the minimum relaxation time, which is
useful not just for determining $\tau^*$, but can also be of use for analysis of timestepping with other schemes.
  \item We have added some discussion of direct comparision of the QI and NSFD schemes. We agree that the benefits
    of spatial adapdivity are less connected with the preceding discussion on timestepping, however we still felt
    that demonstrating the savings of combining AMR with a free surface had some value.
\end{enumerate}

\subsubsection*{Detailed corrections}
\begin{enumerate}
  \item We have reworded the abstract to emphasize time integration of the free surface.
  \item We have cited Jarvis and Peltier (1982) for an early dynamic topography calculation.
  \item We have given an additional reference for the sticky air method.
  \item The surface of Earth is not truly stress free, since there are stresses due to oceans and atmopheres.
    We would argue that ``more closely matches'' is appropriate.
  \item We have cited Popov and Sobolev (2008) for another implicit approach to free surfaces.
  \item We have explicitly given the conditions for free-slip, free surface, and Dirichlet boundary conditions.
  \item We agree that discussion of the discrete mesh is more appropriate in Section 7, and have reorganized accordingly.
  \item This is a good suggestion. We have reorganized Section 3 to put all the information dealing with
    the time evolution of $\Omega(t)$ in one place.
  \item We have introduced a new notation $\hat{\Omega}$ for when we are dealing with a spatially discretized domain.
However, for most of the analysis of Sections 2-5 the equations are agnostic to whether we are using a discretized
or continuous domain, so we have retained the continuous notation when appropriate. We have also tried to clarify
what is meant by the prediction of the shape at a later time $\Delta \Omega$ as an approximate integration of the
domain evolution equation (Eq. 7).
  \item We have changed line 178 to ``become unstable''.
  \item We have clarified the approximations under which the QI stabilization term is identical to
the $M$ matrix in the homogeneous problem. Specifically, they are approximately the same when the slope is small,
and density variations are small compared to the reference value.
  \item We have added a note to this effect.
  \item It is true that our ALE formulation involves assembly of a surface mass matrix, as does the solution of
    the generalized eigenvalue problem. However, not every implementation need have the same approach. That being
    said, it is a good point that the modifications to the Stokes system matrix are overall not very difficult to apply.
  \item We have rephrased this point.
  \item $x$ and $y$ axes are now italicized.
  \item We have clarified this point.
  \item This has been fixed.
  \item We mean Equation (16), which can be reorganized to Equation (18), so either are appropriate here.
  \item We have clarified this point in the text, as well as explored by how much $\tau_\mathrm{min}$ might change.
  \item Equation (19) involves the inverse of $\tau_i$, so finding the largest eigenvalue of this system
is equivalent to finding the minimum relaxation time.
  \item Boundary conditions are now all defined mathematically.
  \item Absolutely right, this is now rephrased.
  \item This is now clarified to an ``$L_2$ projection of $u \cdot n$ onto $\Gamma_F$''.
  \item Changed to say ``cheap''.
  \item Line 267 now says ``acceleration due to gravity''.
  \item L2  $\rightarrow L_2$.
  \item We now use the symbol $E$ for the error. 
  \item Line 472-473 has been rephrased.
  \item We now give more information about the parameters used to run this simulation, including refinement schemes, 
    maximum spatial resolution, and timestepping schemes.
  \item It is true that 500 years is a very small timestep, and that forward Euler timestepping would be stable for this.
We use this timestep because Crameri et. al. (2012) found that a timestep of several hundred years was needed
for the solution to fully converge. The purpose of this test is less to investigate the time integration schemes
than to demonstrate the effectiveness of combining adaptive mesh refinement with a free surface treatment.
To a certain extent we agree that this is not the primary point of the manuscript. However, we still think it
is valuable to demonstrate that our implementation, which is open source and freely available, has this capability.
  \item QI is now defined.
  \item The caption now reads ``significantly fewer unknowns''.
\end{enumerate}

\end{document}

### 2016-06-03 response_to_referee.tex
small edits

@@
 First, we have added a more lengthy discussion of how to numerically determine the minimum relaxation time of
-a given geodynamic simulation, which is the best choice for $\tau^*$. Our favored method
-is using power iteration, which finds the dominant eigenvector/eigenvalue pair of the system.
-Each iteration involves a solution of a Stokes system, and we find good approximation of the minimum
+a given geodynamic simulation, which is the best choice for $\tau^*$ (see Section 6.4). Our favored method
+is using a power iteration, which finds the dominant eigenvector/eigenvalue pair of the system.
+Each iteration involves a solution of a Stokes system, and we find good a approximation for the minimum
 relaxation time in 5-20 iterations, depending on how good the initial guess is.
@@
 or even forward Euler time integration. 
-We provide several numerical examples of using power iteration, including to confirm our analysis
+We provide several numerical examples of using power iteration, which confirm our analysis
 of the effect of the quasi-implicit stabilization term on the spectrum of the system (Equation 29).
@@
 
-Our second major addition is an investigation of the Rayleigh-Taylor instability with a free surface
+Our second major addition is an investigation of the Rayleigh-Taylor instability with a free surface (Section 8.2),
 which was considered in Kaus et. al. (2010). We found that the QI and NSFD schemes are comparably
-accurate at the same adaptive timestepping (with a slight advantage to QI). However, we have found
+accurate the same (adaptive chosen) timestep (with a slight advantage to QI). However, we have found
 that the NSFD scheme is able to take larger stable timesteps (though at the cost of lower accuracy).
@@
 
-We also have specific responses to the ``General Comments'' and ``Detailed Corrections'' sections
+We also have specific responses to the ``General Comments'' and ``Detailed Corrections'' sections of
 May's review, detailed below.
@@
 \begin{enumerate}
-  \item The manuscript is no numbered.
+  \item The manuscript is now numbered.
   \item We have made equation referencing consistent.
@@
   \item We have given an additional reference for the sticky air method.
-  \item The surface of Earth is not truly stress free, since there are stresses due to oceans and atmopheres.
+  \item The surface of Earth is not truly stress free, since there are stresses due to oceans and atmospheres.
     We would argue that ``more closely matches'' is appropriate.

### 2016-06-03 free-surface-paper.tex
small edits

@@
 reference values: $\rho = \rho_0 + \rho^\prime$, $p = p_0 + p^\prime$.
-Using this the hydrostatic reference state (Equation~\eqref{eq:hydrostatic_stokes}) may be subtracted from Equation~\eqref{eq:stokes}.
+Using this, the hydrostatic reference state (Equation~\eqref{eq:hydrostatic_stokes}) may be subtracted from Equation~\eqref{eq:stokes}.
 
@@
 Indeed, analytical solutions for the normal modes are only known for simple geometries and rheologies.
-Instead, they obtain a velocity solution and simply advect the free surface with the local velocity.
+Instead, one typically obtains a velocity solution and then advect the free surface with the local velocity.
 The normal mode solution is instructive, however. 
@@
 If we were to numerically integrate this in time with a forward Euler method, we would find the 
-time-step criterion for stability \citep[e.g.][]{leveque2007finite} to be
+timestep criterion for stability \citep[e.g.][]{leveque2007finite} to be
 \begin{equation}
@@
 For the purposes of stabilization we may approximate $\rho \approx \rho_0$.
-Inserting these two approximations, we find that the stabilization term may be written
-\begin{equation}
--\theta \Delta t \int_{\Gamma_F^n} \rho_0 g ( \mathbf{w \cdot n}) (\mathbf{u \cdot n} )
+Inserting these two approximations, we find that the stabilization term may be written as
+\begin{equation}
+-\theta \Delta t \int_{\Gamma_F^n} \rho_0 g ( \mathbf{w \cdot n}) (\mathbf{u \cdot n} ),
 \label{eq:kmm_stabilization}
@@
 
-In many simulations we expect that the geometry, density and viscosity structure of the model will not change so significantly
+In many simulations we expect that the geometry, density, and viscosity structure of the model will not change so significantly
 that the $\tau^*$ would need to be updated frequently (if ever).

### 2016-06-06 free-surface-paper.tex
Add image of RT benchmark.

@@
 The symbols are for several different runs with different timestepping parameters, using both quasi-implicit (QI) and nonstandard finite difference (NSFD) schemes.
-As the timestepping gets smaller, both schemes converge to the reference solution (though from different directions).}
+As the timestepping gets smaller, both schemes converge to the reference solution (though from different directions).
+(b) Result of the small timestep forward Euler simulation after 5 Myr, including the adaptive mesh.}
 \label{fig:rayleigh_taylor}

### 2016-06-06 figures/rayleigh_taylor.py

@@ -36,9 +36,9 @@ with open(prefix+'output_nsfd_adaptive_cfl_02/log.txt') as f:
 relaxation_times = np.array(relaxation_times[3:])
 
 #Plot rayleigh taylor results
-plt.figure(figsize=(4,4) )
+plt.figure(figsize=(8,4) )
 
-plt.subplot(111)
+plt.subplot(121)
 
 marker = itertools.cycle( ('h', 'v', '^', '*', 'D', 's', 'o') )
 plt.plot( forward_euler[:,0]/1.e6, -forward_euler[:,1]/1.e3, label='FE dt = 500 yr' )
@@ -50,9 +50,21 @@ plt.legend(loc='lower left', prop={'size':10})
 plt.xlim(0.,6.)
 plt.xlabel('Time (Myr)')
 plt.ylabel('Maximum interface depth (km)')
+plt.title('(a)')
+
+plt.subplot(122)
+plt.gca().grid(False)
+result = mpimg.imread('rt_result2.png')
+mpimg.imread('rt_result.png')
+plt.imshow(result, extent=[0,500.,-500., 1.8], aspect='auto')
+plt.xlim(0.,500.)
+plt.ylim(-500., 1.8)
+plt.xlabel('$x$ (km)')
+plt.ylabel('Depth (km)')
+plt.title('(b)')
 
 #plt.show()
-plt.savefig('rayleigh_taylor.pdf')
+plt.savefig('rayleigh_taylor.pdf', dpi=600)
 plt.clf()
 
 #Load in data for testing the effect of different tau choices

### 2016-06-06 figures/rayleigh_taylor_setup.py

@@ -18,8 +18,8 @@ interface = -thickness - amplitude*np.cos(2.*np.pi*x/x[-1.])
 
 #plot fluid
 ax.plot(x,interface, color='k')
-plt.fill_between(x, -L, interface, color='black', alpha=0.3)
-plt.fill_between(x, interface, 0., color='black', alpha=0.6)
+plt.fill_between(x, -L, interface, color='darkgray', alpha=1.0)
+plt.fill_between(x, interface, 0., color='dimgray', alpha=1.0)
 
 
 head_size = 10

### 2016-06-06 figures/rayleigh_taylor.py
Adjust figures a bit.

@@ -62,6 +62,7 @@ plt.ylim(-500., 1.8)
 plt.xlabel('$x$ (km)')
 plt.ylabel('Depth (km)')
 plt.title('(b)')
+plt.tight_layout()
 
 #plt.show()
 plt.savefig('rayleigh_taylor.pdf', dpi=600)
@@ -90,6 +91,7 @@ plt.legend(loc='lower left', prop={'size':10})
 
 plt.title('(a)')
 plt.xlim(0.,6.)
+plt.ylabel('Maximum interface depth (km)')
 plt.xlabel('Time (Myr)')
 
 #Plot relaxation tau with time

### 2016-07-18 free-surface-paper.tex
Fix up references.
Light edits to help with parsing

@@
 \begin{abstract}
-Geodynamic simulations increasingly rely on simulations with a true free surface to 
+Geodynamic simulations increasingly rely on models with a true free surface to 
 investigate questions of dynamic topography, tectonic deformation, gravity perturbations, and
@@
 that manifests as sloshing surface motions, also known as the ``drunken sailor'' instability.
-This instability severely limits stable timestep sizes to those much smaller than could be used
+This instability severely limits stable timestep sizes to those much smaller than can be used
 in geodynamic simulations without a free surface. 
@@
 
+%%BEGINCLIP
+
 %% main text
@@
 Furthermore, the viscosity averaging scheme for areas of transitional
-composition was extremely important, as the subducting slab entrained significant amounts of the sticky air.
+composition was extremely important, as the simulation of the subducting slab entrained significant amounts of the sticky air.
 \citet{crameri2012comparison} conducted comparisons between sticky air and true free surface models, 
@@
 \end{equation}
-For the sake of economy we neglect inhomogeneous stress boundary conditions, though it would be straightforward to include them.
+For the sake of economy, we neglect inhomogeneous stress boundary conditions, though it would be straightforward to include them.
 We also defer discussion of free slip boundary conditions.
@@
 
-In order to better understand the time evolution of the system in Equation~\eqref{eq:final_system}
+In order to better understand the time evolution of the system in Equation~\eqref{eq:final_system},
 we will consider the eigenvalues of a simplified homogeneous system where we neglect variations in density ($\rho^\prime = 0$):
@@
 \begin{equation}
--\int_{\Omega(t)} 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u} ) + \int_{\Omega(t)} p^\prime \nabla \cdot \mathbf{w} 
-+ \int_{\Gamma_F(t)} \mathbf{w} \cdot \mathbf{T} \cdot \mathbf{n} = 0 
+\int_{\Omega(t)} 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u} ) - \int_{\Omega(t)} p^\prime \nabla \cdot \mathbf{w} 
+- \int_{\Gamma_F(t)} \mathbf{w} \cdot \mathbf{T} \cdot \mathbf{n} = 0 
 \label{eq:weak_stokes}
@@
 We will therefore  evaluate the integrals in Equation~\eqref{eq:weak_stokes} 
-over the hydrostatic reference surface and introduce a temporary auxilliary variable $\zeta$ which 
+over the hydrostatic reference surface and introduce a temporary auxiliary variable $\zeta$ which 
 represents the (small) topography of the free surface relative to that reference surface.
@@
 \begin{equation}
-\mathbf{T} \approx \frac{\partial \mathbf{T}}{\partial \mathbf{n} } \cdot \mathbf{n} \zeta = \rho_0 g \zeta \mathbf{I}.
+\mathbf{T} \approx \frac{\partial \mathbf{T}}{\partial \mathbf{n} } \cdot \mathbf{n} \zeta = -\rho_0 g \zeta \mathbf{I}.
 \label{eq:hydrostatic}
@@
 \begin{equation}
--\int_{\Omega(t)} 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u} ) + \int_{\Omega(t)} p^\prime \nabla \cdot \mathbf{w} 
+\int_{\Omega(t)} 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u} ) - \int_{\Omega(t)} p^\prime \nabla \cdot \mathbf{w} 
 + \int_{\Gamma_F(t)} \rho_0 g \zeta  \mathbf{w} \cdot \mathbf{n} = 0.
@@
 \begin{equation}
--\int_{\Omega(t)} 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u}_i ) + \int_{\Omega(t)} p^\prime_i \nabla \cdot \mathbf{w} 
+\int_{\Omega(t)} 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u}_i ) - \int_{\Omega(t)} p^\prime_i \nabla \cdot \mathbf{w} 
 = \tau_i \int_{\Gamma_F(t)} \rho_0 g (\mathbf{u}_i \cdot \mathbf{n} ) (\mathbf{w} \cdot \mathbf{n}).
@@
 \end{equation}
-where $\tau_{\mathrm{min}}$ is this minumum decay time.
+where $\tau_{\mathrm{min}}$ is this minimum decay time.
 The maximum stable timestep is limited by the minimum relaxation timescale.
@@
 solution and black dots show the analytical solution using equation~\eqref{eq:spectrum_shift}.
-In most cases the iteration converges to better than $\sim10^{-5}$ relative accuaracy in about 20 iterations
+In most cases the iteration converges to better than $\sim10^{-5}$ relative accuracy in about 20 iterations
 (and fewer with a good initial guess).
@@
 \end{equation}
-In order for the scheme to be stable, the quantity in brackets must not blow up as it is repeatedly multiplied by itself, or 
+In order for the scheme to be stable, the quantity in brackets must remain bounded as it is repeatedly multiplied by itself, or 
 \begin{equation}
@@
 \end{equation}
-which is simply a statement that the timestep must be postive.
+which is simply a statement that the timestep must be positive.
 
@@
 \tau^* \le 2 \tau_i.
+\label{eq:nsfd_unconditional}
 \end{equation}
@@
 \Delta t \le -\tau^* \log \left(1 - 2 \frac{\tau_i}{\tau^*} \right).
+\label{eq:nsfd_conditional}
 \end{equation}
@@
 \includegraphics[width=0.9\textwidth]{figures/stability_region.pdf}
-\caption{(a) Stability region for the nonstandard finite difference scheme (green).
+\caption{(a) Stability region for the nonstandard finite difference scheme (green, Equations~\eqref{eq:nsfd_unconditional} and~\eqref{eq:nsfd_conditional}).
 On the x-axis is the value of the stabilization timescale, in units of the minimum relaxation time $\tau_{\mathrm{min}}$.
@@
 For $\tau^*\le 2 \tau_{\mathrm{min}}$ the nonstandard finite difference scheme is unconditionally stable.
-(b) Stability region for the quasi-implicit scheme (green).
-Again, the y axis is in units of the minumum relaxation time.
+(b) Stability region for the quasi-implicit scheme (green, Equation~\eqref{eq:cfl_euler_stabilized}).
+Again, the y axis is in units of the minimum relaxation time.
 The x-axis shows the value of the stabilization parameter $\theta$.
 For $\theta = 1$ the quasi-implicit scheme is unconditionally stable.
-In both cases the stability region for the forward Euler scheme is also shown in blue.}
+In both cases the stability region for the forward Euler scheme (Equation~\eqref{eq:cfl_euler}) is also shown in blue.}
 \label{fig:stability_region}
@@
 Otherwise, one can periodically check the evolution and redetermine $\tau_\mathrm{min}$ by 
-solving~\eqref{eq:standard_eigenvalue} during the model run.
-
-In section \label{sec:results} we show an examples the numerical solution of $\tau_\mathrm{min}$
+solving Equation~\eqref{eq:standard_eigenvalue} during the model run.
+
+In Section~\ref{sec:results} we show examples the numerical solution of $\tau_\mathrm{min}$
 and its evolution in time.
@@
 avoiding inverted or otherwise poorly conditioned cells.
-The mesh deformation can be calcluated in many different ways, including algebraic \citep[e.g.][]{thieulot2011fantom} 
+The mesh deformation can be calculated in many different ways, including algebraic \citep[e.g.][]{thieulot2011fantom} 
 and PDE based approaches.
@@
 When discretized, this forms a linear system which can be solved for the mesh velocity $\mathbf{u}_\mathrm{mesh}$ at the 
-free surface. Of course, a new system to solve is not ideal, but this system, being nonzero 
-over only the free surface, is relatively cheap to solve.
+free surface. A new system to solve is not ideal, but this system, being nonzero 
+over only the free surface, is relatively computationally inexpensive to solve.
 
@@
 becomes exact \citep{mickens2002nonstandard}. At this point the magnitude of the error plummets and is no longer 
-a strong function of the timestep. The remaining error is likely due to error in the linear approxmation for 
+a strong function of the timestep. The remaining error is likely due to error in the linear approximation for 
 the analytical solution, the spatial discretization, or the linear solver tolerance.
@@
 We also consider the time evolution of the Rayleigh-Taylor benchmark described in \citet{kaus2010stabilization}.
+In this benchmark a dense, viscous layer with a sinusoidal initial perturbation drips into a less dense, less viscous mantle.
 The benchmark setup is shown in Figure~\ref{fig:rayleigh_taylor_setup}.
@@
 
-In many simulations we expect that the geometry, density, and viscosity structure of the model will not change so significantly
-that the $\tau^*$ would need to be updated frequently (if ever).
+In many simulations we expect that the geometry, density, and viscosity structure of the model will not change
+enough that the $\tau^*$ would need to be updated frequently (if ever).
 
@@
 The blue line serves as a reference solution, and shows the result when the smallest relaxation timescale $\tau_{\mathrm{min}}$ is recalculated every 50 timesteps, and then the stabilization timescale $\tau^*$ is set to that value.
-The red hexagons show the result when we do not update $\tau^*$, instead keeping it at the inital value of $\tau_{\mathrm{min}}$.
+The red hexagons show the result when we do not update $\tau^*$, instead keeping it at the initial value of $\tau_{\mathrm{min}}$.
 The results are indistinguishable from those of the first case.
@@
 solution. In this benchmark, a buoyant blob rises beneath a viscous lid with a free surface, deflecting the
-boundary upwards (for the full setup, see \citet{crameri2012comparison}). 
+boundary upwards, reminiscent of the dynamic topography due to a rising plume (for the full setup, see \citet{crameri2012comparison}). 
 Figure~\ref{fig:amr} shows the convergence of the maximum topography at 3 Myr to its value in a high resolution simulation,
@@
 scheme which is first order accurate in time and is unconditionally stable.
-The nonstandard finite difference scheme is extremely simple to implement, and 
+The nonstandard finite difference scheme is simple to implement, and 
 requires no modifications to the system matrix.
@@
 
+%%ENDCLIP
+
 \section*{Acknowledgements}
@@
 TH is partially supported by NSF grant DMS-1522191.
+Figures were created using the Python packages NumPy and Matplotlib.
 Thanks to Wolfgang Bangerth, Brent Delbridge, and Cedric Thieulot for useful discussions.
 Thanks to Fabio Crameri and Dave May for constructive reviews.
+
 %% The Appendices part is started with the command \appendix;
@@
 %%
+
+\nocite{hunter2007matplotlib}
+\nocite{vanderwalt2011numpy}
+
 \section*{Bibliography}

### 2016-07-18 frontmatter.tex
Add frontmatter.

--- 2016-06-03 free-surface-paper.tex
@@
-%% 
-%% Copyright 2007, 2008, 2009 Elsevier Ltd
-%% 
-%% This file is part of the 'Elsarticle Bundle'.
-%% ---------------------------------------------
-%% 
-%% It may be distributed under the conditions of the LaTeX Project Public
-%% License, either version 1.2 of this license or (at your option) any
-%% later version.  The latest version of this license is in
-%%    http://www.latex-project.org/lppl.txt
-%% and version 1.2 or later is part of all distributions of LaTeX
-%% version 1999/12/01 or later.
-%% 
-%% The list of all files belonging to the 'Elsarticle Bundle' is
-%% given in the file `manifest.txt'.
-%% 
-%% Template article for Elsevier's document class `elsarticle'
-%% with harvard style bibliographic references
-%% SP 2008/03/01
+\chapter{Time integration of free surfaces in viscous flows}
+\label{chap:free_surface}
 
-\documentclass[preprint,12pt,authoryear]{elsarticle}
-
-%% Use the option review to obtain double line spacing
-%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}
-
-%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
-%% for a journal layout:
-%% \documentclass[final,1p,times,authoryear]{elsarticle}
-%% \documentclass[final,1p,times,twocolumn,authoryear]{elsarticle}
-%% \documentclass[final,3p,times,authoryear]{elsarticle}
-%% \documentclass[final,3p,times,twocolumn,authoryear]{elsarticle}
-%% \documentclass[final,5p,times,authoryear]{elsarticle}
-%% \documentclass[final,5p,times,twocolumn,authoryear]{elsarticle}
-
-%% For including figures, graphicx.sty has been loaded in
-%% elsarticle.cls. If you prefer to use the old commands
-%% please give \usepackage{epsfig}
-
-%% The amssymb package provides various useful mathematical symbols
-\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{color, soul}
-\usepackage{url}
-
-\newif\ifdetail
-\detailfalse
-%\detailtrue
-
-%% The amsthm package provides extended theorem environments
-%% \usepackage{amsthm}
-
-%% The lineno packages adds line numbers. Start line numbering with
-%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
-%% for the whole article with \linenumbers.
-\usepackage{lineno}
-
-%% Block of code for fixing corresponding author bug 
-%% in elsarticle template... Don't really understand it
-\makeatletter
-\def\@author#1{\g@addto@macro\elsauthors{\normalsize%
-    \def\baselinestretch{1}%
-    \upshape\authorsep#1\unskip\textsuperscript{%
-      \ifx\@fnmark\@empty\else\unskip\sep\@fnmark\let\sep=,\fi
-      \ifx\@corref\@empty\else\unskip\sep\@corref\let\sep=,\fi
-      }%
-    \def\authorsep{\unskip,\space}%
-    \global\let\@fnmark\@empty
-    \global\let\@corref\@empty  %% Added
-    \global\let\sep\@empty}%
-    \@eadauthor={#1}
-}
-\makeatother
-%% End block of weird code
-
-\journal{Physics of the Earth and Planetary Interiors}
-
-\begin{document}
-
-\begin{frontmatter}
-
-%% Title, authors and addresses
-
-%% use the tnoteref command within \title for footnotes;
-%% use the tnotetext command for theassociated footnote;
-%% use the fnref command within \author or \address for footnotes;
-%% use the fntext command for theassociated footnote;
-%% use the corref command within \author for corresponding author footnotes;
-%% use the cortext command for theassociated footnote;
-%% use the ead command for the email address,
-%% and the form \ead[url] for the home page:
-%% \title{Title\tnoteref{label1}}
-%% \tnotetext[label1]{}
-%% \author{Name\corref{cor1}\fnref{label2}}
-%% \ead{email address}
-%% \ead[url]{home page}
-%% \fntext[label2]{}
-%% \cortext[cor1]{}
-%% \address{Address\fnref{label3}}
-%% \fntext[label3]{}
-
-\title{Stability and accuracy of free surface time integration in viscous flows}
-
-%% use optional labels to link authors explicitly to addresses:
-%% \author[label1,label2]{}
-%% \address[label1]{}
-%% \address[label2]{}
-
-\author{Ian Rose\corref{cor1}\fnref{ref1}}
-\author{Bruce Buffett\fnref{ref1}}
-\author{Timo Heister\fnref{ref2}}
-
-\fntext[ref1]{University of California, Berkeley}
-\fntext[ref2]{Clemson University}
-\cortext[cor1]{Corresponding author, \url{ian.rose@berkeley.edu}}
-
-\address{}
-
-\begin{abstract}
 Geodynamic simulations increasingly rely on simulations with a true free surface to 
@@
 mantle convection software \texttt{ASPECT}.
-\end{abstract}
-
-\begin{keyword}
-Numerical modeling \sep
-Free surface \sep
-Mantle convection \sep
-Geodynamics
-%% keywords here, in the form: keyword \sep keyword
-
-%% PACS codes here, in the form: \PACS code \sep code
-
-%% MSC codes here, in the form: \MSC code \sep code
-%% or \MSC[2008] code \sep code (2000 is the default)
-
-\end{keyword}
-
-\end{frontmatter}
-
-\linenumbers
-
-%% main text
-\section{Introduction}
-\label{sec:intro}
-
-Surface topography in simulations of mantle convection and other geodynamic processes is an important observable,
-allowing insights into Earth's internal density structure, rheology, and geoid perturbations \citep[e.g.][]{richards1984geoid, hager1985lower, baumann2014constraining}.
-Historically, most simulations have been performed with free-slip boundary conditions at the surface, 
-with dynamic topography calculated as a postprocessing step \citep[e.g.][]{jarvis1982mantle, zhong2000role}.
-
-There have been several approaches to treating real free surfaces in geodynamic simulations.
-\citet{zhong1996free} introduced a pseudo-free-surface formulation, where the free surface coordinate was
-treated as an extra variable that was integrated in time. In this formulation, the free surface 
-is approximated on the undeformed Eulerian grid by applying pressure boundary conditions on the reference surface.
-The pressure is determined by a first-order Taylor series approximation of the hydrostatic pressure profile 
-predicted by the surface topography.
-
-A large number of studies have approximated free surfaces in the interior of the domain by using the 
-`sticky air' approximation (see \citet{crameri2012comparison} and references therein). In this approximation there is a low-viscosity, low-density layer in the fluid 
-(termed `air', though its viscosity is much higher) above the free surface, effectively decoupling it from the boundary. 
-Typically a free-slip boundary condition is used above the sticky air, though an open boundary may be preferable \citep{hillebrand2014using}.
-
-Finally, one can use a true free surface, where a stress-free boundary condition is applied on the boundary 
-of the domain. In this case, there can be flow in and out of the domain, so the boundary must move in time.
-A true free surface has mathematical elegance in that the boundary condition of the domain more closely 
-matches the boundary conditions which one is trying to model, but it typically requires a deformable 
-domain with frequent remeshing to avoid ill-conditioned cells.
-
-Recently, the nature of surface boundary conditions have been shown to be important for controlling the 
-dynamics of subduction zone modeling.  In a benchmark study \citet{schmeling2008benchmark} performed extensive testing on the effect
-of a free surface on the sinking of a slab. They found that the nature of the free surface had a large effect 
-on the dynamics of the slab, affecting both the shape and the timing of sinking. Most of the 
-participating codes in that benchmark used the sticky air approximation.
-They found that the specific properties of the sticky air layer controlled the shape and timing of the slab.
-Furthermore, the viscosity averaging scheme for areas of transitional
-composition was extremely important, as the subducting slab entrained significant amounts of the sticky air.
-\citet{crameri2012comparison} conducted comparisons between sticky air and true free surface models, 
-demonstrating a range of parameters for sticky air which can mitigate some of the difficulties that it introduces.
-A study by \citet{quinquis2011role} found that free surface boundary conditions have a large effect on 
-trench migration in a subduction zone, and \citet{crameri2012free} found that a free surface combined 
-with a weak crust is important for producing one-sided subduction zones.
-
-All of the approaches to free surface simulations have been subject to an instability which has been 
-variously termed a ``sloshing,'' or ``drunken sailor'' instability \citep{kaus2010stabilization, duretz2011discretization, kramer2012implicit}. 
-This instability, arising from the large density contrast typical at a free surface (compared with the much smaller 
-internal density contrasts), severely limits the maximum stable timestep for free surface computations.
-Frequently, the maximum stable timestep is several orders of magnitude smaller than that for an otherwise 
-equivalent model with free-slip boundary conditions.
-
-Several studies have attempted to alleviate the timestepping requirements imposed by the sloshing 
-instability. Since the most expensive part of geodynamic simulations is typically the Stokes solve,
-most free surface calculations have preferred to use explicit timestepping methods for the free 
-surface. \citet{kaus2010stabilization} proposed a quasi-implicit scheme which modifies the discretized 
-Stokes matrix, giving it better stability properties.  \citet{popov2008slim3d}, \citet{kramer2012implicit}, and \citet{furuichi2015implicit}
-explored methods for solving the transport of the free surface implicitly.
-
-The paper is organized as follows.
-After introducing the problem in Section~\ref{sec:governing},
-we derive an approach to analyze 
-free surface schemes based on the normal modes in Section~\ref{sec:eigenvalue}
-and Section~\ref{sec:timestepping}.
-In Section~\ref{sec:kmm} we use this approach to look at the
-quasi-implicit stabilization proposed in \citet{kaus2010stabilization}.
-Then, we propose a new time marching scheme for free surface computations with 
-good stability properties (Section~\ref{sec:newscheme}).
-Finally, we describe the implementation of a free surface in the mantle convection software \texttt{ASPECT} in Section~\ref{sec:implementation}, 
-before showing some numerical results and benchmarks in Section~\ref{sec:results}.
-
-\section{Governing equations}
-\label{sec:governing}
-
-We begin with the incompressible momentum conservation equations for creeping incompressible flow:
-\begin{equation}
-\begin{aligned}
-\nabla \cdot \mathbf{T} &= \rho \mathbf{g} \\
-\nabla \cdot \mathbf{u} &= 0,
-\end{aligned}
-\label{eq:momentum_mass_conservation}
-\end{equation}
-where $\mathbf{u}$ is the fluid velocity, $\rho$ is the fluid density, and $\mathbf{g}$ is the force due to gravity.
-$\mathbf{T}$ is the stress tensor for a Newtonian fluid, defined by
-\begin{equation}
-\mathbf{T} = 2 \eta \varepsilon(\mathbf{u}) - p \mathbf{I},
-\label{eq:stress_tensor}
-\end{equation}
-where $\eta$ is the viscosity, $\varepsilon(\mathbf{u}) = \frac{1}{2}(\nabla \mathbf{u} + (\nabla \mathbf{u} )^T )$
-is the strain-rate tensor, and $\mathbf{I}$ is the identity tensor.
-Substituting the stress tensor into Equation~\eqref{eq:momentum_mass_conservation} gets the familiar form of the Stokes equation:
-\begin{equation}
-\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p = \rho \mathbf{g}.
-\label{eq:stokes}
-\end{equation}
-
-For the purposes of this analysis it is useful to define a hydrostatic reference state where the 
-velocity $\mathbf{u}$ is zero:
-\begin{equation}
-- \nabla p_0 = \rho_0 \mathbf{g},
-\label{eq:hydrostatic_stokes}
-\end{equation}
-where $p_0$ is the reference hydrostatic pressure and $\rho_0$ is a reference density profile (which may vary with depth).
-The total pressure and density can then be decomposed into variations about their 
-reference values: $\rho = \rho_0 + \rho^\prime$, $p = p_0 + p^\prime$.
-Using this, the hydrostatic reference state (Equation~\eqref{eq:hydrostatic_stokes}) may be subtracted from Equation~\eqref{eq:stokes}.
-
-This gives rise to the following time dependent, coupled system with unknowns $\mathbf{u}(t)$ and $\mathbf{p}(t)$:
-\begin{equation}
-\begin{aligned}
-\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p^\prime &= \rho^\prime \mathbf{g} \qquad &\text{in } &\Omega(t)\\
-\nabla \cdot \mathbf{u} &= 0  \qquad &\text{in } &\Omega(t)
-\end{aligned}
-\label{eq:final_system}
-\end{equation}
-defined on the bounded, moving domain $\Omega(t)\subset \mathbb{R}^d$ with boundary $\partial \Omega = \Gamma_0 \cup \Gamma_F$
-split into fixed (Dirichlet) and free surface parts $\Gamma_0$ and $\Gamma_F$, respectively.
-The boundary conditions are given by
-\begin{equation}
-\begin{aligned}
-\mathbf{u} &= 0 &\qquad \text{on } &\Gamma_0 \\
-\mathbf{T}\cdot \mathbf{n} &= 0 &\qquad \text{on } &\Gamma_F.
-\end{aligned}
-\end{equation}
-For the sake of economy we neglect inhomogeneous stress boundary conditions, though it would be straightforward to include them.
-We also defer discussion of free slip boundary conditions.
-The domain at time $t$ is defined by advecting a reference configuration $\Omega_0$ by the domain velocity $\mathbf{u}(t)$
-\begin{equation}
- \Omega(t) = \Omega_0 + \int_{t_0}^t \mathbf{u}(t) \text{ d}t.
- \label{eq:domain_evolution}
-\end{equation}
-We defer the discussion of the velocity of the discretized domain to Section~\ref{sec:remeshing}.
-Finally, we denote the displacement at time $t$ by $\zeta = \int_{t_0}^t \mathbf{u}_\mathrm{surface}(t)\cdot \mathbf{n} \text{ d}t$, leading to the evolution
-equation
-\begin{equation}
-\frac{\text{d} \zeta}{\text{d}t} = \mathbf{u \cdot \mathbf{n}} \quad \textrm{on  }  \Gamma_F.
-\label{eq:surface_evolution}
-\end{equation}
-
-\section{Eigenvalue analysis}
-\label{sec:eigenvalue}
-
-In order to better understand the time evolution of the system in Equation~\eqref{eq:final_system}
-we will consider the eigenvalues of a simplified homogeneous system where we neglect variations in density ($\rho^\prime = 0$):
-\begin{equation}
-\begin{aligned}
-\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p^\prime &= 0 \\
-\nabla \cdot \mathbf{u} &= 0.
-\end{aligned}
-\label{eq:homogeneous_stokes}
-\end{equation}
-
-We will proceed with this analysis within a finite element framework, though similar arguments should 
-work for other discrete methods.
-We transform the governing equations into the weak form amenable to finite elements via standard operations \citep[e.g.][]{zienkiewicz1977finite} to get
-
-\begin{equation}
--\int_{\Omega(t)} 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u} ) + \int_{\Omega(t)} p^\prime \nabla \cdot \mathbf{w} 
-+ \int_{\Gamma_F(t)} \mathbf{w} \cdot \mathbf{T} \cdot \mathbf{n} = 0 
-\label{eq:weak_stokes}
-\end{equation}
-\begin{equation}
-\int_{\Omega(t)} q \nabla \cdot \mathbf{u} = 0,
-\label{eq:weak_incompressible}
-\end{equation}
-where $\mathbf{w}$ and $q$ are suitably chosen test functions and the integrals over 
-$\Omega(t)$ and $\Gamma_F(t)$ are over the volume of the domain and the free surface, respectively.
-Note that since the free surface can move, the shape of the domain is a function of time.
-
-The integral over the surface in Equation~\eqref{eq:weak_stokes} accounts for boundary stresses, 
-which should be zero when evaluated on a true free surface.
-Rather than analyzing finite deformation to the free surface (a nonlinear problem),
-we will make the analytically useful approximation of small deformations about the hydrostatic 
-reference surface and analyze their stability.
-We will therefore  evaluate the integrals in Equation~\eqref{eq:weak_stokes} 
-over the hydrostatic reference surface and introduce a temporary auxilliary variable $\zeta$ which 
-represents the (small) topography of the free surface relative to that reference surface.
-On the reference surface the gravity vector is opposite the direction of the surface normal $\mathbf{g} = -g \mathbf{n}$.
-The stress on this reference surface can be approximated by using the first order Taylor series
-expansion of the hydrostatic pressure profile:
-\begin{equation}
-\mathbf{T} \approx \frac{\partial \mathbf{T}}{\partial \mathbf{n} } \cdot \mathbf{n} \zeta = \rho_0 g \zeta \mathbf{I}.
-\label{eq:hydrostatic}
-\end{equation}
-Equation~\eqref{eq:weak_stokes} then becomes
-
-\begin{equation}
--\int_{\Omega(t)} 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u} ) + \int_{\Omega(t)} p^\prime \nabla \cdot \mathbf{w} 
-+ \int_{\Gamma_F(t)} \rho_0 g \zeta  \mathbf{w} \cdot \mathbf{n} = 0.
-\end{equation}
-
-We would like to analyze the time evolution of the normal modes of this system: each mode 
-is the relaxation of topography with a characteristic relaxation time.  
-We denote the normal modes by $\left[ \mathbf{u}_i, p^\prime_i, \zeta_i \right]^T$, with
-relaxation times $\tau_i$, where the subscript corresponds to the $i$th normal mode.
-
-The equations decouple for the normal modes, and Equation~\eqref{eq:surface_evolution} then becomes
-
-\begin{equation}
-\frac {\text{d}}{\text{d} t} \zeta_i(\mathbf{x},t) = \frac{\text{d}}{\text{d}t} \zeta_i(\mathbf{x})e^{-t/\tau_i} = -\frac{\zeta_i(\mathbf{x},t)}{\tau_i} = \mathbf{u}_i \cdot \mathbf{n}.
-\end{equation}
-This can then be used to eliminate $\zeta$ from the Stokes system:
-\begin{equation}
--\int_{\Omega(t)} 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u}_i ) + \int_{\Omega(t)} p^\prime_i \nabla \cdot \mathbf{w} 
-= \tau_i \int_{\Gamma_F(t)} \rho_0 g (\mathbf{u}_i \cdot \mathbf{n} ) (\mathbf{w} \cdot \mathbf{n}).
-\label{eq:weak_eigen}
-\end{equation}
-
-Equations~\eqref{eq:weak_stokes},~\eqref{eq:weak_incompressible}, and~\eqref{eq:weak_eigen}
-are continuous forms of the governing equations which must be discretized in space and time in order to be solved numerically.
-In particular, we must decide at which timestep (or combination of timesteps) we will integrate over the domain $\Omega(t)$.
-We denote discrete times by $t^n$, where $n$ is the timestep number, and $t^{n+1} = t^n+\Delta t$.
-It is most convenient to integrate over the domain at the current timestep $\Omega^n \equiv \Omega(t^n)$ instead of 
-the unknown $\Omega^{n+1} \equiv \Omega(t^{n}+\Delta t)$. This corresponds to an explicit scheme.
-However, as we will see, the implicit scheme of integrating over the (unknown) domain at the next timestep $\Omega^{n+1}$ will
-give a stable timestepping scheme, at the cost of making the problem nonlinear \citep{furuichi2015implicit}.
-
-When these equations are spatially discretized using finite elements \citep[see, e.g.][]{kronbichler2012high} we get
-\begin{equation}
-\begin{bmatrix}
-A & B^T \\
-B & 0 \\
-\end{bmatrix}
-\begin{bmatrix}
-\mathbf{u}^h_i \\
-p^h_i
-\end{bmatrix}
-=
-\tau_i
-\begin{bmatrix}
-M & 0 \\
-0 & 0
-\end{bmatrix}
-\begin{bmatrix}
-\mathbf{u}^h_i \\
-p^h_i
-\end{bmatrix},
-\label{eq:generalized_eigenvalue}
-\end{equation}
-where $\mathbf{u}^h_i$ and $p^h_i$ are finite-dimensional representations of $\mathbf{u}_i$ and $p^\prime_i$,
-and the domain integrals are performed over a discrete approximation of the domain shape
-(such as a linear finite element basis) which we denote by $\hat{\Omega}$.
-$M$ is the discretization of the bilinear form on the right-hand side of Equation~\eqref{eq:weak_eigen}.
-The form of Equation~\eqref{eq:generalized_eigenvalue} does not depend on whether we choose $\hat{\Omega}^n$ or $\hat{\Omega}^{n+1}$
-for the domain integration.
-
-Equation~\eqref{eq:generalized_eigenvalue} is a generalized eigenvalue problem for the normal modes of the system.
-It is rather more difficult to solve than a standard eigenvalue problem because the matrix on the right-hand-side 
-is not invertible. It may, however, be transformed into a standard eigenvalue problem.
-By defining
-\begin{equation}
-\begin{aligned}
-C &= 
-\begin{bmatrix}
-A & B^T \\
-B & 0 \\
-\end{bmatrix} \quad
-D &= 
-\begin{bmatrix}
-M & 0 \\
-0 & 0
-\end{bmatrix} \quad
-\mathbf{y}_i &= 
-\begin{bmatrix}
-\mathbf{u}^h_i \\
-p^h_i
-\end{bmatrix} 
-\end{aligned}
-\label{eq:eigen_substitution}
-\end{equation}
-and then multiplying both sides by $\tau_i^{-1}C^{-1}$ we get
-\begin{equation}
-(C^{-1}D)\mathbf{y}_i = \tau_i^{-1} \mathbf{y}_i.
-\label{eq:standard_eigenvalue}
-\end{equation}
-
-This eigenvalue equation can be solved for the normal modes and relaxation times of the Stokes system with 
-a free surface.
-
-\section{Time integration of the free surface}
-\label{sec:timestepping}
-
-Armed with the normal modes and relaxation times of the Stokes system, we can write down the
-formal solution to Equation~\eqref{eq:surface_evolution}. Let the initial surface topography be 
-represented by a linear combination of its normal modes
-\begin{equation}
-\zeta(\mathbf{x}, t=0) = \displaystyle \sum_i a_i \zeta_i(\mathbf{x}),
-\end{equation}
-then the time evolution of the free surface is given by
-\begin{equation}
-\zeta( \mathbf{x}, t) = \displaystyle \sum_i a_i \zeta_i(\mathbf{x}) e^{-t/\tau_i}.
-\end{equation}
-Of course, most finite element (or finite difference, or finite volume) geodynamic simulations do not resolve 
-the solution and surface topography into its normal modes and integrate those separately. 
-Indeed, analytical solutions for the normal modes are only known for simple geometries and rheologies.
-Instead, one typically obtains a velocity solution and then advect the free surface with the local velocity.
-The normal mode solution is instructive, however. 
-Each mode has the form of a decay equation with characteristic decay time $\tau_i$.
-The decay equation is the archetypical example of a stiff ordinary differential equation.
-If we were to numerically integrate this in time with a forward Euler method, we would find the 
-timestep criterion for stability \citep[e.g.][]{leveque2007finite} to be
-\begin{equation}
-\Delta t  \le 2 \tau_{\mathrm{min}},
-\label{eq:cfl_euler}
-\end{equation}
-where $\tau_{\mathrm{min}}$ is this minumum decay time.
-The maximum stable timestep is limited by the minimum relaxation timescale.
-If a larger timestep than this is taken then those modes will become unstable.
-The modes with the smallest relaxation times are usually those with the largest lengthscales \citep{schubert2001mantle}, 
-and so it will be those which become unstable first, a phenomenon which has been called 
-a ``sloshing'', or ``drunken sailor'' instability \citep{kaus2010stabilization}.
-
-\section{Analysis of quasi-implicit stabilization}
-\label{sec:kmm}
-
-The relaxation timescales for surface topography tend to be considerably shorter than those for 
-convection or tectonic deformation, so the stability requirements for a forward Euler scheme
-are quite onerous.  On the other hand, an implicit time marching scheme requires solving 
-a nonlinear system for the new surface position, or assembling a larger system with surface
-topography unknowns \citep[e.g.][]{kramer2012implicit}.
-
-\citet{kaus2010stabilization} proposed a scheme whereby the body forces on the domain are 
-evaluated on a prediction of the shape of the domain at a later time.
-The continuous weak form of the right-hand-side body forces in is then
-\begin{equation}
-\mathbf{f}_{\mathrm{body}} = \int_{\Omega^n + \Delta \Omega} \rho \mathbf{w} \cdot \mathbf{g}.
-\label{eq:predict}
-\end{equation}
-We can approximate the shape prediction by time integrating the domain evolution equation (Equation~\eqref{eq:domain_evolution}):
-\begin{equation}
-  \Delta \Omega \approx \theta \Delta t \mathbf{u},
-\end{equation}
-where $\theta$ is a free parameter that corresponds to the magnitude of the 
-correction, where zero is no stabilization and one is fully (quasi) implicit.
-
-One can approximately expand the integral in Equation~\eqref{eq:predict} using 
-Reynold's transport theorem to find
-
-\begin{equation}
-\int_{\Omega^n + \Delta \Omega} \rho  \mathbf{w} \cdot \mathbf{g} \approx
-\int_{\Omega^n} \rho  \mathbf{w} \cdot \mathbf{g} + \theta \Delta t \int_{\Gamma_F^n} \rho ( \mathbf{w \cdot g}) (\mathbf{u \cdot n} ).
-\label{eq:kmm}
-\end{equation}
-The volume integral is the same as that of the unstabilized problem, but now we obtained an additional surface integral.
-It has the form of a velocity-dependent surface stress pushing down on the 
-domain, and can be thought of as an artificial viscous damping of the surface.
-Since the term depends on the velocity, it enters the Stokes system matrix as a stabilization term.  
-Empirically, it has been found to be successful at damping instabilities \citep{kaus2010stabilization, quinquis2011role, duretz2011discretization}.
-
-The surface integral in Equation~\eqref{eq:kmm} is almost the same as that on the right hand side of 
-Equation~\eqref{eq:weak_eigen}, which was discretized as the matrix $M$.
-By making two approximations we can connect it with the formalism developed in Section~\ref{sec:eigenvalue}.
-First, when the current position of the surface is near to the reference surface (i.e., the slope is not too steep)
-then gravity is approximately in the direction of the surface normal: $\mathbf{g} \approx -g \mathbf{n}$.
-Furthermore, in Equation~\eqref{eq:weak_eigen} we considered response of the surface in a homogeneous system where there were no lateral density variations,
-while Equation~\eqref{eq:kmm} uses the total density.
-In most geodynamic simulations lateral density variations are small compared to the total density.
-For the purposes of stabilization we may approximate $\rho \approx \rho_0$.
-Inserting these two approximations, we find that the stabilization term may be written as
-\begin{equation}
--\theta \Delta t \int_{\Gamma_F^n} \rho_0 g ( \mathbf{w \cdot n}) (\mathbf{u \cdot n} ),
-\label{eq:kmm_stabilization}
-\end{equation}
-where the integral is now identical to that which discretizes to $M$.
-
-If we discretize the Stokes system with the quasi-implicit stabilization term, we find
-a new generalized eigenvalue problem:
-\begin{equation}
-\begin{bmatrix}
-A + \theta \Delta t M & B^T \\
-B & 0 \\
-\end{bmatrix}
-\begin{bmatrix}
-\mathbf{u}^h_i \\
-p^h_i
-\end{bmatrix}
-=
-\tau^S_i
-\begin{bmatrix}
-M & 0 \\
-0 & 0
-\end{bmatrix}
-\begin{bmatrix}
-\mathbf{u}^h_i \\
-p^h_i
-\end{bmatrix},
-\label{eq:stabilized_generalized_eigenvalue}
-\end{equation}
-where $\tau^S_i$ indicate the eigenvalues of the stabilized system.
-This system may be rearranged:
-\begin{equation}
-\begin{bmatrix}
-A & B^T \\
-B & 0 \\
-\end{bmatrix}
-\begin{bmatrix}
-\mathbf{u}^h_i \\
-p^h_i
-\end{bmatrix}
-=
-\left(\tau^S_i - \theta \Delta t \right)
-\begin{bmatrix}
-M & 0 \\
-0 & 0
-\end{bmatrix}
-\begin{bmatrix}
-\mathbf{u}^h_i \\
-p^h_i
-\end{bmatrix}.
-\label{eq:rearranged_stabilized_generalized_eigenvalue}
-\end{equation}
-
-This is precisely the same generalized eigenvalue problem as Equation~\eqref{eq:generalized_eigenvalue},
-so its eigenvalues must be the same.  This allows us to write the eigenvalues of the stabilized problem in
-terms of those of the unstabilized problem:
-\begin{equation}
-\tau^S_i = \tau_i + \theta \Delta t.
-\label{eq:spectrum_shift}
-\end{equation}
-
-Essentially, the stabilization term lengthens every relaxation time for the system by an amount $\theta \Delta t$.
-This correspondingly lengthens the maximum stable timestep for the forward Euler method:
-\begin{equation}
-\Delta t  \le \frac{2}{1-\theta} \tau_{\mathrm{min}}.
-\label{eq:cfl_euler_stabilized}
-\end{equation}
-Note that as $\theta$ goes to one this scheme should become unconditionally stable,
-but nonlinear effects and discretization errors due to finite deformation of the surface could 
-prevent that stability.
-
-The lengthening of relaxation times due to the quasi-implicit stabilization has an unequal 
-effect on the modes. They are all lengthened by the same amount, which is a much bigger 
-fraction of the total relaxation time for the shorter-time modes than the longer-time ones.
-Therefore the shorter-time modes are effectively damped more.
-This can be seen as an attractive feature of the scheme, since those are the least stable modes,
-though it can result in less accurate time-marching of the longest wavelengths of the system.
-
-\subsection{Numerical determination of $\tau_\mathrm{min}$}
-\label{sec:numerical_determination}
-
-It can be useful to determine the minimum relaxation time $\tau_{\mathrm{min}}$ for a particular
-simulation setup, both to determine an appropriate timestep and as a check on the eigenvalue
-analysis above. As discussed in Section~\ref{sec:eigenvalue}, the generalized eigenvalue problem
-given by Equation~\eqref{eq:stabilized_generalized_eigenvalue} is difficult to solve because the matrix on
-the right-hand-side is not invertible.
-However, we may convert Equation~\eqref{eq:stabilized_generalized_eigenvalue} to a standard eigenvalue problem
-in the same manner as before. Introducing $C_S$ as the Stokes system matrix with the stabilization term
-included, and with $D$ and $\mathbf{y}_i$ the same as in Equation~\eqref{eq:eigen_substitution}, we find
-\begin{equation}
-(C_S^{-1}D)\mathbf{y}_i = \frac{1}{\tau^S_i} \mathbf{y}_i.
-\label{eq:stabilized_standard_eigenvalue}
-\end{equation}
-Several things are of note about this eigenvalue equation which help in solving it.
-The matrix $D$ is a mass matrix which is only nonzero on the free surface,
-so it is mostly empty, and, in general, it is not necessary to explicitly form the
-matrix in order to apply it.
-Second, applying the inverse of $C_S$ amounts to solving the Stokes system, so any
-existing geodynamical code can already do this.
-Third, it is now an equation for the inverse of $\tau^S_i$, and so solving for the 
-minimum relaxation time is now equivalent to finding the \emph{maximum} eigenvalue
-of Equation~\eqref{eq:stabilized_standard_eigenvalue}.
-
-All of these factors mean that power iteration (see, e.g., \citet{golub2012matrix}),
-which finds the dominant eigenvalue/eigenvector pair, is an attractive option for finding $\tau_\mathrm{min}$.
-We have tested the effectiveness of power iteration for finding the minimum relaxation time
-on a problem for which there is an analytical solution.
-We also use it to confirm Equation~\eqref{eq:spectrum_shift}.
-Following \citet{kramer2012implicit} we model the relaxation
-of surface topography in a two-dimensional Cartesian box with an isoviscous fluid.
-The setup is shown in Figure~\ref{fig:benchmark_setup}.
-The eigenvectors of this system are given by infinitesimal sinusoidal perturbations to the free surface
-\begin{equation}
-\zeta(x,0) = \zeta_0 \cos\left( k x \right),
-\label{eq:initial_topography}
-\end{equation}
-where $k = 2 \pi n / L$, and $n$ is a wavenumber which is an integer multiple of $1/2$.
-For an infinitesimal perturbation of this form, the relaxation time $\tau$ is given by
-\begin{equation}
-\tau = \frac{D k + \sinh(D k) \cosh(D k)}{ \sinh^2 (D k) } \frac{2 k \eta}{\rho g},
-\label{eq:analytic_relaxation_time}
-\end{equation}
-where $D$ is the layer depth, $\eta$ is the viscosity, $\rho$ is the density, and $g$ is the acceleration due to gravity.
-
-\begin{figure*}
-\includegraphics[width=0.9\textwidth]{figures/benchmark_setup.pdf}
-\caption{Setup for the power iteration test (Section~\ref{sec:numerical_determination}) and for the free surface relaxation benchmark (Section~\ref{sec:topography_relaxation}).
-A 2D box with an isoviscous fluid has sinusoidal initial topography, with amplitude $\zeta_0$ and wavenumber $n$.
-The box has depth $D$ and length $L$.
-For our tests $\rho = \eta = g = D = L = 1$, and $n=1/2$.
-For the power iteration test there is no initial perturbation ($\zeta_0 = 0.0$), and for the surface relaxation benchmark $\zeta_0 = 0.005$.}
-\label{fig:benchmark_setup}
-\end{figure*}
-The eigenvector with the minimum relaxation time for this system $\tau_{\min}$ corresponds to $n=1/2$ 
-(which, as expected, has the form of the fluid sloshing back and forth).
-Figure~\ref{fig:perturb_spectrum} shows the results of the power iteration, where colored circles indicate the numerical
-solution and black dots show the analytical solution using equation~\eqref{eq:spectrum_shift}.
-In most cases the iteration converges to better than $\sim10^{-5}$ relative accuaracy in about 20 iterations
-(and fewer with a good initial guess).
-
-\begin{figure*}
-\includegraphics[width=0.9\textwidth]{figures/perturb_spectrum.pdf}
-\caption{The effect of quasi-implicit stabilization on the minimum relaxation time of a fluid in a 2D Cartesian box with a free surface (see Figure~\ref{fig:benchmark_setup}), as well as the result of power iteration to find the minimum relaxation time.
-The $x$-axis shows the timestep normalized by minimum relaxation time of the unstabilized problem $\tau_\mathrm{\min}$.
-The $y$-axis shows the stabilized relaxation time $\tau_S$, also normalized by $\tau_\mathrm{min}$.
-Colored circles show the relaxation times found by numerical solution of Equation~\eqref{eq:stabilized_standard_eigenvalue} with different values of the stabilization parameter $\theta$.
-Black dots show the relaxation times calculated using Equation~\eqref{eq:spectrum_shift}.
-The stabilized timescale is lengthened for larger timesteps and for larger values of $\theta$.
-For this test case, power iteration typically attains better than 1\% accuracy in 5-10 iterations.}
-\label{fig:perturb_spectrum}
-\end{figure*}
-
-\section{A novel time-integration scheme}
-\label{sec:newscheme}
-
-The implementation of such a scheme results in a slightly 
-asymmetric matrix, which can be more difficult to solve, requiring either
-changes to the solver/preconditioner, or symmetrization of the stabilization term \citep{kaus2010stabilization}.
-An alternative would be to construct a time-integration scheme that remains stable at larger timesteps.
-The reason that forward Euler integration performs so badly with the decay of topography is that 
-at larger timesteps it overshoots its equilibrium position. This overshoot causes it to lurch back 
-to the \emph{other} side of equilibrium, overshooting even more.
-We would like to construct an explicit scheme that accounts for following properties which we know the system has:
-\begin{itemize}
-\item In the absence of forcing, topography always decreases.
-\item Relaxation of small amplitude topography takes the form of exponential decay.
-\item The decay mode with the shortest relaxation time is the least stable.
-\end{itemize}
-
-\subsection{Nonstandard finite-differences}
-
-A general expression for the evolution of a free surface from time $t^n$ to time $t^{n+1}$ is 
-\begin{equation}
-\mathbf{x}(t^{n+1}) = \mathbf{x}(t^n) + \int_{t^{n}}^{t^{n+1}} \mathbf{u}(t) \text{ d}t,
-\label{eq:time_integration}
-\end{equation}
-where $\mathbf{x}$ is the location of the free surface.  If we approximate $\mathbf{u}(t) \approx \mathbf{u}(t^{n})$, 
-we of course recover the forward Euler scheme.
-However, we can make another choice based on our knowledge of the system behavior. 
-We can approximate $\mathbf{u}(t)$ as
-\begin{equation}
-\mathbf{u}(t) = \mathbf{u}(t^n) e^{-(t-t^n)/\tau^*},
-\label{eq:velocity_decay}
-\end{equation}
-where $\tau^*$ is some as-yet undetermined positive constant which we will refer to as the ``stabilization timescale''.
-This form of $\mathbf{u}$ automatically decays in time, and as we shall see, has much better 
-stability properties than forward Euler integration.
-Using Equation~\eqref{eq:velocity_decay} in Equation~\eqref{eq:time_integration} and integrating, we find
-
-\begin{equation}
-\mathbf{x}(t^{n+1}) = \mathbf{x}(t^n) + \mathbf{u}(t^{n}) \tau^* \left(1-e^{-\Delta t/\tau^*} \right).
-\label{eq:nsfd}
-\end{equation}
-
-The quantity $\tau^*(1-e^{-\Delta t / \tau^*})$ acts as a pseudo-timestep for advecting the free-surface.
-Equation~\eqref{eq:nsfd} is what is known as a nonstandard finite-difference model, based on
-constructing unusual discrete models for differential equation integration.
-The theory has been developed in, among others, a series of papers by
-\citet{mickens1994nonstandard, mickens2002nonstandard, mickens2005dynamic}.
-
-\subsection{Stability of the scheme}
-The parameter $\tau^*$ sets how quickly $\mathbf{u}$ decays in Equation~\eqref{eq:velocity_decay}, and a good 
-choice is crucial for accuracy and stability. A shorter decay time corresponds to more stabilization,
-but if it becomes too short, the surface velocity will become overdamped. In general, we will want 
-to choose $\tau^*$ so that it is as close as possible to the relaxation time of the least stable mode, or $\tau_{\mathrm{min}}$.
-
-To investigate the stability of this scheme we consider a velocity solution comprised of 
-a single normal mode of the system $\mathbf{u} = a_i \mathbf{u}_i$.
-This will decay exponentially with relaxation time $\tau_i$, or
-\begin{equation}
-\frac{\text{d} a_i} {\text{d}t} = - \frac{ a_i }{\tau_i}.
-\end{equation}
-Applying the nonstandard finite difference scheme, we find
-\begin{equation}
-a_i^{n+1} = a_i^{n} \left[ 1 - \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \right].
-\label{eq:recursion}
-\end{equation}
-In order for the scheme to be stable, the quantity in brackets must not blow up as it is repeatedly multiplied by itself, or 
-\begin{equation}
-\left| 1 - \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \right| \le 1.
-\end{equation}
-
-\ifdetail
-Taking the positive value of the absolute value yields the criterion of 
-\begin{equation}
-- \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \le 0,
-\end{equation}
-which can be rearranged to find
-\begin{equation}
-e^{-\Delta t/\tau^*} \le 1,
-\end{equation}
-which is simply a statement that the timestep must be postive.
-
-Taking the negative value of the absolute value yields
-\begin{equation}
-  \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right)  \le 2
-\end{equation}
-\begin{equation}
- 1 - 2 \frac{\tau_i}{\tau^*} \le e^{-\Delta t/\tau^*}
-\end{equation}
-if the left hand side is less than zero, then this is true regardless of step size, or
-\begin{equation}
-\tau^* \le 2 \tau_i
-\end{equation}
-if the left hand side is between zero and one, the expression is more complicated. 
-Taking the log of both sides:
-\begin{equation}
-\Delta t \le -\tau^* \log \left(1 - 2 \frac{\tau_i}{\tau^*} \right).
-\end{equation}
-It is convenient to write this in terms of dimensionless times $\Delta t/ \tau_i$ and $\tau^*/\tau_i$:
-\begin{equation}
-\frac{\Delta t}{\tau_i} \le -\frac{\tau^*}{\tau_i} \log \left(1 - 2 \frac{\tau_i}{\tau^*} \right).
-\end{equation}
-
-\fi
-
-The stability of this scheme is determined by the choice of $\tau^*$.  
-It has a region of unconditional stability, where
-\begin{equation}
-\tau^* \le 2 \tau_i.
-\end{equation}
-Otherwise, the scheme is conditionally stable, with the timestep limited by 
-\begin{equation}
-\Delta t \le -\tau^* \log \left(1 - 2 \frac{\tau_i}{\tau^*} \right).
-\end{equation}
-The stability region is plotted in Figure~\ref{fig:stability_region}.
-
-\begin{figure*}
-\includegraphics[width=0.9\textwidth]{figures/stability_region.pdf}
-\caption{(a) Stability region for the nonstandard finite difference scheme (green).
-On the x-axis is the value of the stabilization timescale, in units of the minimum relaxation time $\tau_{\mathrm{min}}$.
-On the y-axis is the value of the timestep, also in units of the minimum relaxation time.
-For $\tau^*\le 2 \tau_{\mathrm{min}}$ the nonstandard finite difference scheme is unconditionally stable.
-(b) Stability region for the quasi-implicit scheme (green).
-Again, the y axis is in units of the minumum relaxation time.
-The x-axis shows the value of the stabilization parameter $\theta$.
-For $\theta = 1$ the quasi-implicit scheme is unconditionally stable.
-In both cases the stability region for the forward Euler scheme is also shown in blue.}
-\label{fig:stability_region}
-\end{figure*}
-
-\subsection{Accuracy and asymptotics of the scheme}
-It is important to note that even though we derived the nonstandard finite difference scheme assuming a decaying exponential
-for $\mathbf{u}$, it is formally a first-order accurate scheme. As such, it is capable of capturing arbitrary motions 
-of the free surface, while allowing larger timesteps than forward Euler schemes.
-
-Again we may take one of the normal modes as an example and inspect the difference between the nonstandard finite-difference
-scheme and the analytical solution after one timestep:
-
-\begin{equation}
-\begin{aligned}
-a_i(\Delta t) - a_i^{1} &= a_i(0) e^{-\Delta t/\tau_i} - a_i{(0)} \left[ 1 - \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \right] \\
-                        &= a_i{(0)} \left[ e^{-\Delta t/\tau_i} - 1 + \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \right].
-\end{aligned}
-\end{equation}
-Note that when $\tau_i = \tau^*$ the nonstandard finite difference scheme is exact.
-The exponentials may be expanded to find
-\ifdetail
-\begin{equation}
-a_i(\Delta t) - a_i^{1} = - a_i{(0)} \left[ \left(\frac{\Delta t}{\tau_i}\right)^2 - \frac{\tau^*}{\tau_i} \left(\frac{\Delta t }{\tau^* }\right)^2 \right].
-\end{equation}
-\fi
-\begin{equation}
-a_i(\Delta t) - a_i^{1} = {a_i{(0)} } \left( \frac{\Delta t}{\tau_i} \right)^2 \left( 1 - \frac{\tau_i}{\tau^*} \right) + O(\Delta t^3).
-\end{equation}
-The local truncation error is quadratic in $\Delta t$, similar to the forward Euler, and summing this error over 
-many timesteps results in a factor of $\Delta t^{-1}$, demonstrating the first-order global accuracy~\citep[e.g.][]{leveque2007finite}.
-The choice of $\tau^*$ controls the size of the coefficient on the truncation error for the scheme.
-The error for a given mode becomes considerably smaller when $\tau^*$ is close to its natural relaxation times.
-
-It is helpful to take a closer look at the pseudo-timestep introduced in Equation~\eqref{eq:nsfd}: $\tau^*(1-e^{-\Delta t/\tau^*})$.
-As the timestep $\Delta t$ goes to zero, the pseudo timestep approaches $\Delta t$, recovering 
-the forward Euler scheme. However, as $\Delta t$ gets larger, the pseudo-timestep does not 
-increase as quickly, reflecting the decaying nature of the normal modes
-(in fact, the pseudo-timestep is bounded between $\Delta t$ and $\tau^*$).
-Likewise, as the stabilization timescale $\tau^*$ goes to infinity, we also recover the 
-forward Euler scheme, in what is essentially the unstabilized problem. But as $\tau^*$
-goes to zero, the pseudo-timestep also goes to zero. This corresponds to over-stabilizing the
-problem. With too short of a stabilization timescale the free surface is never advected 
-at all (which is a very stable situation, if not very accurate!).
-
-\subsection{Choice of $\tau^*$}
-As discussed above, a full geodynamic simulation will have a spectrum of relaxation times.
-For complete stability, the parameter $\tau^*$ should be chosen such that every mode is stable.
-In practice, this means that a good choice is 
-\begin{equation}
-\tau^* \approx \tau_{\mathrm{min}}.
-\label{eq:tau_choice}
-\end{equation}
-
-Unfortunately, for many models the minimum relaxation time will not be known beforehand. 
-In order to use the nonstandard finite difference scheme, the value of $\tau^*$ would need 
-to be calculated or estimated first.  There are several possible ways to determine this value:
-
-\begin{itemize}
-\item Direct solution of Equation~\eqref{eq:generalized_eigenvalue}. Solution of the whole spectrum is expensive
-when only the minimum relaxation time is required, which means that power iteration on 
-the standard eigenvalue problem (Equation~\eqref{eq:standard_eigenvalue}) can be enough.
-\item Analytical formulae. Several geometry and viscosity model combinations have analytical solutions
-for relaxation spectra. Even if the user's model is not precisely the same (e.g., has some lateral viscosity
-variations), an analytical approximation may be sufficient.
-\item Scaling. In general, we expect the relaxation times to scale with $\tau \sim \frac{\eta}{\rho g L}$,
-where the density, gravity, viscosity, and lengthscale $L$ are all representative values.  A small amount
-of experimentation near this value of $\tau$ can find an appropriate relaxation time.
-\end{itemize}
-
-The first point deserves a bit more attention, since it provides the most general method for determining $\tau_\mathrm{min}$.
-We discussed the numerical determination of $\tau_\mathrm{min}$ in Section~\ref{sec:numerical_determination}
-by solving Equation~\eqref{eq:standard_eigenvalue} via power iteration.
-Each iteration requires a solution of the Stokes system (the $C$ matrix), and we have found
-that $\sim$10 iterations is usually enough to obtain a reasonable estimate of $\tau_\mathrm{min}$.
-This can be done during preprocessing, so over the course of the simulation the cost of determining
-$\tau_\mathrm{min}$ would be negligible.
-
-As a simulation progresses, there is the possibility that its viscosity or density structure
-will change, and as such its minimum relaxation time will evolve.
-This, of course, means that the best choice of $\tau^*$ can change over the course of a simulation.
-If $\tau_\mathrm{min}$ does not change significantly, the initial estimate can be fine.
-Otherwise, one can periodically check the evolution and redetermine $\tau_\mathrm{min}$ by 
-solving~\eqref{eq:standard_eigenvalue} during the model run.
-
-In section \label{sec:results} we show an examples the numerical solution of $\tau_\mathrm{min}$
-and its evolution in time.
-
-\label{sec:dynamic_relaxation}
-
-\section{Implementation in \texttt{ASPECT} }
-\label{sec:implementation}
-We have implemented the ability to run free surface simulations in the mantle convection software \texttt{ASPECT} \citep{aspectweb,kronbichler2012high}. 
-\texttt{ASPECT}, based on the open source, finite element library deal.II \citep{dealII82}, is designed to be highly flexible and modular, with the ability for user-defined rheologies, geometries, and gravity models.
-The implementation of the free surface, therefore, needs to be general enough to work for many combinations of these models, including those 
-which may not have been written yet. In particular, it cannot rely on assumptions regarding the shape of the domain.
-
-Furthermore, \texttt{ASPECT} allows for computations in both 2D and 3D, so the implementation must be sufficiently dimension
-independent to work in both cases. We implement the free surface within an arbitrary Lagrangian-Eulerian (ALE) formulation
-\citep[e.g][]{fullsack1995arbitrary,donea2004encyclopedia}.
-
-\texttt{ASPECT} is a parallelized, distributed memory code with adaptive mesh refinement (see \cite{BBHK10} for details).
-The free surface implementation works with these features.  We store the mesh vertex positions 
-in a fully distributed vector. This vector is continually updated and redistributed across 
-processes upon mesh adaptation (which is handled by the adaptive octree library \texttt{p4est}).
-We also provide adaptive refinement indicators based on being near to the free surface or 
-when the free surface slope is steep to allow for accurate interface tracking.
-
-\subsection{Remeshing}
-\label{sec:remeshing}
-
-The mesh velocity in normal direction at the free surface (with unit normal $\mathbf{n}$) has to be
-consistent with the velocity of the Stokes velocity solution $\mathbf{u}(t)$:
-\begin{equation}
- \mathbf{u}_\mathrm{mesh}(t) \cdot \mathbf{n} = \mathbf{u}(t) \cdot \mathbf{n} \quad \text{on} \quad \Gamma_F
-\end{equation}
-In ALE calculations the internal mesh velocity is undetermined.
-In general, one wants to smoothly deform the mesh so as to preserve its regularity, 
-avoiding inverted or otherwise poorly conditioned cells.
-The mesh deformation can be calcluated in many different ways, including algebraic \citep[e.g.][]{thieulot2011fantom} 
-and PDE based approaches.
-
-We choose to implement remeshing based on solving Laplace's equation for the mesh velocity.
-We solve the equation
-\begin{equation}
-\nabla^2 \mathbf{u}_{\mathrm{mesh}} = 0
-\label{eq:laplacian_smoothing}
-\end{equation}
-subject to the boundary conditions
-\begin{equation}
-\begin{aligned}
-&\mathbf{u}_\mathrm{mesh} &= 0  &\qquad \textrm{on } \Gamma_0. \\
-&\mathbf{u}_\mathrm{mesh} &= \left( \textbf{u} \cdot \textbf{n} \right) \textbf{n} & \qquad \textrm{on } \Gamma_F, \\
-&\mathbf{u}_\mathrm{mesh} \cdot \textbf{n} &= 0  &\qquad \textrm{on } \Gamma_{FS}, \\
-\end{aligned}
-\label{eq:laplacian_bcs}
-\end{equation}
-where $\Gamma_{FS}$ is the part of the boundary with free slip boundary conditions, given by
-\begin{equation}
-\begin{aligned}
- \mathbf{u \cdot n} &= 0 \\ %&\qquad \textrm{on } \Gamma_{FS} \\
- \mathbf{T}\cdot \mathbf{n} - (\mathbf{n}\cdot\mathbf{T} \cdot \mathbf{n})\mathbf{n} &= 0 &\qquad \textrm{on } \Gamma_{FS}.
-\end{aligned}
-\label{eq:free_slip_bcs}
-\end{equation}
-This scheme has the advantage of working for many different domain geometries and combinations of boundary conditions.
-For moderate mesh deformation, the mesh stays smooth and well conditioned, though it breaks down for large deformations.
-
-\subsection{Surface advection}
-With a deformable domain there is the danger that small errors in free surface motion can
-result in poor overall mass conservation in time. In some scenarios, the total volume of the mesh can 
-fluctuate significantly over hundreds or thousands of timesteps.
-Consistency with the Stokes solution requires 
-\begin{equation}
-\mathbf{u}_{\mathrm{mesh}} \cdot \mathbf{n} = \mathbf{u \cdot n}.
-\end{equation}
-
-Unfortunately the normal vectors are not well defined on the mesh vertices, which is 
-where the mesh velocity is defined. One could instead advect the mesh in the direction 
-of the local vertical, or in some weighted average of the cell normals adjacent to a given vertex,
-but we have found that these schemes do not necessarily have good mass conservation 
-properties.
-
-A better approach is to perform an $L_2$ projection of the normal velocity $\mathbf{u}\cdot\mathbf{n}$
-onto the free surface $\Gamma_F$. Multiplying the boundary conditions 
-(Equation~\eqref{eq:laplacian_bcs}) by a test function $\mathbf{w}$ and integrating over the free
-surface part of the boundary, we find
-
-\begin{equation}
-\int_{\Gamma_F} \mathbf{w} \cdot \mathbf{u}_\mathrm{mesh} = 
-\int_{\Gamma_F} \left( \mathbf{w \cdot n } \right) \left( \mathbf{u \cdot n} \right).
-\end{equation}
-When discretized, this forms a linear system which can be solved for the mesh velocity $\mathbf{u}_\mathrm{mesh}$ at the 
-free surface. Of course, a new system to solve is not ideal, but this system, being nonzero 
-over only the free surface, is relatively cheap to solve.
-
-This weak solution to the boundary conditions (Equation~\eqref{eq:laplacian_bcs}) is able to borrow
-the accuracy of the Stokes solve for Equation~\eqref{eq:weak_incompressible}, and we have 
-found it to conserve mass more accurately than algebraic techniques for evaluating the mesh-normal velocity.
-Similar results were found by \citet{fullsack1995arbitrary}.
-
-\section{Numerical results}
-\label{sec:results}
-
-\subsection{Relaxation of sinusoidal topography}
-\label{sec:topography_relaxation}
-
-We demonstrate the convergence of the nonstandard finite difference model by comparison 
-with an analytical solution. 
-As in Section~\ref{sec:numerical_determination} we consider the relaxation
-of sinusoidal surface topography in a two-dimensional Cartesian box with an isoviscous fluid.
-The setup is shown in Figure~\ref{fig:benchmark_setup}.
-
-For an initial topography given by Equation~\eqref{eq:initial_topography}, the relaxation time $\tau$
-is given by Equation~\eqref{eq:analytic_relaxation_time}, and the time evolution of the surface 
-topography is given by
-\begin{equation}
-  \zeta(x,t) = \zeta(x,0) e^{-t/\tau}.
-\end{equation}
-This solution is only valid for infinitesimal topography. However, for small
-initial topography $\zeta_0$ it seems to be sufficiently accurate to test convergence orders up 
-to quadratic \citep{kramer2012implicit, furuichi2015implicit}.
-We estimate the error $E$ by time-integrating the $L_2$ difference between the numerical and analytical
-solutions at the center point of the free surface, over a time interval $T$ (which we choose to be $4\tau$):
-\begin{equation}
-E = \frac{1}{T}\int_0^{T} \lVert \zeta_\mathrm{numeric}(L/2, t) - \zeta_\mathrm{analytic}(L/2, t) \rVert_2 \text{ d}t.
-\end{equation}
-
-Figure~\ref{fig:timestep_convergence}a shows the convergence of the nonstandard finite difference scheme 
-with respect to timestep size. The scheme is first order in time, with improving accuracy as the value of $\tau^*$ approaches
-the relaxation time of the relevant mode $\tau_i$. Interestingly, if $\tau^* = \tau_i$ the advection scheme
-becomes exact \citep{mickens2002nonstandard}. At this point the magnitude of the error plummets and is no longer 
-a strong function of the timestep. The remaining error is likely due to error in the linear approxmation for 
-the analytical solution, the spatial discretization, or the linear solver tolerance.
-
-Figure~\ref{fig:timestep_convergence}b shows the convergence of the quasi-implicit scheme as a function of timestep $\Delta t$.
-When $\theta=0$, it corresponds to the unstabilized forward Euler scheme, and is first order in time. 
-When $\theta=1$ it is also first order in time, but allows for a much larger timestep.  When $\theta=0.5$
-the quasi-implicit guess for the body force is good enough that it actually seems to achieve the second-order 
-convergence of a trapezoidal scheme, though it has not been shown that this extends to more complicated models.
-
-Figure~\ref{fig:tau_sensitivity} shows in more detail the effects of the choice of $\tau^*$. In a narrow region 
-close to the true value of $\tau_\mathrm{min}$ the error becomes very small, but in a broader region nearby 
-the errors are larger. 
-The excellent accuracy when the stabilization timescale $\tau^*$ is close to one of the natural relaxation
-timescales allows for tuning of the scheme to track specific long-wavelength modes, such as 
-those due to rotational or tidal deformation.
-
-\begin{figure*}
-\includegraphics[width=0.95\textwidth]{figures/timestep_convergence.pdf}
-\caption{Convergence tests for the benchmark shown in Figure~\ref{fig:benchmark_setup}.
-(a) Convergence test with timestep size for the nonstandard finite difference scheme.
-Comparison with the slope-one line confirms that it is first-order in time.
-In the case that the stabilization timescale $\tau^*$ is equal to the analytic relaxation time the error becomes very small, as the time integration scheme becomes exact \citep{mickens2002nonstandard}.
-(b) Convergence test with timestep size for the quasi-implicit scheme.
-For $\theta = 1$ and $\theta = 0$ the scheme is first-order accurate (though the latter is just an unstabilized forward-Euler scheme).
-For $\theta=0.5$ the scheme appears second order accurate on this benchmark.}
-\label{fig:timestep_convergence}
-\end{figure*}
-
-\begin{figure*}
-\includegraphics[width=0.9\textwidth]{figures/tau_sensitivity.pdf}
-\caption{Sensitivity of the nonstandard finite difference scheme to $\tau^*$.
-As the relaxation parameter $\tau^*$ approaches the relaxation time of the benchmark case the error reduces.
-The sharp cusp at $\tau^* = \tau_\mathrm{min}$ corresponds to the almost spectral accuracy of the time marching scheme for that case.}
-\label{fig:tau_sensitivity}
-\end{figure*}
-
-\subsection{Rayleigh-Taylor test}
-\label{sec:rayleigh_taylor}
-
-\begin{figure*}
-\includegraphics[width=0.9\textwidth]{figures/rayleigh_taylor_setup.pdf}
-\caption{Setup for the Rayleigh-Taylor test.
-A denser, more viscous layer of lies on top of the mantle.
-The overlying layer has a thickness of 100 km, and a sinusoidal perturbation on its base with and amplitude of 5 km.}
-\label{fig:rayleigh_taylor_setup}
-\end{figure*}
-
-We also consider the time evolution of the Rayleigh-Taylor benchmark described in \citet{kaus2010stabilization}.
-The benchmark setup is shown in Figure~\ref{fig:rayleigh_taylor_setup}.
-In all cases we use adaptive mesh refinement with five global refinement levels and four adaptive refinement levels,
-corresponding to $\sim$1 km cells at the most refined level. We refine according to compositional gradients, as well as
-near to the free surface.
-
-Figure~\ref{fig:rayleigh_taylor} shows the maximum depth of the interface between the dense and light fluid through time.
-The blue line shows the results using forward Euler timestepping with a very small timestep of 500 yrs, and is used as a reference solution.
-Nonstandard finite difference timestepping and quasi-implicit timestepping both converge to the reference solution,
-though from different directions.
-For the nonstandard finite difference simulations we recomputed the value of the minimum relaxation time $\tau_\mathrm{min}$ 
-every 50 timesteps using power iteration, and reset the value of $\tau^*$ to $\tau_\mathrm{min}$.
-
-The behavior of the two schemes is similar, though at the same CFL number quasi-implicit timestepping is slightly more accurate.
-On the other hand, with our implementation we have found that nonstandard finite difference timestepping can allow for larger timesteps 
-than quasi-implicit timestepping while remaining stable.
-
-\begin{figure*}
-\includegraphics[width=0.9\textwidth]{figures/rayleigh_taylor.pdf}
-\caption{Rayleigh-Taylor test.
-(a) Maximum depth of the drip.
-The blue line is the result using forward Euler (FE) integration of the free surface with a small 500 yr timestep, and is used as a reference solution.
-The symbols are for several different runs with different timestepping parameters, using both quasi-implicit (QI) and nonstandard finite difference (NSFD) schemes.
-As the timestepping gets smaller, both schemes converge to the reference solution (though from different directions).}
-\label{fig:rayleigh_taylor}
-\end{figure*}
-
-We also used the Rayleigh-Taylor benchmark to investigate the effect of making a poor choice for $\tau^*$, shown in Figure~\ref{fig:rayleigh_taylor_tau_choice}.
-We include four cases, all using adaptive timestepping with CFL=0.2.
-As a reference solution we include the case where we update $\tau^*$ to be equal to $\tau_\mathrm{min}$ every 50 timesteps,
-allowing for time variations in the viscosity and density structure to change the minimum relaxation time.
-We also show the case where we do not update $\tau^*$ over the course of the simulation, and two cases where
-we use the wrong value by a factor of two in either direction.
-
-Over the course of the Rayleigh-Taylor benchmark, $\tau_\mathrm{min}$ varies by about 25\%, as shown in Figure~\ref{fig:rayleigh_taylor_tau_choice}b.
-The case where we use the initial value of $\tau_\mathrm{min}$ for $\tau^*$ gives nearly identical results,
-whereas the cases where we choose values $\sim$50\% off in either direction give significantly different (though not wildly so) results.
-
-In many simulations we expect that the geometry, density, and viscosity structure of the model will not change so significantly
-that the $\tau^*$ would need to be updated frequently (if ever).
-
-\begin{figure*}
-\includegraphics[width=0.9\textwidth]{figures/rayleigh_taylor_tau_choice.pdf}
-\caption{Effect of the choice of $\tau^*$ on the Rayleigh-Taylor test.
-(a) Maximum depth of the drip.
-In all cases adaptive timestepping is used with CFL=0.2.
-The blue line serves as a reference solution, and shows the result when the smallest relaxation timescale $\tau_{\mathrm{min}}$ is recalculated every 50 timesteps, and then the stabilization timescale $\tau^*$ is set to that value.
-The red hexagons show the result when we do not update $\tau^*$, instead keeping it at the inital value of $\tau_{\mathrm{min}}$.
-The results are indistinguishable from those of the first case.
-The triangles show the results when we make a poor choice of $\tau^*$, by a factor of two in either direction.
-In these cases there is a significant difference from the reference solution.
-(b) Evolution of the minimum relaxation timescale over the course of the simulation.
-It does experience significant changes, but not by orders of magnitude, so it is unsurprising that using the initial value is sufficient for this case.}
-\label{fig:rayleigh_taylor_tau_choice}
-\end{figure*}
-
-\subsection{Mesh adaptivity}
-\label{sec:mesh_adaptivity}
-
-We also investigate the advantages of combining adaptive mesh refinement with free surface computations.
-\citet{crameri2012comparison} performed a community benchmark of a setup for which there is no analytic
-solution. In this benchmark, a buoyant blob rises beneath a viscous lid with a free surface, deflecting the
-boundary upwards (for the full setup, see \citet{crameri2012comparison}). 
-Figure~\ref{fig:amr} shows the convergence of the maximum topography at 3 Myr to its value in a high resolution simulation,
-both with and without adaptive mesh refinement.
-
-For the uniform refinement cases each point is generated by running the benchmark with a different global refinement level.
-For the adaptive case each point is generated by allowing the mesh to be refined according to gradients in density and composition,
-where the maximum refinement level is limited to the same refinement level of the corresponding global refinement simulation.
-In both the uniform and the adaptive cases, the smallest cell size in the most refined case is $\sim$1 km.
-We refine every ten timesteps according to gradients in the density and compositional fields.
-
-We run simulations using both quasi-implicit and nonstandard finite difference schemes for the free surface.
-We use a very small timestep for this test of 500 yr (note that this is significantly smaller than the smallest
-relaxation time of $\sim$ 14 kyr), so most timestepping schemes should be stable and accurate, including forward Euler.
-We choose this step for two reasons: (1) The primary purpose of this test is to investigate the savings
-due to using adaptive mesh refinement with a free surface, and (2) \citet{crameri2012comparison}
-found that a timestep of several hundred years was required to achieve full convergence.
-
-The convergence with and without adaptive mesh refinement have essentially the same behavior, but the adaptive case requires 
-fewer degrees of freedom by a factor of approximately an order of magnitude. More complex models will have
-more detail and so we may be less able to aggressively coarsen them, but we still expect that adaptive mesh refinement 
-will result in significant computational savings.
-
-\begin{figure*}
-\includegraphics[width=0.9\textwidth]{figures/amr.pdf}
-\caption{Convergence with degrees of freedom (DoFs) for the \citet{crameri2012comparison} Case 2 benchmark, for both uniform and adaptive mesh refinement.
-The timestep $\Delta t$ is 500 yr.
-We compare the maximum topography at 3 Myr with its value at high resolution ($\sim$398 m).
-For the adaptive cases we perform mesh refinement according to the sum of density and composition gradients.
-Quasi-implicit (QI) and nonstandard finite difference (NSFD) timestepping converge similarly, but the adaptive mesh refinement simulations require significantly fewer unknowns.
-For NSFD timestepping we used a $\tau^*$ of 14.825 kyr, as determined by an analytical solution of the long-wavelength relaxation time of the free surface for this particular model setup \citep{crameri2012comparison}.}
-\label{fig:amr}
-\end{figure*}
-
-\section{Conclusion}
-We have analyzed stability of free surface boundary conditions in geodynamic simulations and 
-demonstrated the cause of sloshing instabilities using a normal mode analysis.
-This perspective on the problem allowed us to construct an explicit finite difference 
-scheme which is first order accurate in time and is unconditionally stable.
-The nonstandard finite difference scheme is extremely simple to implement, and 
-requires no modifications to the system matrix.
-
-The normal mode perspective on the problem also provides insights into the effect of 
-the quasi-implicit stabilization scheme proposed by \citet{kaus2010stabilization}.
-The relaxation time of each mode is lengthened by an amount $\theta \Delta t$, and the
-maximum allowable timestep is correspondingly lengthened.
-
-It is not clear that the non-standard finite difference scheme is superior to 
-the quasi-implicit scheme. For $\theta=0.5$ the quasi-implicit scheme is more accurate,
-but has a smaller stability region. For $\theta=1$ the two schemes are comparably accurate,
-with a slight advantage to the quasi-implicit scheme. However, we have found that
-the nonstandard finite difference scheme allows for larger stable timesteps, if the modeler 
-is willing to pay the price of reduced accuracy for a particular simulation.
-
-Finally, we have described the implementation of free surface boundary conditions in 
-the open source mantle convection software \texttt{ASPECT}. Both the quasi-implicit 
-scheme and the nonstandard finite difference scheme are available. The implementation is 
-sufficiently general to accommodate many different geometries, rheologies, and 
-gravity models. Furthermore, it runs in parallel and with adaptive mesh refinement.
-
-\section*{Acknowledgements}
-This work was supported through National Science Foundation grant EAR-1246670
-and by the Computational Infrastructure in Geodynamics (CIG), through National Science Foundation grant EAR-0949446. 
-TH is partially supported by NSF grant DMS-1522191.
-Thanks to Wolfgang Bangerth, Brent Delbridge, and Cedric Thieulot for useful discussions.
-Thanks to Fabio Crameri and Dave May for constructive reviews.
-%% The Appendices part is started with the command \appendix;
-%% appendix sections are then done as normal sections
-%% \appendix
-
-%% \section{}
-%% \label{}
-
-%% If you have bibdatabase file and want bibtex to generate the
-%% bibitems, please use
-%%
-\section*{Bibliography}
-\bibliographystyle{elsarticle-harv} 
-\bibliography{free-surface-paper}
-
-%% else use the following coding to input the bibitems directly in the
-%% TeX file.
-
-%\begin{thebibliography}{00}
-
-%% \bibitem[Author(year)]{label}
-%% Text of bibliographic item
-
-%\bibitem[ ()]{}
-
-%\end{thebibliography}
-\end{document}
-
-\endinput
-%%
-%% End of file `elsarticle-template-harv.tex'.

### 2016-07-18 free-surface-paper.tex
Add short figure captions.

@@
 \includegraphics[width=0.9\textwidth]{figures/benchmark_setup.pdf}
-\caption{Setup for the power iteration test (Section~\ref{sec:numerical_determination}) and for the free surface relaxation benchmark (Section~\ref{sec:topography_relaxation}).
+\caption[Setup for the power iteration test (Section~\ref{sec:numerical_determination}) and for the free surface relaxation benchmark (Section~\ref{sec:topography_relaxation}).]{Setup for the power iteration test (Section~\ref{sec:numerical_determination}) and for the free surface relaxation benchmark (Section~\ref{sec:topography_relaxation}).
 A 2D box with an isoviscous fluid has sinusoidal initial topography, with amplitude $\zeta_0$ and wavenumber $n$.
@@
 \includegraphics[width=0.9\textwidth]{figures/perturb_spectrum.pdf}
-\caption{The effect of quasi-implicit stabilization on the minimum relaxation time of a fluid in a 2D Cartesian box with a free surface (see Figure~\ref{fig:benchmark_setup}), as well as the result of power iteration to find the minimum relaxation time.
+\caption[The effect of quasi-implicit stabilization on the minimum relaxation time of a fluid in a 2D Cartesian box with a free surface (see Figure~\ref{fig:benchmark_setup}), as well as the result of power iteration to find the minimum relaxation time.]{The effect of quasi-implicit stabilization on the minimum relaxation time of a fluid in a 2D Cartesian box with a free surface (see Figure~\ref{fig:benchmark_setup}), as well as the result of power iteration to find the minimum relaxation time.
 The $x$-axis shows the timestep normalized by minimum relaxation time of the unstabilized problem $\tau_\mathrm{\min}$.
@@
 \includegraphics[width=0.9\textwidth]{figures/stability_region.pdf}
-\caption{(a) Stability region for the nonstandard finite difference scheme (green, Equations~\eqref{eq:nsfd_unconditional} and~\eqref{eq:nsfd_conditional}).
+\caption[Stability regions for the nonstandard finite difference and quasi-implicit timestepping schemes.]{(a) Stability region for the nonstandard finite difference scheme (green, Equations~\eqref{eq:nsfd_unconditional} and~\eqref{eq:nsfd_conditional}).
 On the x-axis is the value of the stabilization timescale, in units of the minimum relaxation time $\tau_{\mathrm{min}}$.
@@
 (b) Stability region for the quasi-implicit scheme (green, Equation~\eqref{eq:cfl_euler_stabilized}).
-Again, the y axis is in units of the minimum relaxation time.
+Again, the y axis is in units of the minumum relaxation time.
 The x-axis shows the value of the stabilization parameter $\theta$.
@@
 \includegraphics[width=0.95\textwidth]{figures/timestep_convergence.pdf}
-\caption{Convergence tests for the benchmark shown in Figure~\ref{fig:benchmark_setup}.
+\caption[Convergence tests for the topography relaxation benchmark.]{Convergence tests for the benchmark shown in Figure~\ref{fig:benchmark_setup}.
 (a) Convergence test with timestep size for the nonstandard finite difference scheme.
@@
 \includegraphics[width=0.9\textwidth]{figures/tau_sensitivity.pdf}
-\caption{Sensitivity of the nonstandard finite difference scheme to $\tau^*$.
+\caption[Sensitivity of the nonstandard finite difference scheme to $\tau$.]{Sensitivity of the nonstandard finite difference scheme to $\tau^*$.
 As the relaxation parameter $\tau^*$ approaches the relaxation time of the benchmark case the error reduces.
@@
 \includegraphics[width=0.9\textwidth]{figures/rayleigh_taylor_setup.pdf}
-\caption{Setup for the Rayleigh-Taylor test.
+\caption[Setup for the Rayleigh-Taylor test.]{Setup for the Rayleigh-Taylor test.
 A denser, more viscous layer of lies on top of the mantle.
@@
 \includegraphics[width=0.9\textwidth]{figures/rayleigh_taylor.pdf}
-\caption{Rayleigh-Taylor test.
+\caption[Results of the Rayleigh-Taylor test.]{Rayleigh-Taylor test.
 (a) Maximum depth of the drip.
@@
 \includegraphics[width=0.9\textwidth]{figures/rayleigh_taylor_tau_choice.pdf}
-\caption{Effect of the choice of $\tau^*$ on the Rayleigh-Taylor test.
+\caption[Effect of the choice of $\tau^*$ on the Rayleigh-Taylor test.] {Effect of the choice of $\tau^*$ on the Rayleigh-Taylor test.
 (a) Maximum depth of the drip.
@@
 The blue line serves as a reference solution, and shows the result when the smallest relaxation timescale $\tau_{\mathrm{min}}$ is recalculated every 50 timesteps, and then the stabilization timescale $\tau^*$ is set to that value.
-The red hexagons show the result when we do not update $\tau^*$, instead keeping it at the initial value of $\tau_{\mathrm{min}}$.
+The red hexagons show the result when we do not update $\tau^*$, instead keeping it at the inital value of $\tau_{\mathrm{min}}$.
 The results are indistinguishable from those of the first case.
@@
 \includegraphics[width=0.9\textwidth]{figures/amr.pdf}
-\caption{Convergence with degrees of freedom (DoFs) for the \citet{crameri2012comparison} Case 2 benchmark, for both uniform and adaptive mesh refinement.
+\caption[Convergence with degrees of freedom (DoFs) for the \citet{crameri2012comparison} Case 2 benchmark, for both uniform and adaptive mesh refinement.]{Convergence with degrees of freedom (DoFs) for the \citet{crameri2012comparison} Case 2 benchmark, for both uniform and adaptive mesh refinement.
 The timestep $\Delta t$ is 500 yr.

### 2016-07-31 free-surface-paper.tex
Break hidden derivation into appendix.

@@
 \usepackage{url}
+\usepackage{appendix}
 
@@
 \left| 1 - \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \right| \le 1.
-\end{equation}
-
-\ifdetail
-Taking the positive value of the absolute value yields the criterion of 
-\begin{equation}
-- \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \le 0,
-\end{equation}
-which can be rearranged to find
-\begin{equation}
-e^{-\Delta t/\tau^*} \le 1,
-\end{equation}
-which is simply a statement that the timestep must be positive.
-
-Taking the negative value of the absolute value yields
-\begin{equation}
-  \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right)  \le 2
-\end{equation}
-\begin{equation}
- 1 - 2 \frac{\tau_i}{\tau^*} \le e^{-\Delta t/\tau^*}
-\end{equation}
-if the left hand side is less than zero, then this is true regardless of step size, or
-\begin{equation}
-\tau^* \le 2 \tau_i
-\end{equation}
-if the left hand side is between zero and one, the expression is more complicated. 
-Taking the log of both sides:
-\begin{equation}
-\Delta t \le -\tau^* \log \left(1 - 2 \frac{\tau_i}{\tau^*} \right).
-\end{equation}
-It is convenient to write this in terms of dimensionless times $\Delta t/ \tau_i$ and $\tau^*/\tau_i$:
-\begin{equation}
-\frac{\Delta t}{\tau_i} \le -\frac{\tau^*}{\tau_i} \log \left(1 - 2 \frac{\tau_i}{\tau^*} \right).
-\end{equation}
-
-\fi
+\label{eq:stability_condition}
+\end{equation}
 
@@
 \end{equation}
-The stability region is plotted in Figure~\ref{fig:stability_region}.
+The stability region is plotted in Figure~\ref{fig:stability_region}. Equations~\eqref{eq:nsfd_unconditional} and~\eqref{eq:nsfd_conditional}
+are derived in Appendix~\ref{appendix:stability}.
 
@@
 
+\begin{subappendices}
+
+\section{Stability criterion of the NSFD scheme}
+\label{appendix:stability}
+We begin from Equation~\eqref{eq:stability_condition}:
+\begin{equation}
+\left| 1 - \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \right| \le 1.
+\end{equation}
+Timestepping involves repeated multiplication by this quantity. In order for
+the scheme to be stable, this repeated multiplication must not tend to infinity.
+Taking the positive value of the absolute value yields the criterion of 
+\begin{equation}
+- \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \le 0,
+\end{equation}
+which can be rearranged to find
+\begin{equation}
+e^{-\Delta t/\tau^*} \le 1,
+\end{equation}
+which is simply a statement that the timestep must be postive.
+
+Taking the negative value of the absolute value yields
+\begin{equation}
+  \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right)  \le 2
+\end{equation}
+which can be rearranged to find
+\begin{equation}
+ 1 - 2 \frac{\tau_i}{\tau^*} \le e^{-\Delta t/\tau^*}.
+\label{eq:negative_branch}
+\end{equation}
+For positive $\tau_i$ and $\tau^*$ the left hand side is strictly less than one.
+If the left hand side of Equation~\eqref{eq:negative_branch} is less than zero, 
+then the stability criterion is true regardless of step size, which occurs when
+\begin{equation}
+\tau^* \le 2 \tau_i.
+\end{equation}
+This is Equation~\eqref{eq:nsfd_unconditional}.
+If the left hand side of Equation~\eqref{eq:negative_branch} is between zero and one, 
+the expression is more complicated. Taking the log of both sides:
+\begin{equation}
+\Delta t \le -\tau^* \log \left(1 - 2 \frac{\tau_i}{\tau^*} \right).
+\end{equation}
+This is Equation~\eqref{eq:nsfd_conditional}.
+It can be convenient to express this in terms of the dimensionless quantities $\Delta t/ \tau_i$ and $\tau^*/\tau_i$:
+\begin{equation}
+\frac{\Delta t}{\tau_i} \le -\frac{\tau^*}{\tau_i} \log \left(1 - 2 \frac{\tau_i}{\tau^*} \right).
+\end{equation}
+\end{subappendices}
+
 %%ENDCLIP

### 2016-08-01 free-surface-paper.tex
Some minor comments from Nick.

--- 2016-07-18 free-surface-paper.tex
@@
 
-%%BEGINCLIP
-
 %% main text
@@
 \begin{equation}
-\int_{\Omega(t)} 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u} ) - \int_{\Omega(t)} p^\prime \nabla \cdot \mathbf{w} 
-- \int_{\Gamma_F(t)} \mathbf{w} \cdot \mathbf{T} \cdot \mathbf{n} = 0 
+-\int_{\Omega(t)} 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u} ) + \int_{\Omega(t)} p^\prime \nabla \cdot \mathbf{w} 
++ \int_{\Gamma_F(t)} \mathbf{w} \cdot \mathbf{T} \cdot \mathbf{n} = 0 
 \label{eq:weak_stokes}
@@
 We will therefore  evaluate the integrals in Equation~\eqref{eq:weak_stokes} 
-over the hydrostatic reference surface and introduce a temporary auxiliary variable $\zeta$ which 
+over the hydrostatic reference surface and introduce a temporary auxilliary variable $\zeta$ which 
 represents the (small) topography of the free surface relative to that reference surface.
@@
 \begin{equation}
-\mathbf{T} \approx \frac{\partial \mathbf{T}}{\partial \mathbf{n} } \cdot \mathbf{n} \zeta = -\rho_0 g \zeta \mathbf{I}.
+\mathbf{T} \approx \frac{\partial \mathbf{T}}{\partial \mathbf{n} } \cdot \mathbf{n} \zeta = \rho_0 g \zeta \mathbf{I}.
 \label{eq:hydrostatic}
@@
 \begin{equation}
-\int_{\Omega(t)} 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u} ) - \int_{\Omega(t)} p^\prime \nabla \cdot \mathbf{w} 
+-\int_{\Omega(t)} 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u} ) + \int_{\Omega(t)} p^\prime \nabla \cdot \mathbf{w} 
 + \int_{\Gamma_F(t)} \rho_0 g \zeta  \mathbf{w} \cdot \mathbf{n} = 0.
@@
 \begin{equation}
-\int_{\Omega(t)} 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u}_i ) - \int_{\Omega(t)} p^\prime_i \nabla \cdot \mathbf{w} 
+-\int_{\Omega(t)} 2 \eta \varepsilon( \mathbf{w} ) \colon \varepsilon( \mathbf{u}_i ) + \int_{\Omega(t)} p^\prime_i \nabla \cdot \mathbf{w} 
 = \tau_i \int_{\Gamma_F(t)} \rho_0 g (\mathbf{u}_i \cdot \mathbf{n} ) (\mathbf{w} \cdot \mathbf{n}).
@@
 \end{equation}
-where $\tau_{\mathrm{min}}$ is this minimum decay time.
+where $\tau_{\mathrm{min}}$ is this minumum decay time.
 The maximum stable timestep is limited by the minimum relaxation timescale.
@@
 solution and black dots show the analytical solution using equation~\eqref{eq:spectrum_shift}.
-In most cases the iteration converges to better than $\sim10^{-5}$ relative accuracy in about 20 iterations
+In most cases the iteration converges to better than $\sim10^{-5}$ relative accuaracy in about 20 iterations
 (and fewer with a good initial guess).
@@
 \end{equation}
-which is simply a statement that the timestep must be positive.
+which is simply a statement that the timestep must be postive.
 
@@
 avoiding inverted or otherwise poorly conditioned cells.
-The mesh deformation can be calculated in many different ways, including algebraic \citep[e.g.][]{thieulot2011fantom} 
+The mesh deformation can be calcluated in many different ways, including algebraic \citep[e.g.][]{thieulot2011fantom} 
 and PDE based approaches.
@@
 
-%%ENDCLIP
-
 \section*{Acknowledgements}

### 2016-08-01 frontmatter.tex
Remove abstract from frontmatter.

@@
 \label{chap:free_surface}
-
-Geodynamic simulations increasingly rely on simulations with a true free surface to 
-investigate questions of dynamic topography, tectonic deformation, gravity perturbations, and
-global mantle convection. However, implementations of free surface boundary conditions 
-have proven challenging from a standpoint of accuracy, robustness, and stability.
-In particular, time integration of a free surface tends to suffer from a numerical instability
-that manifests as sloshing surface motions, also known as the ``drunken sailor'' instability.
-This instability severely limits stable timestep sizes to those much smaller than could be used
-in geodynamic simulations without a free surface. 
-Several schemes have been proposed in the literature to deal with these instabilities.
-
-Here we analyze the problem of creeping viscous flow with a free surface and discuss the 
-origin of these instabilities. We demonstrate their cause and how existing stabilization 
-schemes work to damp them out.
-We also propose a new scheme for removing instabilities from free surface calculations. 
-It does not require modifications to the system matrix, nor additional variables, but is instead
-an explicit scheme based on nonstandard finite differences.  It relies on a single 
-stabilization parameter which may be identified with the smallest relaxation timescale of the
-free surface.
-
-Finally, we discuss the implementation of a free surface in the open source, community based
-mantle convection software \texttt{ASPECT}.

### 2016-08-03 free-surface-paper.tex
Make an even number of sign errors...
Add appendix

--- 2016-07-31 free-surface-paper.tex
@@
 Equation~\eqref{eq:generalized_eigenvalue} is a generalized eigenvalue problem for the normal modes of the system.
+It has strictly positive eigenvalues (see Appendix~\ref{app:positive_definite}).
 It is rather more difficult to solve than a standard eigenvalue problem because the matrix on the right-hand-side 
@@
 \begin{itemize}
-\item In the absence of forcing, topography always decreases.
-\item Relaxation of small amplitude topography takes the form of exponential decay.
-\item The decay mode with the shortest relaxation time is the least stable.
+\item In the absence of forcing, topography always decreases (Appendix~\ref{app:positive_definite}).
+\item Relaxation of small amplitude topography takes the form of exponential decay (Section~\ref{sec:eigenvalue}).
+\item The decay mode with the shortest relaxation time is the least stable (Section~\ref{sec:timestepping}).
 \end{itemize}
@@
 
+\section{The relaxation spectrum is positive definite}
+\label{app:positive_definite}
+On physical grounds, we expect the relaxation spectrum described in Section~\ref{sec:eigenvalue}
+to have only positive eigenvalues $\tau_i$.
+An imaginary eigenvalue would have an oscillitory component,
+which we do not expect in creeping flows (oscillation being usually due to inertia). 
+A negative eigenvalue would grow exponentially, which would violate conservation of energy.
+Here we demonstrate the positivity of the relaxation time spectrum.
+We could proceed with either the continuous or the discretized version of the governing
+equations, but choose the continuous form here.
+We begin with the homogeneous Stokes equation (Equation~\eqref{eq:stokes}):
+\begin{equation}
+\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p = \rho \mathbf{g}.
+\label{eq:homogeneous_stokes}
+\end{equation}
+We can transform this into the mechanical energy equation by multiplying it by $\mathbf{u}$
+and integrating over the domain:
+\begin{equation}
+\int_{\Omega(t)} \mathbf{u} \cdot \nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \int_{\Omega(t)} \mathbf{u}\cdot \nabla p = 0.
+\end{equation}
+Integrating by parts and applying boundary conditions yields
+\begin{equation}
+-\int_{\Omega(t)} 2 \eta \varepsilon(\mathbf{u}) : \varepsilon(\mathbf{u})  + \int_{\Omega(t)}p \nabla \cdot \mathbf{u} - \int_{\Gamma_F(t)} \mathbf{u} \cdot \mathbf{T} \cdot \mathbf{n} = 0.
+\end{equation}
+Where the integrals correspond to the viscous dissipation, the work done by pressure, and the work done by boundary stresses, respectively.
+Since we are using the incompressible approximation, the second integral is zero.
+On a true free surface the boundary stresses are zero, but we make the same approximation
+used in Equation~\eqref{eq:hydrostatic} and map the stress onto the reference surface, finding
+\begin{equation}
+-\int_{\Omega(t)} 2 \eta \varepsilon(\mathbf{u}) : \varepsilon(\mathbf{u})  + \int_{\Gamma_F(t)} \rho_0 g \zeta \mathbf{u} \cdot \mathbf{n} = 0.
+\end{equation}
+We can now eliminate $\zeta$ from the equation in the same way that was used in Equation~\eqref{eq:weak_eigen},
+thereby introducing the relaxation time $\tau_i$:
+\begin{equation}
+-\int_{\Omega(t)} 2 \eta \varepsilon(\mathbf{u}) : \varepsilon(\mathbf{u})  + \tau_i \int_{\Gamma_F(t)} \rho_0 g (\mathbf{u} \cdot \mathbf{n})^2 = 0.
+\end{equation}
+This equation may be solved for $i\tau_i$:
+\begin{equation}
+\tau_i = \frac{\int_{\Omega(t)} 2 \eta \varepsilon(\mathbf{u}) : \varepsilon(\mathbf{u})}{ \int_{\Gamma_F(t)} \rho_0 g (\mathbf{u} \cdot \mathbf{n})^2}.
+\end{equation}
+The top and bottom of the right hand side are both strictly positive numbers so the $\tau_i$ must also be positive,
+demonstrating the positive definiteness of the system. A similar analysis can be performed for the discrete system.
+There is the possibility for velocities which do not move the free surface at all, in which case the bottom
+integral goes to zero, and the relaxation time becomes infinite. One example of this is the case of shear flow with
+a free upper surface and periodic horizontal boundary conditions. Since we are usually interested in the smallest
+relaxation times, these infinite eigenvalues are not usually a problem.
+
 \section{Stability criterion of the NSFD scheme}

### 2016-08-03 free-surface-paper.tex
Typo.
Attempt to clarify a few things.

@@
 matches the boundary conditions which one is trying to model, but it typically requires a deformable 
-domain with frequent remeshing to avoid ill-conditioned cells.
+domain with frequent remeshing to avoid ill-conditioned cells (i.e., cells that are inverted or have large aspect ratios).
 
@@
 We would like to analyze the time evolution of the normal modes of this system: each mode 
-is the relaxation of topography with a characteristic relaxation time.  
+corresponds to the relaxation of topography with a characteristic wavelength and relaxation time.  
+These times are determined by the model geometry, gravity, viscosity structure, and density structure.
 We denote the normal modes by $\left[ \mathbf{u}_i, p^\prime_i, \zeta_i \right]^T$, with
@@
 for $\mathbf{u}$, it is formally a first-order accurate scheme. As such, it is capable of capturing arbitrary motions 
-of the free surface, while allowing larger timesteps than forward Euler schemes.
+of the free surface, while allowing larger timesteps than the forward Euler scheme.
 
@@
 \end{equation}
-The local truncation error is quadratic in $\Delta t$, similar to the forward Euler, and summing this error over 
-many timesteps results in a factor of $\Delta t^{-1}$, demonstrating the first-order global accuracy~\citep[e.g.][]{leveque2007finite}.
+The local truncation error is quadratic in $\Delta t$, similar to the forward Euler schem.
+Summing this error over many timesteps results in a factor of $\Delta t^{-1}$, which results in a global truncation error that is linear in $\Delta t$ \citep[e.g.][]{leveque2007finite}, demonstrating the first-order accuracy.
 The choice of $\tau^*$ controls the size of the coefficient on the truncation error for the scheme.
@@
 
+The improved understanding of free surface boundary conditions in geodynamic models
+will allow for modelers to make more informed choices about timestepping and free
+surface stabilization techniques for use in simulations related to dynamic topography,
+active tectonics, gravity anomalies, and geomorphologic processes.
+
 \begin{subappendices}
@@
 \end{equation}
-This equation may be solved for $i\tau_i$:
+This equation may be solved for $\tau_i$:
 \begin{equation}

### 2016-08-03 free-surface-paper.tex
Fix up references.

@@
 \section{Introduction}
-\label{sec:intro}
+\label{sec:introduce}
 
@@
 \mathbf{T} \approx \frac{\partial \mathbf{T}}{\partial \mathbf{n} } \cdot \mathbf{n} \zeta = -\rho_0 g \zeta \mathbf{I}.
-\label{eq:hydrostatic}
+\label{eq:hydrostat}
 \end{equation}
@@
 \nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p = \rho \mathbf{g}.
-\label{eq:homogeneous_stokes}
 \end{equation}
@@
 On a true free surface the boundary stresses are zero, but we make the same approximation
-used in Equation~\eqref{eq:hydrostatic} and map the stress onto the reference surface, finding
+used in Equation~\eqref{eq:hydrostat} and map the stress onto the reference surface, finding
 \begin{equation}

### 2016-08-10 free-surface-paper.tex
Spellcheck

--- 2016-08-03 free-surface-paper.tex
@@
 \usepackage{url}
-\usepackage{appendix}
 
@@
 Equation~\eqref{eq:generalized_eigenvalue} is a generalized eigenvalue problem for the normal modes of the system.
-It has strictly positive eigenvalues (see Appendix~\ref{app:positive_definite}).
 It is rather more difficult to solve than a standard eigenvalue problem because the matrix on the right-hand-side 
@@
 \includegraphics[width=0.9\textwidth]{figures/benchmark_setup.pdf}
-\caption[Setup for the power iteration test (Section~\ref{sec:numerical_determination}) and for the free surface relaxation benchmark (Section~\ref{sec:topography_relaxation}).]{Setup for the power iteration test (Section~\ref{sec:numerical_determination}) and for the free surface relaxation benchmark (Section~\ref{sec:topography_relaxation}).
+\caption{Setup for the power iteration test (Section~\ref{sec:numerical_determination}) and for the free surface relaxation benchmark (Section~\ref{sec:topography_relaxation}).
 A 2D box with an isoviscous fluid has sinusoidal initial topography, with amplitude $\zeta_0$ and wavenumber $n$.
@@
 \includegraphics[width=0.9\textwidth]{figures/perturb_spectrum.pdf}
-\caption[The effect of quasi-implicit stabilization on the minimum relaxation time of a fluid in a 2D Cartesian box with a free surface (see Figure~\ref{fig:benchmark_setup}), as well as the result of power iteration to find the minimum relaxation time.]{The effect of quasi-implicit stabilization on the minimum relaxation time of a fluid in a 2D Cartesian box with a free surface (see Figure~\ref{fig:benchmark_setup}), as well as the result of power iteration to find the minimum relaxation time.
+\caption{The effect of quasi-implicit stabilization on the minimum relaxation time of a fluid in a 2D Cartesian box with a free surface (see Figure~\ref{fig:benchmark_setup}), as well as the result of power iteration to find the minimum relaxation time.
 The $x$-axis shows the timestep normalized by minimum relaxation time of the unstabilized problem $\tau_\mathrm{\min}$.
@@
 \begin{itemize}
-\item In the absence of forcing, topography always decreases (Appendix~\ref{app:positive_definite}).
-\item Relaxation of small amplitude topography takes the form of exponential decay (Section~\ref{sec:eigenvalue}).
-\item The decay mode with the shortest relaxation time is the least stable (Section~\ref{sec:timestepping}).
+\item In the absence of forcing, topography always decreases.
+\item Relaxation of small amplitude topography takes the form of exponential decay.
+\item The decay mode with the shortest relaxation time is the least stable.
 \end{itemize}
@@
 \left| 1 - \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \right| \le 1.
-\label{eq:stability_condition}
-\end{equation}
+\end{equation}
+
+\ifdetail
+Taking the positive value of the absolute value yields the criterion of 
+\begin{equation}
+- \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \le 0,
+\end{equation}
+which can be rearranged to find
+\begin{equation}
+e^{-\Delta t/\tau^*} \le 1,
+\end{equation}
+which is simply a statement that the timestep must be positive.
+
+Taking the negative value of the absolute value yields
+\begin{equation}
+  \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right)  \le 2
+\end{equation}
+\begin{equation}
+ 1 - 2 \frac{\tau_i}{\tau^*} \le e^{-\Delta t/\tau^*}
+\end{equation}
+if the left hand side is less than zero, then this is true regardless of step size, or
+\begin{equation}
+\tau^* \le 2 \tau_i
+\end{equation}
+if the left hand side is between zero and one, the expression is more complicated. 
+Taking the log of both sides:
+\begin{equation}
+\Delta t \le -\tau^* \log \left(1 - 2 \frac{\tau_i}{\tau^*} \right).
+\end{equation}
+It is convenient to write this in terms of dimensionless times $\Delta t/ \tau_i$ and $\tau^*/\tau_i$:
+\begin{equation}
+\frac{\Delta t}{\tau_i} \le -\frac{\tau^*}{\tau_i} \log \left(1 - 2 \frac{\tau_i}{\tau^*} \right).
+\end{equation}
+
+\fi
 
@@
 \end{equation}
-The stability region is plotted in Figure~\ref{fig:stability_region}. Equations~\eqref{eq:nsfd_unconditional} and~\eqref{eq:nsfd_conditional}
-are derived in Appendix~\ref{appendix:stability}.
+The stability region is plotted in Figure~\ref{fig:stability_region}.
 
@@
 \includegraphics[width=0.9\textwidth]{figures/stability_region.pdf}
-\caption[Stability regions for the nonstandard finite difference and quasi-implicit timestepping schemes.]{(a) Stability region for the nonstandard finite difference scheme (green, Equations~\eqref{eq:nsfd_unconditional} and~\eqref{eq:nsfd_conditional}).
+\caption{(a) Stability region for the nonstandard finite difference scheme (green, Equations~\eqref{eq:nsfd_unconditional} and~\eqref{eq:nsfd_conditional}).
 On the x-axis is the value of the stabilization timescale, in units of the minimum relaxation time $\tau_{\mathrm{min}}$.
@@
 (b) Stability region for the quasi-implicit scheme (green, Equation~\eqref{eq:cfl_euler_stabilized}).
-Again, the y axis is in units of the minumum relaxation time.
+Again, the y axis is in units of the minimum relaxation time.
 The x-axis shows the value of the stabilization parameter $\theta$.
@@
 \includegraphics[width=0.95\textwidth]{figures/timestep_convergence.pdf}
-\caption[Convergence tests for the topography relaxation benchmark.]{Convergence tests for the benchmark shown in Figure~\ref{fig:benchmark_setup}.
+\caption{Convergence tests for the benchmark shown in Figure~\ref{fig:benchmark_setup}.
 (a) Convergence test with timestep size for the nonstandard finite difference scheme.
@@
 \includegraphics[width=0.9\textwidth]{figures/tau_sensitivity.pdf}
-\caption[Sensitivity of the nonstandard finite difference scheme to $\tau$.]{Sensitivity of the nonstandard finite difference scheme to $\tau^*$.
+\caption{Sensitivity of the nonstandard finite difference scheme to $\tau^*$.
 As the relaxation parameter $\tau^*$ approaches the relaxation time of the benchmark case the error reduces.
@@
 \includegraphics[width=0.9\textwidth]{figures/rayleigh_taylor_setup.pdf}
-\caption[Setup for the Rayleigh-Taylor test.]{Setup for the Rayleigh-Taylor test.
+\caption{Setup for the Rayleigh-Taylor test.
 A denser, more viscous layer of lies on top of the mantle.
@@
 \includegraphics[width=0.9\textwidth]{figures/rayleigh_taylor.pdf}
-\caption[Results of the Rayleigh-Taylor test.]{Rayleigh-Taylor test.
+\caption{Rayleigh-Taylor test.
 (a) Maximum depth of the drip.
@@
 \includegraphics[width=0.9\textwidth]{figures/rayleigh_taylor_tau_choice.pdf}
-\caption[Effect of the choice of $\tau^*$ on the Rayleigh-Taylor test.] {Effect of the choice of $\tau^*$ on the Rayleigh-Taylor test.
+\caption{Effect of the choice of $\tau^*$ on the Rayleigh-Taylor test.
 (a) Maximum depth of the drip.
@@
 \includegraphics[width=0.9\textwidth]{figures/amr.pdf}
-\caption[Convergence with degrees of freedom (DoFs) for the \citet{crameri2012comparison} Case 2 benchmark, for both uniform and adaptive mesh refinement.]{Convergence with degrees of freedom (DoFs) for the \citet{crameri2012comparison} Case 2 benchmark, for both uniform and adaptive mesh refinement.
+\caption{Convergence with degrees of freedom (DoFs) for the \citet{crameri2012comparison} Case 2 benchmark, for both uniform and adaptive mesh refinement.
 The timestep $\Delta t$ is 500 yr.
@@
 
-\begin{subappendices}
-
-\section{The relaxation spectrum is positive definite}
-\label{app:positive_definite}
-On physical grounds, we expect the relaxation spectrum described in Section~\ref{sec:eigenvalue}
-to have only positive eigenvalues $\tau_i$.
-An imaginary eigenvalue would have an oscillitory component,
-which we do not expect in creeping flows (oscillation being usually due to inertia). 
-A negative eigenvalue would grow exponentially, which would violate conservation of energy.
-Here we demonstrate the positivity of the relaxation time spectrum.
-We could proceed with either the continuous or the discretized version of the governing
-equations, but choose the continuous form here.
-We begin with the homogeneous Stokes equation (Equation~\eqref{eq:stokes}):
-\begin{equation}
-\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p = \rho \mathbf{g}.
-\label{eq:homogeneous_stokes}
-\end{equation}
-We can transform this into the mechanical energy equation by multiplying it by $\mathbf{u}$
-and integrating over the domain:
-\begin{equation}
-\int_{\Omega(t)} \mathbf{u} \cdot \nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \int_{\Omega(t)} \mathbf{u}\cdot \nabla p = 0.
-\end{equation}
-Integrating by parts and applying boundary conditions yields
-\begin{equation}
--\int_{\Omega(t)} 2 \eta \varepsilon(\mathbf{u}) : \varepsilon(\mathbf{u})  + \int_{\Omega(t)}p \nabla \cdot \mathbf{u} - \int_{\Gamma_F(t)} \mathbf{u} \cdot \mathbf{T} \cdot \mathbf{n} = 0.
-\end{equation}
-Where the integrals correspond to the viscous dissipation, the work done by pressure, and the work done by boundary stresses, respectively.
-Since we are using the incompressible approximation, the second integral is zero.
-On a true free surface the boundary stresses are zero, but we make the same approximation
-used in Equation~\eqref{eq:hydrostatic} and map the stress onto the reference surface, finding
-\begin{equation}
--\int_{\Omega(t)} 2 \eta \varepsilon(\mathbf{u}) : \varepsilon(\mathbf{u})  + \int_{\Gamma_F(t)} \rho_0 g \zeta \mathbf{u} \cdot \mathbf{n} = 0.
-\end{equation}
-We can now eliminate $\zeta$ from the equation in the same way that was used in Equation~\eqref{eq:weak_eigen},
-thereby introducing the relaxation time $\tau_i$:
-\begin{equation}
--\int_{\Omega(t)} 2 \eta \varepsilon(\mathbf{u}) : \varepsilon(\mathbf{u})  + \tau_i \int_{\Gamma_F(t)} \rho_0 g (\mathbf{u} \cdot \mathbf{n})^2 = 0.
-\end{equation}
-This equation may be solved for $i\tau_i$:
-\begin{equation}
-\tau_i = \frac{\int_{\Omega(t)} 2 \eta \varepsilon(\mathbf{u}) : \varepsilon(\mathbf{u})}{ \int_{\Gamma_F(t)} \rho_0 g (\mathbf{u} \cdot \mathbf{n})^2}.
-\end{equation}
-The top and bottom of the right hand side are both strictly positive numbers so the $\tau_i$ must also be positive,
-demonstrating the positive definiteness of the system. A similar analysis can be performed for the discrete system.
-There is the possibility for velocities which do not move the free surface at all, in which case the bottom
-integral goes to zero, and the relaxation time becomes infinite. One example of this is the case of shear flow with
-a free upper surface and periodic horizontal boundary conditions. Since we are usually interested in the smallest
-relaxation times, these infinite eigenvalues are not usually a problem.
-
-\section{Stability criterion of the NSFD scheme}
-\label{appendix:stability}
-We begin from Equation~\eqref{eq:stability_condition}:
-\begin{equation}
-\left| 1 - \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \right| \le 1.
-\end{equation}
-Timestepping involves repeated multiplication by this quantity. In order for
-the scheme to be stable, this repeated multiplication must not tend to infinity.
-Taking the positive value of the absolute value yields the criterion of 
-\begin{equation}
-- \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \le 0,
-\end{equation}
-which can be rearranged to find
-\begin{equation}
-e^{-\Delta t/\tau^*} \le 1,
-\end{equation}
-which is simply a statement that the timestep must be postive.
-
-Taking the negative value of the absolute value yields
-\begin{equation}
-  \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right)  \le 2
-\end{equation}
-which can be rearranged to find
-\begin{equation}
- 1 - 2 \frac{\tau_i}{\tau^*} \le e^{-\Delta t/\tau^*}.
-\label{eq:negative_branch}
-\end{equation}
-For positive $\tau_i$ and $\tau^*$ the left hand side is strictly less than one.
-If the left hand side of Equation~\eqref{eq:negative_branch} is less than zero, 
-then the stability criterion is true regardless of step size, which occurs when
-\begin{equation}
-\tau^* \le 2 \tau_i.
-\end{equation}
-This is Equation~\eqref{eq:nsfd_unconditional}.
-If the left hand side of Equation~\eqref{eq:negative_branch} is between zero and one, 
-the expression is more complicated. Taking the log of both sides:
-\begin{equation}
-\Delta t \le -\tau^* \log \left(1 - 2 \frac{\tau_i}{\tau^*} \right).
-\end{equation}
-This is Equation~\eqref{eq:nsfd_conditional}.
-It can be convenient to express this in terms of the dimensionless quantities $\Delta t/ \tau_i$ and $\tau^*/\tau_i$:
-\begin{equation}
-\frac{\Delta t}{\tau_i} \le -\frac{\tau^*}{\tau_i} \log \left(1 - 2 \frac{\tau_i}{\tau^*} \right).
-\end{equation}
-\end{subappendices}
-
-%%ENDCLIP
-
 \section*{Acknowledgements}

### 2016-09-04 free-surface-paper.tex
Remove comment.

@@
 \citet{crameri2012comparison} performed a community benchmark of a setup for which there is no analytic
-solution. In this benchmark, a buoyant blob rises beneath a viscous lid with a free surface, deflecting the
-boundary upwards, reminiscent of the dynamic topography due to a rising plume (for the full setup, see \citet{crameri2012comparison}). 
+solution. In this benchmark, a buoyant blob rises beneath a viscous lid with a free surface, deflecting the boundary upwards.
 Figure~\ref{fig:amr} shows the convergence of the maximum topography at 3 Myr to its value in a high resolution simulation,

### 2016-09-04 benchmarks/amr/adaptive_qi/run_models.py
Begin work on automated running of benchmarks.

@@ -33,7 +33,7 @@ refine = range(1,6)
 for r in refine:
   print "Running uniform mesh with %i total refinements\n"%(r+3)
   generate_prm(r)
-  os.system('mpirun /home/ian.rose/aspect/build/aspect tmp.prm')
+  os.system('mpirun -n 4 ../../../aspect/build/aspect tmp.prm')
   outputfile.write("%i %f \n" % (r, grab_topography()) )
   outputfile.flush()
 

### 2016-09-04 benchmarks/amr/uniform_qi/run_models.py

@@ -31,7 +31,7 @@ refine = range(4,9)
 for r in refine:
   print "Running uniform mesh with %i refinements\n"%(r)
   generate_prm(r)
-  os.system('mpirun /home/ian.rose/aspect/build/aspect tmp.prm')
+  os.system('mpirun -n 4 ../../../aspect/build/aspect tmp.prm')
   outputfile.write("%i %f \n" % (r, grab_topography()) )
   outputfile.flush()
 

### 2016-09-04 benchmarks/perturb_spectrum/run_models.py

@@ -48,7 +48,7 @@ for tstep in tsteps:
   for theta in thetas:
     print "Running with time step %f, theta %f\n"%(tstep, theta)
     generate_prm(tstep, theta)
-    os.system('mpirun -n 1 /home/ian/aspect/build/aspect tmp.prm')
+    os.system('mpirun -n 1 ../../aspect/build/aspect tmp.prm')
     analytic_relaxation_time, power_iteration_relaxation_time = get_relaxation_times(tstep, theta)
     outputfile.write("%f %f %f %f \n" % (tstep, theta, analytic_relaxation_time, power_iteration_relaxation_time) )
     outputfile.flush()

### 2016-09-04 benchmarks/tau_sensitivity/run_models.py

@@ -78,7 +78,7 @@ for r in relaxation_times:
   for t in tsteps:
     print "Running with time step %f, relaxation time %f\n"%(t, r)
     generate_prm(t, r)
-    os.system('mpirun -n 4 /home/ian/aspect/build/aspect tmp.prm')
+    os.system('mpirun -n 4 ../../aspect/build/aspect tmp.prm')
     #plot_topography()
     outputfile.write("%f %f %f \n" % (r/tau_s, t/tau_s, calculate_error()) )
     outputfile.flush()

### 2016-09-04 benchmarks/timestep_convergence/run_models.py

@@ -78,7 +78,7 @@ for r in relaxation_times:
   for t in tsteps:
     print "Running with time step %f, relaxation time %f\n"%(t, r)
     generate_prm(t, r)
-    os.system('mpirun -n 4 /home/ian/aspect/build/aspect tmp.prm')
+    os.system('mpirun -n 4 ../../aspect/build/aspect tmp.prm')
 #    plot_topography()
     outputfile.write("%f %f %f \n" % (r/tau_s, t/tau_s, calculate_error()) )
     outputfile.flush()

### 2016-09-04 benchmarks/timestep_convergence_qi/run_models.py

@@ -78,7 +78,7 @@ for theta in thetas:
   for t in tsteps:
     print "Running with time step %f, theta %f\n"%(t, theta)
     generate_prm(t, theta)
-    os.system('mpirun -n 4 /home/ian/aspect/build/aspect tmp.prm')
+    os.system('mpirun -n 4 ../../aspect/build/aspect tmp.prm')
     #plot_topography()
     outputfile.write("%f %f %f \n" % (theta, t/tau_s, calculate_error()) )
     outputfile.flush()

### 2016-09-10 benchmarks/amr/adaptive_nsfd/run_models.py
Update crameri run_models.py scripts.

@@ -13,6 +13,8 @@ def generate_prm( refine ):
       outfile.write('  set Free surface stabilization theta = 0.0\n')
     elif 'Relaxation time' in l:
       outfile.write('  set Relaxation time = 14825.\n')
+    elif 'Use nonstandard finite difference scheme' in l:
+      outfile.write('  set Use nonstandard finite difference scheme = true\n')
     elif 'Time steps between mesh refinement' in l:
       outfile.write('  set Time steps between mesh refinement = 10\n')
     else:

### 2016-09-10 benchmarks/amr/adaptive_qi/run_models.py

@@ -15,6 +15,8 @@ def generate_prm( refine ):
       outfile.write('  set Relaxation time = 0.\n')
     elif 'Time steps between mesh refinement' in l:
       outfile.write('  set Time steps between mesh refinement = 10\n')
+    elif 'Use nonstandard finite difference scheme' in l:
+      outfile.write('  set Use nonstandard finite difference scheme = false\n')
     else:
       outfile.write(l)
   prmfile.close()

### 2016-09-10 benchmarks/amr/uniform_nsfd/run_models.py

@@ -11,8 +11,10 @@ def generate_prm( refine ):
       outfile.write('  set Initial adaptive refinement = %i\n'%(0) )
     elif 'Free surface stabilization theta' in l:
       outfile.write('  set Free surface stabilization theta = 0.0\n')
+    elif 'Use nonstandard finite difference scheme' in l:
+      outfile.write('  set Use nonstandard finite difference scheme = true\n')
     elif 'Relaxation time' in l:
-      outfile.write('  set Relaxation time = 14825.\n')
+      outfile.write('  set Relaxation time = 0.\n')
     else:
       outfile.write(l)
   prmfile.close()

### 2016-09-10 benchmarks/amr/uniform_qi/run_models.py

@@ -13,6 +13,8 @@ def generate_prm( refine ):
       outfile.write('  set Free surface stabilization theta = 0.5\n')
     elif 'Relaxation time' in l:
       outfile.write('  set Relaxation time = 0.\n')
+    elif 'Use nonstandard finite difference scheme' in l:
+      outfile.write('  set Use nonstandard finite difference scheme = false\n')
     else:
       outfile.write(l)
   prmfile.close()

### 2016-09-12 benchmarks/rayleigh_taylor/run_models.py
More updates to RT tests

@@ -7,7 +7,7 @@ def generate_prm( output_dir_name, cfl, max_tstep, use_nsfd, relaxation_time, th
   for l in prmfile.readlines():
     if 'Output directory' in l:
       outfile.write('  set Output directory = %s\n'%(output_dir_name))
-    elif 'CFL numer' in l:
+    elif 'CFL number' in l:
       outfile.write('  set CFL number = %f\n'%(cfl) )
     elif 'Maximum time step' in l:
       outfile.write('  set Maximum time step = %f\n'%(max_tstep) )

### 2016-09-12 figures/rayleigh_taylor.py

@@ -55,7 +55,6 @@ plt.title('(a)')
 plt.subplot(122)
 plt.gca().grid(False)
 result = mpimg.imread('rt_result2.png')
-mpimg.imread('rt_result.png')
 plt.imshow(result, extent=[0,500.,-500., 1.8], aspect='auto')
 plt.xlim(0.,500.)
 plt.ylim(-500., 1.8)

### 2016-09-13 free-surface-paper.tex
Fix sign error.
Address some comments.

@@
 A large number of studies have approximated free surfaces in the interior of the domain by using the 
-`sticky air' approximation (see \citet{crameri2012comparison} and references therein). In this approximation there is a low-viscosity, low-density layer in the fluid 
+`sticky air' approximation (\citet{matsumoto1983numerical}; for a comprehensive review see \citet{crameri2012comparison} and references therein).
+In this approximation there is a low-viscosity, low-density layer in the fluid
 (termed `air', though its viscosity is much higher) above the free surface, effectively decoupling it from the boundary. 
@@
 
-Finally, one can use a true free surface, where a stress-free boundary condition is applied on the boundary 
-of the domain. In this case, there can be flow in and out of the domain, so the boundary must move in time.
+Finally, one can use a true free surface, where a stress-free boundary condition is applied on the surface of the simulation. 
+With this setup there can be flow in and out of the free surface, so it must move in time.
+In many cases this is accomplished through a Lagrangian or arbitrary Lagrangain-Eulerian (ALE) scheme (see Section~\ref{sec:implementation}).
+\citet{duretz2016free} proposed an alternative scheme based on using an Eulerian staggered finite difference mesh
+with a modified stencil at the surface to capture the free surface boundary condition.
+
 A true free surface has mathematical elegance in that the boundary condition of the domain more closely 
@@
 domain with frequent remeshing to avoid ill-conditioned cells. 
+In more complex geodynamic simulations there may be portions of the surface which
+correspond to regions below air (i.e. continental crust) and other portions that lie below water (i.e. oceanic crust).
+In these cases the density contrast at the surface will vary, requiring variations
+in the pressure boundary conditions (though viscous stresses should still be zero).
 
@@
 
-Several studies have attempted to alleviate the timestepping requirements imposed by the sloshing 
+Several studies have attempted to alleviate the timestepping requirements imposed by the drunken sailor 
 instability. Since the most expensive part of geodynamic simulations is typically the Stokes solve,
@@
 \begin{aligned}
-\nabla \cdot \mathbf{T} &= \rho \mathbf{g} \\
+\nabla \cdot \mathbf{T} &= - \rho \mathbf{g} \\
 \nabla \cdot \mathbf{u} &= 0,
@@
 \begin{equation}
-\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p = \rho \mathbf{g}.
+\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p = - \rho \mathbf{g}.
 \label{eq:stokes}
@@
 \begin{equation}
-- \nabla p_0 = \rho_0 \mathbf{g},
+\nabla p_0 = \rho_0 \mathbf{g},
 \label{eq:hydrostatic_stokes}
@@
 \begin{aligned}
-\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p^\prime &= \rho^\prime \mathbf{g} \qquad &\text{in } &\Omega(t)\\
+\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p^\prime &= - \rho^\prime \mathbf{g} \qquad &\text{in } &\Omega(t)\\
 \nabla \cdot \mathbf{u} &= 0  \qquad &\text{in } &\Omega(t)
@@
 We have analyzed stability of free surface boundary conditions in geodynamic simulations and 
-demonstrated the cause of sloshing instabilities using a normal mode analysis.
+demonstrated the cause of the drunken sailor instability using a normal mode analysis.
 This perspective on the problem allowed us to construct an explicit finite difference 

### 2016-09-13 benchmarks/build_plugin.sh
Add simple plugin build script.

@@ -1,0 +1,5 @@
+cd plugin
+rm CMakeCache.txt
+cmake -D ASPECT_DIR=../../aspect/build .
+make
+cd ..

### 2016-11-02 free-surface-paper.tex
Add pseudocode description.

@@
 \subsection{Choice of $\tau^*$}
+\label{sec:tau_choice}
 As discussed above, a full geodynamic simulation will have a spectrum of relaxation times.
@@
 \int_{\Gamma_F} \left( \mathbf{w \cdot n } \right) \left( \mathbf{u \cdot n} \right).
+  \label{eq:mesh_velocity_system}
 \end{equation}
@@
 Similar results were found by \citet{fullsack1995arbitrary}.
+
+\subsection{Summary of the implementation}
+We summarize this section by describing pseudocode for the implementation.
+There are two variations, one using the quasi-implicit scheme and one using nonstandard finite differences.
+
+For the quasi-implicit scheme:
+\begin{enumerate}
+  \item Solve the Stokes system, including the boundary term given in Equation~\eqref{eq:kmm}, 
+    which is a function of timestep $t^n$ and $\theta$.
+  \item Solve for the surface mesh velocity using Equation~\eqref{eq:mesh_velocity_system}.
+  \item Solve for the internal mesh velocity using Equations~\eqref{eq:laplacian_smoothing}~-~\eqref{eq:laplacian_bcs}.
+  \item Advect the mesh forward in time using displacements determined by the forward Euler scheme: $\mathbf{x}(t^{n+1}) = \mathbf{x}(t^n) + \mathbf{u}_\mathrm{mesh}\Delta t$.
+\end{enumerate}
+
+For the nonstandard finite difference scheme:
+\begin{enumerate}
+  \item Estimate $\tau_\mathrm{min}$ using one of the means described in Section~\ref{sec:tau_choice}.
+  \item Solve the Stokes system normally.
+  \item Solve for the surface mesh velocity using Equation~\eqref{eq:mesh_velocity_system}.
+  \item Solve for the internal mesh velocity using Equations~\eqref{eq:laplacian_smoothing}~-~\eqref{eq:laplacian_bcs}.
+  \item Advect the mesh forward in time using displacements determined by the nonstandard finite difference scheme (Equation~\eqref{eq:nsfd}).
+\end{enumerate}
 

### 2016-11-02 new_response_to_referee.tex
Add response letter.

\documentclass[a4paper,12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{csquotes}

%opening
\title{Response to referees}
\author{Ian Rose, Bruce Buffett, and Timo Heister}
\date{}

\begin{document}

\maketitle

We thank Taras Gerya and the anonymous referee for their thoughtful and constructive reviews,
to which we respond inline here.

\section*{ Reviewer 1}

\begin{quotation}
  This paper deals with the (now) notorious issue of the drunken-sailor numerical instability that plagues free-surface calculations particularly where the surface deformation is the primary driver of flow and the flow rates are decaying. The same instability also occurs for stable internal boundaries where there same dynamic arguments apply.
\end{quotation}

\begin{quotation}
There have been a number of mechanisms suggested for "fixing" this instability published in recent years and this MS adds a new take to each of those (and appropriately acknowledges / contextualises them). I think this work is a significant addition to the collection of methods available and I would recommend that the MS be published. There has already been significant to-and-fro in review and this is apparent from the quality of the writing and clarity of the explanation. Nevertheless, every fresh pair of eyes brings some new suggestions and I would make the following comments (which I don't consider binding):
\end{quotation}

\begin{quotation}
I find it helpful to have a pseudo-code explanation of any implementation. There is a lot of derivation but in the end it appears that we have a relatively simple algorithm which pulls together equation (49) and equation (35), but if I am missing any subtle details, then perhaps that is indicating that a description would be helpful.
\end{quotation}

We agree that this is a useful summary of our implementation.
We have added a new section (Section 7.3) which describes a pseduo-code implementation
of both the quasi-implicit and nonstandard finite difference scheme.

\begin{quotation}
I think of the sloshing instability as a physical one in which there is a resonance between the waves on a fluid and the dynamics of the container (fluid sloshing in tankers causing capsizes, for example). So I am not sure that it is a good name for a purely numerical instability even if that is how it has been named by others.
\end{quotation}

This point is well taken. We now note that this instability has been called the ``sloshing'' instability,
but in the bulk of the text we refer to it as the ``drunken sailor'' instability.

\begin{quotation}
Another example I can think of where this stabilization might be useful is at an inflow boundary where the inflow condition is used to balance a global contraint (e.g. volume / mass conservation). An example of where this arises is in modeling of extensional systems where isostasy sets the lower boundary to have a constant (average) pressure through time. There are plenty of spurious modes that can arise in the inflow condition but one instability that is often important is similar to the drunken-sailor and can be supressed by small timesteps.
\end{quotation}
This sounds like an interesting application.
Indeed, there are several applications that exhibit similar instabilities, 
and may be good targets for similar stabilization methods 
(including fluid-structure-interactions and elastoplastic deformation).
That being said, we have not analyzed those applications in this context,
and are unprepared to discuss specific commonalities.
Thus we feel this is outside of the scope of this manuscript.

\begin{quotation}
Following on, we sometimes fix those sorts of problems by under-relaxing the boundary deformation over time with very little mathematical justification for doing so. I would be interested to know if this is a poor-man's equivalent to the solution you propose.
\end{quotation}

We would suggest that under-relaxing boundary deformation is indeed a poor-man's equivalent to our NSFD scheme,
and would love to hear more about this practice.
However, we are not aware of discussion of this in the literature or open-souce codebases, and so can't comment further.

\begin{quotation}
   Your method remains formally 1st order. Are higher-order discretisations of the mesh deformation helpful in this case or do they suffer the same / worse problems ?
\end{quotation}

Our analysis of the instability is purely in the time discretization, 
so higher-order spatial discretizations should suffer from the identical instability.
We have added a note to this effect in the manuscript.

\section*{Reviewer 2}
\begin{quotation}
This is useful and timely paper developing new approach for the free surface treatment with finite elements in geodynamic simulations. This is an important topic in geodynamics since essential need for free surface in modeling regional and global processes has been repeatedly demonstrated (e.g. Crameri, F., Tackley, P.J., Meilick, I., Gerya, T., Kaus, B.J.P. (2012) A free plate surface and weak oceanic crust produce single-sided subduction on Earth. Geophys. Res. Lett., doi:10.1029/2011GL050046. And references therein). The revised version reads well and addressed properly the comments of reviewers. I only noticed few minor points that need some attention (listed below).
Taras Gerya, Zurich, 15.08.2016
\end{quotation}

\begin{quotation}
Introduction. The earliest implementation of sticky air approach was published by Matsumoto, T., Tomoda, Y., 1983. Numerical-simulation of the initiation of subduction at the fracture-zone. J. Phys. Earth 31, 183–194.
Introduction. There is one recent paper about true free surface implementation with staggered grid by T Duretz, DA May, P Yamato  (2016) A free surface capturing discretization for the staggered grid finite difference scheme Geophysical Journal International 204 (3), 1518-1530 
\end{quotation}
We have added these pertinent references.

\begin{quotation}
Line 27. “A true free surface has mathematical elegance in that the boundary condition of the domain more closely matches the boundary conditions which one is trying to model, but it typically requires a deformable domain with frequent remeshing to avoid ill-conditioned cells.” One should mention here, however, possibilities of treating more complex realistic cases of common geodynamic simulations, in which oceanic free surface is covered by water whereas continental regions contact with the athmosphere. Density contrast along such free surface will be non-uniform, which would require some special treatments.
\end{quotation}
This is a good point, and we have added a note to this effect.
\begin{quotation}
Equations 1,3,4,5 – minus is missing in front of RHO*g
\end{quotation}
These are now fixed.

\begin{quotation}
The authors mentioned that they did Schmeling et al. benchmark (2008) but do not show it since it is difficult to interpret and quantify. I would still recommend showing one-two figures with the progress of model evolution and surface development for this model. It will simply show that the developed algorithm does not break when treating more complex plate tectonics-like models with large viscosity contrasts and complex evolution and advection of the free surface.
\end{quotation}

\end{document}

### 2016-11-08 free-surface-paper.tex
Add note about spatial discretizations.
Put pseudocode in a figure.

@@
 \usepackage{url}
+\usepackage{subcaption}
+\usepackage{xcolor}
+\usepackage{mdframed}
 
@@
 
+This stability analysis does not rely on the specific choice of spatial discretization,
+and so should hold equally well for linear finite elements and higher-order elements.
+
 \section{Analysis of quasi-implicit stabilization}
@@
 
-\subsection{Summary of the implementation}
-We summarize this section by describing pseudocode for the implementation.
-There are two variations, one using the quasi-implicit scheme and one using nonstandard finite differences.
-
-For the quasi-implicit scheme:
+Figure~\ref{fig:pseudocode} shows pseudocode descriptions of the free surface implementation described in this section.
+
+\begin{figure}
+\centering
+\begin{subfigure}[b]{0.43\textwidth}
+\begin{mdframed}[backgroundcolor=gray!20]
+{\small
 \begin{enumerate}
@@
 \end{enumerate} 
-
-For the nonstandard finite difference scheme:
+}
+\end{mdframed}
+\caption{Quasi-implicit}
+\end{subfigure}
+\begin{subfigure}[b]{0.43\textwidth}
+\begin{mdframed}[backgroundcolor=gray!20]
+{\small
 \begin{enumerate}
@@
 \end{enumerate}
+\vspace{1.4mm}
+}
+\end{mdframed}
+\caption{Nonstandard finite differences}
+\end{subfigure}
+\caption{Pseudocode descriptions of the our free surface implementations.}
+\label{fig:pseudocode}
+\end{figure}
 

### 2016-11-08 new_response_to_referee.tex
Add some text from Bruce.

@@
 
+We investigated the subduction benchmark of Schmeling et al. 2008, but do not report
+the results because the benchmark is not well designed to test our time-stepping scheme.
+The chief difficulty is sharp corners in the shape of the initial slab. While the initial condition
+is easy to implement, it introduces singularities into the problem. These singularities are
+resolved indirectly by limiting the spatial resolution. Moreover the viscosity contrast in this
+subduction benchmark is only a factor of 100. Relatively little is gained over the (smaller)
+factor of 10 viscosity increase in the Rayleigh-Taylor benchmark.
+
 \end{document}

### 2016-11-11 free-surface-paper.tex
Add Taras to thank yous.
add sentence to abstract

@@
 
-Finally, we discuss the implementation of a free surface in the open source, community based
+Finally, we present numerical results to show the effectiveness of the new approach and discuss the free surface implementation in the open source, community based
 mantle convection software \texttt{ASPECT}.
@@
 Thanks to Wolfgang Bangerth, Brent Delbridge, and Cedric Thieulot for useful discussions.
-Thanks to Fabio Crameri and Dave May for constructive reviews.
+Thanks to Fabio Crameri, Taras Gerya, and Dave May for constructive reviews.
 

### 2016-11-11 free-surface-paper.tex
format pseudocode
small edits

@@
 With this setup there can be flow in and out of the free surface, so it must move in time.
-In many cases this is accomplished through a Lagrangian or arbitrary Lagrangain-Eulerian (ALE) scheme (see Section~\ref{sec:implementation}).
+In many cases this is accomplished through a Lagrangian or arbitrary Lagrangian-Eulerian (ALE) scheme (see Section~\ref{sec:implementation}).
 \citet{duretz2016free} proposed an alternative scheme based on using an Eulerian staggered finite difference mesh
@@
 
+% \begin{figure}
+% \centering
+% \begin{subfigure}[b]{0.43\textwidth}
+% \begin{mdframed}[backgroundcolor=gray!20]
+% {\small
+% \begin{enumerate}
+%   \item Solve the Stokes system, including the boundary term given in Equation~\eqref{eq:kmm}, 
+%     which is a function of timestep $t^n$ and $\theta$.
+%   \item Solve for the surface mesh velocity using Equation~\eqref{eq:mesh_velocity_system}.
+%   \item Solve for the internal mesh velocity using Equations~\eqref{eq:laplacian_smoothing}~-~\eqref{eq:laplacian_bcs}.
+%   \item Advect the mesh forward in time using displacements determined by the forward Euler scheme: $\mathbf{x}(t^{n+1}) = \mathbf{x}(t^n) + \mathbf{u}_\mathrm{mesh}\Delta t$.
+% \end{enumerate} 
+% }
+% \end{mdframed}
+% \caption{Quasi-implicit}
+% \end{subfigure}
+% \begin{subfigure}[b]{0.43\textwidth}
+% \begin{mdframed}[backgroundcolor=gray!20]
+% {\small
+% \begin{enumerate}
+%   \item Estimate $\tau_\mathrm{min}$ using one of the means described in Section~\ref{sec:tau_choice}.
+%   \item Solve the Stokes system normally.
+%   \item Solve for the surface mesh velocity using Equation~\eqref{eq:mesh_velocity_system}.
+%   \item Solve for the internal mesh velocity using Equations~\eqref{eq:laplacian_smoothing}~-~\eqref{eq:laplacian_bcs}.
+%   \item Advect the mesh forward in time using displacements determined by the nonstandard finite difference scheme (Equation~\eqref{eq:nsfd}).
+% \end{enumerate}
+% \vspace{1.4mm}
+% }
+% \end{mdframed}
+% \caption{Nonstandard finite differences}
+% \end{subfigure}
+% \caption{Pseudocode descriptions of the free surface implementations.}
+% \label{fig:pseudocode}
+% \end{figure}
+
 \begin{figure}
 \centering
-\begin{subfigure}[b]{0.43\textwidth}
+
+%\caption
+{Quasi-implicit scheme:}
+
 \begin{mdframed}[backgroundcolor=gray!20]
@@
 \end{mdframed}
-\caption{Quasi-implicit}
-\end{subfigure}
-\begin{subfigure}[b]{0.43\textwidth}
+
+%\caption
+{Nonstandard finite differences:}
+
 \begin{mdframed}[backgroundcolor=gray!20]
@@
 \end{mdframed}
-\caption{Nonstandard finite differences}
-\end{subfigure}
-\caption{Pseudocode descriptions of the our free surface implementations.}
+
+\caption{Pseudocode for a single timestep of the free surface algorithms.}
 \label{fig:pseudocode}
@@
 \begin{figure*}
+\centering
 \includegraphics[width=0.9\textwidth]{figures/rayleigh_taylor.pdf}
@@
 \begin{figure*}
+\centering
 \includegraphics[width=0.9\textwidth]{figures/rayleigh_taylor_tau_choice.pdf}

### 2016-11-11 new_response_to_referee.tex
update response

@@
 We agree that this is a useful summary of our implementation.
-We have added a new section (Section 7.3) which describes a pseduo-code implementation
+We have added Figure 4 which describes a pseudocode implementation
 of both the quasi-implicit and nonstandard finite difference scheme.
@@
 \begin{quotation}
-Another example I can think of where this stabilization might be useful is at an inflow boundary where the inflow condition is used to balance a global contraint (e.g.
+Another example I can think of where this stabilization might be useful is at an inflow boundary where the inflow condition is used to balance a global constraint (e.g.
 volume / mass conservation).
@@
 This sounds like an interesting application.
-Indeed, there are several applications that exhibit similar instabilities, 
+Indeed, there are several situations that exhibit similar instabilities, 
 and may be good targets for similar stabilization methods 
@@
 and would love to hear more about this practice.
-However, we are not aware of discussion of this in the literature or open-souce codebases, and so can't comment further.
+However, we are not aware of discussion of this in the literature or open-source codebases, and so can't comment further or address it in the manuscript.
 

### 2016-11-13 new_response_to_referee.tex
Add note.

@@
 
-Our analysis of the instability is purely in the time discretization, 
+It is true that our method is a first-order method in time.
+Our analysis of the instability is purely for the timestepping,
 so higher-order spatial discretizations should suffer from the identical instability.
